### 2013-11-04 sugarlike.tex
added the .tex file for project proposal - LL Still need LOTSA work!! =)

\documentclass[11pt]{article}

\title{\textbf{Sugarlike: A Language Identification Library for Low Resource Languages}}
\author{Susanne Fertmann, Guy Emerson, Liling Tan}
\date{}

\begin{document}

\maketitle

\section{Introduction}

Sugarlike is a language identification library designed for identifying low resource language given a document\footnote{A document can refer to a webpage, a textfile or even a sentence.}.

\subsection{Previous Work}

Previous work on language identification for low resource languages...
Kevin Schnell's work, blah blah blah... but data not available.

\section{Data}
With regards to the data sources, we propose crawling Twitter posts using a fixed list of hashtags in English and their respective translations. Although the redistribution of our crawled corpus remain questionable, we will distribute the source code that allows researchers to rebuild an update corpus.

\section{Model}

I am not very sure, lol - LL

\subsection{Need for Modularity}

The trained language identification model would remain open and free for use. And more importantly, the model we are building must remain modular in the sense that other research can rebuild an updated Twitter corpus and extend the model simply by either adding new features or new datapoints to the classification model.

\section{Implementation}

Build a crawl for Twitter post.

\end{document}

### 2013-11-05 firstweektask/firstweektask.py
Add test for 2nd method of fistweektask.

@@ -17,5 +17,6 @@ def method3(text):
   print "The language of the sentence is most probably",lang
   pass
 
-# TODO: write test for each method. - LL
+
+# TODO: write test for third method. - S
 # Hey I made a comment - G

### 2013-11-05 firstweektask/test_firstweektask.py

@@ -1,21 +1,68 @@
 import firstweektask as fwt
 import unittest as ut
+import tempfile
+import codecs
+import os
 
 class testfwt (ut.TestCase):
   testsentences = [("This is a foo bar sentence",
                     ["This", "is", "a", "foo", "bar", "sentence"]),                   
                    ("Ich bin schwanger?",
                     ["Ich","bin","schwanger","?"]),
-                   ("Ya, das ist nicht fett.",
-                    ["Ya",",","das","ist","nicht","fett","."]),
-                   ("Das ist baby.",
-                     ["Das","ist","baby","."])
+                   ("Ja, das ist nicht fett.",
+                    ["Ja",",","das","ist","nicht","fett","."]),
+                   ("Das ist ein Baby.",
+                     ["Das","ist", "ein","Baby","."])
                    ]
+  tokencounts = [("This", 1),
+                 ("is", 1),
+                 ("a", 1),
+                 ("foo", 1),
+                 ("bar", 1),
+                 ("sentence", 1),
+                 ("Ich", 1),
+                 ("bin", 1),
+                 ("schwanger", 1),
+                 ("Ja", 1),
+                 ("das", 1),
+                 ("ist", 2),
+                 ("nicht", 1),
+                 ("fett", 1),
+                 ("Das", 1),
+                 ("ein", 1),
+                 ("Baby", 1),
+                 (".", 2),
+                 ("?", 1),
+                 (",", 1)
+                 ]
   
   def testMethod1(self):
     for sent,tokens in self.testsentences:
       result = fwt.method1(sent)
       self.assertEqual(tokens, result)
+      
+  def testMethod2(self):
+    tmpoutfile = tempfile.mkstemp()[1]
+    tmpinfile = tempfile.mkstemp()[1]
+
+    with open('tmpinfile', 'w') as infile:
+      for sent, tokens in self.testsentences:
+        infile.write(sent + '\n')
+     
+    expected_lines = []
+    for token, num in self.tokencounts:
+      expected_lines.append(token + "\t" + str(num))
+                  
+    fwt.method2('tmpinfile', 'tmpoutfile')
+    lines = []
+    with codecs.open('tmpoutfile', 'r') as out:
+      for line in out:
+        lines.append(line.strip('\n'))
+
+    self.assertEqual(set(lines), set(expected_lines))
+    os.remove('tmpinfile')
+    os.remove('tmpoutfile')
+        
       
 if __name__ == "__main__":
   ut.main()

### 2013-11-05 sugarlike.tex
updated the .tex
added more information on the proposal (sugarlike.tex) - Su

@@
 
+%@Su-Ga: I can't think of a name, so i anyhow gave it one "sugarlike". Please feel free to change it
+
 \subsection{Previous Work}
 
-Previous work on language identification for low resource languages...
-Kevin Schnell's work, blah blah blah... but data not available.
+Previous work on language identification has mostly be done high resource languages\footnote{e.g.
+http://shuyo.wordpress.com/2012/02/21/language-detection-for-twitter-with-99-1-accuracy/}.
+One exception is Scannell's (XXXX) indigenous tweets project.
+%where ....
+
+Language identification at the word level is considered as one of the harder task because of the lack of context \{citation needed\}, especially in multilingual documents.
+King and Abney (2013) resolved it in a weakly supervised fashion.
 
 \section{Data}
-With regards to the data sources, we propose crawling Twitter posts using a fixed list of hashtags in English and their respective translations.
+
+%With regards to the data sources, we propose crawling Twitter posts using a fixed list of hashtags in English and their respective translations.
 Although the redistribution of our crawled corpus remain questionable, we will distribute the source code that allows researchers to rebuild an update corpus.
+[we need some proper methodology...
+ahahhaa]
+
+\begin{itemize}
+\item Universal Declaration of Human Rights: almost parallel data,  needs some cleaning in ~300 language
+\item Language Id Corpus (LId): character 3-grams of Crubadan corpus %there's some unknown angeled brackets
+\item Crawl the tweets linked to www.indigenoustweets.com (~150 languages)
+\item Crawl the blogs linked to www.indigenoustweets.com/blogs (~75 languages)
+\end{itemize}
 
@@
 
-I am not very sure, lol - LL
+The resulting system needs to be able to identify (i) the language family and (ii) the low resource language of the document.
+
+\begin{itemize}
+\item Baseline: Baysian model using LId corpus
+\item Main model: to be decided later.
+\end{itemize}
+
+\section{Implementation}
+\itemize{
+\item Data needs to be clean and processable for feature extraction
+\item Choose and implement the language model for language identification
+\item Evaluate the system
+}
 
@@
 
-\section{Implementation}
-
-Build a crawl for Twitter post.
-
 \end{document}

### 2013-11-05 firstweektask/firstweektask.py
Added a third unit test, and fixed some mistakes.

@@ -13,9 +13,9 @@ def method2(infile, outfile):
   return fdist
 
 def method3(text):
-  lang = "Proto-Indo European: English"
-  print "The language of the sentence is most probably",lang
-  pass
+  lang = ("Indo-European","English")
+  print "The language of the sentence is most probably: {1} ({0})".format(lang[0],lang[1])
+  return lang
 
 
 # TODO: write test for third method. - S

### 2013-11-05 firstweektask/test_firstweektask.py

@@ -6,13 +6,17 @@ import os
 
 class testfwt (ut.TestCase):
   testsentences = [("This is a foo bar sentence",
-                    ["This", "is", "a", "foo", "bar", "sentence"]),                   
+                    ["This", "is", "a", "foo", "bar", "sentence"],
+                    ("Indo-European","English")),                   
                    ("Ich bin schwanger?",
-                    ["Ich","bin","schwanger","?"]),
+                    ["Ich","bin","schwanger","?"],
+                    ("Indo-European","German")),
                    ("Ja, das ist nicht fett.",
-                    ["Ja",",","das","ist","nicht","fett","."]),
+                    ["Ja",",","das","ist","nicht","fett","."],
+                    ("Indo-European","German")),
                    ("Das ist ein Baby.",
-                     ["Das","ist", "ein","Baby","."])
+                     ["Das","ist", "ein","Baby","."],
+                    ("Indo-European","German"))
                    ]
   tokencounts = [("This", 1),
                  ("is", 1),
@@ -37,32 +41,38 @@ class testfwt (ut.TestCase):
                  ]
   
   def testMethod1(self):
-    for sent,tokens in self.testsentences:
-      result = fwt.method1(sent)
+    for sentence, tokens, _ in self.testsentences:
+      result = fwt.method1(sentence)
       self.assertEqual(tokens, result)
       
   def testMethod2(self):
     tmpoutfile = tempfile.mkstemp()[1]
     tmpinfile = tempfile.mkstemp()[1]
 
-    with open('tmpinfile', 'w') as infile:
-      for sent, tokens in self.testsentences:
-        infile.write(sent + '\n')
+    with open(tmpinfile, 'w') as infile:
+      for sentence, tokens, _ in self.testsentences:
+        infile.write(sentence + '\n')
      
     expected_lines = []
     for token, num in self.tokencounts:
       expected_lines.append(token + "\t" + str(num))
                   
-    fwt.method2('tmpinfile', 'tmpoutfile')
+    fwt.method2(tmpinfile, tmpoutfile)
     lines = []
-    with codecs.open('tmpoutfile', 'r') as out:
+    with codecs.open(tmpoutfile, 'r') as out:
       for line in out:
         lines.append(line.strip('\n'))
 
     self.assertEqual(set(lines), set(expected_lines))
-    os.remove('tmpinfile')
-    os.remove('tmpoutfile')
-        
+    os.remove(tmpinfile)
[…]

### 2013-11-05 sugarlike.tex
Update proposal, add references.

@@
 \date{}
+
+\usepackage{hyperref}
 
@@
 
-Previous work on language identification has mostly be done high resource languages\footnote{e.g.
-http://shuyo.wordpress.com/2012/02/21/language-detection-for-twitter-with-99-1-accuracy/}.
-One exception is Scannell's (XXXX) indigenous tweets project.
+Previous work on language identification has mostly be done on high resource languages\footnote{e.g.
+\href{http://shuyo.wordpress.com/2012/02/21/language-detection-for-twitter-with-99-1-accuracy/}{http://shuyo.wordpress.com/2012/02/21/language-detection-for-twitter-with-99-1-accuracy/}}.
+One exception is Scannell's indigenous tweets project \footnote{\href{www.indigenoustweets.com}{www.indigenoustweets.com}}.
 %where ....
 
-Language identification at the word level is considered as one of the harder task because of the lack of context \{citation needed\}, especially in multilingual documents.
-King and Abney (2013) resolved it in a weakly supervised fashion.
+Language identification at the word level is considered as one of the harder tasks because of the lack of context \footnote{citation needed}, especially in multilingual documents.
+King and Abney (2013 \cite{king-abney:2013}) resolved it in a weakly supervised fashion.
 
@@
 \begin{itemize}
-\item Universal Declaration of Human Rights: almost parallel data,  needs some cleaning in ~300 language
-\item Language Id Corpus (LId): character 3-grams of Crubadan corpus %there's some unknown angeled brackets
-\item Crawl the tweets linked to www.indigenoustweets.com (~150 languages)
-\item Crawl the blogs linked to www.indigenoustweets.com/blogs (~75 languages)
+\item Universal Declaration of Human Rights: almost parallel data,  needs some cleaning (\~300 languages)
+\item Language Id Corpus (LId): character 3-grams of Crúbadán corpus %there's some unknown angeled brackets
+\item Crawl the tweets linked to www.indigenoustweets.com (\~150 languages)
+\item Crawl the blogs linked to www.indigenoustweets.com/blogs (\~75 languages)
 \end{itemize}
@@
 \item Baseline: Baysian model using LId corpus
-\item Main model: to be decided later.
+\item Main model: to be decided later
 \end{itemize}
@@
 
+\
+\
+\
+\
+\
+
+\bibliography{literature.bib}{}
+\bibliographystyle{plain}
+
 \end{document}

### 2013-11-05 firstweektask/test_firstweektask.py
Put quotes back in

@@ -49,7 +49,7 @@ class testfwt (ut.TestCase):
     tmpoutfile = tempfile.mkstemp()[1]
     tmpinfile = tempfile.mkstemp()[1]
 
-    with open(tmpinfile, 'w') as infile:
+    with open('tmpinfile', 'w') as infile:
       for sentence, tokens, _ in self.testsentences:
         infile.write(sentence + '\n')
      
@@ -57,15 +57,15 @@ class testfwt (ut.TestCase):
     for token, num in self.tokencounts:
       expected_lines.append(token + "\t" + str(num))
                   
-    fwt.method2(tmpinfile, tmpoutfile)
+    fwt.method2('tmpinfile', 'tmpoutfile')
     lines = []
-    with codecs.open(tmpoutfile, 'r') as out:
+    with codecs.open('tmpoutfile', 'r') as out:
       for line in out:
         lines.append(line.strip('\n'))
 
     self.assertEqual(set(lines), set(expected_lines))
-    os.remove(tmpinfile)
-    os.remove(tmpoutfile)
+    os.remove('tmpinfile')
+    os.remove('tmpoutfile')
 
   def testMethod3(self):  # We don't need to test if the language is correct, since the system will never be perfect. 
     for sentence, _, language in self.testsentences:

### 2013-11-05 firstweektask/firstweektask.py
Removed old comments

@@ -17,6 +17,3 @@ def method3(text):
   print "The language of the sentence is most probably: {1} ({0})".format(lang[0],lang[1])
   return lang
 
-
-# TODO: write test for third method. - S
-# Hey I made a comment - G

### 2013-11-05 firstweektask/test_firstweektask.py
fixed the windows file problem

@@ -46,33 +46,41 @@ class testfwt (ut.TestCase):
       self.assertEqual(tokens, result)
       
   def testMethod2(self):
-    tmpoutfile = tempfile.mkstemp()[1]
-    tmpinfile = tempfile.mkstemp()[1]
+    try: 
+      tmpoutfile = tempfile.mkstemp()[1]
+      tmpinfile = tempfile.mkstemp()[1]
+    except: # Just in case, python complains on Windows.
+      tmpoutfile = tempfile.NamedTemporaryFile()
+      tmpinfile = tempfile.NamedTemporaryFile()
+      tmpinfile = tmpinfile.name
+      tmpoutfile = tmpoutfile.name
+      pass
 
-    with open('tmpinfile', 'w') as infile:
-      for sentence, tokens, _ in self.testsentences:
-        infile.write(sentence + '\n')
+    with open(tmpinfile, 'w') as infile:
+      for sent, tokens, _ in self.testsentences:
+        infile.write(sent + '\n')
      
     expected_lines = []
     for token, num in self.tokencounts:
       expected_lines.append(token + "\t" + str(num))
                   
-    fwt.method2('tmpinfile', 'tmpoutfile')
+    fwt.method2(tmpinfile, tmpoutfile)
     lines = []
-    with codecs.open('tmpoutfile', 'r') as out:
+    with codecs.open(tmpoutfile, 'r') as out:
       for line in out:
         lines.append(line.strip('\n'))
 
     self.assertEqual(set(lines), set(expected_lines))
-    os.remove('tmpinfile')
-    os.remove('tmpoutfile')
 
+    os.remove(tmpinfile)
+    os.remove(tmpoutfile)
+    
   def testMethod3(self):  # We don't need to test if the language is correct, since the system will never be perfect. 
     for sentence, _, language in self.testsentences:
       prediction = fwt.method3(sentence)
       self.assertEqual(len(prediction),2)
       self.assertIsInstance(prediction[0],str)
       self.assertIsInstance(prediction[1],str)
-      
+  
 if __name__ == "__main__":
   ut.main()

### 2013-11-05 firstweektask/test_firstweektask.py
fixed windows tempfile without the pass in try-except

@@ -54,7 +54,6 @@ class testfwt (ut.TestCase):
       tmpinfile = tempfile.NamedTemporaryFile()
       tmpinfile = tmpinfile.name
       tmpoutfile = tmpoutfile.name
-      pass
 
     with open(tmpinfile, 'w') as infile:
       for sent, tokens, _ in self.testsentences:

### 2013-11-05 sugarlike.tex
Solve latex layout issues.

@@
 \begin{itemize}
-\item Universal Declaration of Human Rights: almost parallel data,  needs some cleaning (\~300 languages)
+\item Universal Declaration of Human Rights: almost parallel data,  needs some cleaning ($\sim$300 languages)
 \item Language Id Corpus (LId): character 3-grams of Crúbadán corpus %there's some unknown angeled brackets
-\item Crawl the tweets linked to www.indigenoustweets.com (\~150 languages)
-\item Crawl the blogs linked to www.indigenoustweets.com/blogs (\~75 languages)
+\item Crawl the tweets linked to www.indigenoustweets.com ($\sim$150 languages)
+\item Crawl the blogs linked to www.indigenoustweets.com/blogs ($\sim$75 languages)
 \end{itemize}
@@
 \
+\\
 \
-\
-\
-\
+\\
 
-\bibliography{literature.bib}{}
+\bibliography{literature}{}
 \bibliographystyle{plain}

### 2013-11-05 firstweektask/test_firstweektask.py
deleted the linux vs windows try-except

@@ -45,15 +45,11 @@ class testfwt (ut.TestCase):
       result = fwt.method1(sentence)
       self.assertEqual(tokens, result)
       
-  def testMethod2(self):
-    try: 
-      tmpoutfile = tempfile.mkstemp()[1]
-      tmpinfile = tempfile.mkstemp()[1]
-    except: # Just in case, python complains on Windows.
-      tmpoutfile = tempfile.NamedTemporaryFile()
-      tmpinfile = tempfile.NamedTemporaryFile()
-      tmpinfile = tmpinfile.name
-      tmpoutfile = tmpoutfile.name
+  def testMethod2(self):  
+    tmpoutfile = tempfile.NamedTemporaryFile()
+    tmpinfile = tempfile.NamedTemporaryFile()
+    tmpinfile = tmpinfile.name
+    tmpoutfile = tmpoutfile.name
 
     with open(tmpinfile, 'w') as infile:
       for sent, tokens, _ in self.testsentences:

### 2014-01-17 sugarlike/multinomialnaivebayes.py
Added method to naively combine multiple SGT-smoothed models.

@@ -1,16 +1,22 @@
+#!/usr/bin/env python -*- coding: utf-8 -*-
+
+#import sys; sys.path.append('../')
+
 from collections import Counter
 from math import log, exp
 from nltk.probability import SimpleGoodTuringProbDist as SGTdist
 from nltk import FreqDist
 
-def SGT(data, vocab=None):
+import sys; reload(sys); sys.setdefaultencoding("utf-8")
+
+def SGT(data, vocab=None, min=10000):
   '''
   Calculates a multinomial model with Simple Good-Turing smoothing.
   By default, the estimated vocabulary size is twice the observed size, or 10 000, whichever is larger
   Assumes frequencies in the form of a Counter.
   '''
   if vocab == None:
-    vocab = max( 2 * len(data), 10000)
+    vocab = max( 2 * len(data), min)
   return SGTdist(FreqDist(data),vocab)
 
 def SGTestimate(language, input):
@@ -21,7 +27,16 @@ def SGTestimate(language, input):
   result = 0
   for x in input:
     result += language.logprob(x) * input[x]
-  return exp(result)
+  return result
+
+def combined_estimate(classifiers, input):
+  '''
+  Naively combines a set of SGT-smoothed models.
+  '''
+  result = 0
+  for model in classifiers:
+    result += SGTestimate(model, input)
+  return result
 
 def MLE(freq):
   '''
@@ -43,7 +58,7 @@ def MLEestimate(language, input, oov=-float('inf')):
       result += language[x] * input[x]
     except KeyError:  # Out of vocabulary items
       result += oov
-  return exp(result)
+  return result
 
 '''
 train = Counter({'a':1,'b':5,'c':2})
@@ -57,30 +72,21 @@ print MLEestimate(langMLE,test)
 '''
 
 from extractfeature import sentence2ngrams, get_features
-s = "ich bin schwanger"
+s = u"je m'appelle Claude"
 test = Counter(sentence2ngrams(s, with_word_boundary=True))
 print test
 
 trainset = get_features('odin', option='3gram')
 sgt_results = []
-mle_results = []
+#mle_results = []
 
-'''
-german = SGT(trainset['deu'])
-wakawaka = SGT(trainset['wkw'])
 
-for x in test:
-  print x, trainset['deu'][x], SGTestimate(german, Counter({x:1}))
-  print x, trainset['wkw'][x], SGTestimate(wakawaka, Counter({x:1}))
-
[…]

### 2014-01-17 sugarlike/extractfeature.py
edited extractfeature.py to accept 2 letter codes

@@ -28,7 +28,7 @@ def sentence2ngrams(text,n=3, option='char', with_word_boundary=False):
                         for i in text.split()]))
   if option == 'word':
     from nltk.util import ngrams
-    return list(ngrams(text.split(), n))
+    return list(ngrams(text.split(), 1))
   
   if "gram" in option:
     n = int(option[0])
@@ -93,9 +93,12 @@ def crubadan2counters(crubadanfile='crub-131119.zip'):
       path, filename = os.path.split(infile)
       lang = filename.rpartition('.')[0]
       if lang in ISO2LANG:
+        if lang.startswith('fr'):
+          print(lang)
         print ('crubadan', infile, lang, 'Creating feature sets, please be patient...')
         for line in inzipfile.open(infile):
           key, count = line.strip().split(' ')
+          
           if 'words' in path: # Updates wordfreq
             wordfreqs[lang][key] = int(count)
           if 'chars' in path: # Updates charngrams
@@ -103,6 +106,8 @@ def crubadan2counters(crubadanfile='crub-131119.zip'):
       else:
         datalost.add(infile)
   return charngrams, wordfreqs, datalost
+
+crubadan2counters()
 
 def feature_interface(data_source):
   """ Unpickle the feature pickles if exists else create them. """
@@ -121,9 +126,9 @@ def feature_interface(data_source):
   return charngrams, wordfreqs
 
 def get_features(data_source, language=None, option='char', \
-                 with_word_boundary=True):
-  charngs, wordfqs = feature_interface(data_source)
-  """ Get features given the data_source, language and option"""  
+                 with_word_boundary=True):  
+  """ Get features given the data_source, language and option"""
+  charngs, wordfqs = feature_interface(data_source)  
   if option == 'char':
     result = charngs[language] if language else charngs
   elif option == 'word':
@@ -160,7 +165,10 @@ x = get_features('odin',option='3gram')
 for i in x:
   print(i, x[i])
 '''
-
-
-
+'''
+x = get_features('crubadan', option='3gram')
+for i in x:
+  for j in x[i]:
+    print (i,j, x[i][j])
+'''
 

### 2014-01-17 sugarlike/langid.py

@@ -76,3 +76,7 @@ t = 'ich bin schwanger'
 #print sugarlid_gnb(t, option='3gram')[:10]
 #print sugarlid_bnb(t, option='3gram')[:10]
 print sugarlid_cosine(t, option='3gram')[:10]
+print sugarlid_cosine(t, option='2gram')[:10]
+print sugarlid_cosine(t, option='1gram')[:10]
+print sugarlid_cosine(t, option='4gram')[:10]
+print sugarlid_cosine(t, option='5gram')[:10]

### 2014-01-17 universalcorpus/crawlandclean/udhr.py

@@ -112,3 +112,4 @@ def sents(intarfile=parentddir+'/data/udhr/udhr-unicode.tar', \
 def source_sents(intarfile=parentddir+'/data/udhr/udhr-unicode.tar', \
           bysentence=True):
   return sents(intarfile, bysentence)
+

### 2014-01-17 universalcorpus/miniethnologue.py

@@ -8583,6 +8583,272 @@ zho  Chinese  cdo  Min Dong Chinese
 zza  Zaza  diq  Dimli (individual language)
    kiu  Kirmanjki (individual language)""" 
 
+listofwikis = """English Wikipedia  English  English  en  6
+German Wikipedia  German  Latn  de  6
+French Wikipedia  French  Latn  fr  6
+Dutch Wikipedia  Dutch  Latn  nl  6
+Italian Wikipedia  Italian  Latn  it  6
+Spanish Wikipedia  Spanish  Latn  es  6
+Russian Wikipedia  Russian  Cyrl  ru  6
+Swedish Wikipedia  Swedish  Latn  sv  6
+Polish Wikipedia  Polish  Latn  pl  6
+Japanese Wikipedia  Japanese  Jpan  ja  5
+Portuguese Wikipedia  Portuguese  English  pt  6
+Arabic Wikipedia  Arabic  Arab  ar  6
+Chinese Wikipedia  Chinese  Hans/Hant  zh  6
+Ukrainian Wikipedia  Ukrainian  Cyrl  uk  5
+Catalan Wikipedia  Catalan  Latn  ca  5
+Bokmål/Riksmål Wikipedia  Norwegian (Bokmål)  Latn  no  5
+Finnish Wikipedia  Finnish  Latn  fi  5
+Czech Wikipedia  Czech  Latn  cs  5
+Hungarian Wikipedia  Hungarian  Latn  hu  5
+Turkish Wikipedia  Turkish  Latn  tr  5
+Romanian Wikipedia  Romanian  Latn  ro  5
+Swahili Wikipedia  Swahili  English  sw  5
+Korean Wikipedia  Korean  Hang  ko  5
+Kazakh Wikipedia  Kazakh  Cyrl / Latin / Arab  kk  5
+Vietnamese Wikipedia  Vietnamese  Latn  vi  5
+Danish Wikipedia  Danish  Latn  da  5
+Esperanto Wikipedia  Esperanto  Latn  eo  5
+Serbian Wikipedia  Serbian  Cyrl/Latn  sr  5
+Indonesian Wikipedia  Indonesian  Latn  id  5
+Lithuanian Wikipedia  Lithuanian  Latn  lt  5
[…]

### 2014-01-17 sugarlike/extractfeature.py
added "all" features to sentence2ngrams()

@@ -35,6 +35,9 @@ def sentence2ngrams(text,n=3, option='char', with_word_boundary=False):
     return list(chain(*[word2ngrams(i, n, option, with_word_boundary) \
                         for i in text.split()]))
 
+  if "all" in option:
+    return list(chain(*[sentence2ngrams(text,n=i) for i in range(1,5)]))
+
 
 def extract_feature_from_datasource(data_source, outputpath):
   """ Returns a Counter object with the ngrams/word counts. """
@@ -92,9 +95,8 @@ def crubadan2counters(crubadanfile='crub-131119.zip'):
     for infile in inzipfile.namelist():
       path, filename = os.path.split(infile)
       lang = filename.rpartition('.')[0]
+      if '-' in lang: lang = lang.rpartition('-')[0]
       if lang in ISO2LANG:
-        if lang.startswith('fr'):
-          print(lang)
         print ('crubadan', infile, lang, 'Creating feature sets, please be patient...')
         for line in inzipfile.open(infile):
           key, count = line.strip().split(' ')
@@ -107,7 +109,6 @@ def crubadan2counters(crubadanfile='crub-131119.zip'):
         datalost.add(infile)
   return charngrams, wordfreqs, datalost
 
-crubadan2counters()
 
 def feature_interface(data_source):
   """ Unpickle the feature pickles if exists else create them. """

### 2014-01-17 universalcorpus/miniethnologue.py
fixed bugs in wikicode2iso()

@@ -8861,7 +8861,7 @@ def langiso(languagename):
   return getKey(ISO2LANG, languagename)
 
 def wikicode2iso(wikicode):
-  return langiso({i.split()[4]:i.split()[0] \
+  return langiso({i.split('  ')[3]:i.split('  ')[1].lower() \
                   for i in listofwikis.split('\n')}[wikicode].lower())
 
 def macrolang():

### 2014-01-17 universalcorpus/miniethnologue.py
fixed the space bugs in wikicode2iso

@@ -3,6 +3,7 @@
 import codecs
 from collections import defaultdict
 
+# from http://www-01.sil.org/iso639-3/download.asp
 iso6963 = """Id  Print_Name  Inverted_Name
 aaa  Ghotuo  Ghotuo
 aab  Alumu-Tesu  Alumu-Tesu

### 2014-01-17 sugarlike/evaluate.py
added evaluate.py

@@ -1,0 +1,19 @@
+# -*- coding: utf-8 -*-
+
+
+def tenfold(data_source, randseed=0):
+  import random, sys
+  from extractfeature import get_features, sentence2ngrams
+  from universalcorpus import odin, omniglot, udhr
+  random.seed(randseed)
+  corpus = list(locals()[data_source].source_sents())
+  corpus = sorted(iter(corpus), key=lambda k: random.random())
+  totrain = int(len(corpus) * 90 / 100)
+  yield corpus[:totrain] # train data
+  yield corpus[totrain:] # test data
+  
+import sys  
+for i in tenfold('odin'):
+  for j in i:
+    print j
+  

### 2014-01-17 sugarlike/universalcorpus_api.py
fixed typo in final example

@@ -21,12 +21,11 @@ for lang, source, translation in omniglot.phrases():
 for lang, sent in omniglot.source_sents():
   print lang, sent
 
-
 # Example to access UDHR corpus by documents.
 for lang, doc in udhr.documents():
   print lang, doc
 
-# Example to acces UDHR corpy by sentences.
-for lang, sent in udhr.documents():
+# Example to acces UDHR corpus by sentences.
+for lang, sent in udhr.source_sents():
   print lang, sent
   

### 2014-01-18 sugarlike/extractfeature.py
added LANG2ISO for miniethnlogue.py fixed extractfeature such that it throws nothing from crubadan.

@@ -58,7 +58,7 @@ def extract_feature_from_datasource(data_source, outputpath):
     for lang, sent in locals()[data_source].source_sents():
       print (data_source, lang, 'Creating feature sets, please be patient...')
       ##print (sent)
-      if lang in ISO2LANG:
+      if lang in ISO2LANG or lang in MACRO2LANG:
         for n in range(1,5): # Generates 1-5character ngrams.
           charngrams[lang]+= Counter(sentence2ngrams(sent, n, 'char', True))
         wordfreqs[lang]+=Counter(sent.split())
@@ -80,8 +80,8 @@ def crubadan2counters(crubadanfile='crub-131119.zip'):
   Returns the character ngrams, word frequences and list of files that are 
   in crubadan but not listed in the ISO code.
   """
-  import os, zipfile
-  from universalcorpus.miniethnologue import ISO2LANG
+  import os, zipfile, time
+  from universalcorpus.miniethnologue import ISO2LANG, MACRO2LANG, LANG2ISO
   from collections import defaultdict, Counter
   crubadanfile = os.path.dirname(__file__) + \
                      '/../universalcorpus/data/crubadan/' + crubadanfile
@@ -95,7 +95,14 @@ def crubadan2counters(crubadanfile='crub-131119.zip'):
     for infile in inzipfile.namelist():
       path, filename = os.path.split(infile)
       lang = filename.rpartition('.')[0]
-      if '-' in lang: lang = lang.rpartition('-')[0]
+      if '-' in lang: lang = lang.partition('-')[0]
+      if not lang: continue
+      ##start = time.time()
+      if lang not in ISO2LANG:
+        print(lang, LANG2ISO[lang])
+        if LANG2ISO[lang][0] in ISO2LANG: 
+          lang = LANG2ISO[lang][0] 
+      ##print (time.time() -start)
       if lang in ISO2LANG:
         print ('crubadan', infile, lang, 'Creating feature sets, please be patient...')
         for line in inzipfile.open(infile):
@@ -108,7 +115,6 @@ def crubadan2counters(crubadanfile='crub-131119.zip'):
       else:
         datalost.add(infile)
   return charngrams, wordfreqs, datalost
-
 
 def feature_interface(data_source):
   """ Unpickle the feature pickles if exists else create them. """

### 2014-01-18 universalcorpus/crawlandclean/wikipedia.py
added source_sents() to universalcorpus/crawlandclean/wikipedia.py

@@ -123,3 +123,26 @@ def clean_wikipedia(wiki_raw_dir):
 
 #extract_wikipedia('/media/ec609cb5-510c-467e-9655-5e72e99c4153/wikidumps/')
 clean_wikipedia('../data/wikipedia/texts/')
+
+def source_sent(cleanedwikidir):
+  """
+  USAGE:
+  >>> cleanwiki = '/media/alvas/E418A6B618A686E0/xling/cleanedwiki/'
+  >>> for i in source_sent(cleanwiki):
+  >>>   print i
+  
+  NOTE:
+  cleanwiki should be a main directory that contains one directory for each
+  language. And every language directory should contain at least one tarballs. 
+  Regardless of how many files each tarballs contain, it extracts the lines.
+  
+  P/S: I know the nest directory pathing is ugly, but i can't find a simpler
+  way to do this =) 
+  """
+  from utils import read_tarfile
+  for lang in os.listdir(cleanedwikidir):
+    for intarfile in os.listdir(cleanedwikidir+lang):
+      for infile in read_tarfile(cleanedwikidir+lang+"/"+intarfile):
+        with codecs.open(infile,'r','utf8') as fin:
+          for line in fin:
+            yield line.strip()

### 2014-01-18 sugarlike/extractfeature.py
added 'wiki' option to get_features()

@@ -54,7 +54,7 @@ def extract_feature_from_datasource(data_source, outputpath):
   charngrams = defaultdict(Counter)
   wordfreqs = defaultdict(Counter)
 
-  if data_source in ['odin','omniglot','udhr']:
+  if data_source in ['odin','omniglot','udhr','wiki']:
     for lang, sent in locals()[data_source].source_sents():
       print (data_source, lang, 'Creating feature sets, please be patient...')
       ##print (sent)

### 2014-01-18 docs/The Adventures of Sugali.tex
pardon the language =)

\documentclass[11pt]{article}

\title{\textbf{SuGaLi: A Quest to Identify the Languages of the World}}
\author{Susanne Fertmann, Guy Emerson, Liling Tan}
\date{}

\usepackage{hyperref}

\begin{document}

\maketitle

\section{Prologue}

\emph{``And they said, Go to, let us build us a corpus and a software, whose bytes may reach unto heaven; and let us assign a tag for each language, lest we be scattered abroad upon the encodings and data sparsity issues of the whole earth..."} - The Tower of Sugarlike. \\

\noindent It started with the ambition to create a software in the name of Natural Language Processing for Low Resource Language. Three brave scientists pit their wits against the world to show that they can create a software that can identify any language written in the mortal world.

Little did they know that the task perils of bytes that stood before them. Little had they realized the growth of knowledge and potential of what they had created. 
\\

\noindent

\newpage
\section{Bytes by Bytes: Building the Universal Corpus}

\emph{Ashes to ashes, dust to dust, \\
unlike my body, my corpus can't rust. \\
Bytes by bytes, the corpus swells, \\
crawling and cleaning, thinking all is well. \\}
- Songs of Corpora Minstrel \\

\noindent Desperate for data, there was none. The poor scientists had to synthesize one. Searching high and low, the scientists surfed, being extra careful not stepping on copyrights turf.

\section{The Phantom Menace: Encodings}
\emph{``Behold, the encoding is one, and they have all one unicode; and this they begin to do: and now nothing will be restrained from them, which they have imagined to do. Go to, let us go down, and there confound their encodings, that they may not understand one another's text."} - The Tower of Sugarlike.

\section{The Menace Phantom: Language Codes}

\emph{``Behold, the language code is one, and they have all one ISO code; and this they begin to do: and now nothing will be restrained from them, which they have imagined to do. Go to, let us go down, and there confound their ISO codes, that they may not differentiate one another's text."} - The Tower of Sugarlike.

\
\\
\
\\

\bibliography{literature}{}
\bibliographystyle{plain}

\end{document}

### 2014-01-18 docs/The Adventures of Sugali.tex
pardon the language ;P

@@
 
-\emph{``Behold, the language code is one, and they have all one ISO code; and this they begin to do: and now nothing will be restrained from them, which they have imagined to do.
-Go to, let us go down, and there confound their ISO codes, that they may not differentiate one another's text."} - The Tower of Sugarlike.
+\emph{``Behold, the language code is one, and they have all one ISO; and this they begin to do: and now nothing will be restrained from them, which they have imagined to do.
+Go to, let us go down, and there confound their ISOs, that they may not differentiate one another's text."} - The Tower of Sugarlike.
 

### 2014-01-18 sugarlike/extractfeature.py
added tfidf to langid.py

@@ -99,7 +99,7 @@ def crubadan2counters(crubadanfile='crub-131119.zip'):
       if not lang: continue
       ##start = time.time()
       if lang not in ISO2LANG:
-        print(lang, LANG2ISO[lang])
+        print('converting', lang,  'to', LANG2ISO[lang],'...')
         if LANG2ISO[lang][0] in ISO2LANG: 
           lang = LANG2ISO[lang][0] 
       ##print (time.time() -start)

### 2014-01-18 sugarlike/langid.py

@@ -5,54 +5,95 @@ from itertools import chain
 import numpy as np
 from collections import Counter
 
-def features2numpy(data_source, option="3gram"):
-  featureset = get_features(data_source, option=option)
+def features2numpy(data_source, option="3gram", tfidf=False):
+  if tfidf:
+    featureset = tfidfize(data_source, option=option)
+  else:
+    featureset = get_features(data_source, option=option)
+  
   all_features = list(set(chain(*[i.keys() for i in featureset.values()])))
   all_tags = [i for i in featureset]
   data, target = [], []
   for lang in featureset:
-    data.append([featureset[lang][j] for j in all_features])
+    data.append([featureset[lang][j] if j in featureset[lang] else 0 for j in all_features])
     target.append(lang)
     # Sanity check
     ##print [(j,featureset[lang][j]) for j in all_features if featureset[lang][j] > 0]
   return np.array(data), np.array(target), all_features
+
+def tfidfize(data_source, option='3gram'):
+  # see http://timtrueman.com/a-quick-foray-into-linear-algebra-and-python-tf-idf/
+  # see http://scikit-learn.org/stable/modules/preprocessing.html
+  from collections import defaultdict
+  import math, os, io
+  import cPickle as pickle
+
+  tfidf_pickle = ''.join([data_source,'-',option,'-tfidf','.pk'])
+
+  if os.path.exists(tfidf_pickle):
+    with io.open(tfidf_pickle,'rb') as fin:
+      featureset = pickle.load(fin)
+  else:
+    featureset = defaultdict(dict)
+    _featureset = get_features(data_source, option=option)
+    
+    for lang in _featureset:
+      for gram in _featureset[lang]:
+        tf = _featureset[lang][gram] / float(sum(_featureset[lang].values()))
+        idf = math.log(len(_featureset)) / len([i for i in _featureset if gram in _featureset[i]])
+        featureset[lang][gram] = tf * idf
+        print 'Calculating TF-IDF for %s please wait patiently...' % data_source
+        print lang, gram, _featureset[lang][gram], tf, idf, tf * idf
+        
+    with io.open(tfidf_pickle,'wb') as fout:
+      pickle.dump(featureset, fout)
+    
+  return featureset
 
[…]

### 2014-01-19 universalcorpus/test/test_miniethnologue.py
Add test for wikicode2iso.

@@ -1,0 +1,22 @@
+# -*- coding: utf-8 -*-
+
+
+import sys; sys.path.append('../')
+
+import unittest
+from miniethnologue import *
+
+
+
+class TestMiniethnologue(unittest.TestCase):
+  wikicode = {'en':'eng', 'de':'deu', 'zh-min-nan':'nan', 'xyz':None, 'war':'war'} 
+
+
+  def test_wiki2iso(self):
+    for wikicode in self.wikicode.keys():
+      self.assertEquals(self.wikicode[wikicode], wikicode2iso(wikicode))
+    
+
+
+if __name__ == "__main__":
+  unittest.main()

### 2014-01-19 sugarlike/extractfeature.py
added allgrams to sentence2ngrams option

@@ -30,12 +30,12 @@ def sentence2ngrams(text,n=3, option='char', with_word_boundary=False):
     from nltk.util import ngrams
     return list(ngrams(text.split(), 1))
   
-  if "gram" in option:
+  if "gram" in option and "all" not in option:
     n = int(option[0])
     return list(chain(*[word2ngrams(i, n, option, with_word_boundary) \
                         for i in text.split()]))
 
-  if "all" in option:
+  if "allgrams" in option:
     return list(chain(*[sentence2ngrams(text,n=i) for i in range(1,5)]))
 
 

### 2014-01-19 sugarlike/langid.py

@@ -119,9 +119,9 @@ t = 'ich bin schwanger'
 #print sugarlid_gnb(t, option='3gram')[:10]
 #print sugarlid_bnb(t, option='3gram')[:10]
 #print sugarlid_cosine(t, option='char', data_source='odin')[:10]
-x = sugarlid_mnb(t, option='3gram', data_source='udhr', smoothing=0.00001, tfidf=True)[:10]
+x = sugarlid_mnb(t, option='3gram', data_source='crubadan', smoothing=0.00001, tfidf=True)[:10]
 for i in x:
-  print i[0], ISO2LANG[i[1]] 
+  print i[0], i[1], ISO2LANG[i[1]] 
 
 
 #print sugarlid_cosine(t, option='2gram')[:10]

### 2014-01-19 sugarlike/extractfeature.py
removed informal test

@@ -27,17 +27,15 @@ def sentence2ngrams(text,n=3, option='char', with_word_boundary=False):
     return list(chain(*[word2ngrams(i, n, option, with_word_boundary) \
                         for i in text.split()]))
   if option == 'word':
-    from nltk.util import ngrams
-    return list(ngrams(text.split(), 1))
+    return text.split()
   
   if "gram" in option and "all" not in option:
     n = int(option[0])
     return list(chain(*[word2ngrams(i, n, option, with_word_boundary) \
                         for i in text.split()]))
 
-  if "allgrams" in option:
+  if "allgrams" in option or 'all' in option:
     return list(chain(*[sentence2ngrams(text,n=i) for i in range(1,5)]))
-
 
 def extract_feature_from_datasource(data_source, outputpath):
   """ Returns a Counter object with the ngrams/word counts. """
@@ -156,6 +154,14 @@ def get_features(data_source, language=None, option='char', \
   return result if result else print('%s does not have %s features' \
                                      % (data_source, language))
 
+
+'''
+import io
+import cPickle as pickle
+x  = pickle.load(io.open('odin-word.pk','rb'))
+for i in x:
+  print (i,x[i])
+'''
 
 '''
 #Informal Test:

### 2014-01-19 sugarlike/extractfeature.py
added lower feature to crubadan2counters

@@ -73,7 +73,7 @@ def extract_feature_from_datasource(data_source, outputpath):
         
   return charngrams, wordfreqs, datalost
   
-def crubadan2counters(crubadanfile='crub-131119.zip'):
+def crubadan2counters(crubadanfile='crub-131119.zip', lower=False):
   """ 
   Returns the character ngrams, word frequences and list of files that are 
   in crubadan but not listed in the ISO code.
@@ -104,6 +104,7 @@ def crubadan2counters(crubadanfile='crub-131119.zip'):
       if lang in ISO2LANG:
         print ('crubadan', infile, lang, 'Creating feature sets, please be patient...')
         for line in inzipfile.open(infile):
+          if lower: key = key.lower()
           key, count = line.strip().split(' ')
           
           if 'words' in path: # Updates wordfreq

### 2014-01-19 sugarlike/extractfeature.py
added tfidf feature to get_features()

@@ -131,10 +131,37 @@ def feature_interface(data_source):
   
   return charngrams, wordfreqs
 
+def tfidfize(_featureset, data_source, option):
+  """ Normalized the feature with TF-IDF."""
+  from collections import defaultdict
+  import math, os, io
+  import cPickle as pickle
+  
+  tfidf_pickle = ''.join([data_source,'-',option,'-tfidf','.pk'])
+  
+  if os.path.exists(tfidf_pickle): # If feature already tfidfized before.
+    with io.open(tfidf_pickle,'rb') as fin:
+      fs = pickle.load(fin)
+  else:
+    fs = defaultdict(dict) 
+    for lang in _featureset:
+      for gram in _featureset[lang]:
+        tf = _featureset[lang][gram] / float(sum(_featureset[lang].values()))
+        idf = math.log(len(_featureset)) / len([i for i in _featureset if gram in _featureset[i]])
+        fs[lang][gram] = tf * idf
+        print('Calculating TF-IDF for %s please wait patiently...' % data_source)
+        print (lang, gram, _featureset[lang][gram], tf, idf, tf*idf)
+        
+    with io.open(tfidf_pickle,'wb') as fout:
+      pickle.dump(fs, fout)
+      
+  return fs
+    
 def get_features(data_source, language=None, option='char', \
-                 with_word_boundary=True):  
+                 with_word_boundary=True, tfidf=False):  
   """ Get features given the data_source, language and option"""
-  charngs, wordfqs = feature_interface(data_source)  
+  charngs, wordfqs = feature_interface(data_source)
+  
   if option == 'char':
     result = charngs[language] if language else charngs
   elif option == 'word':
@@ -149,23 +176,23 @@ def get_features(data_source, language=None, option='char', \
   if option == None:
     return charngs, wordfqs
   
-  if not with_word_boundary: #TODO: 
+  if tfidf:
+    result = tfidfize(result,data_source, option)
+  
+  if not with_word_boundary: #TODO: to filter features with "<" or ">" 
     pass
   
   return result if result else print('%s does not have %s features' \
                                      % (data_source, language))
 
-
-'''
-import io
-import cPickle as pickle
-x  = pickle.load(io.open('odin-word.pk','rb'))
-for i in x:
-  print (i,x[i])
-'''
-
 '''
 #Informal Test:
+
+x = get_features('odin',option='char', tfidf=True) # get_features() with tfidf
+for i in x:
[…]

### 2014-02-28 docs/eacl-2014-styles/uniwac2.tex
Universal Corpus writeup for ComputEL workshop

\documentclass[11pt]{article}
\usepackage{eacl2014}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{color}
\usepackage[hidelinks]{hyperref}
\usepackage{enumitem}

\usepackage{listings}
\usepackage{fancyvrb}

\usepackage{color}
\definecolor{gray}{rgb}{0.4,0.4,0.4}
\definecolor{darkblue}{rgb}{0.0,0.0,0.6}
\definecolor{cyan}{rgb}{0.0,0.6,0.6}
\definecolor{orange}{rgb}{1,0.5,0}

\lstset{
  basicstyle=\ttfamily,
  columns=fullflexible,
  showstringspaces=false,
  commentstyle=\color{gray}\upshape
}

\lstdefinelanguage{XML}
{
  morestring=[b]",
  morestring=[s]{>}{<},
  morecomment=[s]{<?}{?>},
  stringstyle=\color{black},
  identifierstyle=\color{darkblue},
  keywordstyle=\color{cyan},
  morekeywords={xmlns,version,type}% list your attributes here
}

\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

\title{uniWaC: Compiling the Foundation Text for the WaCky Universal Corpus}

\author{Susanne Fertmann\textsuperscript{$\alpha$}, Guy Emerson\textsuperscript{$\beta$}, Liling Tan\textsuperscript{$\alpha$}, Alexis Palmer\textsuperscript{$\alpha$} and Michaela Regneri\textsuperscript{$\alpha$} \\
  Universit{\"a}t des Saarlandes\textsuperscript{$\alpha$} / Campus, 66123 Saarbrücken, Germany \\
  Deutsches Forschungszentrum f{\"u}r Künstliche Intelligenz (DFKI)\textsuperscript{$\beta$} / \\ Stuhlsatzenhausweg 3, 66123 Saarbrücken, Germany \\
  {\tt s9sufert@stud.uni-saarland.de, emerson@coli.uni-saarland.de,} \\
  {\tt liling.tan@uni-saarland.de}}

\date{}

\begin{document}
\maketitle
\begin{abstract}
A grand aim of linguistics is to seek a universal theory of human language and the opposing relativist view of linguistics seek to prove otherwise. To accomplish either goal, the compilation of a Universal Corpus with significant data from a large variety of languages is necessary. The Web contains vast amounts of linguistic data however it is largely saturated by English content (\textgreater50\%). Nonetheless, there are websites that contains small amount of data in a large number of languages. To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contains linguistic data from multiple languages and merge them with existing corpora collections.This paper describes the progress of compiling, cleaning and standardizing the foundation texts for the Universal Corpus (\texttt{uniWaC}) and briefly describes the ongoing work to merge our compilation with existing corpora collections.
\end{abstract}

\section{Introduction}
Abney and Bird (2010) posed a grand challenge to build a Universal Corpus that will include as many of the world's languages as possible, in a consistent structure that permits crosslingual processing and analysis. Ideally, the Universal Corpus would be a complete digitization of every human language. To empirically prove or disprove the existence of a universal theory of human languages, a Universal Corpus is necessary.

Currently multilingual corpora efforts have limited coverage in number of languages and the number of language families the corpora represents; for instance, the OPUS corpus covers over 90 languages (Tiedemann, 2012), i.e. 1.27\% of the total number of languages in the world; Leipzig Corpora
Collection\footnote{corpora.uni-leipzig.de} contains corpora in 230 languages, 3.24\% of 7105 living languages (Biemann et al. 2007). Even corpora that boast of linguistic diversity lack in language families’ coverage; e.g. the linguistically diverse NTU-Multilingual Corpus covers only 7 out of 136 language families (Tan and Bond, 2011). The quintessential solution towards a Universal Corpus is to merge all existing corpora and provide an ubiquitous access interface. To compile the foundation text for the Universal Corpus (uniWaC), we crawled and cleaned web data that contains multilingual texts and merge them with existing corpora collections to form the foundation text for the Universal Corpus.

\subsection{WaCky Corpora}

Baroni and Bernardini (2004) introduced the BootCaT toolkit to bootstrap specialized corpora and terms from the web. Then, Sharoff (2006) built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus. Later, Ferraresi (2007) introduced and discussed the notion of compiling Web as Corpus \emph{"properly"} with the inaugural ukWaC using web-crawled data from a list of seed words. Thereafter, the Crúbadán corpora was created for a large number of under-resourced languages (Scannell, 2007). Following which the deWaC (German), itWaC (Italian) and frWaC (French) were compiled (Baroni et al. 2008).

Subsequently, Brunello (2009) stressed the notion of building free corpora from the web using documents released under Creative Commons licenses. To build better quality WaCky corpora, Versley and Panchenko (2012) introduced the notion content-sensitive boilerplate detection in cleaning web corpora.

The initial wave of monolingual WaCky corpora (Baroni et al. 2008) inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus (Dickinson et al. 2013) and the Feed Corpus (Minocha et al. 2013). Meanwhile, Web-driven corpora compilations also focused on building large  multilingual resources, e.g. Leipzig Corpora Collection (\texttt{LCC}) (Biemann et al. 2007), the COrpora from the Web (\texttt{COW}) project (Schaefer and Bildhauer, 2012).

Different from previous WaCs built under the Web as Corpus kool ynitiative (WaCky), we did not perform seed query web-searching or web crawling to achieve a balanced resource. Instead the foundation text for uniWaC was built using dedicated site crawling and/or site-dependent cleaning from four data sources, viz. (i) \emph{Online Database of Interlinear Text} (ODIN), (ii) \emph{Omniglot} website, (iii) \emph{Universal Declaration of Human Rights} (UDHR) and (iv) \emph{Wikipedia} dumps. Our emphasis for the uniWaC is based on based on \textbf{\emph{universality}}; covering as many languages as possible is the first priority.

\section{Crawling and Cleaning}

Although data size matters in general NLP (Banko and Brill, 2001), \emph{universality} is the utmost priority in NLP for the Universal Corpus initiative (aka Human Language Project; Abney and Bird, 2010; 2011). Hence, the uniWaC approach to data collection is based on focused/dedicated crawling unlike previous WaC word/URL list seeding. We manually selected sites that contains texts from a large variety of languages and wrote scripts to crawl and clean specific pages from the different data sources. The crawling and cleaning was done simultaneously.

The resulting files were saved as plaintext file with \texttt{UTF-8} encoding. One directory for each data source and one file per language. The parallel data (from ODIN and Omniglot) are saved as tab-delimited files, one line per sentence, one column per language. The monolingual data (from UDHR and Wikipedia) are saved as newline delimited plaintext file, each line denoting a paragraph or sentence and each document separated by an empty line. The files are saved with the following naming convention to allow easy access, the filename contains the data source and its ISO 639-3 code, e.g. \texttt{odin-iii.txt} contains the \emph{ODIN} data for the \emph{Nuosu} language.

\subsection{ODIN}

The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinerized Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags. Each XML file is saved under a filename that matches the respective language code. While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.

\begin{minipage}{\columnwidth}

a. \quad o lesu mai \\
\indent \qquad 2sg return here \\
\indent \qquad `\emph{You return here.}' \\

\centerline{Figure 1: IGT adhering to Leipzig Glossing Rule.}
\end{minipage}

\begin{minipage}{\columnwidth}
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>a. o lesu mai</line>
    <line>2sg return here</line>
    <line>`You return here.'</line>
  </example>
</igit>
\end{lstlisting} 
\centerline{Figure 2: An IGT in ODIN's XML format.}
\end{minipage}
\\ \\
\noindent The IGTs from a web document usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss. From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2. Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
\begin{minipage}{\columnwidth}
\begin{lstlisting}[language=XML]
<igt>
  <example>
    <line>(69) na-Na-tmi-kwalca-t 
    (Foley 1991)</line>
    <line>3sgA-1sgO-say-rise-PERF
    </line>
    <line>`She woke me up' 
    (by verbal action)</line>
  </example>
</igit>
\end{lstlisting} 
\smallskip
\centerline{Figure 3: Another IGT in ODIN's XML format.}
\end{minipage}
\\ \\
\noindent The primary problem in extracting the source text is lack of consistency of the IGTs from the ODIN data. For instance, Figure 2 uses an alphabet bullet convention, i.e. `\emph{a.}', whereas the IGT in Figure 3 uses a bracketed number indexing convention, i.e. `\emph{(69)}'. In addition, the source line in Figure 3 included the citation to the example, which would be considered as noise to corpus.

Lewis and Xia (2010) noted that it is not uncommon for IGT found in documents to deviant from one uniform convention; when compiling the ODIN data, they attempted a regex based approached and reported 59\% F-score. Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness':
\\ \\
\begin{minipage}{\columnwidth}
\begin{itemize}
\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
\begin{itemize}
\item[(i)]
\begin{Verbatim}
^(?\s?\w{1,5}\s*[):.]\s*
\end{Verbatim}
\item[(ii)] 
\begin{Verbatim}
[\[\(].{1,}[\]\)]
\end{Verbatim}
\end{itemize}
\item \emph{Cleanest}: Accepts only source lines without any punctuation.
\end{itemize}
\end{minipage}
\\ \\
\noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages. The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 693-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.

\subsection{Omniglot}

The Omniglot crawling and cleaning required more tact and after manual inspection of the www.omniglot.com website, only the \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} pages were consistent in their HTML markups that allows easy cleaning. We wrote a crawler script that downloads pages with that contains the following in their URLs:

\begin{itemize}[noitemsep]
\item \url{www.omniglot.com/language/phrases/*} 
\item \url{www.omniglot.com/babel/*}
\end{itemize}

\noindent The \emph{`Useful foreign phrases'} page contains parallel phrases in embedded within the \texttt{<th><tr>...</tr><th>} tags. The English phrase and foreign language phrase are encoded separately with \texttt{<td>...</td>} tags. Figure 4 shows an instance of an Omniglot \emph{Useful foreign phrase} page. Since the HTML page was designed as WYSIWYG, several pages have non-textual elements within the \texttt{<tr>...</tr>} tags for aesthetics. For example, the non-breaking space, \texttt{<td>\&nbsp;</td>} in Fig. 4; the extracted text was properly converted into plaintext format using the HTMLParser\footnote{http://docs.python.org/2/library/htmlparser.html} module, during which the non textual elements would have been cleaned.

The same process of crawling URLs and cleaning was performed on the \emph{`Tower of Babel'} pages and the only difference was that the texts were embedded in a \texttt{<ol><li>...</li></ol>} tags instead.

\begin{minipage}{\columnwidth}
\begin{lstlisting}[language=XML]
<th>
  ...
  <tr>
    <td>I don't understand</td>
    <td>No appo cumpresu nutta</td>
    <td>&nbsp;</td>
  </tr>
  <tr>
    <td>I don't understand</td>
    <td>Non d'isco</td>
    <td>&nbsp;</td>
  </tr>
  ...
<th>
\end{lstlisting} 
\centerline{Figure 4: A \emph{Useful Foreign Phrase} page}
\end{minipage}
\smallskip

\noindent The primary problem of the Omniglot data is that to the only information about the language of the data on the page is given in the title of the HTML and its URL. Additionally, only the name of the language was given not the ISO 639-3 code. Even though Omniglot pages contain language meta data in XHTML markup attributes, most pages are encoded with the \texttt{lang="en"} attribute because the pages are catered for English speaking audience and the boilerplates (e.g. navigation bar, site banner and footer) on the pages are in English. This causes a problem in file storage and access since it is necessary to save the files in the naming convention, e.g. \texttt{omniglot-smo.txt} but the language code is not found any where on the page. To resolve the problem of standardizing, we have written a conversion script using the code to language mappings from \url{http://www-01.sil.org/iso639-3/iso-639-3.tab}

\subsection{Universal Declaration of Human Rights}

Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format. Originally, we tried to resolve the encoding problems by automatically identifying the encoding and decoding the textfiles using the \texttt{libmagic} library\footnote{http://linux.die.net/man/3/libmagic} and then re-encoding them to \texttt{UTF-8} whenever possible.

However the better solution was found when we chanced upon the UDHR site\footnote{http://unicode.org/udhr/d/} from the \url{unicode.org} domain. The UDHR textfiles found on the site was free of encoding problem but there was a boilerplate at the start of each file which records it metadata. It was simply clean by skipping the first four lines of each file.

Although the redistribution of previous compiled data is not scientifically attractive but redistribution/recompilation of data to improve the usability and quality of the data is worth mentioning. For instance, the SETimes corpus first compiled by Tyers and Alperen (2010) was cleaned and redistributed by Ljubesic and Agic (2013).

\subsection{Wikipedia Dumps}

Tapping on the crowd-sourced Wikipedia articles for NLP forms a crucial part of developing data-driven NLP tools and application ({\color{red}{citation neeeded}}). To automatically download the Wikipedia dumps, we have scripted the \texttt{wget} with ISO 639-2 language code as a variable (\texttt{\$I}) in a shell script, as such:

\smallskip
\noindent \texttt{\$ wget http://download.wikimedia.
org/\$Iwiki/latest/itwiki-latest-
pages-articles.xml.bz2}
\medskip

\noindent One major issue with using the Wikipedia dump is the sheer size and extracting text from a glut of Wikipedia markup. To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.

After converting them into textfiles, we used the following regexes to delete residual Wikipedia markups and magic words\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.

\section{Access and Copyrights}

\section{Representation and Universality}
This is we report the counts and size of corpus.

\section{Ongoing Work}

Currently, the uniWaC project is continually adding more data from the LCC Leipzig Corpora. Finally, the first phase of the uniWaC will end with the implementation of unsupervised tokenizer (e.g. Chung and Gildea, 2009) such that initial experiments on crosslingual analysis can be carried out to advance the grand aim of a unified linguistic theory.

% TODO: use proper bibtex !!!!
\section*{References}
Steven Abney and Steven Bird. 2010. The Human Language Project: Building a Universal Corpus of the World’s Languages. In Proceedings of the 48th Annual Meeting of the Association for Computational Linguistics. Uppsala, Sweden.
\\\\
Steven Abney and Steven Bird. 2011. Towards a data model for the Universal Corpus. In Proceedings of the 4th Workshop on Building and Using Comparable Corpora: Comparable Corpora and the Web. Portland, Oregon.
\\\\
Michele Banko and Eric Brill. 2001. Michele Banko and Eric Brill. 2001. Scaling to very very large corpora for natural language disambiguation. In Proceedings of the 39th Annual Meeting on Association for Computational Linguistics (ACL '01). Association for Computational Linguistics, Stroudsburg, PA, USA, 26-33.
\\\\
Marco Baroni, Silvia Bernardini, Adriano Ferraresi and Eros Zanchetta. (2008). The WaCky wide web: A collection of very large linguistically processed web-crawled corpora. Language Resources and Evaluation, 43(3), 209–231.
\\\\
Tagyoung Chung and Daniel Gildea. 2009. Unsupervised tokenization for machine translation. In Proceedings of the 2009 Conference on Empirical Methods in Natural Language Processing: Volume 2 - Volume 2 (EMNLP '09), Vol. 2. Association for Computational Linguistics, Stroudsburg, PA, USA, 718-726.
\\\\
Adriano Ferraresi. 2007. Building a very large corpus of English obtained by Web crawling: ukWaC. Master Thesis, University of Bologna
\\\\
Jörg Tiedemann. 2012. Parallel Data, Tools and Interfaces in OPUS. In Proceedings of the 8th International Conference on Language Resources and Evaluation. Istanbul, Turkey.
\\\\
Liling Tan and Francis Bond. 2011. Building and Annotating the Linguistically Diverse NTU-MC (NTU-Multilingual Corpus). In Proceedings of The 25th Pacific Asia Conference on Language, Information, and Computation. Singapore, Singapore.

\end{document}

### 2014-02-28 docs/eacl-2014-styles/eacl2014.tex
added eacl tex styles and bib

%
% File eacl2014.tex
%
% Contact g.bouma@rug.nl yannick.parmentier@univ-orleans.fr
%
% Based on the instruction file for ACL 2013 
% which in turns was based on the instruction files for previous 
% ACL and EACL conferences

%% Based on the instruction file for EACL 2006 by Eneko Agirre and Sergi Balari
%% and that of ACL 2008 by Joakim Nivre and Noah Smith

\documentclass[11pt]{article}
\usepackage{eacl2014}
\usepackage{times}
\usepackage{url}
\usepackage{latexsym}
\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to

\title{Instructions for EACL-2014 Proceedings}

\author{First Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  {\tt email@domain} \\\And
  Second Author \\
  Affiliation / Address line 1 \\
  Affiliation / Address line 2 \\
  Affiliation / Address line 3 \\
  {\tt email@domain} \\}

\date{}

\begin{document}
\maketitle
\begin{abstract}
  This document contains the instructions for preparing a camera-ready
  manuscript for the proceedings of EACL-2014. The document itself
  conforms to its own specifications, and is therefore an example of
  what your manuscript should look like. These instructions should be
  used for both papers submitted for review and for final versions of
  accepted papers.  Authors are asked to conform to all the directions
  reported in this document.
\end{abstract}

\section{Credits}

This document has been adapted from the instructions for earlier
(E)ACL-proceedings, including those for ACL-2013 by Roberto Navigli
and Jing-Shing Chang, those from ACL-2012 by Maggie Li and Michael
White, those from ACL-2010 by Jing-Shing Chang and Philipp Koehn,
those for ACL-2008 by Johanna D. Moore, Simone Teufel, James Allan,
and Sadaoki Furui, those for EACL-2006 by Eneko Agirre and Sergi
Balari, those for ACL-2005 by Hwee Tou Ng and Kemal Oflazer, those for
ACL-2002 by Eugene Charniak and Dekang Lin, and earlier ACL and EACL
formats. Those versions were written by several people, including John
Chen, Henry S. Thompson and Donald Walker. Additional elements were
taken from the formatting instructions of the {\em International Joint
  Conference on Artificial Intelligence}.

\section{Introduction}
% das: removed reference to PostScript

The following instructions are directed to authors of papers submitted
to EACL-2014 or accepted for publication in its proceedings. All
authors are required to adhere to these specifications. Authors are
required to provide a Portable Document Format (PDF) version of their
papers. \textbf{The proceedings will be printed on A4 paper}. Authors
from countries in which access to word-processing systems is limited
should contact the publication chairs Gosse Bouma ({\small {\tt
  g.bouma@rug.nl}}) and Yannick Parmentier ({\small {\tt
  yannick.parmentier@univ-orleans.fr}}) as soon as possible.

\section{General Instructions}

Manuscripts must be in two-column format.  Exceptions to the
two-column format include the title, authors' names and complete
addresses, which must be centered at the top of the first page, and
any full-width figures or tables (see the guidelines in
Subsection~\ref{ssec:fonts}). {\bf Type single-spaced.}  Start all
pages directly under the top margin. See the guidelines later
regarding formatting the first page (Subsection~\ref{ssec:first}).
The manuscript should be printed single-sided and its length should
not exceed the maximum page limit described in
Section~\ref{sec:length}.  {\bf Do not number the pages.}

\subsection{Electronically-available resources}

EACL-2014 provides this description in \LaTeX2e ({\small {\tt
    eacl2014.tex}}) and PDF format ({\small {\tt eacl2014.pdf}}),
along with the \LaTeX2e style file used to format it ({\small {\tt
    eacl2014.sty}}) and an ACL bibliography style ({\small {\tt
    acl.bst}}) in case you want to use the Bib\TeX\ reference management
software. These files are all available at
\url{http://www.eacl2014.org}. A Microsoft Word template file ({\small
  {\tt eacl2014.docx}}) is also available at the same URL. We strongly
recommend the use of these style files, which have been appropriately
tailored for the EACL-2014 proceedings. If you have an option, we
recommend that you use the \LaTeX2e version. \textbf{If you will be
  using the Microsoft Word template, we suggest that you anonymize
  your source file so that the pdf produced does not retain your
  identity.}  This can be done by removing any personal information
from your source document properties.

\subsection{Format of Electronic Manuscript}
\label{sect:pdf}

For the production of the electronic manuscript you must use Adobe's
Portable Document Format (PDF). 
%% Add instructions about using pdflatex
This format can be generated directly from \LaTeX2e files or from
postscript ones. On Linux/Unix-like systems, you can use {\tt
  pdflatex} to generate a PDF file from \LaTeX2e files, or {\tt
  ps2pdf} to convert from postscript to PDF. 
%%
In Microsoft Windows, you can use Adobe's Distiller, or if you have
\texttt{cygwin} installed, you can use \texttt{dvipdf} or
\texttt{ps2pdf}. Note that some word processing programs generate PDF
which may not include all the necessary fonts (esp. tree diagrams,
symbols). When you print or create the PDF file, there is usually an
option in your printer setup to include none, all or just non-standard
fonts.  Please make sure that you select the option of including ALL
the fonts. {\em Before sending it, test your PDF by printing it from a
  computer different from the one where it was created.} Moreover,
some word processors may generate very large postscript/PDF files,
where each page is rendered as an image. Such images may reproduce
poorly. In this case, try alternative ways to obtain the postscript
and/or PDF. One way on some systems is to install a driver for a
postscript printer, send your document to the printer specifying
``Output to a file'', then convert the file to PDF.
Please keep in mind that it is of utmost importance to use \textbf{A4 format}.

%%Removed by YP (use of the \special command above instead):
%% It is of utmost importance to specify the \textbf{A4 format} (21 cm x
%% 29.7 cm) when formatting the paper. 
%% When working with {\tt dvips}, for instance, one should specify {\tt -t a4}.

Print-outs of the PDF file on A4 paper should be identical to the
hardcopy version. If you cannot meet the above requirements about the
production of your electronic submission, please contact the
publication chairs above as soon as possible.

\subsection{Layout}
\label{ssec:layout}

Format manuscripts two columns to a page, in the manner these
instructions are formatted. The exact dimensions for a page on A4
paper are:

\begin{itemize}
\item Left and right margins: 2.5 cm
\item Top margin: 2.5 cm
\item Bottom margin: 2.5 cm
\item Column width: 7.7 cm
\item Column height: 24.7 cm
\item Gap between columns: 0.6 cm
\end{itemize}

\noindent Papers should not be submitted on any other paper size.
 If you cannot meet the above requirements about the production of your electronic submission, 
please contact the publication chairs above as soon as possible.

\subsection{Fonts}
\label{ssec:fonts}
For reasons of uniformity, Adobe's {\bf Times Roman} font should be
used. In \LaTeX2e{} this is accomplished by putting

\begin{quote}
\begin{verbatim}
\usepackage{times}
\usepackage{latexsym}
\end{verbatim}
\end{quote}
in the preamble. If Times Roman is unavailable, use {\bf Computer
  Modern Roman} (\LaTeX2e{}'s default).  Note that the latter is about
  10\% less dense than Adobe's Times Roman font.

\begin{table}[h]
\begin{center}
\begin{tabular}{|l|rl|}
\hline \bf Type of Text & \bf Font Size & \bf Style \\ \hline
paper title & 15 pt & bold \\
author names & 12 pt & bold \\
author affiliation & 12 pt & \\
the word ``Abstract'' & 12 pt & bold \\
section titles & 12 pt & bold \\
document text & 11 pt  &\\
captions & 11 pt & \\
abstract text & 10 pt & \\
bibliography & 10 pt & \\
footnotes & 9 pt & \\
\hline
\end{tabular}
\end{center}
\caption{\label{font-table} Font guide. }
\end{table}

\subsection{The First Page}
\label{ssec:first}

Center the title, author's name(s) and affiliation(s) across both
columns. Do not use footnotes for affiliations. Do not include the
paper ID number assigned during the submission process. Use the
two-column format only when you begin the abstract.

{\bf Title}: Place the title centered at the top of the first page, in
a 15-point bold font. (For a complete guide to font sizes and styles, 
see Table~\ref{font-table}) Long titles should be typed on two lines without
a blank line intervening. Approximately, put the title at 2.5 cm from
the top of the page, followed by a blank line, then the author's
names(s), and the affiliation on the following line. Do not use only
initials for given names (middle initials are allowed). Do not format surnames
in all capitals (e.g., use ``Schlangen'' not ``SCHLANGEN'').
Do not format title and section headings in all capitals as well
except for proper names (such as ``BLEU'') that are conventionally
in all capitals.
The affiliation should contain the author's complete address, and if
possible, an electronic mail address. Leave about 2 cm between the
affiliation and the body of the first page.
The title, author names and addresses should be completely
identical to those entered to the electronical paper submission
website in order to maintain the consistency of author information
among all publications of the conference.

{\bf Abstract}: Type the abstract at the beginning of the first
column. The width of the abstract text should be smaller than the
width of the columns for the text in the body of the paper by about
0.6 cm on each side. Center the word {\bf Abstract} in a 12 point bold
font above the body of the abstract. The abstract should be a concise
summary of the general thesis and conclusions of the paper. It should
be no longer than 200 words. The abstract text should be in 10 point font.

{\bf Text}: Begin typing the main body of the text immediately after
the abstract, observing the two-column format as shown in 
the present document. Do not include page numbers.

{\bf Indent} when starting a new paragraph. Use 11 points for text and 
subsection headings, 12 points for section headings and 15 points for
the title.

\subsection{Sections}

{\bf Headings}: Type and label section and subsection headings in the
style shown on the present document.  Use numbered sections (Arabic
numerals) in order to facilitate cross references. Number subsections
with the section number and the subsection number separated by a dot,
in Arabic numerals. Do not number subsubsections.

{\bf Citations}: Citations within the text appear
in parentheses as~\cite{Gusfield:97} or, if the author's name appears in
the text itself, as Gusfield~\shortcite{Gusfield:97}. 
Append lowercase letters to the year in cases of ambiguity.  
Treat double authors as in~\cite{Aho:72}, but write as
 in~\cite{Chandra:81} when more than two authors are involved. Collapse multiple citations as
in~\cite{Gusfield:97,Aho:72}. Also refrain from using full citations as sentence constituents. We
suggest that instead of
\begin{quote}
  ``\cite{Gusfield:97} showed that ...''
\end{quote}
you use
\begin{quote}
``Gusfield \shortcite{Gusfield:97}   showed that ...''
\end{quote}

If you are using the provided \LaTeX{} and Bib\TeX{} style files, you
can use the command \verb|\newcite| to get ``author (year)'' citations.

As reviewing will be double-blind, the submitted version of the papers should not include the
authors' names and affiliations. Furthermore, self-references that
reveal the author's identity, e.g.,
\begin{quote}
``We previously showed \cite{Gusfield:97} ...''  
\end{quote}
should be avoided. Instead, use citations such as 
\begin{quote}
``Gusfield \shortcite{Gusfield:97}
previously showed ... ''
\end{quote}

\textbf{Please do not  use anonymous citations} and  do not include acknowledgements 
when submitting your papers. Papers that do not conform
to these requirements may be rejected without review.

\textbf{References}: Gather the full set of references together under
the heading {\bf References}; place the section before any Appendices,
unless they contain references. Arrange the references alphabetically
by first author, rather than by order of occurrence in the text.
Provide as complete a citation as possible, using a consistent format,
such as the one for {\em Computational Linguistics\/} or the one in the 
{\em Publication Manual of the American 
Psychological Association\/}~\cite{APA:83}.  Use of full names for
authors rather than initials is preferred.  A list of abbreviations
for common computer science journals can be found in the ACM 
{\em Computing Reviews\/}~\cite{ACM:83}.

The \LaTeX{} and Bib\TeX{} style files provided roughly fit the
American Psychological Association format, allowing regular citations, 
short citations and multiple citations as described above.

{\bf Appendices}: Appendices, if any, directly follow the text and the
references (but see above).  Letter them in sequence and provide an
informative title: {\bf Appendix A. Title of Appendix}.

\textbf{Acknowledgement} section should appear in accepted manuscripts only. It should go as a last section immediately
before the references.  Do not number the acknowledgement section.

\subsection{Footnotes}

{\bf Footnotes}: Put footnotes at the bottom of the page and use 9
points text. They may be numbered or referred to by asterisks or other
symbols.\footnote{This is how a footnote should appear.} Footnotes
should be separated from the text by a line.\footnote{Note the line
separating the footnotes from the text.}

\subsection{Graphics}

{\bf Illustrations}: Place figures, tables, and photographs in the
paper near where they are first discussed, rather than at the end, if
possible.  Wide illustrations may run across both columns.  Color
illustrations are allowed, provided you have verified that  
they will be understandable when printed in black ink.

{\bf Captions}: Provide a caption for every illustration; number each one
sequentially in the form:  ``Figure 1. Caption of the Figure.'' ``Table 1.
Caption of the Table.''  Type the captions of the figures and 
tables below the body, using 11 point text.

\section{Translation of non-English Terms}

It is also advised to supplement non-English characters and terms
with appropriate transliterations and/or translations
since not all readers understand all such characters and terms.
Inline transliteration or translation can be represented in
the order of: original-form transliteration ``translation''.

\section{Length of Submission}
\label{sec:length}

Long papers may consist of up to 8 pages of content (excluding
references) and short papers may consist of up to 
four (4) pages plus 2 pages for references in the proceedings.  Papers that do not conform to the
specified length and formatting requirements are subject to be
rejected without review.

\section{Other Issues}

Papers that had software and/or dataset submitted for the review
process should also submit it with the camera-ready paper. Besides,
the software and/or dataset should not be anonymous.

Please note that the publications of EACL-2014 will be publicly
available at ACL Anthology (http://aclweb.org/anthology-new/) on April
19th, 2014, one week before the start of the conference. Since some of
the authors may have plans to file patents related to their papers in
the conference, we are reminding authors that April 19th, 2014 may be
considered to be the official publication date, instead of the opening
day of the conference.

\section*{Acknowledgments}

Do not number the acknowledgment section. Do not include this section
when submitting your paper for review.

% If you use BibTeX with a bib file named eacl2014.bib, 
% you should add the following two lines:
\bibliographystyle{acl}
\bibliography{eacl2014}

% Otherwise you can include your references as follows:
%% \begin{thebibliography}{}

%% \bibitem[\protect\citename{Aho and Ullman}1972]{Aho:72}
%% Alfred~V. Aho and Jeffrey~D. Ullman.
%% \newblock 1972.
%% \newblock {\em The Theory of Parsing, Translation and Compiling}, volume~1.
%% \newblock Prentice-{Hall}, Englewood Cliffs, NJ.

%% \bibitem[\protect\citename{{American Psychological Association}}1983]{APA:83}
%% {American Psychological Association}.
%% \newblock 1983.
%% \newblock {\em Publications Manual}.
%% \newblock American Psychological Association, Washington, DC.

%% \bibitem[\protect\citename{{Association for Computing Machinery}}1983]{ACM:83}
%% {Association for Computing Machinery}.
%% \newblock 1983.
%% \newblock {\em Computing Reviews}, 24(11):503--512.

%% \bibitem[\protect\citename{Chandra \bgroup et al.\egroup }1981]{Chandra:81}
%% Ashok~K. Chandra, Dexter~C. Kozen, and Larry~J. Stockmeyer.
%% \newblock 1981.
%% \newblock Alternation.
%% \newblock {\em Journal of the Association for Computing Machinery},
%%   28(1):114--133.

%% \bibitem[\protect\citename{Gusfield}1997]{Gusfield:97}
%% Dan Gusfield.
%% \newblock 1997.
%% \newblock {\em Algorithms on Strings, Trees and Sequences}.
%% \newblock Cambridge University Press, Cambridge, UK.

%% \end{thebibliography}

\end{document}

### 2014-02-28 docs/acl-2014/paper.tex
Continued work on paper.

--- 2014-02-28 docs/eacl-2014-styles/uniwac2.tex
@@
 \documentclass[11pt]{article}
-\usepackage{eacl2014}
+\usepackage{acl2014}
 \usepackage{times}
@@
 \usepackage{latexsym}
-\usepackage{amsmath}
-\usepackage{color}
-\usepackage[hidelinks]{hyperref}
+
+\bibliographystyle{acl}
+
 \usepackage{enumitem}
-
 \usepackage{listings}
 \usepackage{fancyvrb}
-
 \usepackage{color}
@@
 
-\special{papersize=210mm,297mm} % to avoid having to use "-t a4" with dvips 
-%\setlength\titlebox{6.5cm}  % You can expand the title box if you really have to
+\setlength\titlebox{5cm}
 
-\title{uniWaC: Compiling the Foundation Text for the WaCky Universal Corpus}
+\title{Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages}
 
-\author{Susanne Fertmann\textsuperscript{$\alpha$}, Guy Emerson\textsuperscript{$\beta$}, Liling Tan\textsuperscript{$\alpha$}, Alexis Palmer\textsuperscript{$\alpha$} and Michaela Regneri\textsuperscript{$\alpha$} \\
-  Universit{\"a}t des Saarlandes\textsuperscript{$\alpha$} / Campus, 66123 Saarbrücken, Germany \\
-  Deutsches Forschungszentrum f{\"u}r Künstliche Intelligenz (DFKI)\textsuperscript{$\beta$} / \\ Stuhlsatzenhausweg 3, 66123 Saarbrücken, Germany \\
+\author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
+  Universit{\"a}t des Saarlandes \\
+  Campus, 66123 Saarbr{\"u}cken, Germany \\
   {\tt s9sufert@stud.uni-saarland.de, emerson@coli.uni-saarland.de,} \\
@@
 \begin{abstract}
-A grand aim of linguistics is to seek a universal theory of human language and the opposing relativist view of linguistics seek to prove otherwise.
-To accomplish either goal, the compilation of a Universal Corpus with significant data from a large variety of languages is necessary.
-The Web contains vast amounts of linguistic data however it is largely saturated by English content (\textgreater50\%).
-Nonetheless, there are websites that contains small amount of data in a large number of languages.
-To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contains linguistic data from multiple languages and merge them with existing corpora collections.This paper describes the progress of compiling, cleaning and standardizing the foundation texts for the Universal Corpus (\texttt{uniWaC}) and briefly describes the ongoing work to merge our compilation with existing corpora collections.
+The study of endangered languages is limited by a lack of data.
+Existing corpora collections are limited in the range of languages covered, in standardisation, or in machine-readability.
+This makes the situation even worse for the computational linguist, especially one who would like to take a cross-linguistic or typological approach.
+We first survey existing efforts to compile cross-linguistic resources, then describe our own approach and give an example application - language clustering.
+To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contain data from a large number languages, and converted them into a standardised form consistent with the guidelines set out by Abney and Bird (2010).
+The resulting corpus is more easily-accessible and machine-readable than any of the underlying data sources, and represents a significant base corpus for researchers to draw on and add to in the future.
 \end{abstract}
@@
 \section{Introduction}
-Abney and Bird (2010) posed a grand challenge to build a Universal Corpus that will include as many of the world's languages as possible, in a consistent structure that permits crosslingual processing and analysis.
-Ideally, the Universal Corpus would be a complete digitization of every human language.
-To empirically prove or disprove the existence of a universal theory of human languages, a Universal Corpus is necessary.
+
+\newcite{abneybird2010} posed the grand challenge of building a Universal Corpus, including all of the world's languages, in a consistent structure.
+
+\section{Related Work}
 
@@
 
-\subsection{Wikipedia Dumps}
+\subsection{Wikipedia}
 
@@
 
-\section{Access and Copyrights}
+\subsection{Leipzig Corpora Collection}
+
+Do we want to mention this?
+
+\section{Copyright Issues}
+
+Creative Commons Licence, etc.
 
 \section{Representation and Universality}
-This is we report the counts and size of corpus.
 
-\section{Ongoing Work}
+Number of languages, size of corpus, etc.
 
-Currently, the uniWaC project is continually adding more data from the LCC Leipzig Corpora.
-Finally, the first phase of the uniWaC will end with the implementation of unsupervised tokenizer (e.g.
-Chung and Gildea, 2009) such that initial experiments on crosslingual analysis can be carried out to advance the grand aim of a unified linguistic theory.
+\section{Language Clustering}
 
-% TODO: use proper bibtex !!!!
-\section*{References}
-Steven Abney and Steven Bird.
-2010.
-The Human Language Project: Building a Universal Corpus of the World’s Languages.
-In Proceedings of the 48th Annual Meeting of the Association for Computational Linguistics.
-Uppsala, Sweden.
-\\\\
-Steven Abney and Steven Bird.
-2011.
-Towards a data model for the Universal Corpus.
-In Proceedings of the 4th Workshop on Building and Using Comparable Corpora: Comparable Corpora and the Web.
-Portland, Oregon.
-\\\\
-Michele Banko and Eric Brill.
-2001.
-Michele Banko and Eric Brill.
-2001.
-Scaling to very very large corpora for natural language disambiguation.
-In Proceedings of the 39th Annual Meeting on Association for Computational Linguistics (ACL '01).
-Association for Computational Linguistics, Stroudsburg, PA, USA, 26-33.
-\\\\
-Marco Baroni, Silvia Bernardini, Adriano Ferraresi and Eros Zanchetta.
-(2008).
-The WaCky wide web: A collection of very large linguistically processed web-crawled corpora.
-Language Resources and Evaluation, 43(3), 209–231.
-\\\\
-Tagyoung Chung and Daniel Gildea.
-2009.
-Unsupervised tokenization for machine translation.
-In Proceedings of the 2009 Conference on Empirical Methods in Natural Language Processing: Volume 2 - Volume 2 (EMNLP '09), Vol.
-2.
-Association for Computational Linguistics, Stroudsburg, PA, USA, 718-726.
-\\\\
-Adriano Ferraresi. 2007. Building a very large corpus of English obtained by Web crawling: ukWaC. Master Thesis, University of Bologna
-\\\\
-Jörg Tiedemann. 2012. Parallel Data, Tools and Interfaces in OPUS. In Proceedings of the 8th International Conference on Language Resources and Evaluation. Istanbul, Turkey.
-\\\\
-Liling Tan and Francis Bond.
-2011.
-Building and Annotating the Linguistically Diverse NTU-MC (NTU-Multilingual Corpus).
-In Proceedings of The 25th Pacific Asia Conference on Language, Information, and Computation.
-Singapore, Singapore.
+This would be an example application, if we have time...
+
+\section{Future Work}
+
+Other corpora we could work on...
+
+\section{Conclusion}
+
+Summarise what we've done!
+
+\bibliography{references}
 

### 2014-03-01 docs/acl-2014/paper.tex
I'm going to keep committing regularly, in case anyone else is also editing.

@@
 \bibliographystyle{acl}
+
+\usepackage[utf8]{inputenc}
 
@@
 \author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
-  Universit{\"a}t des Saarlandes \\
-  Campus, 66123 Saarbr{\"u}cken, Germany \\
+  Universität des Saarlandes \\
+  Campus, 66123 Saarbrücken, Germany \\
   {\tt s9sufert@stud.uni-saarland.de, emerson@coli.uni-saarland.de,} \\
@@
 
+7105 living languages.
+
 \section{Related Work}
 
-Currently multilingual corpora efforts have limited coverage in number of languages and the number of language families the corpora represents; for instance, the OPUS corpus covers over 90 languages (Tiedemann, 2012), i.e.
-1.27\% of the total number of languages in the world; Leipzig Corpora
-Collection\footnote{corpora.uni-leipzig.de} contains corpora in 230 languages, 3.24\% of 7105 living languages (Biemann et al.
-2007).
-Even corpora that boast of linguistic diversity lack in language families’ coverage; e.g.
-the linguistically diverse NTU-Multilingual Corpus covers only 7 out of 136 language families (Tan and Bond, 2011).
-The quintessential solution towards a Universal Corpus is to merge all existing corpora and provide an ubiquitous access interface.
-To compile the foundation text for the Universal Corpus (uniWaC), we crawled and cleaned web data that contains multilingual texts and merge them with existing corpora collections to form the foundation text for the Universal Corpus.
+[...]
+Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
+For instance, the OPUS corpus (Tiedemann, 2012) covers over 90 languages, only 1.27\% of the world's languages; The Leipzig Corpora Collection\footnote{corpora.uni-leipzig.de} (???Biemann et al., 2007) contains corpora and dictionaries in 230 languages, 3.24\% of the world's languages.
+Even corpora that boast of linguistic diversity include only a small number of language families, e.g.
+the linguistically diverse NTU-Multilingual Corpus (Tan and Bond, 2011) covers only 7 out of 136 language families.
+Producing a universal corpus requires the linguistic community to merge existing corpora and provide an ubiquitous access interface.
+To compile the foundation text for the Universal Corpus (uniWaC), we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
 
@@
 
-\section{Crawling and Cleaning}
+\section{Data Structure}
+
+\newcite{abneybird2011} describe the data structure in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
+In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
+
+However, we disagree that analysed texts should be accommodated in this way.
+If such a universal corpus is to succeed at all, it must enjoy support from the entire linguistics community, and doing this requires complete theory-neutrality.
+Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
+Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
+More fundamentally, they assume that the corpus should be segmented into words, but Haspelmath [ref] argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
+
+It is not within the scope of this paper to resolve these theoretical concerns.
+Instead, we suggest that the data structure for a universal corpus should be even more lightweight than Abney and Bird suggest.
+We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
+To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource language as a convenient surrogate of meaning.
+By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
+We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules [ref].
+The only difference here is that a text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
+The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies cannot be raised.
+We admit that highly detailed annotations, such as in treebanks and dependency banks, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
+
+\section{Data Sources}
 
@@
 \texttt{odin-iii.txt} contains the \emph{ODIN} data for the \emph{Nuosu} language.
+
+API which allows access according to Abney and Bird.
+However, maintaining a plain text file ensures that the data will be easily accessible, even if the surrounding infrastructure changes.
+[Seven dimensions]
+
+Our resulting corpus runs the full gamut of texts outlined by Abney and Bird, from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN). ...
+
+In the follow subsections, we describe the preprocessing required to convert each data source into a standardised form.
 

### 2014-03-01 docs/acl-2014/paper.tex
Converted all citations to bibtex.

@@
 We first survey existing efforts to compile cross-linguistic resources, then describe our own approach and give an example application - language clustering.
-To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contain data from a large number languages, and converted them into a standardised form consistent with the guidelines set out by Abney and Bird (2010).
+To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contain data from a large number languages, and converted them into a standardised form consistent with the guidelines set out by \newcite{abney2010universal}.
 The resulting corpus is more easily-accessible and machine-readable than any of the underlying data sources, and represents a significant base corpus for researchers to draw on and add to in the future.
@@
 
-\newcite{abneybird2010} posed the grand challenge of building a Universal Corpus, including all of the world's languages, in a consistent structure.
+\newcite{abney2010universal} posed the grand challenge of building a Universal Corpus, including all of the world's languages, in a consistent structure.
 
@@
 Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
-For instance, the OPUS corpus (Tiedemann, 2012) covers over 90 languages, only 1.27\% of the world's languages; The Leipzig Corpora Collection\footnote{corpora.uni-leipzig.de} (???Biemann et al., 2007) contains corpora and dictionaries in 230 languages, 3.24\% of the world's languages.
+For instance, the OPUS corpus \cite{tiedemann2012opus} covers over 90 languages, only 1.27\% of the world's languages; The Leipzig Corpora Collection\footnote{corpora.uni-leipzig.de} \cite{biemann2007leipzig} contains corpora and dictionaries in 230 languages, 3.24\% of the world's languages.
 Even corpora that boast of linguistic diversity include only a small number of language families, e.g.
-the linguistically diverse NTU-Multilingual Corpus (Tan and Bond, 2011) covers only 7 out of 136 language families.
+the linguistically diverse NTU-Multilingual Corpus \cite{tan2011ntu} covers only 7 out of 136 language families.
 Producing a universal corpus requires the linguistic community to merge existing corpora and provide an ubiquitous access interface.
-To compile the foundation text for the Universal Corpus (uniWaC), we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
+To compile the foundation text for the Universal Corpus, we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
 
-\subsection{WaCky Corpora}
+\subsection{Web as Corpus}
 
-Baroni and Bernardini (2004) introduced the BootCaT toolkit to bootstrap specialized corpora and terms from the web.
-Then, Sharoff (2006) built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
-Later, Ferraresi (2007) introduced and discussed the notion of compiling Web as Corpus \emph{"properly"} with the inaugural ukWaC using web-crawled data from a list of seed words.
-Thereafter, the Crúbadán corpora was created for a large number of under-resourced languages (Scannell, 2007).
-Following which the deWaC (German), itWaC (Italian) and frWaC (French) were compiled (Baroni et al.
-2008).
+\newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialized corpora and terms from the web.
+Then, \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
+Later, \newcite{ferraresi2008ukwac} introduced and discussed the notion of compiling Web as Corpus \emph{"properly"} with the inaugural ukWaC using web-crawled data from a list of seed words.
+Thereafter, the Crúbadán corpora was created for a large number of under-resourced languages \cite{scannell2007crubadan}.
+Following which the deWaC (German), itWaC (Italian) and frWaC (French) were compiled \cite{baroni2009wacky}.
 
-Subsequently, Brunello (2009) stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
-To build better quality WaCky corpora, Versley and Panchenko (2012) introduced the notion content-sensitive boilerplate detection in cleaning web corpora.
+Subsequently, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
+To build better quality WaCky corpora, \newcite{versley2012quality} introduced the notion content-sensitive boilerplate detection in cleaning web corpora.
 
-The initial wave of monolingual WaCky corpora (Baroni et al.
-2008) inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus (Dickinson et al.
-2013) and the Feed Corpus (Minocha et al.
-2013).
+The initial wave of monolingual WaCky corpora \cite{baroni2008wacky} inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus \cite{dickinson2010korean} and the Feed Corpus \cite{minocha2013feed}.
 Meanwhile, Web-driven corpora compilations also focused on building large  multilingual resources, e.g.
-Leipzig Corpora Collection (\texttt{LCC}) (Biemann et al.
-2007), the COrpora from the Web (\texttt{COW}) project (Schaefer and Bildhauer, 2012).
+Leipzig Corpora Collection (\texttt{LCC}) \cite{biemann2007leipzig}, the COrpora from the Web (\texttt{COW}) project \cite{schaefer2012cow}.
 
@@
 
-\newcite{abneybird2011} describe the data structure in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
+\newcite{abney2011data} describe the data structure in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
 In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
@@
 Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
-More fundamentally, they assume that the corpus should be segmented into words, but Haspelmath [ref] argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
+More fundamentally, they assume that the corpus should be segmented into words, but \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
 
@@
 By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules [ref].
-The only difference here is that a text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
+We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules.\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} The only difference here is that a text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
 The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies cannot be raised.
@@
 
-Although data size matters in general NLP (Banko and Brill, 2001), \emph{universality} is the utmost priority in NLP for the Universal Corpus initiative (aka Human Language Project; Abney and Bird, 2010; 2011).
+Although data size matters in general NLP \cite{banko2001scaling}, \emph{universality} is the utmost priority in NLP for the Universal Corpus initiative (aka Human Language Project; Abney and Bird, 2010; 2011).
 Hence, the uniWaC approach to data collection is based on focused/dedicated crawling unlike previous WaC word/URL list seeding.
@@
 \\ \\
-\noindent The primary problem in extracting the source text is lack of consistency of the IGTs from the ODIN data.
-For instance, Figure 2 uses an alphabet bullet convention, i.e.
-`\emph{a.}', whereas the IGT in Figure 3 uses a bracketed number indexing convention, i.e.
-`\emph{(69)}'.
-In addition, the source line in Figure 3 included the citation to the example, which would be considered as noise to corpus.
+\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
+In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexiing elements varies.
+In addition, the source line in Figure 3 includes a citation for the example, which introduces noise to the corpus.
 
-Lewis and Xia (2010) noted that it is not uncommon for IGT found in documents to deviant from one uniform convention; when compiling the ODIN data, they attempted a regex based approached and reported 59\% F-score.
+Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from a single convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
 Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness':
@@
 Although the redistribution of previous compiled data is not scientifically attractive but redistribution/recompilation of data to improve the usability and quality of the data is worth mentioning.
-For instance, the SETimes corpus first compiled by Tyers and Alperen (2010) was cleaned and redistributed by Ljubesic and Agic (2013).
+For instance, the SETimes corpus first compiled by Tyers and Alperen (2010) was cleaned and redistributed by \newcite{ljubesic2013ner}.
+[This is not mentioned in the reference...]
 

### 2014-03-01 docs/acl-2014/paper.tex
More work on paper.

@@
 Subsequently, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
-To build better quality WaCky corpora, \newcite{versley2012quality} introduced the notion content-sensitive boilerplate detection in cleaning web corpora.
+To build better quality WaCky corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection in cleaning web corpora.
 
-The initial wave of monolingual WaCky corpora \cite{baroni2008wacky} inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus \cite{dickinson2010korean} and the Feed Corpus \cite{minocha2013feed}.
+The initial wave of monolingual WaCky corpora \cite{baroni2009wacky} inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus \cite{dickinson2010korean} and the Feed Corpus \cite{minocha2013feed}.
 Meanwhile, Web-driven corpora compilations also focused on building large  multilingual resources, e.g.
-Leipzig Corpora Collection (\texttt{LCC}) \cite{biemann2007leipzig}, the COrpora from the Web (\texttt{COW}) project \cite{schaefer2012cow}.
+The Leipzig Corpora Collection (\texttt{LCC}) \cite{biemann2007leipzig}, and the COrpora from the Web (\texttt{COW}) project \cite{schaefer2012cow}.
 
-Different from previous WaCs built under the Web as Corpus kool ynitiative (WaCky), we did not perform seed query web-searching or web crawling to achieve a balanced resource.
-Instead the foundation text for uniWaC was built using dedicated site crawling and/or site-dependent cleaning from four data sources, viz.
+Unlike previous WaCs built under the Web as Corpus kool ynitiative (WaCky), we did not perform seed query web-searching or web crawling to achieve a balanced resource.
+Instead, the foundation text for uniWaC was built using dedicated site crawling and/or site-dependent cleaning from four data sources, viz.
 (i) \emph{Online Database of Interlinear Text} (ODIN), (ii) \emph{Omniglot} website, (iii) \emph{Universal Declaration of Human Rights} (UDHR) and (iv) \emph{Wikipedia} dumps.
@@
 
-\newcite{abney2011data} describe the data structure in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
+\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
 In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
@@
 The resulting files were saved as plaintext file with \texttt{UTF-8} encoding.
-One directory for each data source and one file per language.
+We used one directory for each data source and one file per language.
 The parallel data (from ODIN and Omniglot) are saved as tab-delimited files, one line per sentence, one column per language.
@@
 \texttt{odin-iii.txt} contains the \emph{ODIN} data for the \emph{Nuosu} language.
+[File structure is not relevant.]
 
@@
 
-The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinerized Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
+ODIN (The Online Database of Interlinear Text) is a repository of IGTs extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
+Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation; however, the data requires further processing to bring it in line with the proposed format for a universal corpus.
+
+The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
 Each XML file is saved under a filename that matches the respective language code.
 While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.
+[Huh?
+Should this really happen?
+- the codes are given on their website and look valid to me...]
 
@@
 \indent \qquad 2sg return here \\
+
 \indent \qquad `\emph{You return here.}' \\
-
-\centerline{Figure 1: IGT adhering to Leipzig Glossing Rule.}
+\centerline{Figure 1: IGT from ODIN.}
 \end{minipage}
@@
   <example>
-    <line>a. o lesu mai</line>
+    <line>21 a. o lesu mai</line>
     <line>2sg return here</line>
@@
 \end{lstlisting} 
-\centerline{Figure 2: An IGT in ODIN's XML format.}
+\centerline{Figure 2: Fijian IGT from ODIN.}
 \end{minipage}
@@
     <line>(69) na-Na-tmi-kwalca-t 
-    (Foley 1991)</line>
+    Yimas (Foley 1991)</line>
     <line>3sgA-1sgO-say-rise-PERF
@@
 \smallskip
-\centerline{Figure 3: Another IGT in ODIN's XML format.}
+\centerline{Figure 3: Yimas IGT from ODIN.}
 \end{minipage}
@@
 \noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
-In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexiing elements varies.
-In addition, the source line in Figure 3 includes a citation for the example, which introduces noise to the corpus.
+In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
+In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.
 
-Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from a single convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness':
+Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
+Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness': [Have we considered that the number of elements in the source line and gloss line should match...?]
 \\ \\
@@
 \end{itemize}
-\item \emph{Cleanest}: Accepts only source lines without any punctuation.
+\item \emph{Cleanest}: Only source lines without punctuation.
 \end{itemize}
@@
 \\ \\
+
 \noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
-The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 693-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
+The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
 
@@
 
-The Omniglot crawling and cleaning required more tact and after manual inspection of the www.omniglot.com website, only the \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} pages were consistent in their HTML markups that allows easy cleaning.
-We wrote a crawler script that downloads pages with that contains the following in their URLs:
+The Omniglot website\footnote{htp://www.omniglot.com/} is an online encyclopedia of writing systems and languages.
+We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
+
+[The following discussion is far too technical.]
 
@@
 
-\noindent The primary problem of the Omniglot data is that to the only information about the language of the data on the page is given in the title of the HTML and its URL.
-Additionally, only the name of the language was given not the ISO 639-3 code.
-Even though Omniglot pages contain language meta data in XHTML markup attributes, most pages are encoded with the \texttt{lang="en"} attribute because the pages are catered for English speaking audience and the boilerplates (e.g.
-navigation bar, site banner and footer) on the pages are in English.
-This causes a problem in file storage and access since it is necessary to save the files in the naming convention, e.g.
-\texttt{omniglot-smo.txt} but the language code is not found any where on the page.
-To resolve the problem of standardizing, we have written a conversion script using the code to language mappings from \url{http://www-01.sil.org/iso639-3/iso-639-3.tab}
+\noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
+To resolve this issue, we converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
 
@@
 
+The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages.
 Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
-Originally, we tried to resolve the encoding problems by automatically identifying the encoding and decoding the textfiles using the \texttt{libmagic} library\footnote{http://linux.die.net/man/3/libmagic} and then re-encoding them to \texttt{UTF-8} whenever possible.
+Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d/}, which are free of encoding issues.
+The first four lines of each file records metadata, and the rest is the translation of the UDHR.
 
-However the better solution was found when we chanced upon the UDHR site\footnote{http://unicode.org/udhr/d/} from the \url{unicode.org} domain.
-The UDHR textfiles found on the site was free of encoding problem but there was a boilerplate at the start of each file which records it metadata.
-It was simply clean by skipping the first four lines of each file.
-
-Although the redistribution of previous compiled data is not scientifically attractive but redistribution/recompilation of data to improve the usability and quality of the data is worth mentioning.
+[This paragraph seems completely incongruous] Although the redistribution of previous compiled data is not scientifically attractive but redistribution/recompilation of data to improve the usability and quality of the data is worth mentioning.
 For instance, the SETimes corpus first compiled by Tyers and Alperen (2010) was cleaned and redistributed by \newcite{ljubesic2013ner}.
@@
 Tapping on the crowd-sourced Wikipedia articles for NLP forms a crucial part of developing data-driven NLP tools and application ({\color{red}{citation neeeded}}).
-To automatically download the Wikipedia dumps, we have scripted the \texttt{wget} with ISO 639-2 language code as a variable (\texttt{\$I}) in a shell script, as such:
+[It might be popular, but it's not crucial...] To automatically download the Wikipedia dumps, we have scripted the \texttt{wget} with ISO 639-2 language code as a variable (\texttt{\$I}) in a shell script, as such: [This is too technical]
 
@@
 
-After converting them into textfiles, we used the following regexes to delete residual Wikipedia markups and magic words\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
+After converting them into textfiles, we used the following regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
 

### 2014-03-02 docs/acl-2014/paper.tex
Further work on paper.

@@
 \begin{abstract}
-The study of endangered languages is limited by a lack of data.
+The study of endangered languages is constrained by a lack of data.
 Existing corpora collections are limited in the range of languages covered, in standardisation, or in machine-readability.
@@
 
-\newcite{abney2010universal} posed the grand challenge of building a Universal Corpus, including all of the world's languages, in a consistent structure.
+XXX living languages are documented in Ethnologue [ref], but [ref] estimated that by the end of the century, as many as XX\% of these will go extinct.
+However, only a fraction of these are well documented, and even fewer have machine-readable resources [ref](Maxwell and Hughes, 2006?), so the time to work on compiling these resources is now.
 
-7105 living languages.
+Although there have been some previous attempts to produce cross-linguistic resources, there are few which both cover a wide range of languages, and are also machine-readable.
+We survey existing efforts in section \ref{sec:related}, and discuss their limitations in more detail.
 
-\section{Related Work}
+Abney and Bird \shortcite{abney2010universal,abney2011data} posed the grand challenge of building a Universal Corpus, calling it the Human Language Project.
+Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
+They propose a specific data structure, which we describe and discuss in section \ref{sec:structure}.
+We have accepted their challenge, and have begun converting existing resources into a format consistent with their specifications for a universal corpus.
+We have drawn on four [or five?] web sources, cleaning and standardising them as described in section \ref{sec:sources}, to produce a seed corpus for the Human Language Project.
+In sections \ref{sec:stats} and \ref{sec:copyright}, we respetively give a summary of the data contained, and discuss copyright and distribution.
 
-[...]
+[Note: remove this section if we don't do the work!] We believe the resulting corpus is the first of its kind: large enough and consistent enough to allow language processing on a grand scale.
+In section \ref{sec:cluster}, we give an example application of this corpus: language clustering.
+We use the frequencies of character n-grams and words to estimate the similarity of two languages.
+Despite this approach being hightly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
+Finally, we discuss future work in secion \ref{sec:future}, and conclude in section \ref{sec:conclusion}.
+
+\section{Related Work} \label{sec:related}
+
+[There is a lot more stuff to be added here...]
 Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
@@
 
-\newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialized corpora and terms from the web.
-Then, \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
-Later, \newcite{ferraresi2008ukwac} introduced and discussed the notion of compiling Web as Corpus \emph{"properly"} with the inaugural ukWaC using web-crawled data from a list of seed words.
-Thereafter, the Crúbadán corpora was created for a large number of under-resourced languages \cite{scannell2007crubadan}.
-Following which the deWaC (German), itWaC (Italian) and frWaC (French) were compiled \cite{baroni2009wacky}.
+In recent years, there has been increasing interesting in crawling websites to produce corpora.
+For example, \newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialised corpora, and \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
+\newcite{ferraresi2008ukwac} produced ukWaC, a large-scale English corpus, and \cite{baroni2009wacky} applied a similar method to produce corpora in German, Italian, and French.
+To build better quality web corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection for cleaning data.
 
-Subsequently, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
-To build better quality WaCky corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection in cleaning web corpora.
+However, despite the amount of text they include, such corpora are usually limited in the number of languages represented.
+Some projects have emerged which strive for language diversity, such as the Leipzig Corpora Collection (LCC) \cite{biemann2007leipzig}, and COrpora from the Web (COW) \cite{schaefer2012cow}.
+The LCC currently offers free download of corpora in 117 languages, and dictionaries in many others, bringing the total number of languages up to 230.
+COW includes corpora in [XX] languages.
+These collections are certainly commendable, but they currently still fall short of universality - 230 languages represents only 3.3% of the world's languages [check figure].
 
-The initial wave of monolingual WaCky corpora \cite{baroni2009wacky} inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus \cite{dickinson2010korean} and the Feed Corpus \cite{minocha2013feed}.
-Meanwhile, Web-driven corpora compilations also focused on building large  multilingual resources, e.g.
-The Leipzig Corpora Collection (\texttt{LCC}) \cite{biemann2007leipzig}, and the COrpora from the Web (\texttt{COW}) project \cite{schaefer2012cow}.
+\newcite{scannell2007crubadan} describes the Crúbadán Project, an attempt to crawl the web for data in a much larger range of languages, including endangered ones.
+At the time of writing, the number of languages represented has grown to [XXX] However, due to copyright issues, not all of the data is freely available.
+Motivated by similar copyright concerns, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
 
-Unlike previous WaCs built under the Web as Corpus kool ynitiative (WaCky), we did not perform seed query web-searching or web crawling to achieve a balanced resource.
-Instead, the foundation text for uniWaC was built using dedicated site crawling and/or site-dependent cleaning from four data sources, viz.
-(i) \emph{Online Database of Interlinear Text} (ODIN), (ii) \emph{Omniglot} website, (iii) \emph{Universal Declaration of Human Rights} (UDHR) and (iv) \emph{Wikipedia} dumps.
-Our emphasis for the uniWaC is based on based on \textbf{\emph{universality}}; covering as many languages as possible is the first priority.
+Unlike the above efforts, we have not performed seed-query web-searching or web-crawling to achieve a balanced resource.
+Instead, we have focused on a small number of sources which contain data in a wide variety of languages, as described in section \ref{sec:sources}.
+However, these sources represent a variety of data types which a linguist might encounter, ranging from single-language text with significant non-linguistic markup (Wikipedia) to structured data with detailed linguistic annotations (ODIN).
 
-\section{Data Structure}
+\section{Data Structure} \label{sec:structure}
 
-\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
+\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level [add more detail]; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
 In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
@@
 However, we disagree that analysed texts should be accommodated in this way.
-If such a universal corpus is to succeed at all, it must enjoy support from the entire linguistics community, and doing this requires complete theory-neutrality.
+If such a universal corpus is to succeed at all, it must enjoy support from a substantial part, if not all, of the linguistics community, and doing this requires theory-neutrality.
 Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
@@
 We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
-To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource language as a convenient surrogate of meaning.
+To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource reference language as a convenient surrogate of meaning.
 By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules.\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} The only difference here is that a text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
-The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies cannot be raised.
-We admit that highly detailed annotations, such as in treebanks and dependency banks, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
+We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules,\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} and as practised by documentary linguists.
 
-\section{Data Sources}
+Then, only difference between analysed and aligned texts is that the text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
+The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies are difficult to raise.
+We admit that highly detailed annotations, such as trees and dependency graphs, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
+To make it indicate what reference language is being used, we propose prefixing \texttt{gloss-} to the language code of the reference language - for example \texttt{gloss-eng} and \texttt{gloss-eng} if a text has been glossed into English or Spanish, respectively.
+Using a prefix rather than a suffix makes it clear that this is not a subvariety of the given language.
 
-Although data size matters in general NLP \cite{banko2001scaling}, \emph{universality} is the utmost priority in NLP for the Universal Corpus initiative (aka Human Language Project; Abney and Bird, 2010; 2011).
-Hence, the uniWaC approach to data collection is based on focused/dedicated crawling unlike previous WaC word/URL list seeding.
-We manually selected sites that contains texts from a large variety of languages and wrote scripts to crawl and clean specific pages from the different data sources.
-The crawling and cleaning was done simultaneously.
+Finally, it is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
+[ref - seven dimensions] describes a set of best practices for maintaining portability of digital information.
+[more detail] For these reasons, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
+An API then allows access to this data according to the guidelines of Abney and Bird, who remain agnostic as to the specific form of data storage.
+If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.
 
-The resulting files were saved as plaintext file with \texttt{UTF-8} encoding.
-We used one directory for each data source and one file per language.
-The parallel data (from ODIN and Omniglot) are saved as tab-delimited files, one line per sentence, one column per language.
-The monolingual data (from UDHR and Wikipedia) are saved as newline delimited plaintext file, each line denoting a paragraph or sentence and each document separated by an empty line.
-The files are saved with the following naming convention to allow easy access, the filename contains the data source and its ISO 639-3 code, e.g.
-\texttt{odin-iii.txt} contains the \emph{ODIN} data for the \emph{Nuosu} language.
-[File structure is not relevant.]
+\section{Data Sources} \label{sec:sources}
 
-API which allows access according to Abney and Bird.
-However, maintaining a plain text file ensures that the data will be easily accessible, even if the surrounding infrastructure changes.
-[Seven dimensions]
+Although data size matters in general NLP \cite{banko2001scaling} [is this the best reference?], \emph{universality} is the utmost priority for a universal corpus.
+We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured):
 
-Our resulting corpus runs the full gamut of texts outlined by Abney and Bird, from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN). ...
+\begin{enumerate}
+\item The Online Database of Interlinear Text (ODIN)
+\item The Omniglot website
+\item The Universal Declaration of Human Rights (UDHR)
+\item Wikipedia
+\item The Leipzig Corpora Collection (LCC) [do we include this?]
+\end{enumerate}
 
-In the follow subsections, we describe the preprocessing required to convert each data source into a standardised form.
+In the following subsections, we describe each source, and the processing required to convert the data into a standardised form.
+Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia and LCC) to parallel text (UDHR and Omniglot) to IGTs (ODIN).
 
@@
 
-\begin{minipage}{\columnwidth}
+\begin{minipage}{\columnwidth} %Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
 \begin{lstlisting}[language=XML]
@@
 Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness': [Have we considered that the number of elements in the source line and gloss line should match...?]
+Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness':
+
+[This hasn't been implemented, but I think we could try this: Split source and gloss lines into tokens according to whitespace.
+Check the source has more tokens.
+Count the number of hyphens and equal signs in each token (since they denote morphemes).
+Match the gloss line with a substring of the source line, according to numbers of hyphens and equal signs.
+If there's exactly one match, keep it, and discard tokens at the ends of the line.
+If there's more than one match, look for other punctuation in the initial and final tokens.]
 \\ \\
@@
 \noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
-To resolve this issue, we converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
+To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
 
@@
 The first four lines of each file records metadata, and the rest is the translation of the UDHR.
-
-[This paragraph seems completely incongruous] Although the redistribution of previous compiled data is not scientifically attractive but redistribution/recompilation of data to improve the usability and quality of the data is worth mentioning.
-For instance, the SETimes corpus first compiled by Tyers and Alperen (2010) was cleaned and redistributed by \newcite{ljubesic2013ner}.
-[This is not mentioned in the reference...]
+This dataset was extremely clean, and simply required segmentation into sentences.
 
@@
 
-After converting them into textfiles, we used the following regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
+After converting them into textfiles, we used the following [missing information] regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
 
@@
 
-\section{Copyright Issues}
+\section{Copyright Issues} \label{sec:copyright}
 
@@
 
-\section{Representation and Universality}
+\section{Representation and Universality} \label{sec:stats}
 
@@
 
-\section{Language Clustering}
+\section{Language Clustering} \label{sec:cluster}
 
@@
 
-\section{Future Work}
+\section{Future Work} \label{sec:future}
 
@@
 
-\section{Conclusion}
+\section{Conclusion} \label{sec:conclusion}
 
-Summarise what we've done!
+In this paper, we have described the creation of a foundation text for a universal corpus, following the guidelines of Abney and Bird (2010; 2011).
+To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia, and the LCC.
+The resulting corpus is more more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices of [ref - seven dimensions].
+[Give a summary statistic of size.] To demonstrate the utility of this resource, we used the data to perform language clustering, which gave promising results.
+We believe that this is the first corpus of its kind, which we hope will act as a seed corpus for the Human Language Project.
 

### 2014-03-02 docs/acl-2014/paper.tex
Added references.

@@
 
-XXX living languages are documented in Ethnologue [ref], but [ref] estimated that by the end of the century, as many as XX\% of these will go extinct.
-However, only a fraction of these are well documented, and even fewer have machine-readable resources [ref](Maxwell and Hughes, 2006?), so the time to work on compiling these resources is now.
+At the time of writing, 7105 living languages are documented in Ethnologue\footnote{http://www.ethnologue.com/}, but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century.
+However, only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
 
@@
 Finally, it is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
-[ref - seven dimensions] describes a set of best practices for maintaining portability of digital information.
-[more detail] For these reasons, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
-An API then allows access to this data according to the guidelines of Abney and Bird, who remain agnostic as to the specific form of data storage.
+\newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining "seven dimensions" along which this can vary.
+[more detail?] Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
+We have written an API to allow access to this data according to the guidelines of Abney and Bird, who remain agnostic as to the specific form of data storage.
 If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.
@@
 
+We have made all our resources publically available.\footnote{To maintain anonymity, we have removed details, but, if accepted for publication, instructions will be included in the print version of this paper.}
+
 \section{Representation and Universality} \label{sec:stats}
@@
 
-Other corpora we could work on...
+Specific suggestions on how to work on resources mentioned in section \ref{sec:related}...
 
@@
 To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia, and the LCC.
-The resulting corpus is more more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices of [ref - seven dimensions].
+The resulting corpus is more more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}.
 [Give a summary statistic of size.] To demonstrate the utility of this resource, we used the data to perform language clustering, which gave promising results.

### 2014-03-02 docs/acl-2014/paper.tex
Work on paper: Wikipedia section, emails, typos.

@@
   Campus, 66123 Saarbrücken, Germany \\
-  {\tt s9sufert@stud.uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-  {\tt liling.tan@uni-saarland.de}}
+  {\tt liling.tan@uni-saarland.de, emerson@coli.uni-saarland.de,} \\
+  {\tt susfert@coli.uni-saarland.de}}
 
@@
 We have drawn on four [or five?] web sources, cleaning and standardising them as described in section \ref{sec:sources}, to produce a seed corpus for the Human Language Project.
-In sections \ref{sec:stats} and \ref{sec:copyright}, we respetively give a summary of the data contained, and discuss copyright and distribution.
+In sections \ref{sec:stats} and \ref{sec:copyright}, we respectively give a summary of the data contained, and discuss copyright and distribution.
 
@@
 We use the frequencies of character n-grams and words to estimate the similarity of two languages.
-Despite this approach being hightly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
-Finally, we discuss future work in secion \ref{sec:future}, and conclude in section \ref{sec:conclusion}.
+Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
+Finally, we discuss future work in section \ref{sec:future}, and conclude in section \ref{sec:conclusion}.
 
@@
 
-Tapping on the crowd-sourced Wikipedia articles for NLP forms a crucial part of developing data-driven NLP tools and application ({\color{red}{citation neeeded}}).
-[It might be popular, but it's not crucial...] To automatically download the Wikipedia dumps, we have scripted the \texttt{wget} with ISO 639-2 language code as a variable (\texttt{\$I}) in a shell script, as such: [This is too technical]
+Tapping on the crowd-sourced Wikipedia articles for NLP forms a crucial part of developing data-driven NLP tools and application ({\color{red}{citation needed}}).
+[It might be popular, but it's not crucial...] To automatically download the Wikipedia dumps, we used the wp-download\footnote{https://github.com/babilen/wp-download} tool, adapted to an expanded set of languages.
 
-\smallskip
-\noindent \texttt{\$ wget http://download.wikimedia.
-org/\$Iwiki/latest/itwiki-latest-
-pages-articles.xml.bz2}
-\medskip
+%To automatically download the Wikipedia dumps, we have scripted the \texttt{wget} with ISO 639-2 language code as a variable (\texttt{\$I}) in a shell script, as such: [This is too technical][Such script has not been implemented]
 
-\noindent One major issue with using the Wikipedia dump is the sheer size and extracting text from a glut of Wikipedia markup.
-To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
+%\smallskip
+%\noindent \texttt{\$ wget http://download.wikimedia.
+%org/\$Iwiki/latest/itwiki-latest-
+%pages-articles.xml.bz2}
+%\medskip
 
-After converting them into textfiles, we used the following [missing information] regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
+\noindent One major issue with using the Wikipedia dumps is the sheer size and extracting text from a glut of Wikipedia markup.
+To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor\footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
+After the conversion into textfiles, we used the following [missing information] regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
+
+Wikipedia uses its own set of language codes, most of which are in ISO 639-1 or ISO 639-2/3. We automatically converted all of them into ISO~639-3.
 
@@
 To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia, and the LCC.
-The resulting corpus is more more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}.
+The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}.
 [Give a summary statistic of size.] To demonstrate the utility of this resource, we used the data to perform language clustering, which gave promising results.

### 2014-03-03 docs/acl-2014/paper.tex
added copyright section

--- 2014-03-02 docs/acl-2014/paper.tex
@@
 \begin{abstract}
-The study of endangered languages is constrained by a lack of data.
+The study of endangered languages is limited by a lack of data.
 Existing corpora collections are limited in the range of languages covered, in standardisation, or in machine-readability.
@@
 
-XXX living languages are documented in Ethnologue [ref], but [ref] estimated that by the end of the century, as many as XX\% of these will go extinct.
-However, only a fraction of these are well documented, and even fewer have machine-readable resources [ref](Maxwell and Hughes, 2006?), so the time to work on compiling these resources is now.
+\newcite{abney2010universal} posed the grand challenge of building a Universal Corpus, including all of the world's languages, in a consistent structure.
 
-Although there have been some previous attempts to produce cross-linguistic resources, there are few which both cover a wide range of languages, and are also machine-readable.
-We survey existing efforts in section \ref{sec:related}, and discuss their limitations in more detail.
+7105 living languages.
 
-Abney and Bird \shortcite{abney2010universal,abney2011data} posed the grand challenge of building a Universal Corpus, calling it the Human Language Project.
-Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
-They propose a specific data structure, which we describe and discuss in section \ref{sec:structure}.
-We have accepted their challenge, and have begun converting existing resources into a format consistent with their specifications for a universal corpus.
-We have drawn on four [or five?] web sources, cleaning and standardising them as described in section \ref{sec:sources}, to produce a seed corpus for the Human Language Project.
-In sections \ref{sec:stats} and \ref{sec:copyright}, we respetively give a summary of the data contained, and discuss copyright and distribution.
+\section{Related Work}
 
-[Note: remove this section if we don't do the work!] We believe the resulting corpus is the first of its kind: large enough and consistent enough to allow language processing on a grand scale.
-In section \ref{sec:cluster}, we give an example application of this corpus: language clustering.
-We use the frequencies of character n-grams and words to estimate the similarity of two languages.
-Despite this approach being hightly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
-Finally, we discuss future work in secion \ref{sec:future}, and conclude in section \ref{sec:conclusion}.
-
-\section{Related Work} \label{sec:related}
-
-[There is a lot more stuff to be added here...]
+[...]
 Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
@@
 
-In recent years, there has been increasing interesting in crawling websites to produce corpora.
-For example, \newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialised corpora, and \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
-\newcite{ferraresi2008ukwac} produced ukWaC, a large-scale English corpus, and \cite{baroni2009wacky} applied a similar method to produce corpora in German, Italian, and French.
-To build better quality web corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection for cleaning data.
+\newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialized corpora and terms from the web.
+Then, \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
+Later, \newcite{ferraresi2008ukwac} introduced and discussed the notion of compiling Web as Corpus \emph{"properly"} with the inaugural ukWaC using web-crawled data from a list of seed words.
+Thereafter, the Crúbadán corpora was created for a large number of under-resourced languages \cite{scannell2007crubadan}.
+Following which the deWaC (German), itWaC (Italian) and frWaC (French) were compiled \cite{baroni2009wacky}.
 
-However, despite the amount of text they include, such corpora are usually limited in the number of languages represented.
-Some projects have emerged which strive for language diversity, such as the Leipzig Corpora Collection (LCC) \cite{biemann2007leipzig}, and COrpora from the Web (COW) \cite{schaefer2012cow}.
-The LCC currently offers free download of corpora in 117 languages, and dictionaries in many others, bringing the total number of languages up to 230.
-COW includes corpora in [XX] languages.
-These collections are certainly commendable, but they currently still fall short of universality - 230 languages represents only 3.3% of the world's languages [check figure].
+Subsequently, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
+To build better quality WaCky corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection in cleaning web corpora.
 
-\newcite{scannell2007crubadan} describes the Crúbadán Project, an attempt to crawl the web for data in a much larger range of languages, including endangered ones.
-At the time of writing, the number of languages represented has grown to [XXX] However, due to copyright issues, not all of the data is freely available.
-Motivated by similar copyright concerns, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
+The initial wave of monolingual WaCky corpora \cite{baroni2009wacky} inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus \cite{dickinson2010korean} and the Feed Corpus \cite{minocha2013feed}.
+Meanwhile, Web-driven corpora compilations also focused on building large  multilingual resources, e.g.
+The Leipzig Corpora Collection (\texttt{LCC}) \cite{biemann2007leipzig}, and the COrpora from the Web (\texttt{COW}) project \cite{schaefer2012cow}.
 
-Unlike the above efforts, we have not performed seed-query web-searching or web-crawling to achieve a balanced resource.
-Instead, we have focused on a small number of sources which contain data in a wide variety of languages, as described in section \ref{sec:sources}.
-However, these sources represent a variety of data types which a linguist might encounter, ranging from single-language text with significant non-linguistic markup (Wikipedia) to structured data with detailed linguistic annotations (ODIN).
+Unlike previous WaCs built under the Web as Corpus kool ynitiative (WaCky), we did not perform seed query web-searching or web crawling to achieve a balanced resource.
+Instead, the foundation text for uniWaC was built using dedicated site crawling and/or site-dependent cleaning from four data sources, viz.
+(i) \emph{Online Database of Interlinear Text} (ODIN), (ii) \emph{Omniglot} website, (iii) \emph{Universal Declaration of Human Rights} (UDHR) and (iv) \emph{Wikipedia} dumps.
+Our emphasis for the uniWaC is based on based on \textbf{\emph{universality}}; covering as many languages as possible is the first priority.
 
-\section{Data Structure} \label{sec:structure}
+\section{Data Structure}
 
-\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level [add more detail]; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
+\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
 In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
@@
 However, we disagree that analysed texts should be accommodated in this way.
-If such a universal corpus is to succeed at all, it must enjoy support from a substantial part, if not all, of the linguistics community, and doing this requires theory-neutrality.
+If such a universal corpus is to succeed at all, it must enjoy support from the entire linguistics community, and doing this requires complete theory-neutrality.
 Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
@@
 We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
-To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource reference language as a convenient surrogate of meaning.
+To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource language as a convenient surrogate of meaning.
 By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules,\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} and as practised by documentary linguists.
+We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules.\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} The only difference here is that a text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
+The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies cannot be raised.
+We admit that highly detailed annotations, such as in treebanks and dependency banks, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
 
-Then, only difference between analysed and aligned texts is that the text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
-The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies are difficult to raise.
-We admit that highly detailed annotations, such as trees and dependency graphs, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
-To make it indicate what reference language is being used, we propose prefixing \texttt{gloss-} to the language code of the reference language - for example \texttt{gloss-eng} and \texttt{gloss-eng} if a text has been glossed into English or Spanish, respectively.
-Using a prefix rather than a suffix makes it clear that this is not a subvariety of the given language.
+\section{Data Sources}
 
-Finally, it is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
-[ref - seven dimensions] describes a set of best practices for maintaining portability of digital information.
-[more detail] For these reasons, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
-An API then allows access to this data according to the guidelines of Abney and Bird, who remain agnostic as to the specific form of data storage.
-If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.
+Although data size matters in general NLP \cite{banko2001scaling}, \emph{universality} is the utmost priority in NLP for the Universal Corpus initiative (aka Human Language Project; Abney and Bird, 2010; 2011).
+Hence, the uniWaC approach to data collection is based on focused/dedicated crawling unlike previous WaC word/URL list seeding.
+We manually selected sites that contains texts from a large variety of languages and wrote scripts to crawl and clean specific pages from the different data sources.
+The crawling and cleaning was done simultaneously.
 
-\section{Data Sources} \label{sec:sources}
+The resulting files were saved as plaintext file with \texttt{UTF-8} encoding.
+We used one directory for each data source and one file per language.
+The parallel data (from ODIN and Omniglot) are saved as tab-delimited files, one line per sentence, one column per language.
+The monolingual data (from UDHR and Wikipedia) are saved as newline delimited plaintext file, each line denoting a paragraph or sentence and each document separated by an empty line.
+The files are saved with the following naming convention to allow easy access, the filename contains the data source and its ISO 639-3 code, e.g.
+\texttt{odin-iii.txt} contains the \emph{ODIN} data for the \emph{Nuosu} language.
+[File structure is not relevant.]
 
-Although data size matters in general NLP \cite{banko2001scaling} [is this the best reference?], \emph{universality} is the utmost priority for a universal corpus.
-We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured):
+API which allows access according to Abney and Bird.
+However, maintaining a plain text file ensures that the data will be easily accessible, even if the surrounding infrastructure changes.
+[Seven dimensions]
 
-\begin{enumerate}
-\item The Online Database of Interlinear Text (ODIN)
-\item The Omniglot website
-\item The Universal Declaration of Human Rights (UDHR)
-\item Wikipedia
-\item The Leipzig Corpora Collection (LCC) [do we include this?]
-\end{enumerate}
+Our resulting corpus runs the full gamut of texts outlined by Abney and Bird, from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN). ...
 
-In the following subsections, we describe each source, and the processing required to convert the data into a standardised form.
-Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia and LCC) to parallel text (UDHR and Omniglot) to IGTs (ODIN).
+In the follow subsections, we describe the preprocessing required to convert each data source into a standardised form.
 
@@
 
-\begin{minipage}{\columnwidth} %Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
+\begin{minipage}{\columnwidth}
 \begin{lstlisting}[language=XML]
@@
 Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness':
-
-[This hasn't been implemented, but I think we could try this: Split source and gloss lines into tokens according to whitespace.
-Check the source has more tokens.
-Count the number of hyphens and equal signs in each token (since they denote morphemes).
-Match the gloss line with a substring of the source line, according to numbers of hyphens and equal signs.
-If there's exactly one match, keep it, and discard tokens at the ends of the line.
-If there's more than one match, look for other punctuation in the initial and final tokens.]
+Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness': [Have we considered that the number of elements in the source line and gloss line should match...?]
 \\ \\
@@
 
-\noindent The \emph{`Useful foreign phrases'} page contains parallel phrases in embedded within the \texttt{<th><tr>...</tr><th>} tags.
-The English phrase and foreign language phrase are encoded separately with \texttt{<td>...</td>} tags.
-Figure 4 shows an instance of an Omniglot \emph{Useful foreign phrase} page.
-Since the HTML page was designed as WYSIWYG, several pages have non-textual elements within the \texttt{<tr>...</tr>} tags for aesthetics.
-For example, the non-breaking space, \texttt{<td>\&nbsp;</td>} in Fig.
-4; the extracted text was properly converted into plaintext format using the HTMLParser\footnote{http://docs.python.org/2/library/htmlparser.html} module, during which the non textual elements would have been cleaned.
-
-The same process of crawling URLs and cleaning was performed on the \emph{`Tower of Babel'} pages and the only difference was that the texts were embedded in a \texttt{<ol><li>...</li></ol>} tags instead.
-
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<th>
-  ...
-  <tr>
-    <td>I don't understand</td>
-    <td>No appo cumpresu nutta</td>
-    <td>&nbsp;</td>
-  </tr>
-  <tr>
-    <td>I don't understand</td>
-    <td>Non d'isco</td>
-    <td>&nbsp;</td>
-  </tr>
-  ...
-<th>
-\end{lstlisting} 
-\centerline{Figure 4: A \emph{Useful Foreign Phrase} page}
-\end{minipage}
-\smallskip
-
 \noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
-To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
+To resolve this issue, we converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
 
@@
 The first four lines of each file records metadata, and the rest is the translation of the UDHR.
-This dataset was extremely clean, and simply required segmentation into sentences.
 
@@
 
-Tapping on the crowd-sourced Wikipedia articles for NLP forms a crucial part of developing data-driven NLP tools and application ({\color{red}{citation neeeded}}).
-[It might be popular, but it's not crucial...] To automatically download the Wikipedia dumps, we have scripted the \texttt{wget} with ISO 639-2 language code as a variable (\texttt{\$I}) in a shell script, as such: [This is too technical]
-
-\smallskip
-\noindent \texttt{\$ wget http://download.wikimedia.
-org/\$Iwiki/latest/itwiki-latest-
-pages-articles.xml.bz2}
-\medskip
-
+Wikipedia contains 30.8 million articles in 287 languages and the text dump can be downloaded from \url{www.dumps.wikimedia.org}.
 \noindent One major issue with using the Wikipedia dump is the sheer size and extracting text from a glut of Wikipedia markup.
 To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
+After converting them into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
 
-After converting them into textfiles, we used the following [missing information] regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
+\section{Copyright Issues}
 
-\subsection{Leipzig Corpora Collection}
+Corpora copyright remains a grey area and a hobglobin for corpora compilation and distribution (Kilgariff, 2002).
+Often, free access to language data online does not imply the right to download, retain, remix and redistribute.
 
-Do we want to mention this?
+ODIN refers itself as a database of IGTs that are returned as a result of querying from the source documents.
+As such, only the links to the documents and the appropriate citations and not the copies of the source documents are maintained on the ODIN site (Lewis and Xia, 2010).
+However, the ODIN developers have left the users with no specific license/copyrights restriction for readaptation or redistribution.
 
-\section{Copyright Issues} \label{sec:copyright}
+The Omniglot phrases are free to be used for non-commercial purposes as stated in the FAQ page but the rights of the Tower of Babel snippets from different bible sources remains iffy.
+For example, Omniglot cited www.bibles.org as one of its sources for the Tower of Babel chapter but the bibles.org site stated in fine print that `\emph{Except expressly permitted herein, you may not modify, copy, distribute, decompile, disassemble, reverse engineer, create derivative works, or otherwise use or manipulate the Site or any of its content without our prior written consent}'\footnote{http://bibles.org/pages/legal\#copyright}.
 
-Creative Commons Licence, etc.
+The UDHR copyrights remains questionable.
+Although users are free to use, copy, modify, merge, publish and distribute Unicode Data Files, there is no explicit ownership of the UDHR textfiles from the Unicode website.
+However, the original source of the UDHR documents from the Office of the High Commissioner for Human Rights clearly states that `\emph{None of the materials provided on this web site may be used, reproduced or transmitted, in whole or in part}'\footnote{http://www.un.org/en/aboutun/copyright/}.
 
-\section{Representation and Universality} \label{sec:stats}
+The Wikipedia texts are licensed under Creative Commons Attribution Share-like\footnote{http://creativecommons.org/licenses/by-sa/3.0/legalcode}, which allows redistribution and adaptation.
+Hence the clean Wikipedia texts we have compiled can be freely redistributed without infringing any copyrights.
+
+%%
+%To avoid any copyrights infringement, ...
+After surveying the rights for so many days, there is no way to redistribute the corpus unless we jumble up the sentences, but if we do that, the "document" level will be lost and that will not be a "good" resource to do NLP for LRL since data is already sparse.
+%%
+
+\section{Representation and Universality}
 
@@
 
-\section{Language Clustering} \label{sec:cluster}
+\section{Language Clustering}
 
@@
 
-\section{Future Work} \label{sec:future}
+\section{Future Work}
 
@@
 
-\section{Conclusion} \label{sec:conclusion}
+\section{Conclusion}
 
-In this paper, we have described the creation of a foundation text for a universal corpus, following the guidelines of Abney and Bird (2010; 2011).
-To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia, and the LCC.
-The resulting corpus is more more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices of [ref - seven dimensions].
-[Give a summary statistic of size.] To demonstrate the utility of this resource, we used the data to perform language clustering, which gave promising results.
-We believe that this is the first corpus of its kind, which we hope will act as a seed corpus for the Human Language Project.
+Summarise what we've done!
 

### 2014-03-03 universalcorpus/crawlandclean/omniglot.py

@@ -84,6 +84,7 @@ def get_phrases(with_mp3=False,testing=False):
       # Get name of language in the particular language.
       langname2 = bs(str(soup.findAll('th')[1])).text
       all_phrases = defaultdict(list)
+      
       # Each <tr>...</tr> is a phrase in the table.
       phrasetable = soup.findAll('tr')
       for phrases in phrasetable:

### 2014-03-04 docs/acl-2014/paper.tex
Merged previous versions.
Added comments about copyright.

--- 2014-03-02 docs/acl-2014/paper.tex
@@
 
-At the time of writing, 7105 living languages are documented in Ethnologue\footnote{http://www.ethnologue.com/}, but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century.
+At the time of writing, 7105 living languages are documented in Ethnologue\footnote{http://www.ethnologue.com}, but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century.
 However, only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
@@
 - the codes are given on their website and look valid to me...]
+\\ \\
+\begin{minipage}{\columnwidth}
+
+\quad 21 a.\quad o lesu mai \\
+\indent \qquad\qquad 2sg return here \\
+\indent \qquad\qquad `\emph{You return here.}' \\
+\centerline{Figure 1: Fijian IGT from ODIN.}
+\end{minipage}
+
+%%
+% Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
+% Also, some of the lines go outside of the column!!
+%%
 
 \begin{minipage}{\columnwidth}
-
-a. \quad o lesu mai \\
-\indent \qquad 2sg return here \\
-
-\indent \qquad `\emph{You return here.}' \\
-\centerline{Figure 1: IGT from ODIN.}
-\end{minipage}
-
-\begin{minipage}{\columnwidth} %Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
 \begin{lstlisting}[language=XML]
@@
   </example>
-</igit>
+</igt>
 \end{lstlisting} 
-\centerline{Figure 2: Fijian IGT from ODIN.}
+\centerline{Figure 2: Fijian IGT in ODIN's XML format.}
 \end{minipage}
 \\ \\
-\noindent The IGTs from a web document usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
+\noindent The IGTs from an academic publication usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
 From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2.
@@
 \smallskip
-\centerline{Figure 3: Yimas IGT from ODIN.}
+\centerline{Figure 3: Yimas IGT in ODIN's XML format.}
 \end{minipage}
@@
 
-The Omniglot website\footnote{htp://www.omniglot.com/} is an online encyclopedia of writing systems and languages.
+The Omniglot website\footnote{htp://www.omniglot.com} is an online encyclopedia of writing systems and languages.
 We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
-
-[The following discussion is far too technical.]
+The urls for these resources are:
 
@@
 \end{itemize}
-
-\noindent The \emph{`Useful foreign phrases'} page contains parallel phrases in embedded within the \texttt{<th><tr>...</tr><th>} tags.
-The English phrase and foreign language phrase are encoded separately with \texttt{<td>...</td>} tags.
-Figure 4 shows an instance of an Omniglot \emph{Useful foreign phrase} page.
-Since the HTML page was designed as WYSIWYG, several pages have non-textual elements within the \texttt{<tr>...</tr>} tags for aesthetics.
-For example, the non-breaking space, \texttt{<td>\&nbsp;</td>} in Fig.
-4; the extracted text was properly converted into plaintext format using the HTMLParser\footnote{http://docs.python.org/2/library/htmlparser.html} module, during which the non textual elements would have been cleaned.
-
-The same process of crawling URLs and cleaning was performed on the \emph{`Tower of Babel'} pages and the only difference was that the texts were embedded in a \texttt{<ol><li>...</li></ol>} tags instead.
-
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<th>
-  ...
-  <tr>
-    <td>I don't understand</td>
-    <td>No appo cumpresu nutta</td>
-    <td>&nbsp;</td>
-  </tr>
-  <tr>
-    <td>I don't understand</td>
-    <td>Non d'isco</td>
-    <td>&nbsp;</td>
-  </tr>
-  ...
-<th>
-\end{lstlisting} 
-\centerline{Figure 4: A \emph{Useful Foreign Phrase} page}
-\end{minipage}
-\smallskip
 
@@
 Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
-Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d/}, which are free of encoding issues.
+Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d}, which are free of encoding issues.
 The first four lines of each file records metadata, and the rest is the translation of the UDHR.
@@
 
-Tapping on the crowd-sourced Wikipedia articles for NLP forms a crucial part of developing data-driven NLP tools and application ({\color{red}{citation needed}}).
-[It might be popular, but it's not crucial...] To automatically download the Wikipedia dumps, we used the wp-download\footnote{https://github.com/babilen/wp-download} tool, adapted to an expanded set of languages.
+At the time of writing, Wikipedia contains 30.8 million articles in 287 languages, which provides a sizeable amount of clean monolingual text in a fairly wide range of languages.
+Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.
+One major issue with using the Wikipedia dump extracting text from a glut of Wikipedia markup.
+To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
+After conversion into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
 
-%To automatically download the Wikipedia dumps, we have scripted the \texttt{wget} with ISO 639-2 language code as a variable (\texttt{\$I}) in a shell script, as such: [This is too technical][Such script has not been implemented]
-
-%\smallskip
-%\noindent \texttt{\$ wget http://download.wikimedia.
-%org/\$Iwiki/latest/itwiki-latest-
-%pages-articles.xml.bz2}
-%\medskip
-
-\noindent One major issue with using the Wikipedia dumps is the sheer size and extracting text from a glut of Wikipedia markup.
-To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor\footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
-After the conversion into textfiles, we used the following [missing information] regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
-
-Wikipedia uses its own set of language codes, most of which are in ISO 639-1 or ISO 639-2/3. We automatically converted all of them into ISO~639-3.
-
-\subsection{Leipzig Corpora Collection}
-
-Do we want to mention this?
+Wikipedia uses its own set of language codes, most of which are in ISO~639-1 or ISO~639-2/3.
+We converted all of them into ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{http://meta.wikimedia.org/wiki/Special\_language\_codes}
 
@@
 
-Creative Commons Licence, etc.
+Corpora copyright remains a grey area and a hobglobin for corpora compilation and distribution (Kilgarriff, 2002)[@Li, please add this reference to the bibliography file].
+Often, free access to language data online does not imply the right to download, retain, remix and redistribute.
 
-We have made all our resources publically available.\footnote{To maintain anonymity, we have removed details, but, if accepted for publication, instructions will be included in the print version of this paper.}
+ODIN refers itself as a database of IGTs that are returned as a result of querying from the source documents.
+As such, only the links to the documents and the appropriate citations and not the copies of the source documents are maintained on the ODIN site.
+However, the ODIN developers have left the users with no specific license/copyrights restriction for readaptation or redistribution.
+
+The Omniglot phrases are free to be used for non-commercial purposes as stated in the FAQ page\footnote{http://www.omniglot.com/faqs.htm} but the rights of the Tower of Babel snippets from different bible sources remains iffy.
+For example, Omniglot cited www.bibles.org as one of its sources for the Tower of Babel chapter but the bibles.org site stated in fine print that `\emph{Except expressly permitted herein, you may not modify, copy, distribute, decompile, disassemble, reverse engineer, create derivative works, or otherwise use or manipulate the Site or any of its content without our prior written consent}'\footnote{http://bibles.org/pages/legal\#copyright}.
+[Ga: actually, about copying information, Omniglot says: \emph{'If you would like to copy or re-use any of the language-related articles, please ask the authors for permission.
+Contact details of the authors can be found at the bottom of the articles in most cases, if they're not there, please contact me, and I'll try to find them for you.'} - I suspect Adger has been given permission for each translation, and I think we would have to ask every single author if we want to re-distribute the data.
+For the tower of babel translations, this could end up a nightmare...]
+
+Although users are free to use, copy, modify, merge, publish and distribute Unicode Data Files, [Ga: Do we have a reference for this?
+Particularly distribution] there is no explicit ownership of the UDHR textfiles from the Unicode website.
+However, the original source of the UDHR documents from the Office of the High Commissioner for Human Rights clearly states that `\emph{None of the materials provided on this web site may be used, reproduced or transmitted, in whole or in part, in any form or by any means, electronic or mechanical, including photocopying, recording or the use of any information storage and retrieval system, except as provided for in the Terms and Conditions of Use of United Nations Web Sites, without permission in writing from the publisher.}'\footnote{http://www.un.org/en/aboutun/copyright/}.
+[Ga: Actually, it's not quite so bad - if we look at the terms and conditions, it says: \emph{'The United Nations grants permission to Users to visit the Site and to download and copy the information, documents and materials (collectively, “Materials”) from the Site for the User’s personal, non-commercial use, without any right to resell or redistribute them or to compile or create derivative works therefrom, subject to the terms and conditions outlined below, and also subject to more specific restrictions that may apply to specific Material within this Site.'}\footnote{http://www.un.org/en/aboutun/terms/} Specifically for the UDHR, it says: \emph{'If UDHR translations or materials are reproduced, users should make reference to this website as a source by providing a link.'}\footnote{http://www.ohchr.org/EN/UDHR/Pages/Introduction.aspx}  So, it looks like we're okay to use the UDHR as we currently are, but if we want to redistribute it, we need to get their permission in writing.]
+
+The Wikipedia texts are licensed under Creative Commons Attribution Share-like\footnote{http://creativecommons.org/licenses/by-sa/3.0/legalcode}, which allows redistribution and adaptation.
+Hence the clean Wikipedia texts we have compiled can be freely redistributed without infringing any copyrights.
+We have made this resource publically available.\footnote{To maintain anonymity, we do not give details of how to access the data, but, if accepted for publication, instructions will be included in the print version of this paper.}
+
+%%
+%To avoid any copyrights infringement, ...
+After surveying the rights for so many days, there is no way to redistribute the corpus unless we jumble up the sentences, but if we do that, the "document" level will be lost and that will not be a "good" resource to do NLP for LRL since data is already sparse.
+%%
 

### 2014-03-16 sugarlike/data_integrity.py
Avoid None sentences in omniglot. Using milk instead of sklearn for clustering.

@@ -34,6 +34,14 @@ def feature_count():
   pass
 
 
-check_data_integrity('odin', remove=False)
-check_data_integrity('omniglot', remove=False)
-check_data_integrity('udhr', remove=False)
+#check_data_integrity('odin', remove=False)
+#check_data_integrity('omniglot', remove=False)
+#check_data_integrity('udhr', remove=False)
+
+import os
+parentddir = os.path.abspath(os.path.join(os.path.dirname(__file__), os.path.pardir))
+import sys; sys.path.append(parentddir)
+from universalcorpus.miniethnologue import ISO2LANG
+from universalcorpus import odin
+print len(set([lang for lang, igts in odin.igts() if lang in ISO2LANG]))
+  

### 2014-03-16 sugarlike/evaluate.py

@@ -58,6 +58,7 @@ def normalise(featurevector, length=1):
   """
   Normalises a feature vector to a specific length.
   Zero vectors are left zero.
+  NOTE: featurevector will be updated with normalized count.
   """
   try:
     norm = length / sqrt(sum([x**2 for x in featurevector.values()]))
@@ -164,16 +165,16 @@ def evaluator(data_source, option="all", model="cosine", tfidf=False, with_word_
     if model == "cosine":
       print "Normalising to unit length..."
       for lang in featureset:
-        normalise(featureset[lang])
+        normalise(featureset[lang]) # Updates featureset.
       if tfidf:
         print "Calculating tf-idf..."
-        tfidfize(featureset)
+        tfidfize(featureset) # Updates featureset.
     
     elif model == "cosine-combined":
       print "Normalising and re-weighting components..."
       for lang in featureset:
         for i in range(6):
-          normalise(featureset[lang][i], weight[i])
+          normalise(featureset[lang][i], weight[i]) # Updates featureset.
     
     print "Evaluating..."
     fold_results = Counter()  # Records the number of times the correct language is at a specific rank 

### 2014-03-16 universalcorpus/crawlandclean/omniglot.py

@@ -213,7 +213,7 @@ def phrases(intarfile=parentddir+'/data/omniglot/omniglotphrases.tar', \
     with codecs.open(infile,'r','utf8') as fin:
       for line in fin.readlines():
         sentence, translation = line.strip().split('\t')
-        if onlysource:
+        if onlysource and sentence:
           yield language, sentence
         else:
           yield language, sentence, translation

### 2014-03-16 universalcorpus/langcluster.py

@@ -2,58 +2,104 @@
 
 from collections import Counter, defaultdict
 from itertools import chain
+from operator import itemgetter
 from threading import Thread
 from Queue import Queue
+import time, os, ast, random
 
 import numpy as np
+import pandas as pd
 from sklearn import cluster
+import milk
 
 from crawlandclean import odin, omniglot, udhr
+from utils import normalize
 
-def word2ngrams(text, n=3, exact=True):
-  """ Convert text into character ngrams. """
-  return [text[i:i+n] for i in range(0, len(text)+1-n)]
+random.seed(0)
+
+def word2ngrams(text, n=3):
+  """ Convert word into character ngrams. """
+  return [text[i:i+n] for i in range(len(text)-n+1)]
+
+def sent2ngrams(text, n=3):
+  """ Convert sentence into character ngrams. """
+  return list(chain(*[word2ngrams(i,n) for i in text.lower().split()]))
 
 def generate_ngrams(data_source):
   ngrams = defaultdict(Counter)
   for lang, sent in globals()[data_source].source_sents():
-    ngrams[lang].update(list(chain(*[word2ngrams(i) 
-                                     for i in sent.lower().split()])))
+    for num in range(2,4):
+      _counter = normalize(Counter(sent2ngrams(sent,num)))
+      if _counter:
+        ##print data_source, num, _counter
+        ngrams[lang]+= _counter
+    x = normalize(Counter(sent.split()))
+    if not x: print sent.split(), Counter(sent.split()) 
+    ngrams[lang]+=normalize(Counter(sent.split()))
+  print data_source
   return ngrams
 
-data_sources = ['odin','omniglot','udhr']
-
-
-# Parallel processing hocus-pocus. 
 def wrapper(func, arg, queue):
   """" Wrapper class for multi-threaded functions """
   queue.put(func(arg))
 
-q1, q2, q3 = Queue(), Queue(), Queue()
-Thread(target=wrapper, args=(generate_ngrams, data_sources[0], q1)).start() 
-Thread(target=wrapper, args=(generate_ngrams, data_sources[1], q2)).start() 
-Thread(target=wrapper, args=(generate_ngrams, data_sources[2], q2)).start()
-odin_ngrams = q1.get(); omniglot_ngrams = q2.get(); udhr_ngrams = q3.get()
+def load_ngram_array():
+  if os.path.exists('language.tag') and \
+  os.path.exists('tocluster.data.npy') and \
[…]

### 2014-03-16 universalcorpus/utils.py

@@ -41,3 +41,18 @@ class Suppress_print(object):
     sys.stdout = StringIO.StringIO()
   def __exit__(self, *args):
     sys.stdout = self.stdout
+    
+def normalize(featurevector, length=1):
+  """
+  Normalises a feature vector to a specific length.
+  Zero vectors are left zero.
+  NOTE: featurevector will be updated with normalized count.
+  """
+  from math import sqrt
+  try:
+    norm = length / sqrt(sum([x*x for x in featurevector.values()]))
+  except ZeroDivisionError:
+    return
+  for feat in featurevector:
+    featurevector[feat] *= norm
+  return featurevector

### 2014-03-16 universalcorpus/crawlandclean/udhr.py
prevents yielding of None.

@@ -101,7 +101,8 @@ def documents(intarfile=parentddir+'/data/udhr/udhr-unicode.tar', \
     with codecs.open(infile,'r','utf8') as fin:
       if bysentence:
         for sentence in fin.readlines():
-          yield language, sentence
+          if sentence:
+            yield language, sentence
       else:
         yield language, fin.read()
         

### 2014-03-16 universalcorpus/langcluster.py

@@ -34,7 +34,7 @@ def generate_ngrams(data_source):
         ##print data_source, num, _counter
         ngrams[lang]+= _counter
     x = normalize(Counter(sent.split()))
-    if not x: print sent.split(), Counter(sent.split()) 
+    if x == None: print data_source, sent.split(), normalize(Counter(sent.split()))
     ngrams[lang]+=normalize(Counter(sent.split()))
   print data_source
   return ngrams

### 2014-03-17 universalcorpus/__init__.py
adding features to ethnologue wrapper

@@ -1,4 +1,4 @@
 # -*- coding: utf-8 -*-
 
 import unicodize, miniethnologue
-from crawlandclean import odin, omniglot, udhr, wikipedia
+from crawlandclean import odin, omniglot, udhr, wikipedia, ethnologue

### 2014-03-17 universalcorpus/crawlandclean/__init__.py

@@ -1,3 +1,3 @@
 # -*- coding: utf-8 -*-
 
-import odin, omniglot, udhr, wikipedia
+import odin, omniglot, udhr, wikipedia,  ethnologue

### 2014-03-17 universalcorpus/crawlandclean/ethnologue.py

@@ -1,13 +1,15 @@
 # -*- coding: utf-8 -*-
 
-import codecs, urllib2, re, os
+import codecs, urllib2, re, os, sys
 from collections import defaultdict
 import cPickle as pickle
+
+parentddir = os.path.abspath(os.path.join(os.path.dirname(__file__), os.path.pardir))
 
 ETHNOLOGUE_DOMAIN = "http://www.ethnologue.com/"
 LANG_FAMILY_TAG = '''<div class="views-field views-field-name">'''
 LANG_TAG = '''<span class="views-field views-field-field-iso-639-3">'''
-ETHNO_DIR = "../data/ethnologue/"
+ETHNO_DIR = parentddir+"/data/ethnologue/"
 
 class Tree(defaultdict):
   """
@@ -106,7 +108,7 @@ def get_language_families():
           isocode = re.findall(r'\[([^]]*)\]',lang)[0]
           geo = re.findall(r'\([^)]*\)',lang)[0].rpartition(" ")[2][:-1]
           ##print isocode, langname , geo
-          lang_fams[langfamily].append((isocode, langname , geo))
+          lang_fams[langfamily].append((isocode, langname , geo, langfamlink))
   return lang_fams
 
 def load_language_families():
@@ -129,15 +131,43 @@ def load_language_families():
   with codecs.open(ETHNO_DIR+'languagefamilies.pk','rb') as fin2: 
     return pickle.load(fin2)
 
-# TODO:
-def get_langinfo(isocode):
-  """ Gets language information from Ethnologue. """
-  page = "http://www.ethnologue.com/language/"+isocode
-
-# TODO:
-def langiso():
-  """ [in]: Language name, [out]: isocode. """
-  pass
-
-langfam = get_language_families()
-print len(langfam)
+def lang_families():
+  from bs4 import BeautifulSoup as bs
+  langfamfile = 'languagefamilies_with_info.pk'
+  if not os.path.exists(ETHNO_DIR+langfamfile):
+    ethno = {'pop':"field field-name-field-population field-type-"+\
+             "text-with-summary field-label-inline clearfix",
+             'altnames':"field field-name-field-alternate-names field-type-"+\
+             "text-long field-label-inline clearfix",
+             'loc':"field field-name-field-region field-type-"+\
+             "text-with-summary field-label-inline clearfix",
+             'status':"field field-name-language-status field-type-"+\
+             "ds field-label-inline clearfix"}
[…]

### 2014-03-17 universalcorpus/crawlandclean/omniglot.py

@@ -214,7 +214,7 @@ def phrases(intarfile=parentddir+'/data/omniglot/omniglotphrases.tar', \
       for line in fin.readlines():
         sentence, translation = line.strip().split('\t')
         if onlysource and sentence:
-          yield language, sentence
+          yield language, sentence.strip()
         else:
           yield language, sentence, translation
 

### 2014-03-17 universalcorpus/crawlandclean/udhr.py

@@ -102,7 +102,7 @@ def documents(intarfile=parentddir+'/data/udhr/udhr-unicode.tar', \
       if bysentence:
         for sentence in fin.readlines():
           if sentence:
-            yield language, sentence
+            yield language, sentence.strip()
       else:
         yield language, fin.read()
         

### 2014-03-17 universalcorpus/langcluster.py

@@ -34,8 +34,7 @@ def generate_ngrams(data_source):
         ##print data_source, num, _counter
         ngrams[lang]+= _counter
     x = normalize(Counter(sent.split()))
-    if x == None: print data_source, sent.split(), normalize(Counter(sent.split()))
-    ngrams[lang]+=normalize(Counter(sent.split()))
+    if x: ngrams[lang]+=x
   print data_source
   return ngrams
 

### 2014-03-17 docs/liling.tex
added section on clustering.

\documentclass[11pt]{article}
\title{\textbf{universal corpus}}
\date{}
\begin{document}
\maketitle

\section{Detecting Similar Languages}

When working on low-resource or endangered languages, computational and documentary linguists face the issue of lack of resources and knowledge about the language. Often, having knowledge about related or similar language(s) provides useful lexical, syntactic or morphological minimal pairs across languages and also helps in bootstrapping language models in NLP (Yarowsky
and Ngai, 2001; Xia and Lewis, 2007).

The task of identifying related/similar language is also known as ``language classification" where (i) languages are mapped onto language families trees depending on their historical ancestry often measured by comparing string distance from a fix lexical list across languages, aka. \emph{genetic classification} or \emph{lexicostatistics} (e.g. Swadesh, 1952; Starostin, 2010) or (ii) languages are grouped by their degrees of similarity with regards to their typological features, aka. \emph{typological classification} (e.g. Daume, 2009; Georgi et al. 2010).

To exemplify the use of the universal corpus for research on low-resource languages, we experimented with automatic detection of similar languages using K-means clustering with character ngrams and word unigrams features. Each language is represented by a vector of character bigrams and trigrams and word unigram features from the universal corpus. 
\newline \newline
\noindent \textbf{SHOW FANCY FIGURE OF CLUSTER}
\newline \newline
\textbf{Evaluate the cluster (quantitatively/qualitatively)}

\end{document}

### 2014-03-17 docs/computEL/clean.tex
Added new directory with individual files, for easier co-authoring. Also pretty major reorganization.

--- 2014-03-02 docs/acl-2014/paper.tex
@@
-\documentclass[11pt]{article}
-\usepackage{acl2014}
-\usepackage{times}
-\usepackage{url}
-\usepackage{latexsym}
+\section{Something about cleaning data, merging data, formatting, standardization}
 
-\bibliographystyle{acl}
+\subsection{Details for individual data sources (or perhaps rather individual topics}
 
-\usepackage[utf8]{inputenc}
+\textcolor{blue}{[Import things from previous section]}
 
-\usepackage{enumitem}
-\usepackage{listings}
-\usepackage{fancyvrb}
-\usepackage{color}
-\definecolor{gray}{rgb}{0.4,0.4,0.4}
-\definecolor{darkblue}{rgb}{0.0,0.0,0.6}
-\definecolor{cyan}{rgb}{0.0,0.6,0.6}
-\definecolor{orange}{rgb}{1,0.5,0}
+\subsection{Universal Corpus and Data Structure} \label{sec:structure}
 
-\lstset{
-  basicstyle=\ttfamily,
-  columns=fullflexible,
-  showstringspaces=false,
-  commentstyle=\color{gray}\upshape
-}
-
-\lstdefinelanguage{XML}
-{
-  morestring=[b]",
-  morestring=[s]{>}{<},
-  morecomment=[s]{<?}{?>},
-  stringstyle=\color{black},
-  identifierstyle=\color{darkblue},
-  keywordstyle=\color{cyan},
-  morekeywords={xmlns,version,type}% list your attributes here
-}
-
-\setlength\titlebox{5cm}
-
-\title{Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages}
-
-\author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
-  Universität des Saarlandes \\
-  Campus, 66123 Saarbrücken, Germany \\
-  {\tt liling.tan@uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-  {\tt susfert@coli.uni-saarland.de}}
-
-\date{}
-
-\begin{document}
-\maketitle
-\begin{abstract}
-The study of endangered languages is constrained by a lack of data.
-Existing corpora collections are limited in the range of languages covered, in standardisation, or in machine-readability.
-This makes the situation even worse for the computational linguist, especially one who would like to take a cross-linguistic or typological approach.
-We first survey existing efforts to compile cross-linguistic resources, then describe our own approach and give an example application - language clustering.
-To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contain data from a large number languages, and converted them into a standardised form consistent with the guidelines set out by \newcite{abney2010universal}.
-The resulting corpus is more easily-accessible and machine-readable than any of the underlying data sources, and represents a significant base corpus for researchers to draw on and add to in the future.
-\end{abstract}
-
-\section{Introduction}
-
-At the time of writing, 7105 living languages are documented in Ethnologue\footnote{http://www.ethnologue.com/}, but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century.
-However, only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
-
-Although there have been some previous attempts to produce cross-linguistic resources, there are few which both cover a wide range of languages, and are also machine-readable.
-We survey existing efforts in section \ref{sec:related}, and discuss their limitations in more detail.
-
-Abney and Bird \shortcite{abney2010universal,abney2011data} posed the grand challenge of building a Universal Corpus, calling it the Human Language Project.
-Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
-They propose a specific data structure, which we describe and discuss in section \ref{sec:structure}.
-We have accepted their challenge, and have begun converting existing resources into a format consistent with their specifications for a universal corpus.
-We have drawn on four [or five?] web sources, cleaning and standardising them as described in section \ref{sec:sources}, to produce a seed corpus for the Human Language Project.
-In sections \ref{sec:stats} and \ref{sec:copyright}, we respectively give a summary of the data contained, and discuss copyright and distribution.
-
-[Note: remove this section if we don't do the work!] We believe the resulting corpus is the first of its kind: large enough and consistent enough to allow language processing on a grand scale.
-In section \ref{sec:cluster}, we give an example application of this corpus: language clustering.
-We use the frequencies of character n-grams and words to estimate the similarity of two languages.
-Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
-Finally, we discuss future work in section \ref{sec:future}, and conclude in section \ref{sec:conclusion}.
-
-\section{Related Work} \label{sec:related}
-
-[There is a lot more stuff to be added here...]
-Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
-For instance, the OPUS corpus \cite{tiedemann2012opus} covers over 90 languages, only 1.27\% of the world's languages; The Leipzig Corpora Collection\footnote{corpora.uni-leipzig.de} \cite{biemann2007leipzig} contains corpora and dictionaries in 230 languages, 3.24\% of the world's languages.
-Even corpora that boast of linguistic diversity include only a small number of language families, e.g.
-the linguistically diverse NTU-Multilingual Corpus \cite{tan2011ntu} covers only 7 out of 136 language families.
-Producing a universal corpus requires the linguistic community to merge existing corpora and provide an ubiquitous access interface.
-To compile the foundation text for the Universal Corpus, we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
-
-\subsection{Web as Corpus}
-
-In recent years, there has been increasing interesting in crawling websites to produce corpora.
-For example, \newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialised corpora, and \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
-\newcite{ferraresi2008ukwac} produced ukWaC, a large-scale English corpus, and \cite{baroni2009wacky} applied a similar method to produce corpora in German, Italian, and French.
-To build better quality web corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection for cleaning data.
-
-However, despite the amount of text they include, such corpora are usually limited in the number of languages represented.
-Some projects have emerged which strive for language diversity, such as the Leipzig Corpora Collection (LCC) \cite{biemann2007leipzig}, and COrpora from the Web (COW) \cite{schaefer2012cow}.
-The LCC currently offers free download of corpora in 117 languages, and dictionaries in many others, bringing the total number of languages up to 230.
-COW includes corpora in [XX] languages.
-These collections are certainly commendable, but they currently still fall short of universality - 230 languages represents only 3.3% of the world's languages [check figure].
-
-\newcite{scannell2007crubadan} describes the Crúbadán Project, an attempt to crawl the web for data in a much larger range of languages, including endangered ones.
-At the time of writing, the number of languages represented has grown to [XXX] However, due to copyright issues, not all of the data is freely available.
-Motivated by similar copyright concerns, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
-
-Unlike the above efforts, we have not performed seed-query web-searching or web-crawling to achieve a balanced resource.
-Instead, we have focused on a small number of sources which contain data in a wide variety of languages, as described in section \ref{sec:sources}.
-However, these sources represent a variety of data types which a linguist might encounter, ranging from single-language text with significant non-linguistic markup (Wikipedia) to structured data with detailed linguistic annotations (ODIN).
-
-\section{Data Structure} \label{sec:structure}
+\textcolor{blue}{[Perhaps this comes elsewhere?]}
 
@@
 If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.
-
-\section{Data Sources} \label{sec:sources}
-
-Although data size matters in general NLP \cite{banko2001scaling} [is this the best reference?], \emph{universality} is the utmost priority for a universal corpus.
-We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured):
-
-\begin{enumerate}
-\item The Online Database of Interlinear Text (ODIN)
-\item The Omniglot website
-\item The Universal Declaration of Human Rights (UDHR)
-\item Wikipedia
-\item The Leipzig Corpora Collection (LCC) [do we include this?]
-\end{enumerate}
-
-In the following subsections, we describe each source, and the processing required to convert the data into a standardised form.
-Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia and LCC) to parallel text (UDHR and Omniglot) to IGTs (ODIN).
-
-\subsection{ODIN}
-
-ODIN (The Online Database of Interlinear Text) is a repository of IGTs extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
-Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation; however, the data requires further processing to bring it in line with the proposed format for a universal corpus.
-
-The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
-Each XML file is saved under a filename that matches the respective language code.
-While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.
-[Huh?
-Should this really happen?
-- the codes are given on their website and look valid to me...]
-
-\begin{minipage}{\columnwidth}
-
-a. \quad o lesu mai \\
-\indent \qquad 2sg return here \\
-
-\indent \qquad `\emph{You return here.}' \\
-\centerline{Figure 1: IGT from ODIN.}
-\end{minipage}
-
-\begin{minipage}{\columnwidth} %Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>21 a. o lesu mai</line>
-    <line>2sg return here</line>
-    <line>`You return here.'</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\centerline{Figure 2: Fijian IGT from ODIN.}
-\end{minipage}
-\\ \\
-\noindent The IGTs from a web document usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
-From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2.
-Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>(69) na-Na-tmi-kwalca-t 
-    Yimas (Foley 1991)</line>
-    <line>3sgA-1sgO-say-rise-PERF
-    </line>
-    <line>`She woke me up' 
-    (by verbal action)</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\smallskip
-\centerline{Figure 3: Yimas IGT from ODIN.}
-\end{minipage}
-\\ \\
-\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
-In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
-In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.
-
-Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness':
-
-[This hasn't been implemented, but I think we could try this: Split source and gloss lines into tokens according to whitespace.
-Check the source has more tokens.
-Count the number of hyphens and equal signs in each token (since they denote morphemes).
-Match the gloss line with a substring of the source line, according to numbers of hyphens and equal signs.
-If there's exactly one match, keep it, and discard tokens at the ends of the line.
-If there's more than one match, look for other punctuation in the initial and final tokens.]
-\\ \\
-\begin{minipage}{\columnwidth}
-\begin{itemize}
-\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
-\begin{itemize}
-\item[(i)]
-\begin{Verbatim}
-^(?\s?\w{1,5}\s*[):.]\s*
-\end{Verbatim}
-\item[(ii)] 
-\begin{Verbatim}
-[\[\(].{1,}[\]\)]
-\end{Verbatim}
-\end{itemize}
-\item \emph{Cleanest}: Only source lines without punctuation.
-\end{itemize}
-\end{minipage}
-\\ \\
-
-\noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
-The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
-
-\subsection{Omniglot}
-
-The Omniglot website\footnote{htp://www.omniglot.com/} is an online encyclopedia of writing systems and languages.
-We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
-
-[The following discussion is far too technical.]
-
-\begin{itemize}[noitemsep]
-\item \url{www.omniglot.com/language/phrases/*} 
-\item \url{www.omniglot.com/babel/*}
-\end{itemize}
-
-\noindent The \emph{`Useful foreign phrases'} page contains parallel phrases in embedded within the \texttt{<th><tr>...</tr><th>} tags.
-The English phrase and foreign language phrase are encoded separately with \texttt{<td>...</td>} tags.
-Figure 4 shows an instance of an Omniglot \emph{Useful foreign phrase} page.
-Since the HTML page was designed as WYSIWYG, several pages have non-textual elements within the \texttt{<tr>...</tr>} tags for aesthetics.
-For example, the non-breaking space, \texttt{<td>\&nbsp;</td>} in Fig.
-4; the extracted text was properly converted into plaintext format using the HTMLParser\footnote{http://docs.python.org/2/library/htmlparser.html} module, during which the non textual elements would have been cleaned.
-
-The same process of crawling URLs and cleaning was performed on the \emph{`Tower of Babel'} pages and the only difference was that the texts were embedded in a \texttt{<ol><li>...</li></ol>} tags instead.
-
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<th>
-  ...
-  <tr>
-    <td>I don't understand</td>
-    <td>No appo cumpresu nutta</td>
-    <td>&nbsp;</td>
-  </tr>
-  <tr>
-    <td>I don't understand</td>
-    <td>Non d'isco</td>
-    <td>&nbsp;</td>
-  </tr>
-  ...
-<th>
-\end{lstlisting} 
-\centerline{Figure 4: A \emph{Useful Foreign Phrase} page}
-\end{minipage}
-\smallskip
-
-\noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
-To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
-
-\subsection{Universal Declaration of Human Rights}
-
-The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages.
-Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
-Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d/}, which are free of encoding issues.
-The first four lines of each file records metadata, and the rest is the translation of the UDHR.
-This dataset was extremely clean, and simply required segmentation into sentences.
-
-\subsection{Wikipedia}
-
-Tapping on the crowd-sourced Wikipedia articles for NLP forms a crucial part of developing data-driven NLP tools and application ({\color{red}{citation needed}}).
-[It might be popular, but it's not crucial...] To automatically download the Wikipedia dumps, we used the wp-download\footnote{https://github.com/babilen/wp-download} tool, adapted to an expanded set of languages.
-
-%To automatically download the Wikipedia dumps, we have scripted the \texttt{wget} with ISO 639-2 language code as a variable (\texttt{\$I}) in a shell script, as such: [This is too technical][Such script has not been implemented]
-
-%\smallskip
-%\noindent \texttt{\$ wget http://download.wikimedia.
-%org/\$Iwiki/latest/itwiki-latest-
-%pages-articles.xml.bz2}
-%\medskip
-
-\noindent One major issue with using the Wikipedia dumps is the sheer size and extracting text from a glut of Wikipedia markup.
-To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor\footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
-After the conversion into textfiles, we used the following [missing information] regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
-
-Wikipedia uses its own set of language codes, most of which are in ISO 639-1 or ISO 639-2/3. We automatically converted all of them into ISO~639-3.
-
-\subsection{Leipzig Corpora Collection}
-
-Do we want to mention this?
-
-\section{Copyright Issues} \label{sec:copyright}
-
-Creative Commons Licence, etc.
-
-We have made all our resources publically available.\footnote{To maintain anonymity, we have removed details, but, if accepted for publication, instructions will be included in the print version of this paper.}
-
-\section{Representation and Universality} \label{sec:stats}
-
-Number of languages, size of corpus, etc.
-
-\section{Language Clustering} \label{sec:cluster}
-
-This would be an example application, if we have time...
-
-\section{Future Work} \label{sec:future}
-
-Specific suggestions on how to work on resources mentioned in section \ref{sec:related}...
-
-\section{Conclusion} \label{sec:conclusion}
-
-In this paper, we have described the creation of a foundation text for a universal corpus, following the guidelines of Abney and Bird (2010; 2011).
-To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia, and the LCC.
-The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}.
-[Give a summary statistic of size.] To demonstrate the utility of this resource, we used the data to perform language clustering, which gave promising results.
-We believe that this is the first corpus of its kind, which we hope will act as a seed corpus for the Human Language Project.
-
-\bibliography{references}
-
-\end{document}

### 2014-03-17 docs/computEL/intro.tex
Added new directory with individual files, for easier co-authoring. Also pretty major reorganization.

--- 2014-03-03 docs/acl-2014/paper.tex
@@
-\documentclass[11pt]{article}
-\usepackage{acl2014}
-\usepackage{times}
-\usepackage{url}
-\usepackage{latexsym}
+\section{Introduction} \label{sec:intro}
 
-\bibliographystyle{acl}
+At the time of writing, 7105 living languages are documented in Ethnologue\footnote{http://www.ethnologue.com}, but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century.
+However, only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
 
-\usepackage[utf8]{inputenc}
+Although there have been some previous attempts to produce cross-linguistic resources, there are few which both cover a wide range of languages, and are also machine-readable.
+We survey existing efforts in section \ref{sec:related}, and discuss their limitations in more detail.
 
-\usepackage{enumitem}
-\usepackage{listings}
-\usepackage{fancyvrb}
-\usepackage{color}
-\definecolor{gray}{rgb}{0.4,0.4,0.4}
-\definecolor{darkblue}{rgb}{0.0,0.0,0.6}
-\definecolor{cyan}{rgb}{0.0,0.6,0.6}
-\definecolor{orange}{rgb}{1,0.5,0}
+Abney and Bird \shortcite{abney2010universal,abney2011data} posed the grand challenge of building a Universal Corpus, calling it the Human Language Project.
+Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
+They propose a specific data structure, which we describe and discuss in section \ref{sec:structure}.
+We have accepted their challenge, and have begun converting existing resources into a format consistent with their specifications for a universal corpus.
+We have drawn on four [or five?] web sources, cleaning and standardising them as described in section \ref{sec:sources}, to produce a seed corpus for the Human Language Project.
+In sections \ref{sec:stats} and \ref{sec:copyright}, we respectively give a summary of the data contained, and discuss copyright and distribution.
+\textcolor{blue}{[needs rewriting following reorganization.]}
 
-\lstset{
-  basicstyle=\ttfamily,
-  columns=fullflexible,
-  showstringspaces=false,
-  commentstyle=\color{gray}\upshape
-}
-
-\lstdefinelanguage{XML}
-{
-  morestring=[b]",
-  morestring=[s]{>}{<},
-  morecomment=[s]{<?}{?>},
-  stringstyle=\color{black},
-  identifierstyle=\color{darkblue},
-  keywordstyle=\color{cyan},
-  morekeywords={xmlns,version,type}% list your attributes here
-}
-
-\setlength\titlebox{5cm}
-
-\title{Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages}
-
-\author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
-  Universität des Saarlandes \\
-  Campus, 66123 Saarbrücken, Germany \\
-  {\tt s9sufert@stud.uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-  {\tt liling.tan@uni-saarland.de}}
-
-\date{}
-
-\begin{document}
-\maketitle
-\begin{abstract}
-The study of endangered languages is limited by a lack of data.
-Existing corpora collections are limited in the range of languages covered, in standardisation, or in machine-readability.
-This makes the situation even worse for the computational linguist, especially one who would like to take a cross-linguistic or typological approach.
-We first survey existing efforts to compile cross-linguistic resources, then describe our own approach and give an example application - language clustering.
-To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contain data from a large number languages, and converted them into a standardised form consistent with the guidelines set out by \newcite{abney2010universal}.
-The resulting corpus is more easily-accessible and machine-readable than any of the underlying data sources, and represents a significant base corpus for researchers to draw on and add to in the future.
-\end{abstract}
-
-\section{Introduction}
-
-\newcite{abney2010universal} posed the grand challenge of building a Universal Corpus, including all of the world's languages, in a consistent structure.
-
-7105 living languages.
-
-\section{Related Work}
-
-[...]
-Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
-For instance, the OPUS corpus \cite{tiedemann2012opus} covers over 90 languages, only 1.27\% of the world's languages; The Leipzig Corpora Collection\footnote{corpora.uni-leipzig.de} \cite{biemann2007leipzig} contains corpora and dictionaries in 230 languages, 3.24\% of the world's languages.
-Even corpora that boast of linguistic diversity include only a small number of language families, e.g.
-the linguistically diverse NTU-Multilingual Corpus \cite{tan2011ntu} covers only 7 out of 136 language families.
-Producing a universal corpus requires the linguistic community to merge existing corpora and provide an ubiquitous access interface.
-To compile the foundation text for the Universal Corpus, we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
-
-\subsection{Web as Corpus}
-
-\newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialized corpora and terms from the web.
-Then, \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
-Later, \newcite{ferraresi2008ukwac} introduced and discussed the notion of compiling Web as Corpus \emph{"properly"} with the inaugural ukWaC using web-crawled data from a list of seed words.
-Thereafter, the Crúbadán corpora was created for a large number of under-resourced languages \cite{scannell2007crubadan}.
-Following which the deWaC (German), itWaC (Italian) and frWaC (French) were compiled \cite{baroni2009wacky}.
-
-Subsequently, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
-To build better quality WaCky corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection in cleaning web corpora.
-
-The initial wave of monolingual WaCky corpora \cite{baroni2009wacky} inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus \cite{dickinson2010korean} and the Feed Corpus \cite{minocha2013feed}.
-Meanwhile, Web-driven corpora compilations also focused on building large  multilingual resources, e.g.
-The Leipzig Corpora Collection (\texttt{LCC}) \cite{biemann2007leipzig}, and the COrpora from the Web (\texttt{COW}) project \cite{schaefer2012cow}.
-
-Unlike previous WaCs built under the Web as Corpus kool ynitiative (WaCky), we did not perform seed query web-searching or web crawling to achieve a balanced resource.
-Instead, the foundation text for uniWaC was built using dedicated site crawling and/or site-dependent cleaning from four data sources, viz.
-(i) \emph{Online Database of Interlinear Text} (ODIN), (ii) \emph{Omniglot} website, (iii) \emph{Universal Declaration of Human Rights} (UDHR) and (iv) \emph{Wikipedia} dumps.
-Our emphasis for the uniWaC is based on based on \textbf{\emph{universality}}; covering as many languages as possible is the first priority.
-
-\section{Data Structure}
-
-\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
-In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
-
-However, we disagree that analysed texts should be accommodated in this way.
-If such a universal corpus is to succeed at all, it must enjoy support from the entire linguistics community, and doing this requires complete theory-neutrality.
-Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
-Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
-More fundamentally, they assume that the corpus should be segmented into words, but \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
-
-It is not within the scope of this paper to resolve these theoretical concerns.
-Instead, we suggest that the data structure for a universal corpus should be even more lightweight than Abney and Bird suggest.
-We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
-To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource language as a convenient surrogate of meaning.
-By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules.\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} The only difference here is that a text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
-The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies cannot be raised.
-We admit that highly detailed annotations, such as in treebanks and dependency banks, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
-
-\section{Data Sources}
-
-Although data size matters in general NLP \cite{banko2001scaling}, \emph{universality} is the utmost priority in NLP for the Universal Corpus initiative (aka Human Language Project; Abney and Bird, 2010; 2011).
-Hence, the uniWaC approach to data collection is based on focused/dedicated crawling unlike previous WaC word/URL list seeding.
-We manually selected sites that contains texts from a large variety of languages and wrote scripts to crawl and clean specific pages from the different data sources.
-The crawling and cleaning was done simultaneously.
-
-The resulting files were saved as plaintext file with \texttt{UTF-8} encoding.
-We used one directory for each data source and one file per language.
-The parallel data (from ODIN and Omniglot) are saved as tab-delimited files, one line per sentence, one column per language.
-The monolingual data (from UDHR and Wikipedia) are saved as newline delimited plaintext file, each line denoting a paragraph or sentence and each document separated by an empty line.
-The files are saved with the following naming convention to allow easy access, the filename contains the data source and its ISO 639-3 code, e.g.
-\texttt{odin-iii.txt} contains the \emph{ODIN} data for the \emph{Nuosu} language.
-[File structure is not relevant.]
-
-API which allows access according to Abney and Bird.
-However, maintaining a plain text file ensures that the data will be easily accessible, even if the surrounding infrastructure changes.
-[Seven dimensions]
-
-Our resulting corpus runs the full gamut of texts outlined by Abney and Bird, from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN). ...
-
-In the follow subsections, we describe the preprocessing required to convert each data source into a standardised form.
-
-\subsection{ODIN}
-
-ODIN (The Online Database of Interlinear Text) is a repository of IGTs extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
-Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation; however, the data requires further processing to bring it in line with the proposed format for a universal corpus.
-
-The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
-Each XML file is saved under a filename that matches the respective language code.
-While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.
-[Huh?
-Should this really happen?
-- the codes are given on their website and look valid to me...]
-
-\begin{minipage}{\columnwidth}
-
-a. \quad o lesu mai \\
-\indent \qquad 2sg return here \\
-
-\indent \qquad `\emph{You return here.}' \\
-\centerline{Figure 1: IGT from ODIN.}
-\end{minipage}
-
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>21 a. o lesu mai</line>
-    <line>2sg return here</line>
-    <line>`You return here.'</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\centerline{Figure 2: Fijian IGT from ODIN.}
-\end{minipage}
-\\ \\
-\noindent The IGTs from a web document usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
-From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2.
-Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>(69) na-Na-tmi-kwalca-t 
-    Yimas (Foley 1991)</line>
-    <line>3sgA-1sgO-say-rise-PERF
-    </line>
-    <line>`She woke me up' 
-    (by verbal action)</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\smallskip
-\centerline{Figure 3: Yimas IGT from ODIN.}
-\end{minipage}
-\\ \\
-\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
-In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
-In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.
-
-Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness': [Have we considered that the number of elements in the source line and gloss line should match...?]
-\\ \\
-\begin{minipage}{\columnwidth}
-\begin{itemize}
-\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
-\begin{itemize}
-\item[(i)]
-\begin{Verbatim}
-^(?\s?\w{1,5}\s*[):.]\s*
-\end{Verbatim}
-\item[(ii)] 
-\begin{Verbatim}
-[\[\(].{1,}[\]\)]
-\end{Verbatim}
-\end{itemize}
-\item \emph{Cleanest}: Only source lines without punctuation.
-\end{itemize}
-\end{minipage}
-\\ \\
-
-\noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
-The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
-
-\subsection{Omniglot}
-
-The Omniglot website\footnote{htp://www.omniglot.com/} is an online encyclopedia of writing systems and languages.
-We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
-
-[The following discussion is far too technical.]
-
-\begin{itemize}[noitemsep]
-\item \url{www.omniglot.com/language/phrases/*} 
-\item \url{www.omniglot.com/babel/*}
-\end{itemize}
-
-\noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
-To resolve this issue, we converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
-
-\subsection{Universal Declaration of Human Rights}
-
-The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages.
-Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
-Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d/}, which are free of encoding issues.
-The first four lines of each file records metadata, and the rest is the translation of the UDHR.
-
-\subsection{Wikipedia}
-
-Wikipedia contains 30.8 million articles in 287 languages and the text dump can be downloaded from \url{www.dumps.wikimedia.org}.
-\noindent One major issue with using the Wikipedia dump is the sheer size and extracting text from a glut of Wikipedia markup.
-To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
-After converting them into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
-
-\section{Copyright Issues}
-
-Corpora copyright remains a grey area and a hobglobin for corpora compilation and distribution (Kilgariff, 2002).
-Often, free access to language data online does not imply the right to download, retain, remix and redistribute.
-
-ODIN refers itself as a database of IGTs that are returned as a result of querying from the source documents.
-As such, only the links to the documents and the appropriate citations and not the copies of the source documents are maintained on the ODIN site (Lewis and Xia, 2010).
-However, the ODIN developers have left the users with no specific license/copyrights restriction for readaptation or redistribution.
-
-The Omniglot phrases are free to be used for non-commercial purposes as stated in the FAQ page but the rights of the Tower of Babel snippets from different bible sources remains iffy.
-For example, Omniglot cited www.bibles.org as one of its sources for the Tower of Babel chapter but the bibles.org site stated in fine print that `\emph{Except expressly permitted herein, you may not modify, copy, distribute, decompile, disassemble, reverse engineer, create derivative works, or otherwise use or manipulate the Site or any of its content without our prior written consent}'\footnote{http://bibles.org/pages/legal\#copyright}.
-
-The UDHR copyrights remains questionable.
-Although users are free to use, copy, modify, merge, publish and distribute Unicode Data Files, there is no explicit ownership of the UDHR textfiles from the Unicode website.
-However, the original source of the UDHR documents from the Office of the High Commissioner for Human Rights clearly states that `\emph{None of the materials provided on this web site may be used, reproduced or transmitted, in whole or in part}'\footnote{http://www.un.org/en/aboutun/copyright/}.
-
-The Wikipedia texts are licensed under Creative Commons Attribution Share-like\footnote{http://creativecommons.org/licenses/by-sa/3.0/legalcode}, which allows redistribution and adaptation.
-Hence the clean Wikipedia texts we have compiled can be freely redistributed without infringing any copyrights.
-
-%%
-%To avoid any copyrights infringement, ...
-After surveying the rights for so many days, there is no way to redistribute the corpus unless we jumble up the sentences, but if we do that, the "document" level will be lost and that will not be a "good" resource to do NLP for LRL since data is already sparse.
-%%
-
-\section{Representation and Universality}
-
-Number of languages, size of corpus, etc.
-
-\section{Language Clustering}
-
-This would be an example application, if we have time...
-
-\section{Future Work}
-
-Other corpora we could work on...
-
-\section{Conclusion}
-
-Summarise what we've done!
-
-\bibliography{references}
-
-\end{document}
+[Note: remove this section if we don't do the work!] We believe the resulting corpus is the first of its kind: large enough and consistent enough to allow language processing on a grand scale.
+In section \ref{sec:cluster}, we give an example application of this corpus: language clustering.
+We use the frequencies of character n-grams and words to estimate the similarity of two languages.
+Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
+Finally, we discuss future work in section \ref{sec:future}, and conclude in section \ref{sec:conclusion}.

### 2014-03-17 docs/computEL/paper.tex
Added new directory with individual files, for easier co-authoring. Also pretty major reorganization.

--- 2014-03-03 docs/acl-2014/paper.tex
@@
 
-\section{Introduction}
-
-\newcite{abney2010universal} posed the grand challenge of building a Universal Corpus, including all of the world's languages, in a consistent structure.
-
-7105 living languages.
-
-\section{Related Work}
-
-[...]
-Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
-For instance, the OPUS corpus \cite{tiedemann2012opus} covers over 90 languages, only 1.27\% of the world's languages; The Leipzig Corpora Collection\footnote{corpora.uni-leipzig.de} \cite{biemann2007leipzig} contains corpora and dictionaries in 230 languages, 3.24\% of the world's languages.
-Even corpora that boast of linguistic diversity include only a small number of language families, e.g.
-the linguistically diverse NTU-Multilingual Corpus \cite{tan2011ntu} covers only 7 out of 136 language families.
-Producing a universal corpus requires the linguistic community to merge existing corpora and provide an ubiquitous access interface.
-To compile the foundation text for the Universal Corpus, we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
-
-\subsection{Web as Corpus}
-
-\newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialized corpora and terms from the web.
-Then, \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
-Later, \newcite{ferraresi2008ukwac} introduced and discussed the notion of compiling Web as Corpus \emph{"properly"} with the inaugural ukWaC using web-crawled data from a list of seed words.
-Thereafter, the Crúbadán corpora was created for a large number of under-resourced languages \cite{scannell2007crubadan}.
-Following which the deWaC (German), itWaC (Italian) and frWaC (French) were compiled \cite{baroni2009wacky}.
-
-Subsequently, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
-To build better quality WaCky corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection in cleaning web corpora.
-
-The initial wave of monolingual WaCky corpora \cite{baroni2009wacky} inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus \cite{dickinson2010korean} and the Feed Corpus \cite{minocha2013feed}.
-Meanwhile, Web-driven corpora compilations also focused on building large  multilingual resources, e.g.
-The Leipzig Corpora Collection (\texttt{LCC}) \cite{biemann2007leipzig}, and the COrpora from the Web (\texttt{COW}) project \cite{schaefer2012cow}.
-
-Unlike previous WaCs built under the Web as Corpus kool ynitiative (WaCky), we did not perform seed query web-searching or web crawling to achieve a balanced resource.
-Instead, the foundation text for uniWaC was built using dedicated site crawling and/or site-dependent cleaning from four data sources, viz.
-(i) \emph{Online Database of Interlinear Text} (ODIN), (ii) \emph{Omniglot} website, (iii) \emph{Universal Declaration of Human Rights} (UDHR) and (iv) \emph{Wikipedia} dumps.
-Our emphasis for the uniWaC is based on based on \textbf{\emph{universality}}; covering as many languages as possible is the first priority.
-
-\section{Data Structure}
-
-\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
-In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
-
-However, we disagree that analysed texts should be accommodated in this way.
-If such a universal corpus is to succeed at all, it must enjoy support from the entire linguistics community, and doing this requires complete theory-neutrality.
-Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
-Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
-More fundamentally, they assume that the corpus should be segmented into words, but \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
-
-It is not within the scope of this paper to resolve these theoretical concerns.
-Instead, we suggest that the data structure for a universal corpus should be even more lightweight than Abney and Bird suggest.
-We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
-To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource language as a convenient surrogate of meaning.
-By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules.\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} The only difference here is that a text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
-The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies cannot be raised.
-We admit that highly detailed annotations, such as in treebanks and dependency banks, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
-
-\section{Data Sources}
-
-Although data size matters in general NLP \cite{banko2001scaling}, \emph{universality} is the utmost priority in NLP for the Universal Corpus initiative (aka Human Language Project; Abney and Bird, 2010; 2011).
-Hence, the uniWaC approach to data collection is based on focused/dedicated crawling unlike previous WaC word/URL list seeding.
-We manually selected sites that contains texts from a large variety of languages and wrote scripts to crawl and clean specific pages from the different data sources.
-The crawling and cleaning was done simultaneously.
-
-The resulting files were saved as plaintext file with \texttt{UTF-8} encoding.
-We used one directory for each data source and one file per language.
-The parallel data (from ODIN and Omniglot) are saved as tab-delimited files, one line per sentence, one column per language.
-The monolingual data (from UDHR and Wikipedia) are saved as newline delimited plaintext file, each line denoting a paragraph or sentence and each document separated by an empty line.
-The files are saved with the following naming convention to allow easy access, the filename contains the data source and its ISO 639-3 code, e.g.
-\texttt{odin-iii.txt} contains the \emph{ODIN} data for the \emph{Nuosu} language.
-[File structure is not relevant.]
-
-API which allows access according to Abney and Bird.
-However, maintaining a plain text file ensures that the data will be easily accessible, even if the surrounding infrastructure changes.
-[Seven dimensions]
-
-Our resulting corpus runs the full gamut of texts outlined by Abney and Bird, from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN). ...
-
-In the follow subsections, we describe the preprocessing required to convert each data source into a standardised form.
-
-\subsection{ODIN}
-
-ODIN (The Online Database of Interlinear Text) is a repository of IGTs extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
-Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation; however, the data requires further processing to bring it in line with the proposed format for a universal corpus.
-
-The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
-Each XML file is saved under a filename that matches the respective language code.
-While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.
-[Huh?
-Should this really happen?
-- the codes are given on their website and look valid to me...]
-
-\begin{minipage}{\columnwidth}
-
-a. \quad o lesu mai \\
-\indent \qquad 2sg return here \\
-
-\indent \qquad `\emph{You return here.}' \\
-\centerline{Figure 1: IGT from ODIN.}
-\end{minipage}
-
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>21 a. o lesu mai</line>
-    <line>2sg return here</line>
-    <line>`You return here.'</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\centerline{Figure 2: Fijian IGT from ODIN.}
-\end{minipage}
-\\ \\
-\noindent The IGTs from a web document usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
-From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2.
-Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>(69) na-Na-tmi-kwalca-t 
-    Yimas (Foley 1991)</line>
-    <line>3sgA-1sgO-say-rise-PERF
-    </line>
-    <line>`She woke me up' 
-    (by verbal action)</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\smallskip
-\centerline{Figure 3: Yimas IGT from ODIN.}
-\end{minipage}
-\\ \\
-\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
-In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
-In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.
-
-Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness': [Have we considered that the number of elements in the source line and gloss line should match...?]
-\\ \\
-\begin{minipage}{\columnwidth}
-\begin{itemize}
-\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
-\begin{itemize}
-\item[(i)]
-\begin{Verbatim}
-^(?\s?\w{1,5}\s*[):.]\s*
-\end{Verbatim}
-\item[(ii)] 
-\begin{Verbatim}
-[\[\(].{1,}[\]\)]
-\end{Verbatim}
-\end{itemize}
-\item \emph{Cleanest}: Only source lines without punctuation.
-\end{itemize}
-\end{minipage}
-\\ \\
-
-\noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
-The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
-
-\subsection{Omniglot}
-
-The Omniglot website\footnote{htp://www.omniglot.com/} is an online encyclopedia of writing systems and languages.
-We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
-
-[The following discussion is far too technical.]
-
-\begin{itemize}[noitemsep]
-\item \url{www.omniglot.com/language/phrases/*} 
-\item \url{www.omniglot.com/babel/*}
-\end{itemize}
-
-\noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
-To resolve this issue, we converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
-
-\subsection{Universal Declaration of Human Rights}
-
-The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages.
-Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
-Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d/}, which are free of encoding issues.
-The first four lines of each file records metadata, and the rest is the translation of the UDHR.
-
-\subsection{Wikipedia}
-
-Wikipedia contains 30.8 million articles in 287 languages and the text dump can be downloaded from \url{www.dumps.wikimedia.org}.
-\noindent One major issue with using the Wikipedia dump is the sheer size and extracting text from a glut of Wikipedia markup.
-To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
-After converting them into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
-
-\section{Copyright Issues}
-
-Corpora copyright remains a grey area and a hobglobin for corpora compilation and distribution (Kilgariff, 2002).
-Often, free access to language data online does not imply the right to download, retain, remix and redistribute.
-
-ODIN refers itself as a database of IGTs that are returned as a result of querying from the source documents.
-As such, only the links to the documents and the appropriate citations and not the copies of the source documents are maintained on the ODIN site (Lewis and Xia, 2010).
-However, the ODIN developers have left the users with no specific license/copyrights restriction for readaptation or redistribution.
-
-The Omniglot phrases are free to be used for non-commercial purposes as stated in the FAQ page but the rights of the Tower of Babel snippets from different bible sources remains iffy.
-For example, Omniglot cited www.bibles.org as one of its sources for the Tower of Babel chapter but the bibles.org site stated in fine print that `\emph{Except expressly permitted herein, you may not modify, copy, distribute, decompile, disassemble, reverse engineer, create derivative works, or otherwise use or manipulate the Site or any of its content without our prior written consent}'\footnote{http://bibles.org/pages/legal\#copyright}.
-
-The UDHR copyrights remains questionable.
-Although users are free to use, copy, modify, merge, publish and distribute Unicode Data Files, there is no explicit ownership of the UDHR textfiles from the Unicode website.
-However, the original source of the UDHR documents from the Office of the High Commissioner for Human Rights clearly states that `\emph{None of the materials provided on this web site may be used, reproduced or transmitted, in whole or in part}'\footnote{http://www.un.org/en/aboutun/copyright/}.
-
-The Wikipedia texts are licensed under Creative Commons Attribution Share-like\footnote{http://creativecommons.org/licenses/by-sa/3.0/legalcode}, which allows redistribution and adaptation.
-Hence the clean Wikipedia texts we have compiled can be freely redistributed without infringing any copyrights.
-
-%%
-%To avoid any copyrights infringement, ...
-After surveying the rights for so many days, there is no way to redistribute the corpus unless we jumble up the sentences, but if we do that, the "document" level will be lost and that will not be a "good" resource to do NLP for LRL since data is already sparse.
-%%
-
-\section{Representation and Universality}
-
-Number of languages, size of corpus, etc.
-
-\section{Language Clustering}
-
-This would be an example application, if we have time...
-
-\section{Future Work}
-
-Other corpora we could work on...
-
-\section{Conclusion}
-
-Summarise what we've done!
+\input{intro}
+\input{relwork}
+\input{data}
+\input{clean}
+\input{cluster}
+\input{corpus}
+\input{concl}
 

### 2014-03-17 docs/computEL/relwork.tex
Added new directory with individual files, for easier co-authoring. Also pretty major reorganization.

--- 2014-03-03 docs/acl-2014/paper.tex
@@
-\documentclass[11pt]{article}
-\usepackage{acl2014}
-\usepackage{times}
-\usepackage{url}
-\usepackage{latexsym}
+\section{Related Work} \label{sec:related}
 
-\bibliographystyle{acl}
-
-\usepackage[utf8]{inputenc}
-
-\usepackage{enumitem}
-\usepackage{listings}
-\usepackage{fancyvrb}
-\usepackage{color}
-\definecolor{gray}{rgb}{0.4,0.4,0.4}
-\definecolor{darkblue}{rgb}{0.0,0.0,0.6}
-\definecolor{cyan}{rgb}{0.0,0.6,0.6}
-\definecolor{orange}{rgb}{1,0.5,0}
-
-\lstset{
-  basicstyle=\ttfamily,
-  columns=fullflexible,
-  showstringspaces=false,
-  commentstyle=\color{gray}\upshape
-}
-
-\lstdefinelanguage{XML}
-{
-  morestring=[b]",
-  morestring=[s]{>}{<},
-  morecomment=[s]{<?}{?>},
-  stringstyle=\color{black},
-  identifierstyle=\color{darkblue},
-  keywordstyle=\color{cyan},
-  morekeywords={xmlns,version,type}% list your attributes here
-}
-
-\setlength\titlebox{5cm}
-
-\title{Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages}
-
-\author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
-  Universität des Saarlandes \\
-  Campus, 66123 Saarbrücken, Germany \\
-  {\tt s9sufert@stud.uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-  {\tt liling.tan@uni-saarland.de}}
-
-\date{}
-
-\begin{document}
-\maketitle
-\begin{abstract}
-The study of endangered languages is limited by a lack of data.
-Existing corpora collections are limited in the range of languages covered, in standardisation, or in machine-readability.
-This makes the situation even worse for the computational linguist, especially one who would like to take a cross-linguistic or typological approach.
-We first survey existing efforts to compile cross-linguistic resources, then describe our own approach and give an example application - language clustering.
-To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contain data from a large number languages, and converted them into a standardised form consistent with the guidelines set out by \newcite{abney2010universal}.
-The resulting corpus is more easily-accessible and machine-readable than any of the underlying data sources, and represents a significant base corpus for researchers to draw on and add to in the future.
-\end{abstract}
-
-\section{Introduction}
-
-\newcite{abney2010universal} posed the grand challenge of building a Universal Corpus, including all of the world's languages, in a consistent structure.
-
-7105 living languages.
-
-\section{Related Work}
-
-[...]
+[There is a lot more stuff to be added here...]
 Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
@@
 To compile the foundation text for the Universal Corpus, we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
-
-\subsection{Web as Corpus}
-
-\newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialized corpora and terms from the web.
-Then, \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
-Later, \newcite{ferraresi2008ukwac} introduced and discussed the notion of compiling Web as Corpus \emph{"properly"} with the inaugural ukWaC using web-crawled data from a list of seed words.
-Thereafter, the Crúbadán corpora was created for a large number of under-resourced languages \cite{scannell2007crubadan}.
-Following which the deWaC (German), itWaC (Italian) and frWaC (French) were compiled \cite{baroni2009wacky}.
-
-Subsequently, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
-To build better quality WaCky corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection in cleaning web corpora.
-
-The initial wave of monolingual WaCky corpora \cite{baroni2009wacky} inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus \cite{dickinson2010korean} and the Feed Corpus \cite{minocha2013feed}.
-Meanwhile, Web-driven corpora compilations also focused on building large  multilingual resources, e.g.
-The Leipzig Corpora Collection (\texttt{LCC}) \cite{biemann2007leipzig}, and the COrpora from the Web (\texttt{COW}) project \cite{schaefer2012cow}.
-
-Unlike previous WaCs built under the Web as Corpus kool ynitiative (WaCky), we did not perform seed query web-searching or web crawling to achieve a balanced resource.
-Instead, the foundation text for uniWaC was built using dedicated site crawling and/or site-dependent cleaning from four data sources, viz.
-(i) \emph{Online Database of Interlinear Text} (ODIN), (ii) \emph{Omniglot} website, (iii) \emph{Universal Declaration of Human Rights} (UDHR) and (iv) \emph{Wikipedia} dumps.
-Our emphasis for the uniWaC is based on based on \textbf{\emph{universality}}; covering as many languages as possible is the first priority.
-
-\section{Data Structure}
-
-\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
-In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
-
-However, we disagree that analysed texts should be accommodated in this way.
-If such a universal corpus is to succeed at all, it must enjoy support from the entire linguistics community, and doing this requires complete theory-neutrality.
-Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
-Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
-More fundamentally, they assume that the corpus should be segmented into words, but \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
-
-It is not within the scope of this paper to resolve these theoretical concerns.
-Instead, we suggest that the data structure for a universal corpus should be even more lightweight than Abney and Bird suggest.
-We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
-To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource language as a convenient surrogate of meaning.
-By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules.\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} The only difference here is that a text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
-The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies cannot be raised.
-We admit that highly detailed annotations, such as in treebanks and dependency banks, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
-
-\section{Data Sources}
-
-Although data size matters in general NLP \cite{banko2001scaling}, \emph{universality} is the utmost priority in NLP for the Universal Corpus initiative (aka Human Language Project; Abney and Bird, 2010; 2011).
-Hence, the uniWaC approach to data collection is based on focused/dedicated crawling unlike previous WaC word/URL list seeding.
-We manually selected sites that contains texts from a large variety of languages and wrote scripts to crawl and clean specific pages from the different data sources.
-The crawling and cleaning was done simultaneously.
-
-The resulting files were saved as plaintext file with \texttt{UTF-8} encoding.
-We used one directory for each data source and one file per language.
-The parallel data (from ODIN and Omniglot) are saved as tab-delimited files, one line per sentence, one column per language.
-The monolingual data (from UDHR and Wikipedia) are saved as newline delimited plaintext file, each line denoting a paragraph or sentence and each document separated by an empty line.
-The files are saved with the following naming convention to allow easy access, the filename contains the data source and its ISO 639-3 code, e.g.
-\texttt{odin-iii.txt} contains the \emph{ODIN} data for the \emph{Nuosu} language.
-[File structure is not relevant.]
-
-API which allows access according to Abney and Bird.
-However, maintaining a plain text file ensures that the data will be easily accessible, even if the surrounding infrastructure changes.
-[Seven dimensions]
-
-Our resulting corpus runs the full gamut of texts outlined by Abney and Bird, from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN). ...
-
-In the follow subsections, we describe the preprocessing required to convert each data source into a standardised form.
-
-\subsection{ODIN}
-
-ODIN (The Online Database of Interlinear Text) is a repository of IGTs extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
-Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation; however, the data requires further processing to bring it in line with the proposed format for a universal corpus.
-
-The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
-Each XML file is saved under a filename that matches the respective language code.
-While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.
-[Huh?
-Should this really happen?
-- the codes are given on their website and look valid to me...]
-
-\begin{minipage}{\columnwidth}
-
-a. \quad o lesu mai \\
-\indent \qquad 2sg return here \\
-
-\indent \qquad `\emph{You return here.}' \\
-\centerline{Figure 1: IGT from ODIN.}
-\end{minipage}
-
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>21 a. o lesu mai</line>
-    <line>2sg return here</line>
-    <line>`You return here.'</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\centerline{Figure 2: Fijian IGT from ODIN.}
-\end{minipage}
-\\ \\
-\noindent The IGTs from a web document usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
-From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2.
-Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>(69) na-Na-tmi-kwalca-t 
-    Yimas (Foley 1991)</line>
-    <line>3sgA-1sgO-say-rise-PERF
-    </line>
-    <line>`She woke me up' 
-    (by verbal action)</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\smallskip
-\centerline{Figure 3: Yimas IGT from ODIN.}
-\end{minipage}
-\\ \\
-\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
-In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
-In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.
-
-Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness': [Have we considered that the number of elements in the source line and gloss line should match...?]
-\\ \\
-\begin{minipage}{\columnwidth}
-\begin{itemize}
-\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
-\begin{itemize}
-\item[(i)]
-\begin{Verbatim}
-^(?\s?\w{1,5}\s*[):.]\s*
-\end{Verbatim}
-\item[(ii)] 
-\begin{Verbatim}
-[\[\(].{1,}[\]\)]
-\end{Verbatim}
-\end{itemize}
-\item \emph{Cleanest}: Only source lines without punctuation.
-\end{itemize}
-\end{minipage}
-\\ \\
-
-\noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
-The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
-
-\subsection{Omniglot}
-
-The Omniglot website\footnote{htp://www.omniglot.com/} is an online encyclopedia of writing systems and languages.
-We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
-
-[The following discussion is far too technical.]
-
-\begin{itemize}[noitemsep]
-\item \url{www.omniglot.com/language/phrases/*} 
-\item \url{www.omniglot.com/babel/*}
-\end{itemize}
-
-\noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
-To resolve this issue, we converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
-
-\subsection{Universal Declaration of Human Rights}
-
-The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages.
-Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
-Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d/}, which are free of encoding issues.
-The first four lines of each file records metadata, and the rest is the translation of the UDHR.
-
-\subsection{Wikipedia}
-
-Wikipedia contains 30.8 million articles in 287 languages and the text dump can be downloaded from \url{www.dumps.wikimedia.org}.
-\noindent One major issue with using the Wikipedia dump is the sheer size and extracting text from a glut of Wikipedia markup.
-To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
-After converting them into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
-
-\section{Copyright Issues}
-
-Corpora copyright remains a grey area and a hobglobin for corpora compilation and distribution (Kilgariff, 2002).
-Often, free access to language data online does not imply the right to download, retain, remix and redistribute.
-
-ODIN refers itself as a database of IGTs that are returned as a result of querying from the source documents.
-As such, only the links to the documents and the appropriate citations and not the copies of the source documents are maintained on the ODIN site (Lewis and Xia, 2010).
-However, the ODIN developers have left the users with no specific license/copyrights restriction for readaptation or redistribution.
-
-The Omniglot phrases are free to be used for non-commercial purposes as stated in the FAQ page but the rights of the Tower of Babel snippets from different bible sources remains iffy.
-For example, Omniglot cited www.bibles.org as one of its sources for the Tower of Babel chapter but the bibles.org site stated in fine print that `\emph{Except expressly permitted herein, you may not modify, copy, distribute, decompile, disassemble, reverse engineer, create derivative works, or otherwise use or manipulate the Site or any of its content without our prior written consent}'\footnote{http://bibles.org/pages/legal\#copyright}.
-
-The UDHR copyrights remains questionable.
-Although users are free to use, copy, modify, merge, publish and distribute Unicode Data Files, there is no explicit ownership of the UDHR textfiles from the Unicode website.
-However, the original source of the UDHR documents from the Office of the High Commissioner for Human Rights clearly states that `\emph{None of the materials provided on this web site may be used, reproduced or transmitted, in whole or in part}'\footnote{http://www.un.org/en/aboutun/copyright/}.
-
-The Wikipedia texts are licensed under Creative Commons Attribution Share-like\footnote{http://creativecommons.org/licenses/by-sa/3.0/legalcode}, which allows redistribution and adaptation.
-Hence the clean Wikipedia texts we have compiled can be freely redistributed without infringing any copyrights.
-
-%%
-%To avoid any copyrights infringement, ...
-After surveying the rights for so many days, there is no way to redistribute the corpus unless we jumble up the sentences, but if we do that, the "document" level will be lost and that will not be a "good" resource to do NLP for LRL since data is already sparse.
-%%
-
-\section{Representation and Universality}
-
-Number of languages, size of corpus, etc.
-
-\section{Language Clustering}
-
-This would be an example application, if we have time...
-
-\section{Future Work}
-
-Other corpora we could work on...
-
-\section{Conclusion}
-
-Summarise what we've done!
-
-\bibliography{references}
-
-\end{document}

### 2014-03-17 universalcorpus/crawlandclean/ethnologue.py
added language families pickle.

@@ -134,6 +134,7 @@ def load_language_families():
 def lang_families():
   from bs4 import BeautifulSoup as bs
   langfamfile = 'languagefamilies_with_info.pk'
+  fout_tmp = codecs.open('langfam.tmp','w','utf8')
   if not os.path.exists(ETHNO_DIR+langfamfile):
     ethno = {'pop':"field field-name-field-population field-type-"+\
              "text-with-summary field-label-inline clearfix",
@@ -149,7 +150,10 @@ def lang_families():
       for j in _lf[i]:
         site = bs(urllib2.urlopen('https://www.ethnologue.com/language/'+j).read().decode('utf8'))
         name = site.find('title').text.replace(' | Ethnologue','').strip()
-        pop = site.find(attrs={"class": ethno['pop']}).find(attrs={"class": 'even'}).text.strip()
+        try:
+          pop = site.find(attrs={"class": ethno['pop']}).find(attrs={"class": 'even'}).text.strip()
+        except AttributeError:
+          pop = None  
         status = site.find(attrs={"class": ethno['status']}).find(attrs={"class": 'even'}).text.strip()
         try:
           loc = site.find(attrs={"class": ethno['loc']}).find(attrs={"class": 'even'}).text.strip()
@@ -159,15 +163,83 @@ def lang_families():
           altnames = site.find(attrs={"class": ethno['altnames']}).find(attrs={"class": 'even'}).text.strip()
         except AttributeError:
           altnames = None
-        lf[i].append((i,name, pop, altnames, loc, status))
-        print j, i, status
+        lf[i].append((j, name, pop, altnames, loc, status))
+        print>>fout_tmp, "\t".join(map(str,[j, i, name, pop, altnames, loc, status]))
     with codecs.open(ETHNO_DIR+langfamfile,'wb') as fout:
       pickle.dump(lf, fout)
   
   with codecs.open(ETHNO_DIR+langfamfile,'wb') as fin2:
     return pickle.load(lf, fin2)
   
-lang_families()
-    
-    
+##lang_families()
+
+def download_lang_families():
+  _lf = {k:[j[0] for j in v] for k,v in load_language_families().items()}
+  for i in _lf:
+    for j in _lf[i]:
+      if os.path.exists(parentddir+"/data/ethnologue/languages/"+j+'.ethno'):
+        print i
+      try:
+        site = urllib2.urlopen('https://www.ethnologue.com/language/' \
+                                  +j).read().decode('utf8')
[…]

### 2014-03-17 universalcorpus/langcluster.py

@@ -27,9 +27,9 @@ def sent2ngrams(text, n=3):
 
 def generate_ngrams(data_source):
   ngrams = defaultdict(Counter)
-  for lang, sent in globals()[data_source].source_sents():
-    for num in range(2,4):
-      _counter = normalize(Counter(sent2ngrams(sent,num)))
+  for num in range(2,4):
+    for lang, sent in globals()[data_source].source_sents():
+      _counter = normalize(Counter(sent2ngrams(sent,num))) ## TODO: normalize all ngrams in the 
       if _counter:
         ##print data_source, num, _counter
         ngrams[lang]+= _counter
@@ -76,7 +76,8 @@ def load_ngram_array():
       languages.append(lang)
       x = [ng[i] for i in all_features]
       data.append(x)
-    data = pd.lib.to_object_array(data).astype(float)
+    
+    data = pd.lib.to_object_array(data).astype(float) # converts counters to array.
     print>>open('language.tag','w'), languages
     np.save('tocluster.data',data)
     print>>codecs.open('all_features','w'), " ".join(all_features)
@@ -102,3 +103,14 @@ print set(clusters), len(set(clusters))
 print clusters
 print languages
 print all_features
+
+
+def distance(counter1, counter2):
+  """
+  Calculates the dot product, assuming NORMALIZED.
+  """
+  dp = 0
+  for x in counter1.keys().intersection(counter2.keys()):
+    dp += counter1[x]*counter2[x]
+  return dp
+

### 2014-03-17 universalcorpus/miniethnologue.py

@@ -1,6 +1,6 @@
 # -*- coding: utf-8 -*-
 
-import codecs
+import codecs, re
 from collections import defaultdict
 
 # from http://www-01.sil.org/iso639-3/iso-639-3_Name_Index.tab
@@ -11156,7 +11156,6 @@ zun  Zuni  Shiwi'ma  zun  USA  2  19280  160996  ZUN  -  -  -  ifb ifk  en  no  
 zyj  Youjiang Zhuang  -  zyj  China  6  853  5544  -  -  -  -  en  en  no  -  Fri Oct 18 00:05:31 CDT 2013  -  Tai-Kadai, Kam-Tai, Tai, Northern
 zza  Zaza  -  zza  Turkey  93  1126852  6643343  -  -  -  -  kmr ote"""
 
-
 def getKey(dic, value):
   return [k for k,v in sorted(dic.items()) if value in v]
 
@@ -11169,6 +11168,13 @@ for i in retired_iso6393.split('\n'):
   line = i.split('  ')
   code = line[0]; langname = line[1]
   ISO2LANG[code].append(langname.lower())
+
+RETIRED2ISO = defaultdict(list)
+for i in retired_iso6393.split('\n'):
+  line = i.split('  ')
+  oldcode = line[0]
+  newcode = [line[3]] if line[3] else re.findall(r'\[([^]]*)\]',line[4])
+  RETIRED2ISO[oldcode] = newcode
   
 for i in crubadantable.split('\n'):
   line = i.split('  '); code = line[3]

### 2014-03-17 docs/computEL/concl.tex

--- 2014-03-03 docs/acl-2014/paper.tex
@@
-\documentclass[11pt]{article}
-\usepackage{acl2014}
-\usepackage{times}
-\usepackage{url}
-\usepackage{latexsym}
+\section{Future Work} \label{sec:future}
 
-\bibliographystyle{acl}
+Specific suggestions on how to work on resources mentioned in section \ref{sec:related}...
 
-\usepackage[utf8]{inputenc}
+\section{Conclusion} \label{sec:conclusion}
 
-\usepackage{enumitem}
-\usepackage{listings}
-\usepackage{fancyvrb}
-\usepackage{color}
-\definecolor{gray}{rgb}{0.4,0.4,0.4}
-\definecolor{darkblue}{rgb}{0.0,0.0,0.6}
-\definecolor{cyan}{rgb}{0.0,0.6,0.6}
-\definecolor{orange}{rgb}{1,0.5,0}
-
-\lstset{
-  basicstyle=\ttfamily,
-  columns=fullflexible,
-  showstringspaces=false,
-  commentstyle=\color{gray}\upshape
-}
-
-\lstdefinelanguage{XML}
-{
-  morestring=[b]",
-  morestring=[s]{>}{<},
-  morecomment=[s]{<?}{?>},
-  stringstyle=\color{black},
-  identifierstyle=\color{darkblue},
-  keywordstyle=\color{cyan},
-  morekeywords={xmlns,version,type}% list your attributes here
-}
-
-\setlength\titlebox{5cm}
-
-\title{Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages}
-
-\author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
-  Universität des Saarlandes \\
-  Campus, 66123 Saarbrücken, Germany \\
-  {\tt s9sufert@stud.uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-  {\tt liling.tan@uni-saarland.de}}
-
-\date{}
-
-\begin{document}
-\maketitle
-\begin{abstract}
-The study of endangered languages is limited by a lack of data.
-Existing corpora collections are limited in the range of languages covered, in standardisation, or in machine-readability.
-This makes the situation even worse for the computational linguist, especially one who would like to take a cross-linguistic or typological approach.
-We first survey existing efforts to compile cross-linguistic resources, then describe our own approach and give an example application - language clustering.
-To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contain data from a large number languages, and converted them into a standardised form consistent with the guidelines set out by \newcite{abney2010universal}.
-The resulting corpus is more easily-accessible and machine-readable than any of the underlying data sources, and represents a significant base corpus for researchers to draw on and add to in the future.
-\end{abstract}
-
-\section{Introduction}
-
-\newcite{abney2010universal} posed the grand challenge of building a Universal Corpus, including all of the world's languages, in a consistent structure.
-
-7105 living languages.
-
-\section{Related Work}
-
-[...]
-Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
-For instance, the OPUS corpus \cite{tiedemann2012opus} covers over 90 languages, only 1.27\% of the world's languages; The Leipzig Corpora Collection\footnote{corpora.uni-leipzig.de} \cite{biemann2007leipzig} contains corpora and dictionaries in 230 languages, 3.24\% of the world's languages.
-Even corpora that boast of linguistic diversity include only a small number of language families, e.g.
-the linguistically diverse NTU-Multilingual Corpus \cite{tan2011ntu} covers only 7 out of 136 language families.
-Producing a universal corpus requires the linguistic community to merge existing corpora and provide an ubiquitous access interface.
-To compile the foundation text for the Universal Corpus, we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
-
-\subsection{Web as Corpus}
-
-\newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialized corpora and terms from the web.
-Then, \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
-Later, \newcite{ferraresi2008ukwac} introduced and discussed the notion of compiling Web as Corpus \emph{"properly"} with the inaugural ukWaC using web-crawled data from a list of seed words.
-Thereafter, the Crúbadán corpora was created for a large number of under-resourced languages \cite{scannell2007crubadan}.
-Following which the deWaC (German), itWaC (Italian) and frWaC (French) were compiled \cite{baroni2009wacky}.
-
-Subsequently, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
-To build better quality WaCky corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection in cleaning web corpora.
-
-The initial wave of monolingual WaCky corpora \cite{baroni2009wacky} inspired more domain specific WaCky corpora, such as the Korean Web Learner's Corpus \cite{dickinson2010korean} and the Feed Corpus \cite{minocha2013feed}.
-Meanwhile, Web-driven corpora compilations also focused on building large  multilingual resources, e.g.
-The Leipzig Corpora Collection (\texttt{LCC}) \cite{biemann2007leipzig}, and the COrpora from the Web (\texttt{COW}) project \cite{schaefer2012cow}.
-
-Unlike previous WaCs built under the Web as Corpus kool ynitiative (WaCky), we did not perform seed query web-searching or web crawling to achieve a balanced resource.
-Instead, the foundation text for uniWaC was built using dedicated site crawling and/or site-dependent cleaning from four data sources, viz.
-(i) \emph{Online Database of Interlinear Text} (ODIN), (ii) \emph{Omniglot} website, (iii) \emph{Universal Declaration of Human Rights} (UDHR) and (iv) \emph{Wikipedia} dumps.
-Our emphasis for the uniWaC is based on based on \textbf{\emph{universality}}; covering as many languages as possible is the first priority.
-
-\section{Data Structure}
-
-\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
-In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
-
-However, we disagree that analysed texts should be accommodated in this way.
-If such a universal corpus is to succeed at all, it must enjoy support from the entire linguistics community, and doing this requires complete theory-neutrality.
-Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
-Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
-More fundamentally, they assume that the corpus should be segmented into words, but \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
-
-It is not within the scope of this paper to resolve these theoretical concerns.
-Instead, we suggest that the data structure for a universal corpus should be even more lightweight than Abney and Bird suggest.
-We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
-To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource language as a convenient surrogate of meaning.
-By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules.\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} The only difference here is that a text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
-The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies cannot be raised.
-We admit that highly detailed annotations, such as in treebanks and dependency banks, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
-
-\section{Data Sources}
-
-Although data size matters in general NLP \cite{banko2001scaling}, \emph{universality} is the utmost priority in NLP for the Universal Corpus initiative (aka Human Language Project; Abney and Bird, 2010; 2011).
-Hence, the uniWaC approach to data collection is based on focused/dedicated crawling unlike previous WaC word/URL list seeding.
-We manually selected sites that contains texts from a large variety of languages and wrote scripts to crawl and clean specific pages from the different data sources.
-The crawling and cleaning was done simultaneously.
-
-The resulting files were saved as plaintext file with \texttt{UTF-8} encoding.
-We used one directory for each data source and one file per language.
-The parallel data (from ODIN and Omniglot) are saved as tab-delimited files, one line per sentence, one column per language.
-The monolingual data (from UDHR and Wikipedia) are saved as newline delimited plaintext file, each line denoting a paragraph or sentence and each document separated by an empty line.
-The files are saved with the following naming convention to allow easy access, the filename contains the data source and its ISO 639-3 code, e.g.
-\texttt{odin-iii.txt} contains the \emph{ODIN} data for the \emph{Nuosu} language.
-[File structure is not relevant.]
-
-API which allows access according to Abney and Bird.
-However, maintaining a plain text file ensures that the data will be easily accessible, even if the surrounding infrastructure changes.
-[Seven dimensions]
-
-Our resulting corpus runs the full gamut of texts outlined by Abney and Bird, from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN). ...
-
-In the follow subsections, we describe the preprocessing required to convert each data source into a standardised form.
-
-\subsection{ODIN}
-
-ODIN (The Online Database of Interlinear Text) is a repository of IGTs extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
-Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation; however, the data requires further processing to bring it in line with the proposed format for a universal corpus.
-
-The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
-Each XML file is saved under a filename that matches the respective language code.
-While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.
-[Huh?
-Should this really happen?
-- the codes are given on their website and look valid to me...]
-
-\begin{minipage}{\columnwidth}
-
-a. \quad o lesu mai \\
-\indent \qquad 2sg return here \\
-
-\indent \qquad `\emph{You return here.}' \\
-\centerline{Figure 1: IGT from ODIN.}
-\end{minipage}
-
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>21 a. o lesu mai</line>
-    <line>2sg return here</line>
-    <line>`You return here.'</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\centerline{Figure 2: Fijian IGT from ODIN.}
-\end{minipage}
-\\ \\
-\noindent The IGTs from a web document usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
-From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2.
-Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>(69) na-Na-tmi-kwalca-t 
-    Yimas (Foley 1991)</line>
-    <line>3sgA-1sgO-say-rise-PERF
-    </line>
-    <line>`She woke me up' 
-    (by verbal action)</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\smallskip
-\centerline{Figure 3: Yimas IGT from ODIN.}
-\end{minipage}
-\\ \\
-\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
-In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
-In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.
-
-Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness': [Have we considered that the number of elements in the source line and gloss line should match...?]
-\\ \\
-\begin{minipage}{\columnwidth}
-\begin{itemize}
-\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
-\begin{itemize}
-\item[(i)]
-\begin{Verbatim}
-^(?\s?\w{1,5}\s*[):.]\s*
-\end{Verbatim}
-\item[(ii)] 
-\begin{Verbatim}
-[\[\(].{1,}[\]\)]
-\end{Verbatim}
-\end{itemize}
-\item \emph{Cleanest}: Only source lines without punctuation.
-\end{itemize}
-\end{minipage}
-\\ \\
-
-\noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
-The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
-
-\subsection{Omniglot}
-
-The Omniglot website\footnote{htp://www.omniglot.com/} is an online encyclopedia of writing systems and languages.
-We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
-
-[The following discussion is far too technical.]
-
-\begin{itemize}[noitemsep]
-\item \url{www.omniglot.com/language/phrases/*} 
-\item \url{www.omniglot.com/babel/*}
-\end{itemize}
-
-\noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
-To resolve this issue, we converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
-
-\subsection{Universal Declaration of Human Rights}
-
-The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages.
-Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
-Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d/}, which are free of encoding issues.
-The first four lines of each file records metadata, and the rest is the translation of the UDHR.
-
-\subsection{Wikipedia}
-
-Wikipedia contains 30.8 million articles in 287 languages and the text dump can be downloaded from \url{www.dumps.wikimedia.org}.
-\noindent One major issue with using the Wikipedia dump is the sheer size and extracting text from a glut of Wikipedia markup.
-To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
-After converting them into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
-
-\section{Copyright Issues}
-
-Corpora copyright remains a grey area and a hobglobin for corpora compilation and distribution (Kilgariff, 2002).
-Often, free access to language data online does not imply the right to download, retain, remix and redistribute.
-
-ODIN refers itself as a database of IGTs that are returned as a result of querying from the source documents.
-As such, only the links to the documents and the appropriate citations and not the copies of the source documents are maintained on the ODIN site (Lewis and Xia, 2010).
-However, the ODIN developers have left the users with no specific license/copyrights restriction for readaptation or redistribution.
-
-The Omniglot phrases are free to be used for non-commercial purposes as stated in the FAQ page but the rights of the Tower of Babel snippets from different bible sources remains iffy.
-For example, Omniglot cited www.bibles.org as one of its sources for the Tower of Babel chapter but the bibles.org site stated in fine print that `\emph{Except expressly permitted herein, you may not modify, copy, distribute, decompile, disassemble, reverse engineer, create derivative works, or otherwise use or manipulate the Site or any of its content without our prior written consent}'\footnote{http://bibles.org/pages/legal\#copyright}.
-
-The UDHR copyrights remains questionable.
-Although users are free to use, copy, modify, merge, publish and distribute Unicode Data Files, there is no explicit ownership of the UDHR textfiles from the Unicode website.
-However, the original source of the UDHR documents from the Office of the High Commissioner for Human Rights clearly states that `\emph{None of the materials provided on this web site may be used, reproduced or transmitted, in whole or in part}'\footnote{http://www.un.org/en/aboutun/copyright/}.
-
-The Wikipedia texts are licensed under Creative Commons Attribution Share-like\footnote{http://creativecommons.org/licenses/by-sa/3.0/legalcode}, which allows redistribution and adaptation.
-Hence the clean Wikipedia texts we have compiled can be freely redistributed without infringing any copyrights.
-
-%%
-%To avoid any copyrights infringement, ...
-After surveying the rights for so many days, there is no way to redistribute the corpus unless we jumble up the sentences, but if we do that, the "document" level will be lost and that will not be a "good" resource to do NLP for LRL since data is already sparse.
-%%
-
-\section{Representation and Universality}
-
-Number of languages, size of corpus, etc.
-
-\section{Language Clustering}
-
-This would be an example application, if we have time...
-
-\section{Future Work}
-
-Other corpora we could work on...
-
-\section{Conclusion}
-
-Summarise what we've done!
-
-\bibliography{references}
-
-\end{document}
+In this paper, we have described the creation of a foundation text for a universal corpus, following the guidelines of Abney and Bird (2010; 2011).
+To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia, and the LCC.
+The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}.
+[Give a summary statistic of size.] To demonstrate the utility of this resource, we used the data to perform language clustering, which gave promising results.
+We believe that this is the first corpus of its kind, which we hope will act as a seed corpus for the Human Language Project.

### 2014-03-17 docs/computEL/data.tex

--- 2014-03-04 docs/acl-2014/paper.tex
@@
-\documentclass[11pt]{article}
-\usepackage{acl2014}
-\usepackage{times}
-\usepackage{url}
-\usepackage{latexsym}
+\section{The Data}\label{sec:data}
 
-\bibliographystyle{acl}
+\subsection{Representation and Universality} \label{sec:stats}
 
-\usepackage[utf8]{inputenc}
+Number of languages, size of corpus, etc.
 
-\usepackage{enumitem}
-\usepackage{listings}
-\usepackage{fancyvrb}
-\usepackage{color}
-\definecolor{gray}{rgb}{0.4,0.4,0.4}
-\definecolor{darkblue}{rgb}{0.0,0.0,0.6}
-\definecolor{cyan}{rgb}{0.0,0.6,0.6}
-\definecolor{orange}{rgb}{1,0.5,0}
-
-\lstset{
-  basicstyle=\ttfamily,
-  columns=fullflexible,
-  showstringspaces=false,
-  commentstyle=\color{gray}\upshape
-}
-
-\lstdefinelanguage{XML}
-{
-  morestring=[b]",
-  morestring=[s]{>}{<},
-  morecomment=[s]{<?}{?>},
-  stringstyle=\color{black},
-  identifierstyle=\color{darkblue},
-  keywordstyle=\color{cyan},
-  morekeywords={xmlns,version,type}% list your attributes here
-}
-
-\setlength\titlebox{5cm}
-
-\title{Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages}
-
-\author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
-  Universität des Saarlandes \\
-  Campus, 66123 Saarbrücken, Germany \\
-  {\tt liling.tan@uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-  {\tt susfert@coli.uni-saarland.de}}
-
-\date{}
-
-\begin{document}
-\maketitle
-\begin{abstract}
-The study of endangered languages is constrained by a lack of data.
-Existing corpora collections are limited in the range of languages covered, in standardisation, or in machine-readability.
-This makes the situation even worse for the computational linguist, especially one who would like to take a cross-linguistic or typological approach.
-We first survey existing efforts to compile cross-linguistic resources, then describe our own approach and give an example application - language clustering.
-To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contain data from a large number languages, and converted them into a standardised form consistent with the guidelines set out by \newcite{abney2010universal}.
-The resulting corpus is more easily-accessible and machine-readable than any of the underlying data sources, and represents a significant base corpus for researchers to draw on and add to in the future.
-\end{abstract}
-
-\section{Introduction}
-
-At the time of writing, 7105 living languages are documented in Ethnologue\footnote{http://www.ethnologue.com}, but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century.
-However, only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
-
-Although there have been some previous attempts to produce cross-linguistic resources, there are few which both cover a wide range of languages, and are also machine-readable.
-We survey existing efforts in section \ref{sec:related}, and discuss their limitations in more detail.
-
-Abney and Bird \shortcite{abney2010universal,abney2011data} posed the grand challenge of building a Universal Corpus, calling it the Human Language Project.
-Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
-They propose a specific data structure, which we describe and discuss in section \ref{sec:structure}.
-We have accepted their challenge, and have begun converting existing resources into a format consistent with their specifications for a universal corpus.
-We have drawn on four [or five?] web sources, cleaning and standardising them as described in section \ref{sec:sources}, to produce a seed corpus for the Human Language Project.
-In sections \ref{sec:stats} and \ref{sec:copyright}, we respectively give a summary of the data contained, and discuss copyright and distribution.
-
-[Note: remove this section if we don't do the work!] We believe the resulting corpus is the first of its kind: large enough and consistent enough to allow language processing on a grand scale.
-In section \ref{sec:cluster}, we give an example application of this corpus: language clustering.
-We use the frequencies of character n-grams and words to estimate the similarity of two languages.
-Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
-Finally, we discuss future work in section \ref{sec:future}, and conclude in section \ref{sec:conclusion}.
-
-\section{Related Work} \label{sec:related}
-
-[There is a lot more stuff to be added here...]
-Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
-For instance, the OPUS corpus \cite{tiedemann2012opus} covers over 90 languages, only 1.27\% of the world's languages; The Leipzig Corpora Collection\footnote{corpora.uni-leipzig.de} \cite{biemann2007leipzig} contains corpora and dictionaries in 230 languages, 3.24\% of the world's languages.
-Even corpora that boast of linguistic diversity include only a small number of language families, e.g.
-the linguistically diverse NTU-Multilingual Corpus \cite{tan2011ntu} covers only 7 out of 136 language families.
-Producing a universal corpus requires the linguistic community to merge existing corpora and provide an ubiquitous access interface.
-To compile the foundation text for the Universal Corpus, we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
-
-\subsection{Web as Corpus}
-
-In recent years, there has been increasing interesting in crawling websites to produce corpora.
-For example, \newcite{baroni2004bootcat} introduced the BootCaT toolkit to bootstrap specialised corpora, and \newcite{sharoff2006search} built corpora in various languages by issuing queries for random combinations of frequent words to create a balanced corpus not unlike the British National Corpus.
-\newcite{ferraresi2008ukwac} produced ukWaC, a large-scale English corpus, and \cite{baroni2009wacky} applied a similar method to produce corpora in German, Italian, and French.
-To build better quality web corpora, \newcite{versley2012quality} introduced the notion of content-sensitive boilerplate detection for cleaning data.
-
-However, despite the amount of text they include, such corpora are usually limited in the number of languages represented.
-Some projects have emerged which strive for language diversity, such as the Leipzig Corpora Collection (LCC) \cite{biemann2007leipzig}, and COrpora from the Web (COW) \cite{schaefer2012cow}.
-The LCC currently offers free download of corpora in 117 languages, and dictionaries in many others, bringing the total number of languages up to 230.
-COW includes corpora in [XX] languages.
-These collections are certainly commendable, but they currently still fall short of universality - 230 languages represents only 3.3% of the world's languages [check figure].
-
-\newcite{scannell2007crubadan} describes the Crúbadán Project, an attempt to crawl the web for data in a much larger range of languages, including endangered ones.
-At the time of writing, the number of languages represented has grown to [XXX] However, due to copyright issues, not all of the data is freely available.
-Motivated by similar copyright concerns, \newcite{brunello2009free} stressed the notion of building free corpora from the web using documents released under Creative Commons licenses.
-
-Unlike the above efforts, we have not performed seed-query web-searching or web-crawling to achieve a balanced resource.
-Instead, we have focused on a small number of sources which contain data in a wide variety of languages, as described in section \ref{sec:sources}.
-However, these sources represent a variety of data types which a linguist might encounter, ranging from single-language text with significant non-linguistic markup (Wikipedia) to structured data with detailed linguistic annotations (ODIN).
-
-\section{Data Structure} \label{sec:structure}
-
-\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level [add more detail]; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
-In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
-
-However, we disagree that analysed texts should be accommodated in this way.
-If such a universal corpus is to succeed at all, it must enjoy support from a substantial part, if not all, of the linguistics community, and doing this requires theory-neutrality.
-Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
-Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
-More fundamentally, they assume that the corpus should be segmented into words, but \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
-
-It is not within the scope of this paper to resolve these theoretical concerns.
-Instead, we suggest that the data structure for a universal corpus should be even more lightweight than Abney and Bird suggest.
-We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
-To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource reference language as a convenient surrogate of meaning.
-By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules,\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} and as practised by documentary linguists.
-
-Then, only difference between analysed and aligned texts is that the text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
-The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies are difficult to raise.
-We admit that highly detailed annotations, such as trees and dependency graphs, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
-To make it indicate what reference language is being used, we propose prefixing \texttt{gloss-} to the language code of the reference language - for example \texttt{gloss-eng} and \texttt{gloss-eng} if a text has been glossed into English or Spanish, respectively.
-Using a prefix rather than a suffix makes it clear that this is not a subvariety of the given language.
-
-Finally, it is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
-\newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining "seven dimensions" along which this can vary.
-[more detail?] Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
-We have written an API to allow access to this data according to the guidelines of Abney and Bird, who remain agnostic as to the specific form of data storage.
-If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.
-
-\section{Data Sources} \label{sec:sources}
+\subsection{Data Sources} \label{sec:sources}
 
@@
 
-\subsection{ODIN}
+\subsubsection{ODIN}
 
@@
 \\ \\
+
 \noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
@@
 
-\subsection{Omniglot}
+\subsubsection{Omniglot}
 
@@
 
-\subsection{Universal Declaration of Human Rights}
+\subsubsection{Universal Declaration of Human Rights}
 
@@
 
-\subsection{Wikipedia}
+\subsubsection{Wikipedia}
 
@@
 We converted all of them into ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{http://meta.wikimedia.org/wiki/Special\_language\_codes}
-
-\section{Copyright Issues} \label{sec:copyright}
-
-Corpora copyright remains a grey area and a hobglobin for corpora compilation and distribution (Kilgarriff, 2002)[@Li, please add this reference to the bibliography file].
-Often, free access to language data online does not imply the right to download, retain, remix and redistribute.
-
-ODIN refers itself as a database of IGTs that are returned as a result of querying from the source documents.
-As such, only the links to the documents and the appropriate citations and not the copies of the source documents are maintained on the ODIN site.
-However, the ODIN developers have left the users with no specific license/copyrights restriction for readaptation or redistribution.
-
-The Omniglot phrases are free to be used for non-commercial purposes as stated in the FAQ page\footnote{http://www.omniglot.com/faqs.htm} but the rights of the Tower of Babel snippets from different bible sources remains iffy.
-For example, Omniglot cited www.bibles.org as one of its sources for the Tower of Babel chapter but the bibles.org site stated in fine print that `\emph{Except expressly permitted herein, you may not modify, copy, distribute, decompile, disassemble, reverse engineer, create derivative works, or otherwise use or manipulate the Site or any of its content without our prior written consent}'\footnote{http://bibles.org/pages/legal\#copyright}.
-[Ga: actually, about copying information, Omniglot says: \emph{'If you would like to copy or re-use any of the language-related articles, please ask the authors for permission.
-Contact details of the authors can be found at the bottom of the articles in most cases, if they're not there, please contact me, and I'll try to find them for you.'} - I suspect Adger has been given permission for each translation, and I think we would have to ask every single author if we want to re-distribute the data.
-For the tower of babel translations, this could end up a nightmare...]
-
-Although users are free to use, copy, modify, merge, publish and distribute Unicode Data Files, [Ga: Do we have a reference for this?
-Particularly distribution] there is no explicit ownership of the UDHR textfiles from the Unicode website.
-However, the original source of the UDHR documents from the Office of the High Commissioner for Human Rights clearly states that `\emph{None of the materials provided on this web site may be used, reproduced or transmitted, in whole or in part, in any form or by any means, electronic or mechanical, including photocopying, recording or the use of any information storage and retrieval system, except as provided for in the Terms and Conditions of Use of United Nations Web Sites, without permission in writing from the publisher.}'\footnote{http://www.un.org/en/aboutun/copyright/}.
-[Ga: Actually, it's not quite so bad - if we look at the terms and conditions, it says: \emph{'The United Nations grants permission to Users to visit the Site and to download and copy the information, documents and materials (collectively, “Materials”) from the Site for the User’s personal, non-commercial use, without any right to resell or redistribute them or to compile or create derivative works therefrom, subject to the terms and conditions outlined below, and also subject to more specific restrictions that may apply to specific Material within this Site.'}\footnote{http://www.un.org/en/aboutun/terms/} Specifically for the UDHR, it says: \emph{'If UDHR translations or materials are reproduced, users should make reference to this website as a source by providing a link.'}\footnote{http://www.ohchr.org/EN/UDHR/Pages/Introduction.aspx}  So, it looks like we're okay to use the UDHR as we currently are, but if we want to redistribute it, we need to get their permission in writing.]
-
-The Wikipedia texts are licensed under Creative Commons Attribution Share-like\footnote{http://creativecommons.org/licenses/by-sa/3.0/legalcode}, which allows redistribution and adaptation.
-Hence the clean Wikipedia texts we have compiled can be freely redistributed without infringing any copyrights.
-We have made this resource publically available.\footnote{To maintain anonymity, we do not give details of how to access the data, but, if accepted for publication, instructions will be included in the print version of this paper.}
-
-%%
-%To avoid any copyrights infringement, ...
-After surveying the rights for so many days, there is no way to redistribute the corpus unless we jumble up the sentences, but if we do that, the "document" level will be lost and that will not be a "good" resource to do NLP for LRL since data is already sparse.
-%%
-
-\section{Representation and Universality} \label{sec:stats}
-
-Number of languages, size of corpus, etc.
-
-\section{Language Clustering} \label{sec:cluster}
-
-This would be an example application, if we have time...
-
-\section{Future Work} \label{sec:future}
-
-Specific suggestions on how to work on resources mentioned in section \ref{sec:related}...
-
-\section{Conclusion} \label{sec:conclusion}
-
-In this paper, we have described the creation of a foundation text for a universal corpus, following the guidelines of Abney and Bird (2010; 2011).
-To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia, and the LCC.
-The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}.
-[Give a summary statistic of size.] To demonstrate the utility of this resource, we used the data to perform language clustering, which gave promising results.
-We believe that this is the first corpus of its kind, which we hope will act as a seed corpus for the Human Language Project.
-
-\bibliography{references}
-
-\end{document}

### 2014-03-17 docs/computEL/liling.tex

--- 2014-03-17 docs/liling.tex
@@
 
-\section{Detecting Similar Languages}
-
-When working on low-resource or endangered languages, computational and documentary linguists face the issue of lack of resources and knowledge about the language.
-Often, having knowledge about related or similar language(s) provides useful lexical, syntactic or morphological minimal pairs across languages and also helps in bootstrapping language models in NLP (Yarowsky
-and Ngai, 2001; Xia and Lewis, 2007).
-
-The task of identifying related/similar language is also known as ``language classification" where (i) languages are mapped onto language families trees depending on their historical ancestry often measured by comparing string distance from a fix lexical list across languages, aka.
-\emph{genetic classification} or \emph{lexicostatistics} (e.g.
-Swadesh, 1952; Starostin, 2010) or (ii) languages are grouped by their degrees of similarity with regards to their typological features, aka.
-\emph{typological classification} (e.g.
-Daume, 2009; Georgi et al.
-2010).
-
-To exemplify the use of the universal corpus for research on low-resource languages, we experimented with automatic detection of similar languages using K-means clustering with character ngrams and word unigrams features.
-Each language is represented by a vector of character bigrams and trigrams and word unigram features from the universal corpus.
-\newline \newline
-\noindent \textbf{SHOW FANCY FIGURE OF CLUSTER}
-\newline \newline
-\textbf{Evaluate the cluster (quantitatively/qualitatively)}
-
 \end{document}

### 2014-03-17 universalcorpus/counting.py
added counting.py

@@ -1,0 +1,59 @@
+# -*- coding: utf-8 -*-
+
+from crawlandclean import udhr, omniglot, odin, ethnologue
+from collections import Counter, defaultdict
+from miniethnologue import ISO2LANG, MACRO2LANG, RETIRED2ISO
+##from crawlandclean import ethnologue
+
+dead = {"osp":"Old Spanish", "odt":"Old Dutch", "goh": "Old High German",
+        "got":"Gothic","wlm":"Middle Welsh","oge":"Old Georgian",
+        "tpn":"Tupinamba","ojp":"Old Japanese","sga":"Old Irish",
+        "hit":"Hittite","tkm":"Takelma","dum":"Middle Dutch",
+        "fro":"Old French","nci":"Classical Nahuatl","gmh":"Middle High German",
+        "mxi":"Mozarabic"}
+macro_split = {"nob":"Norwegian, Bokmaal","nno":"Norwegian Nynorsk"}
+constructed = {"ido":"Ido","tlh":"Klingon","tzl":"Talossan","jbo":"Lojban",
+               "ina":"Interlingua"}
+
+language_families = ethnologue.language_families()
+for i in [odin, omniglot, udhr]:
+  langs = [j for j in i.languages() if j in ISO2LANG]
+  macros = {j:MACRO2LANG[j] for j in langs if j in MACRO2LANG}
+  retired = {j:RETIRED2ISO[j] for j in langs if j in RETIRED2ISO}
+  splits = [j for j in langs if j in macro_split.keys()] 
+  died = [j for j in langs if j in dead.keys()]
+  con = [j for j in langs if j in constructed.keys()]
+  
+  isolangs = list(set(langs)-set(macros.keys())-set(retired.keys())-\
+                  set(dead.keys())-set(macro_split.keys())-\
+                  set(constructed.keys()))
+                  
+  families = list(set([language_families[j][0][0] for j in isolangs]))
+  
+  endangerment = defaultdict(list)
+  for j in isolangs:
+    ed = language_families[j][0][-1].split()[0]
+    endangerment[ed].append(j)
+  
+  ##print len(langs), len(macros), len(retired), len(families)
+  ##print len(splits), len(died), len(con)
+  ##print splits
+  ##print retired
+  ##print died
+  '''
+  for e in sorted(endangerment):
+    for i in families:
+      print e, len(endangerment[e])
+  print len(set(endangerment['10']))
+  print
+  '''
+  families2endanger = defaultdict(list)
[…]

### 2014-03-18 docs/computEL/relwork.tex
Update relwork.tex
Added related work. Still needs polishing.

@@
 
-[There is a lot more stuff to be added here...]
-Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
-For instance, the OPUS corpus \cite{tiedemann2012opus} covers over 90 languages, only 1.27\% of the world's languages; The Leipzig Corpora Collection\footnote{corpora.uni-leipzig.de} \cite{biemann2007leipzig} contains corpora and dictionaries in 230 languages, 3.24\% of the world's languages.
-Even corpora that boast of linguistic diversity include only a small number of language families, e.g.
-the linguistically diverse NTU-Multilingual Corpus \cite{tan2011ntu} covers only 7 out of 136 language families.
-Producing a universal corpus requires the linguistic community to merge existing corpora and provide an ubiquitous access interface.
-To compile the foundation text for the Universal Corpus, we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
+In this section, we review existing efforts to compile multilingual machine-readable resources.
+Although some commercial resources are available, we restrict attention to freely accessible data, as this will be most relevant to documentary linguistics and computational linguists.
+All figures given below were correct at the time of writing, but it must be borne in mind that most resources discussed below are constantly growing.
+
+\subsection{Traditional Archives}
+
+Many archives exist to store the wealth of traditional resources produced by the documentary linguistics community.
+Such documents are increasingly being digitised, or produced in a digital form, and there are a number of archives which now offer free online access to their data.
+
+Some archives aim for a universal scope, including languages from all parts of the world.  Of particular note are:
+
+The Language Archive, maintained by the Max Planck Institute of Psycholinguistics, which includes DoBeS (Dokumentation Bedrohter Sprachen), among many other projects.
+Collection Pangloss, maintained by LACITO (Langues et Civilisations à Tradition Orale).
+The Endangered Languages Archive (ELAR), maintained by the School of Oriental and African Studies, London, which forms one part of the Hans Rausing Endangered Languages Project (HRELP).
+
+Regional archives include:
+
+AILLA (Archive of Indigenous Languages of Latin America), at the University of Texas at Austin. (Not in OLAC)
+AIATSIS collections (Australian Institute of Aboriginal and Torres Strait Islander Studies) (Not in OLAC)
+California Language Archive, managed by the Survey of California and Other Indian Languages
+ANLA (Alaska Native Language Archive), at the University of Alaska Fairbanks.
+PARADISEC (Pacific and Regional Archive for Digital Sources in Endangered Cultures)
+
+However, there are two main problems common to all of the above data sources.
+Firstly, the data is not always machine readable.
+For instance, in their own words, ANLA consists of "mostly paper records".
+Even where the data is available digitally, these often take the form of scanned images, or audio files.
+While both can provide invaluable information, they are extremely difficult to process with a computer, requiring an impractical level of pre-processing of the image or audio, before linguistic analysis can begin.
+Even textual data, which avoids these issues, may not be available in a machine-readable form, being stored as pdfs or
+
+Secondly, when data is machine readable, the format can vary wildly.
+This makes automated processing difficult, especially if one is not aware of the details of each project.
+Even when metadata standards and encodings agree, there can be idiosyncractic markup or non-linguistic information, such as labels for speakers in the transcript of a conversation.
+
+We can see that there is still much work to be done by individual researchers in digitising and standardising linguistic data, and it is outside of the scope of this paper to attempt this for the above archives.
+Guidelines for producing new materials are available from the E-MELD project (Electronic Metastructure for Endangered Languages Data), which specifically aimed to deal with the expanding number of standards for linguistic data.
+It gives best practice recommendations, illustrated with eleven case studies, and provides input tools which link to the GOLD ontology language, and the OLAC metadata set.
+Further recommendations are given by \newcite{bird2003port}, who describes seven dimensions along which the portability of linguistic data can vary.
+Various tools are available from The Language Archive at the Max Planck Institute for Psycholinguistics.
+
+Many of these archives are part of the Open Language Archive Community (OLAC), a subcommunity of the Open Archives Initiative.
+OLAC maintains a metadata standard, based on the 15-element Dublin Core, which allows a user to search through all participating archives in a unified fashion.
+However, centralising access to disparate resources, while of course extremely helpful, does not solve the problem of inconsistent standards.
+Indeed, it can become difficult even to answer simple questions like "how many languages are represented?"
+
+We do of course acknowledge that such resources have their purpose - for instance, there can be plenty of material to work with for a researcher who is looking to find data on a specific language, who is willing to put in a little time browsing through different catalogues, and who only needs human-readable data.
+However, for large-scale machine processing, they leave much to be desired.
+
+\subsection{Generic corpora collections}
+
+There are some corpora collections which do not take endangered languages to be a main aim, but which nonetheless cover an increasing number of languages.
+
+MetaShare (Multilingual Europe Technology Alliance) provides data in a little over 100 languages.
+While language codes are used, they have not been standardised, so that multiple codes are used for the same language.
+
+Linguistic Data Consortium (LDC) and the European Language Resources Association (ELRA) both offer data in multiple languages.
+However, while large in size, they cover only a limited number of languages.
+Furthermore, the corpora they contain are stored separately, making it difficult to access data according to language.
+
+It seems likely that resources which do not specifically aim to cover endangered languages will continue to be skewed heavily towards a handful of high-resource languages.
+
+\subsection{Parallel Corpora}
+
+The Machine Translation community has assembled a number of parallel corpora, which are crucial for statistical machine translation.
+The OPUS corpus \cite{tiedemann2012opus} subsumes a number of other well-known parallel corpora, such as Europarl, and covers documents from 350 languages, with various language pairs.
+However, not all of the data has been manually checked, which introduces noise.
+
+\subsection{Web Corpora}
+
+There has been increasing interest in deriving corpora from the web, due to the promise of large amounts of data.
+The majority of web corpora are however aimed at either one or a small number of languages, which is perhaps to be expected, given that the majority of online text is written in a handful of high-resource languages.
+Nonetheless, there have been a few efforts to apply the same methods to a wider range of languages.
+
+HC Corpora currently provides download of corpora in 68 different language varieties, which vary in size from 2M to 150M words.
+The corpora are thus of a respectable size, but only 1% of the world's languages are represented.
+A further difficulty is that languages are named, without the corresponding ISO codes.
+This problem may be small when all 68 varieties are well-known, but will become a serious concern if this corpus collection is to grow and include smaller languages.
+
+The Leipzig Corpora Collection (LCC)\footnote{http://corpora.uni-leipzig.de} \cite{biemann2007leipzig} provides download of corpora in 117 languages, and dictionaries in a number of others, bringing the total number of represented languages up to 230.
+The corpora are large, readily available, in plain text, and labelled with ISO codes.
+
+The Crúbadán Project aims to crawl the web for text in low-resource languages, and data is currently available for 1872 languages.
+This represents a significant portion of the world's languages; unfortunately, due to copyright restrictions, only lists of n-grams and their frequencies are publically available, not the texts themselves.
+While the breadth of languages covered makes this a useful resource for cross-linguistic research, the lack of actual texts means that only a limited range of applications are possible with this data.
+
+While the above efforts are commendable, and allow cross-linguistic analysis to a certain extent, there is still a long way to go to producing a universal corpus.
+
+\subsection{Cross-linguistic projects}
+
+Responding to the call to document and preserve the world's languages, highly cross-linguistic projects have sprung up, striving towards the aim of universality.
+Of particular note are the Endangered Languages Project, and the Rosetta Project.
+
+However, as much as these projects are to be praised for their commitment to universality, it is difficult to use their data to perform large-scale NLP.
+A researcher looking for a data on a specific language can easily navigate to the language in question, and download the available resources - but trying to access all the data can be awkward.
+While the sleek online interface used by the Endangered Languages Project may make it more accessible to those unfamiliar with language processing, it is not ideal for a computational linguistics researcher.

### 2014-03-19 universalcorpus/counting.py
added evaluator for langcluster.py

@@ -11,7 +11,7 @@ dead = {"osp":"Old Spanish", "odt":"Old Dutch", "goh": "Old High German",
         "hit":"Hittite","tkm":"Takelma","dum":"Middle Dutch",
         "fro":"Old French","nci":"Classical Nahuatl","gmh":"Middle High German",
         "mxi":"Mozarabic"}
-macro_split = {"nob":"Norwegian, Bokmaal","nno":"Norwegian Nynorsk"}
+macro_split = {"nob":"Norwegian, Bokmaal","nno":"Norwegian Nynorsk"} # nor.
 constructed = {"ido":"Ido","tlh":"Klingon","tzl":"Talossan","jbo":"Lojban",
                "ina":"Interlingua"}
 

### 2014-03-19 universalcorpus/crawlandclean/ethnologue.py

@@ -238,8 +238,12 @@ def language_families():
   langfam = pickle.load(codecs.open(ETHNO_DIR+\
                                   'languagefamilies_with_info.pk','rb'))
   return langfam
-'''
-x = language_families()
-for i in x:
-  print i, x[i][0]
-'''
+
+langfam = language_families()
+ISO2FAMILY = {i:langfam[i][0][0] for i in langfam}
+FAMILIES2ISO = defaultdict(list)
+for i in langfam:
+  FAMILIES2ISO[langfam[i][0][0]].append(i)
+
+
+

### 2014-03-19 universalcorpus/langcluster.py

@@ -1,19 +1,22 @@
 # -*- coding: utf-8 -*-
 
 from collections import Counter, defaultdict
-from itertools import chain
+from itertools import chain, product
 from operator import itemgetter
 from threading import Thread
 from Queue import Queue
-import time, os, ast, random
+import time, os, ast, random,codecs
 
 import numpy as np
 import pandas as pd
+import h5py
 from sklearn import cluster
 import milk
+from math import sqrt, acos
+import cPickle as pickle
 
-from crawlandclean import odin, omniglot, udhr
-from utils import normalize
+from crawlandclean import odin, omniglot, udhr, ethnologue
+from miniethnologue import ISO2LANG, MACRO2LANG, RETIRED2ISO
 
 random.seed(0)
 
@@ -24,19 +27,34 @@ def word2ngrams(text, n=3):
 def sent2ngrams(text, n=3):
   """ Convert sentence into character ngrams. """
   return list(chain(*[word2ngrams(i,n) for i in text.lower().split()]))
+    
+def normalize(featurevector, length=1):
+  """
+  Normalises a feature vector to a specific length.
+  Zero vectors are left zero.
+  NOTE: featurevector will be updated with normalized count.
+  """
+  from math import sqrt
+  try:
+    ##print length, type(length)
+    ##print featurevector.values()
+    norm = length / sqrt(sum([x*x for x in featurevector.values()]))
+  except ZeroDivisionError:
+    return featurevector
+  for feat in featurevector:
+    featurevector[feat] *= norm
+  return featurevector
 
 def generate_ngrams(data_source):
-  ngrams = defaultdict(Counter)
-  for num in range(2,4):
-    for lang, sent in globals()[data_source].source_sents():
-      _counter = normalize(Counter(sent2ngrams(sent,num))) ## TODO: normalize all ngrams in the 
-      if _counter:
-        ##print data_source, num, _counter
-        ngrams[lang]+= _counter
-    x = normalize(Counter(sent.split()))
-    if x: ngrams[lang]+=x
-  print data_source
-  return ngrams
+  twograms = defaultdict(Counter) 
+  threegrams = defaultdict(Counter)
+  unigrams = defaultdict(Counter)
[…]

### 2014-03-19 universalcorpus/utils.py

@@ -41,18 +41,3 @@ class Suppress_print(object):
     sys.stdout = StringIO.StringIO()
   def __exit__(self, *args):
     sys.stdout = self.stdout
-    
-def normalize(featurevector, length=1):
-  """
-  Normalises a feature vector to a specific length.
-  Zero vectors are left zero.
-  NOTE: featurevector will be updated with normalized count.
-  """
-  from math import sqrt
-  try:
-    norm = length / sqrt(sum([x*x for x in featurevector.values()]))
-  except ZeroDivisionError:
-    return
-  for feat in featurevector:
-    featurevector[feat] *= norm
-  return featurevector

### 2014-03-19 docs/computEL/liling.tex
added countings for wikipedia

@@
 
+\section{Corpus Coverage and Universality, something, something numbers...}
+
+\begin{table}
+\centering
+    \begin{tabular}{l|cc}
+    ~         & \#Languages & \#Families \\ \hline
+    ODIN      & 1218       & 101       \\
+    Omniglot  & 134        & 21        \\
+    UDHR      & 355        & 47        \\
+    Wikipedia & 209        & 22       \\
+    \end{tabular}
+\caption{Corpus Coverage}
+\end{table}
+
+According to ethnologue
+Total Living languages: 7105
+Total Living language families: 147
+
 \end{document}

### 2014-03-19 universalcorpus/counting.py

@@ -10,11 +10,12 @@ dead = {"osp":"Old Spanish", "odt":"Old Dutch", "goh": "Old High German",
         "tpn":"Tupinamba","ojp":"Old Japanese","sga":"Old Irish",
         "hit":"Hittite","tkm":"Takelma","dum":"Middle Dutch",
         "fro":"Old French","nci":"Classical Nahuatl","gmh":"Middle High German",
-        "mxi":"Mozarabic"}
+        "mxi":"Mozarabic", "ang": "Old English"}
 macro_split = {"nob":"Norwegian, Bokmaal","nno":"Norwegian Nynorsk"} # nor.
 constructed = {"ido":"Ido","tlh":"Klingon","tzl":"Talossan","jbo":"Lojban",
-               "ina":"Interlingua"}
+               "ina":"Interlingua","nov":"Novial"}
 
+alllangs = []
 language_families = ethnologue.language_families()
 for i in [odin, omniglot, udhr]:
   langs = [j for j in i.languages() if j in ISO2LANG]
@@ -27,33 +28,39 @@ for i in [odin, omniglot, udhr]:
   isolangs = list(set(langs)-set(macros.keys())-set(retired.keys())-\
                   set(dead.keys())-set(macro_split.keys())-\
                   set(constructed.keys()))
-                  
+  alllangs+=isolangs
+  
+print alllangs
+  
+  
+'''                
   families = list(set([language_families[j][0][0] for j in isolangs]))
   
   endangerment = defaultdict(list)
   for j in isolangs:
     ed = language_families[j][0][-1].split()[0]
     endangerment[ed].append(j)
+'''  
   
+
   ##print len(langs), len(macros), len(retired), len(families)
   ##print len(splits), len(died), len(con)
   ##print splits
   ##print retired
   ##print died
-  '''
-  for e in sorted(endangerment):
+
+'''for e in sorted(endangerment):
     for i in families:
       print e, len(endangerment[e])
-  print len(set(endangerment['10']))
-  print
-  '''
+  
   families2endanger = defaultdict(list)
   for j in isolangs:
     fam = language_families[j][0][0]
     ed = language_families[j][0][-1].split()[0]
-    print language_families[j]
-    print ed, fam
-    
-x = list(set([v[0][0] for k,v in language_families.items()]))
-print x
-print len(x)
+    ##print language_families[j]
+    ##print ed, fam
+
+  x = list(set([v[0][0] for k,v in language_families.items()]))
+  print x
+  print len(x)
+  '''

### 2014-03-19 universalcorpus/crawlandclean/ethnologue.py

@@ -202,8 +202,13 @@ ethno = {'pop':"field field-name-field-population field-type-"+\
              "text-with-summary field-label-inline clearfix",
              'status':"field field-name-language-status field-type-"+\
              "ds field-label-inline clearfix"}
+
+
 import udhr, odin, omniglot
-langs_we_have = sorted(set(udhr.languages() + odin.languages() + omniglot.languages()))
+from miniethnologue import WIKI2ISO, ISO2LANG
+
+wikilangs = [WIKI2ISO[i] for i in WIKI2ISO if WIKI2ISO[i] and WIKI2ISO[i] in ISO2LANG]
+langs_we_have = sorted(set(udhr.languages() + odin.languages() + omniglot.languages() + wikilangs))
 
 lf = defaultdict(list)
 for i in _lf:
@@ -233,6 +238,8 @@ for i in _lf:
 with codecs.open(ETHNO_DIR+langfamfile,'wb') as fout:
   pickle.dump(lf, fout)
 '''
+
+
 
 def language_families():
   langfam = pickle.load(codecs.open(ETHNO_DIR+\

### 2014-03-19 universalcorpus/crawlandclean/wikipedia.py

@@ -150,3 +150,13 @@ def source_sents(cleanedwikidir=parentddir+"/data/wikipedia/clean/"):
 for i,j in source_sents():
   print i,j
 '''
+            
+def langs():
+  from miniethnologue import WIKI2ISO, ISO2LANG
+  from crawlandclean import ethnologue
+  wikilangs = [WIKI2ISO[i] for i in WIKI2ISO if WIKI2ISO[i] and WIKI2ISO[i] in ISO2LANG]
+  
+  language_families = ethnologue.language_families()
+  
+  return set(language_families.keys()).difference(wikilangs)
+  

### 2014-03-19 universalcorpus/miniethnologue.py

@@ -11238,9 +11238,25 @@ def macrolang():
     else:
       lang = i.strip().split('  ')[0]
     macro2lang[lastmacro].append(lang)
+  macro2lang['zha'].append('Zhuang')
   return macro2lang
     
 MACRO2LANG = macrolang()
+
+CONLANG = {"ido":"Ido","tlh":"Klingon","tzl":"Talossan","jbo":"Lojban",
+          "ina":"Interlingua","nov":"Novial", "arc":"Imperial Aramaic",
+          "got":"Gothic","ile":"Interlingue","vol":"Volapük"}
+
+DEADLANG = {"osp":"Old Spanish", "odt":"Old Dutch", "goh": "Old High German",
+        "got":"Gothic","wlm":"Middle Welsh","oge":"Old Georgian",
+        "tpn":"Tupinamba","ojp":"Old Japanese","sga":"Old Irish",
+        "hit":"Hittite","tkm":"Takelma","dum":"Middle Dutch",
+        "fro":"Old French","nci":"Classical Nahuatl","gmh":"Middle High German",
+        "mxi":"Mozarabic", "ang": "Old English", "lzh": "Literary Chinese"}
+
+
+
+  
 
 ##print len(MACRO2LANG)
 #for i in ISO2LANG:

### 2014-03-19 docs/computEL/liling.tex
added combined counts

@@
     UDHR      				& 355        & 47        \\
-    Wikipedia & 209        & 22       \\
+    Wikipedia 				& 209        & 22       \\ \hline
+    \textbf{Combined}	& 1347			 & 106 
     \end{tabular}

### 2014-03-19 docs/computEL/liling.tex
added #tokens and size in table

--- 2014-03-19 docs/computEL/liling.tex
@@
 
-\begin{table}
+\begin{table}[h!]
 \centering
-    \begin{tabular}{l|cc}
-    ~         & \#Languages & \#Families \\ \hline
-    ODIN      & 1218       & 101       \\
-    Omniglot  & 134        & 21        \\
-    UDHR      & 355        & 47        \\
-    Wikipedia & 209        & 22       \\
+    \begin{tabular}{l|rrrr}
+    ~         				& \#Languages & \#Families 	&\#tokens		& Size	\\ \hline
+    ODIN      				& 1,217      & 101       		& 58,556		& 39 MB		\\
+    Omniglot  				& 134        & 21        		&	6,749			& 677 KB	\\
+    UDHR      				& 355        & 47        		&	33,117		& 5.2 MB	\\
+    Wikipedia 				& 209        & 22       		&						& 37 GB		\\ \hline
+    \textbf{Combined}	& 1,347			 & 106 
     \end{tabular}
@@
 
+6,749
+33,117
+
 \end{document}

### 2014-03-19 universalcorpus/counting.py

@@ -4,6 +4,7 @@ from crawlandclean import udhr, omniglot, odin, ethnologue
 from collections import Counter, defaultdict
 from miniethnologue import ISO2LANG, MACRO2LANG, RETIRED2ISO
 ##from crawlandclean import ethnologue
+from itertools import chain
 
 dead = {"osp":"Old Spanish", "odt":"Old Dutch", "goh": "Old High German",
         "got":"Gothic","wlm":"Middle Welsh","oge":"Old Georgian",
@@ -29,19 +30,18 @@ for i in [odin, omniglot, udhr]:
                   set(dead.keys())-set(macro_split.keys())-\
                   set(constructed.keys()))
   alllangs+=isolangs
-  
-print alllangs
-  
-  
-'''                
   families = list(set([language_families[j][0][0] for j in isolangs]))
   
   endangerment = defaultdict(list)
   for j in isolangs:
     ed = language_families[j][0][-1].split()[0]
     endangerment[ed].append(j)
-'''  
-  
+    
+  #for e in sorted(endangerment):
+  #  print i, e, len(endangerment[e])
+
+  print sum([len(i) for i in endangerment.values()]), len(isolangs), len(langs)
+    
 
   ##print len(langs), len(macros), len(retired), len(families)
   ##print len(splits), len(died), len(con)

### 2014-03-19 docs/computEL/liling.tex
updated tex on clustering results

@@
 
-According to ethnologue
-Total Living languages: 7105
-Total Living language families: 147
+\noindent
+According to ethnologue \\
+Total Living languages: 7105 \\
+Total Living language families: 147 \\
 
-6,749
-33,117
+\begin{table}[h!]
+    \begin{tabular}{l|ccc|ccc}
+    ~        & complete & ~       & ~       & ward    & ~       & ~       \\ \cline{2-7}
+    ~        & precision & recall       & f-score       & precision    & recall       & f-score      \\ \hline
+    distance & 0.0614   & 0.8565 & \textbf{0.1099} & 0.06140  & 0.8565 & \textbf{0.1099} \\
+    maxclust & 0.1925  & \textbf{0.0927} & 0.0692 & \textbf{0.1963} & 0.0905 & 0.0686  \\
+    \end{tabular}
+\caption{Comparing of clustering algorithms where number of clusters is set to genetic grouping on Ethnologue.}
+\end{table}
+
+\noindent
+see full table in http://goo.gl/bs2vPo
 

### 2014-03-19 universalcorpus/langcluster.py

@@ -143,8 +143,6 @@ def distance(c1, c2):
     return 0
   return acos(dimsum)
 
-
-
 if not os.path.exists('ngram.data'):
   data = load_ngram_array()
   with open('ngram.data','wb') as ngramout:
@@ -209,39 +207,49 @@ matrix = pd.lib.to_object_array(matrix).astype(float)
 
 import scipy
 import matplotlib.pylab as plt
-x = scipy.cluster.hierarchy.linkage(matrix, method='weighted')
-y = scipy.cluster.hierarchy.dendrogram(x)
-plt.savefig('tree.png')
 
-z = scipy.cluster.hierarchy.fcluster(x,148,'maxclust')
+methods = 'single complete average weighted centroid ward'.split()
+criterion = 'inconsistent distance maxclust'.split()
 
+fouteval = open('cluster.eval.outputs','w')
 
-lang2clusters = {}
-cluster2langs = defaultdict(list)
-for i,j in zip(z.tolist(), labels):
-  lang2clusters[j] = i
-  cluster2langs[i].append(j)
-
-fouti = open('induce-clusters','w')
-precisions, recalls, fscores = [], [], []
-for i in labels:
-  gold_class = ethnologue.FAMILIES2ISO[ethnologue.ISO2FAMILY[i]]
-  induced_cluster = set(cluster2langs[lang2clusters[i]])
-  print>>fouti, i
-  print>>fouti, induced_cluster
-  print>>fouti, gold_class
-  print>>fouti, "\n"
-  overlap = len(induced_cluster.intersection(gold_class))
-  rec = overlap /float(len(gold_class))
-  prec = overlap /float(len(induced_cluster))
-  precisions.append(prec)
-  recalls.append(rec)
-  fscores.append(2*prec*rec/float(prec/rec))
+for me, cr in product(methods,criterion):
+  x = scipy.cluster.hierarchy.linkage(matrix, method=me)
+  y = scipy.cluster.hierarchy.dendrogram(x)
+  plt.savefig('tree.png')
   
-def avg(l):
-  return sum(l)/float(len(l))
-
-print avg(precisions), avg(recalls), avg(fscores)
+  z = scipy.cluster.hierarchy.fcluster(x,148,cr)
+  
+  lang2clusters = {}
+  cluster2langs = defaultdict(list)
+  for cluster_num, lang_labels in zip(z.tolist(), labels):
+    lang2clusters[lang_labels] = cluster_num
+    cluster2langs[cluster_num].append(lang_labels)
+    
[…]

### 2014-03-19 docs/computEL/clean.tex
Mostly reorganizational edits

@@
-\section{Something about cleaning data, merging data, formatting, standardization}
+\section{Data source case studies}
 
-\subsection{Details for individual data sources (or perhaps rather individual topics}
+In this section we describe in detail the work required to merge each of our four data sources into a single common source. \textcolor{blue}{[Universal Corpus???]}
 
-\textcolor{blue}{[Import things from previous section]}
+In the following subsections, we describe each source, and the processing required to convert the data into a standardised form.
 
-\subsection{Universal Corpus and Data Structure} \label{sec:structure}
+\subsection{UDHR}
 
-\textcolor{blue}{[Perhaps this comes elsewhere?]}
+The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages.
+Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
+Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d}, which are free of encoding issues.
+The first four lines of each file records metadata, and the rest is the translation of the UDHR.
+This dataset was extremely clean, and simply required segmentation into sentences.
 
-\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level [add more detail]; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
-In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
+\subsection{Wikipedia}
 
-However, we disagree that analysed texts should be accommodated in this way.
-If such a universal corpus is to succeed at all, it must enjoy support from a substantial part, if not all, of the linguistics community, and doing this requires theory-neutrality.
-Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
-Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
-More fundamentally, they assume that the corpus should be segmented into words, but \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
+At the time of writing, Wikipedia contains 30.8 million articles in 287 languages, which provides a sizeable amount of clean monolingual text in a fairly wide range of languages.
+Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.
+One major issue with using the Wikipedia dump extracting text from a glut of Wikipedia markup.
+To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
+After conversion into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
 
-It is not within the scope of this paper to resolve these theoretical concerns.
-Instead, we suggest that the data structure for a universal corpus should be even more lightweight than Abney and Bird suggest.
-We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
-To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource reference language as a convenient surrogate of meaning.
-By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules,\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} and as practised by documentary linguists.
+Wikipedia uses its own set of language codes, most of which are in ISO~639-1 or ISO~639-2/3.
+We converted all of them into ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{http://meta.wikimedia.org/wiki/Special\_language\_codes}
 
-Then, only difference between analysed and aligned texts is that the text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
-The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies are difficult to raise.
-We admit that highly detailed annotations, such as trees and dependency graphs, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
-To make it indicate what reference language is being used, we propose prefixing \texttt{gloss-} to the language code of the reference language - for example \texttt{gloss-eng} and \texttt{gloss-eng} if a text has been glossed into English or Spanish, respectively.
-Using a prefix rather than a suffix makes it clear that this is not a subvariety of the given language.
+\subsection{Omniglot}
 
-Finally, it is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
-\newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining "seven dimensions" along which this can vary.
-[more detail?] Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
-We have written an API to allow access to this data according to the guidelines of Abney and Bird, who remain agnostic as to the specific form of data storage.
-If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.
+The Omniglot website\footnote{htp://www.omniglot.com} is an online encyclopedia of writing systems and languages.
+We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
+The urls for these resources are:
+
+\begin{itemize}[noitemsep]
+\item \url{www.omniglot.com/language/phrases/*} 
+\item \url{www.omniglot.com/babel/*}
+\end{itemize}
+
+\noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
+To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
+
+\subsection{ODIN}
+
+ODIN (The Online Database of Interlinear Text) is a repository of IGTs extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
+Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation; however, the data requires further processing to bring it in line with the proposed format for a universal corpus.
+
+The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
+Each XML file is saved under a filename that matches the respective language code.
+While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.
+[Huh?
+Should this really happen?
+- the codes are given on their website and look valid to me...]
+\\ \\
+\begin{minipage}{\columnwidth}
+
+\quad 21 a.\quad o lesu mai \\
+\indent \qquad\qquad 2sg return here \\
+\indent \qquad\qquad `\emph{You return here.}' \\
+\centerline{Figure 1: Fijian IGT from ODIN.}
+\end{minipage}
+
+%%
+% Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
+% Also, some of the lines go outside of the column!!
+%%
+
+\begin{minipage}{\columnwidth}
+\begin{lstlisting}[language=XML]
+<igt>
+  <example>
+    <line>21 a. o lesu mai</line>
+    <line>2sg return here</line>
+    <line>`You return here.'</line>
+  </example>
+</igt>
+\end{lstlisting} 
+\centerline{Figure 2: Fijian IGT in ODIN's XML format.}
+\end{minipage}
+\\ \\
+\noindent The IGTs from an academic publication usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
+From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2.
+Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
+\begin{minipage}{\columnwidth}
+\begin{lstlisting}[language=XML]
+<igt>
+  <example>
+    <line>(69) na-Na-tmi-kwalca-t 
+    Yimas (Foley 1991)</line>
+    <line>3sgA-1sgO-say-rise-PERF
+    </line>
+    <line>`She woke me up' 
+    (by verbal action)</line>
+  </example>
+</igit>
+\end{lstlisting} 
+\smallskip
+\centerline{Figure 3: Yimas IGT in ODIN's XML format.}
+\end{minipage}
+\\ \\
+
+\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
+In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
+In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.
+
+Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
+Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness':
+
+[This hasn't been implemented, but I think we could try this: Split source and gloss lines into tokens according to whitespace.
+Check the source has more tokens.
+Count the number of hyphens and equal signs in each token (since they denote morphemes).
+Match the gloss line with a substring of the source line, according to numbers of hyphens and equal signs.
+If there's exactly one match, keep it, and discard tokens at the ends of the line.
+If there's more than one match, look for other punctuation in the initial and final tokens.]
+\\ \\
+\begin{minipage}{\columnwidth}
+\begin{itemize}
+\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
+\begin{itemize}
+\item[(i)]
+\begin{Verbatim}
+^(?\s?\w{1,5}\s*[):.]\s*
+\end{Verbatim}
+\item[(ii)] 
+\begin{Verbatim}
+[\[\(].{1,}[\]\)]
+\end{Verbatim}
+\end{itemize}
+\item \emph{Cleanest}: Only source lines without punctuation.
+\end{itemize}
+\end{minipage}
+\\ \\
+
+\noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
+The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.

### 2014-03-19 docs/computEL/cluster.tex
Mostly reorganizational edits

--- 2014-03-17 docs/liling.tex
@@
-\documentclass[11pt]{article}
-\title{\textbf{universal corpus}}
-\date{}
-\begin{document}
-\maketitle
-
-\section{Detecting Similar Languages}
+\section{Detecting Similar Languages} \label{sec:cluster}
 
@@
 
-\end{document}
+\begin{table*}[h!]
+\begin{centering}
+
+    \begin{tabular}{l|ccc|ccc}
+    ~        & complete & ~       & ~       & ward    & ~       & ~       \\ \cline{2-7}
+    ~        & precision & recall       & f-score       & precision    & recall       & f-score      \\ \hline
+    distance & 0.0614   & 0.8565 & \textbf{0.1099} & 0.06140  & 0.8565 & \textbf{0.1099} \\
+    maxclust & 0.1925  & \textbf{0.0927} & 0.0692 & \textbf{0.1963} & 0.0905 & 0.0686  \\
+    \end{tabular}
+\caption{Comparing of clustering algorithms where number of clusters is set to genetic grouping on Ethnologue.}
+\end{centering}
+\end{table*}
+
+\noindent
+see full table in http://goo.gl/bs2vPo

### 2014-03-19 docs/computEL/data.tex
Mostly reorganizational edits

@@
 Number of languages, size of corpus, etc.
+
+\begin{table}
+\centering
+    \begin{tabular}{l|cc}
+    ~         				& \#Languages & \#Families \\ \hline
+    ODIN      				& 1218       & 101       \\
+    Omniglot  				& 134        & 21        \\
+    UDHR      				& 355        & 47        \\
+    Wikipedia 				& 209        & 22       \\ \hline
+    \textbf{Combined}	& 1347			 & 106 
+    \end{tabular}
+\caption{Corpus Coverage}
+\end{table}
+
+According to ethnologue
+Total Living languages: 7105
+Total Living language families: 147
+
+\begin{table}[h!]
+\centering
+    \begin{tabular}{l|rr|rr}
+    ~         				& \#Languages & \#Families 	&\#tokens		& Size	\\ \hline
+    ODIN      				& 1,217      & 101       		& 58,556		& 39 MB		\\
+    Omniglot  				& 134        & 21        		&	6,749			& 677 KB	\\
+    UDHR      				& 355        & 47        		&	33,117		& 5.2 MB	\\
+    Wikipedia 				& 209        & 22       		&						& 37 GB		\\ \hline
+    \textbf{Combined}	& 1,347			 & 106 
+    \end{tabular}
+\caption{Corpus Coverage}
+\end{table}
 
@@
 
-In the following subsections, we describe each source, and the processing required to convert the data into a standardised form.
 Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia and LCC) to parallel text (UDHR and Omniglot) to IGTs (ODIN).
@@
 
-ODIN (The Online Database of Interlinear Text) is a repository of IGTs extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
-Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation; however, the data requires further processing to bring it in line with the proposed format for a universal corpus.
-
-The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
-Each XML file is saved under a filename that matches the respective language code.
-While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.
-[Huh?
-Should this really happen?
-- the codes are given on their website and look valid to me...]
-\\ \\
-\begin{minipage}{\columnwidth}
-
-\quad 21 a.\quad o lesu mai \\
-\indent \qquad\qquad 2sg return here \\
-\indent \qquad\qquad `\emph{You return here.}' \\
-\centerline{Figure 1: Fijian IGT from ODIN.}
-\end{minipage}
-
-%%
-% Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
-% Also, some of the lines go outside of the column!!
-%%
-
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>21 a. o lesu mai</line>
-    <line>2sg return here</line>
-    <line>`You return here.'</line>
-  </example>
-</igt>
-\end{lstlisting} 
-\centerline{Figure 2: Fijian IGT in ODIN's XML format.}
-\end{minipage}
-\\ \\
-\noindent The IGTs from an academic publication usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
-From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2.
-Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>(69) na-Na-tmi-kwalca-t 
-    Yimas (Foley 1991)</line>
-    <line>3sgA-1sgO-say-rise-PERF
-    </line>
-    <line>`She woke me up' 
-    (by verbal action)</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\smallskip
-\centerline{Figure 3: Yimas IGT in ODIN's XML format.}
-\end{minipage}
-\\ \\
-
-\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
-In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
-In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.
-
-Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness':
-
-[This hasn't been implemented, but I think we could try this: Split source and gloss lines into tokens according to whitespace.
-Check the source has more tokens.
-Count the number of hyphens and equal signs in each token (since they denote morphemes).
-Match the gloss line with a substring of the source line, according to numbers of hyphens and equal signs.
-If there's exactly one match, keep it, and discard tokens at the ends of the line.
-If there's more than one match, look for other punctuation in the initial and final tokens.]
-\\ \\
-\begin{minipage}{\columnwidth}
-\begin{itemize}
-\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
-\begin{itemize}
-\item[(i)]
-\begin{Verbatim}
-^(?\s?\w{1,5}\s*[):.]\s*
-\end{Verbatim}
-\item[(ii)] 
-\begin{Verbatim}
-[\[\(].{1,}[\]\)]
-\end{Verbatim}
-\end{itemize}
-\item \emph{Cleanest}: Only source lines without punctuation.
-\end{itemize}
-\end{minipage}
-\\ \\
-
-\noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
-The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
-
 \subsubsection{Omniglot}
-
-The Omniglot website\footnote{htp://www.omniglot.com} is an online encyclopedia of writing systems and languages.
-We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
-The urls for these resources are:
-
-\begin{itemize}[noitemsep]
-\item \url{www.omniglot.com/language/phrases/*} 
-\item \url{www.omniglot.com/babel/*}
-\end{itemize}
-
-\noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
-To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
 
@@
 
-The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages.
-Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
-Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d}, which are free of encoding issues.
-The first four lines of each file records metadata, and the rest is the translation of the UDHR.
-This dataset was extremely clean, and simply required segmentation into sentences.
-
 \subsubsection{Wikipedia}
 
-At the time of writing, Wikipedia contains 30.8 million articles in 287 languages, which provides a sizeable amount of clean monolingual text in a fairly wide range of languages.
-Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.
-One major issue with using the Wikipedia dump extracting text from a glut of Wikipedia markup.
-To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
-After conversion into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
+\subsection{Universal Corpus and Data Structure} \label{sec:structure}
 
-Wikipedia uses its own set of language codes, most of which are in ISO~639-1 or ISO~639-2/3.
-We converted all of them into ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{http://meta.wikimedia.org/wiki/Special\_language\_codes}
+\textcolor{blue}{[Perhaps this comes elsewhere?]}
+
+\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level [add more detail]; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
+In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
+
+However, we disagree that analysed texts should be accommodated in this way.
+If such a universal corpus is to succeed at all, it must enjoy support from a substantial part, if not all, of the linguistics community, and doing this requires theory-neutrality.
+Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
+Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
+More fundamentally, they assume that the corpus should be segmented into words, but \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
+
+It is not within the scope of this paper to resolve these theoretical concerns.
+Instead, we suggest that the data structure for a universal corpus should be even more lightweight than Abney and Bird suggest.
+We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
+To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource reference language as a convenient surrogate of meaning.
+By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
+We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules,\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} and as practised by documentary linguists.
+
+Then, only difference between analysed and aligned texts is that the text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
+The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies are difficult to raise.
+We admit that highly detailed annotations, such as trees and dependency graphs, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
+To make it indicate what reference language is being used, we propose prefixing \texttt{gloss-} to the language code of the reference language - for example \texttt{gloss-eng} and \texttt{gloss-eng} if a text has been glossed into English or Spanish, respectively.
+Using a prefix rather than a suffix makes it clear that this is not a subvariety of the given language.
+
+Finally, it is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
+\newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining "seven dimensions" along which this can vary.
+[more detail?] Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
+We have written an API to allow access to this data according to the guidelines of Abney and Bird, who remain agnostic as to the specific form of data storage.
+If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.

### 2014-03-19 docs/computEL/intro.tex
Mostly reorganizational edits

@@
 
-At the time of writing, 7105 living languages are documented in Ethnologue\footnote{http://www.ethnologue.com}, but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century.
+At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century.
 However, only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
+
+\textcolor{blue}{[move \& rewrite]}
 

### 2014-03-19 docs/computEL/paper.tex
minor revisions to main file
Mostly reorganizational edits

@@
 
-\author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
-  Universität des Saarlandes \\
-  Campus, 66123 Saarbrücken, Germany \\
-  {\tt liling.tan@uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-  {\tt susfert@coli.uni-saarland.de}}
+% \author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
+%   Universität des Saarlandes \\
+%   Campus, 66123 Saarbrücken, Germany \\
+%   {\tt liling.tan@uni-saarland.de, emerson@coli.uni-saarland.de,} \\
+%   {\tt susfert@coli.uni-saarland.de}}
 
@@
 \begin{abstract}
+\textcolor{blue}{[revisit title, revisit abstract]}
 The study of endangered languages is constrained by a lack of data.
@@
 \input{relwork}
+%\input{liling}
 \input{data}

### 2014-03-19 docs/computEL/intro.tex
Major edits to intro

@@
 
-\textcolor{blue}{[move \& rewrite]}
+Several years ago, Abney and Bird \shortcite{abney2010universal,abney2011data} posed the challenge of building a Universal Corpus (UC), calling it the Human Language Project.
+Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
+The challenge was issued to the computational linguistics (CL) community, from the perspective that the data management, language processing, and machine learning tools\footnote{\textcolor{blue}{am I missing anything?}} well-known in CL must be brought to bear on the problems of documentary linguistics (DL), if we are to make any serious progress toward building such a resource.
+The UC as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages (LRLs), which in turn drives developments in automated analysis, supporting the work of DL and increasing the efficiency of the language documentation process.
+\textcolor{blue}{[cite our AL work?
+is this too strong?
+also, run-on sentence.
+also, perhaps mention this workshop, which has precisely these goals]}
 
-Although there have been some previous attempts to produce cross-linguistic resources, there are few which both cover a wide range of languages, and are also machine-readable.
-We survey existing efforts in section \ref{sec:related}, and discuss their limitations in more detail.
+We have accepted this challenge and have begun converting existing resources into a format consistent with Abney and Bird's specifications for a universal corpus.
+We aim for a collection of resources that includes data: (a) from as many languages as possible, and (b) in a format which is both in accordance with best practice archiving recommendations and readily accessible as training data for machine learning and for other computational methods.
+\textcolor{blue}{[kind of clunky]} Of course there are many relevant efforts toward producing cross-linguistic resources, which we survey in section \ref{sec:related}.
+To the best of our knowledge, though, no existing effort meets these two desiderata to the extent of our corpus, which we will refer to throughout as the Human Language Project Seed Corpus (HLPSC).
+\textcolor{blue}{[other naming ideas?]}
 
-Abney and Bird \shortcite{abney2010universal,abney2011data} posed the grand challenge of building a Universal Corpus, calling it the Human Language Project.
+To produce the HLPSC, we have drawn on four web sources, described in section \ref{sec:sources}.
+To bring the four resources into a single common format and data structure (section \ref{sec:structure}), each required different degrees and types of cleaning and standardisation.
+We describe the steps required in detail in section \ref{sec:case_studies}, presenting each resource as a separate case study.
+By doing so we hope to make future resource conversion efforts more efficient, as they can be guided by the lessons we learned in assembling our seed corpus.
+
+We believe the resulting corpus, which at present covers 1374 languages from 106 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing.
+\textcolor{blue}{[START HERE]}
+
+% Abney and Bird \shortcite{abney2010universal,abney2011data} posed the grand challenge of building a Universal Corpus, calling it the Human Language Project.
 Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.

### 2014-03-20 universalcorpus/langcluster.py
cleanup the code a little.

@@ -1,22 +1,24 @@
 # -*- coding: utf-8 -*-
 
+import cPickle as pickle
+import time, os, ast, random,codecs
+
+from math import sqrt, acos
 from collections import Counter, defaultdict
 from itertools import chain, product
 from operator import itemgetter
 from threading import Thread
 from Queue import Queue
-import time, os, ast, random,codecs
 
 import numpy as np
 import pandas as pd
-import h5py
-from sklearn import cluster
-import milk
-from math import sqrt, acos
-import cPickle as pickle
+##from sklearn import cluster; import milk
+import scipy
+import matplotlib.pylab as plt
 
 from crawlandclean import odin, omniglot, udhr, ethnologue
 from miniethnologue import ISO2LANG, MACRO2LANG, RETIRED2ISO
+
 
 random.seed(0)
 
@@ -36,8 +38,6 @@ def normalize(featurevector, length=1):
   """
   from math import sqrt
   try:
-    ##print length, type(length)
-    ##print featurevector.values()
     norm = length / sqrt(sum([x*x for x in featurevector.values()]))
   except ZeroDivisionError:
     return featurevector
@@ -45,15 +45,29 @@ def normalize(featurevector, length=1):
     featurevector[feat] *= norm
   return featurevector
 
+dead = {"osp":"Old Spanish", "odt":"Old Dutch", "goh": "Old High German",
+        "got":"Gothic","wlm":"Middle Welsh","oge":"Old Georgian",
+        "tpn":"Tupinamba","ojp":"Old Japanese","sga":"Old Irish",
+        "hit":"Hittite","tkm":"Takelma","dum":"Middle Dutch",
+        "fro":"Old French","nci":"Classical Nahuatl","gmh":"Middle High German",
+        "mxi":"Mozarabic"}
+macro_split = {"nob":"Norwegian, Bokmaal","nno":"Norwegian Nynorsk"} # nor.
+constructed = {"ido":"Ido","tlh":"Klingon","tzl":"Talossan","jbo":"Lojban",
+               "ina":"Interlingua"}
+
+living_languages = set(ISO2LANG.keys()) - set(RETIRED2ISO.keys()) - \
+set(dead.keys()) - set(constructed.keys()) - set(MACRO2LANG.keys()) - \
+set(macro_split.keys())
+
 def generate_ngrams(data_source):
   twograms = defaultdict(Counter) 
   threegrams = defaultdict(Counter)
[…]

### 2014-03-20 docs/computEL/intro.tex
continuing with intro

@@
 By doing so we hope to make future resource conversion efforts more efficient, as they can be guided by the lessons we learned in assembling our seed corpus.
+To that end, many of the resources described in \ref{sec:related} are perhaps candidates for inclusion in the next stage of building a UC.
 
 We believe the resulting corpus, which at present covers 1374 languages from 106 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing.
-\textcolor{blue}{[START HERE]}
+To test this belief, we use the HLPSC in a test application (\ref{sec:cluster}): the task of language clustering.
+We use surface-level features extracted from our data to estimate the similarity of two languages.
 

### 2014-03-20 docs/computEL/data.tex
Update data.tex
Added discussion of heatmap.

@@
 
-Number of languages, size of corpus, etc.
+According to Ethnologue, there are 7105 living languages, and 147 living language families.
+Across all our data sources, we manage to cover 1347 languages in 106 families, which represents 19.0\% of the world's languages.
+To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) in figure \textcolor{blue}{[insert reference here when it's ready]}.
+The values in each cell have been colored according to proportion of languages represented, with green indicated good coverage and red poor.
+It's interesting to note that vigorous languages (6a) are poorly represented across all data sources, and worse than more endangered categories.
+In terms of language documentation, vigorous languages are less urgent goals than those in categories 6b and up, but this highlights an unexpected gap in linguistic resources.
 
-\begin{table}
-\centering
-    \begin{tabular}{l|cc}
-    ~         				& \#Languages & \#Families \\ \hline
-    ODIN      				& 1218       & 101       \\
-    Omniglot  				& 134        & 21        \\
-    UDHR      				& 355        & 47        \\
-    Wikipedia 				& 209        & 22       \\ \hline
-    \textbf{Combined}	& 1347			 & 106 
-    \end{tabular}
-\caption{Corpus Coverage}
-\end{table}
-
-According to ethnologue
-Total Living languages: 7105
-Total Living language families: 147
+%\begin{table}
+%\centering
+%    \begin{tabular}{l|cc}
+%    ~         				& \#Languages & \#Families \\ \hline
+%    ODIN      				& 1218       & 101       \\
+%    Omniglot  				& 134        & 21        \\
+%    UDHR      				& 355        & 47        \\
+%    Wikipedia 				& 209        & 22       \\ \hline
+%    \textbf{Combined}	& 1347			 & 106 
+%    \end{tabular}
+%\caption{Corpus Coverage}
+%\end{table}
 
@@
 \caption{Corpus Coverage}
+\label{table:corpus}
 \end{table}
+
+\textcolor{blue}{[insert heatmap here]}
 
@@
 
-Although data size matters in general NLP \cite{banko2001scaling} [is this the best reference?], \emph{universality} is the utmost priority for a universal corpus.
+Although data size matters in general NLP \cite{banko2001scaling} \textcolor{blue}{[is this the best reference?]}, \emph{universality} is the utmost priority for a universal corpus.
 We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured):
@@
 \item Wikipedia
-\item The Leipzig Corpora Collection (LCC) [do we include this?]
 \end{enumerate}

### 2014-03-20 docs/computEL/intro.tex
minor intro edits

@@
 To test this belief, we use the HLPSC in a test application (\ref{sec:cluster}): the task of language clustering.
-We use surface-level features extracted from our data to estimate the similarity of two languages.
+We use surface-level features (frequencies of character n-grams and words) extracted from our data to estimate the similarity of two languages.
+\textcolor{blue}{[START HERE WITH CLEARER DESCRIPTION OF CLUSTERING RESULTS]}
+Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
+Finally, we discuss future work in section \ref{sec:future}, and conclude in section \ref{sec:conclusion}.
 
@@
 
-[Note: remove this section if we don't do the work!] We believe the resulting corpus is the first of its kind: large enough and consistent enough to allow language processing on a grand scale.
+% [Note: remove this section if we don't do the work!] We believe the resulting corpus is the first of its kind: large enough and consistent enough to allow language processing on a grand scale.
 In section \ref{sec:cluster}, we give an example application of this corpus: language clustering.
 We use the frequencies of character n-grams and words to estimate the similarity of two languages.
-Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
-Finally, we discuss future work in section \ref{sec:future}, and conclude in section \ref{sec:conclusion}.

### 2014-03-20 docs/computEL/clean.tex
Major edits to intro
converted  minipages to figures

@@
-\section{Data source case studies}
+\section{Data source case studies} \label{sec:case_studies}
 
@@
 - the codes are given on their website and look valid to me...]
-\\ \\
-\begin{minipage}{\columnwidth}
 
+\begin{figure}
 \quad 21 a.\quad o lesu mai \\
@@
 \indent \qquad\qquad `\emph{You return here.}' \\
-\centerline{Figure 1: Fijian IGT from ODIN.}
-\end{minipage}
+\caption{Fijian IGT from ODIN.} \label{fig:odin_fijian}
+\end{figure}
+
+% \begin{minipage}{\columnwidth}
+% \quad 21 a.\quad o lesu mai \\
+% \indent \qquad\qquad 2sg return here \\
+% \indent \qquad\qquad `\emph{You return here.}' \\
+% %\caption{Fijian IGT from ODIN.} 
+% \end{minipage}
 
@@
 
-\begin{minipage}{\columnwidth}
+\begin{figure}
 \begin{lstlisting}[language=XML]
@@
 \end{lstlisting} 
-\centerline{Figure 2: Fijian IGT in ODIN's XML format.}
-\end{minipage}
-\\ \\
+\caption{Fijian IGT in ODIN's XML format.} \label{fig:odin_fijian_xml}
+\end{figure}
+
+% \begin{minipage}{\columnwidth}
+% \begin{lstlisting}[language=XML]
+% <igt>
+%   <example>
+%     <line>21 a. o lesu mai</line>
+%     <line>2sg return here</line>
+%     <line>`You return here.'</line>
+%   </example>
+% </igt>
+% \end{lstlisting} 
+% \centerline{Figure 2: Fijian IGT in ODIN's XML format.}
+% \end{minipage}
+
 \noindent The IGTs from an academic publication usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
@@
 Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
-\begin{minipage}{\columnwidth}
+
+\begin{figure}
 \begin{lstlisting}[language=XML]
@@
 \smallskip
-\centerline{Figure 3: Yimas IGT in ODIN's XML format.}
-\end{minipage}
-\\ \\
+\caption{Yimas IGT in ODIN's XML format.}\label{fig:odin_yimas_xml}
+\end{figure}
+
+% \begin{minipage}{\columnwidth}
+% \begin{lstlisting}[language=XML]
+% <igt>
+%   <example>
+%     <line>(69) na-Na-tmi-kwalca-t 
+%     Yimas (Foley 1991)</line>
+%     <line>3sgA-1sgO-say-rise-PERF
+%     </line>
+%     <line>`She woke me up' 
+%     (by verbal action)</line>
+%   </example>
+% </igit>
+% \end{lstlisting} 
+% \smallskip
+% \centerline{Figure 3: Yimas IGT in ODIN's XML format.}
+% \end{minipage}
 
@@
 If there's more than one match, look for other punctuation in the initial and final tokens.]
-\\ \\
-\begin{minipage}{\columnwidth}
+
+\begin{figure}
 \begin{itemize}
@@
 \end{itemize}
-\end{minipage}
-\\ \\
+\caption{Do we want this still as a figure?}
+\end{figure}
+
+% \begin{minipage}{\columnwidth}
+% \begin{itemize}
+% \item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
+% \begin{itemize}
+% \item[(i)]
+% \begin{Verbatim}
+% ^(?\s?\w{1,5}\s*[):.]\s*
+% \end{Verbatim}
+% \item[(ii)] 
+% \begin{Verbatim}
+% [\[\(].{1,}[\]\)]
+% \end{Verbatim}
+% \end{itemize}
+% \item \emph{Cleanest}: Only source lines without punctuation.
+% \end{itemize}
+% \end{minipage}
 

### 2014-03-20 docs/computEL/data.tex
converted  minipages to figures
Update data.tex

@@
 
+\subsection{Universal Corpus and Data Structure} \label{sec:structure}
+
+Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}.
+Aligned texts consist of multiple parallel documents, aligned at the document, sentence, or word level.
+The collection of parallel documents as a whole is assigned a unique identifier, with each individual document labelled by this identifier and the corresponding language code.
+A monolingual document is simply one for which the document ID only has a single language code associated with it.
+Sentences and words within documents are also assigned identifiers (relative to the document), if alignment has been performed at that level.
+
+Analyses texts, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.
+This is stored as a table, where each row represents a word, and each column a type of information.
+They propose the following columns: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
+
+Out of our data sources, three can be straightforwardly represented in their aligned text structure.
+However, ODIN contains richer annotations, which are in fact difficult to fit into their proposal, which we discuss in section \ref{sec:odin} below.
+
 \subsection{Data Sources} \label{sec:sources}
 
-Although data size matters in general NLP \cite{banko2001scaling} \textcolor{blue}{[is this the best reference?]}, \emph{universality} is the utmost priority for a universal corpus.
+Although data size matters in general NLP, \emph{universality} is the utmost priority for a universal corpus.
 We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured):
@@
 
-Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia and LCC) to parallel text (UDHR and Omniglot) to IGTs (ODIN).
+Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN).
+We describe each source in the following subsections.
 
-\subsubsection{ODIN}
+\subsubsection{Universal Declaration of Human Rights}
+
+The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations in 1948, and represents the first global expression of human rights.
+It consists of 30 articles, amounting to about four pages of text.
+This is a useful document for CL, since it has been translated into a wide variety of languages, providing a highly parallel text.
+
+\subsubsection{Wikipedia}
+
+Wikipedia\footnote{url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability.
+At the time of writing, it contained 30.8 million articles in 287 languages, which provides a sizeable amount of clean monolingual text in a fairly wide range of languages.
+Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.
 
@@
 
-\subsubsection{Universal Declaration of Human Rights}
+The Omniglot website\footnote{url{http://www.omniglot.com}} is an online encyclopedia of writing systems and languages.
+We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
+The urls for these resources are:
 
-\subsubsection{Wikipedia}
+\begin{itemize}[noitemsep]
+\item \url{www.omniglot.com/language/phrases/*} 
+\item \url{www.omniglot.com/babel/*}
+\end{itemize}
 
-\subsection{Universal Corpus and Data Structure} \label{sec:structure}
+\subsubsection{ODIN} \label{sec:odin}
 
-\textcolor{blue}{[Perhaps this comes elsewhere?]}
+ODIN (The Online Database of Interlinear Text) is a repository of interlinear glossed texts (IGTs) extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
+Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation.
+An IGT consists of three lines: (i) the source, a sentence in a target language, (ii) the gloss, an analysis of each element in the source, and (iii) the translation, done at the sentence level.
+The gloss line can additionally include a number of linguistic terms, which means that the gloss can more properly described as being written in a metalanguage, rather than a natural language.
+In ODIN, translations are into English, and glosses are written in an English-based metalanguage.
+An accepted set of guidelines are given by the Leipzig Glossing Rules,\footnote{\url{http://www.eva.mpg.de/lingua/resources/glossing-rules.php}}, where morphemes within words are separated by hyphens (or equal signs, for clitics), and the same number of hyphens should appear in each word of the source and gloss.
 
-\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level [add more detail]; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
-In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
+In contrast, the proposed data structure for the universal corpus is aligned at the word level, and includes a specific list of relevant features which should be used to annotate words.
+When we try to adapt IGTs into this format, we run into certain problems.
+Firstly, there is the problem that the most fundamental unit of analysis according to the Leipzig Glossing Rules is the morpheme, not the word.
+Ideally, we should encode this information explicitly in a universal corpus, assigning a unique identifier to each morpheme (instead of, or in addition to each word).
+Indeed, \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which undermines the central position of words in the proposed data structure.
 
-However, we disagree that analysed texts should be accommodated in this way.
-If such a universal corpus is to succeed at all, it must enjoy support from a substantial part, if not all, of the linguistics community, and doing this requires theory-neutrality.
-Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
-Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
-More fundamentally, they assume that the corpus should be segmented into words, but \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
+Secondly, it is unclear how to represent the gloss.
+Since the gloss line is written in a natural language, we cannot treat it as a simple translation.
+However, it is not straightforward to incorporate it into the proposed structure for analysed texts, either.
+We could simply always map the gloss to the \textsc{gloss} field, but this would run against the suggestion that this field should "contain symbols from controlled vocabularies".
+Indeed, this information might be more suited to the \textsc{morph, pos,} or \textsc{rel} fields, but it would be difficult to automatically process ODIN to determine which morphemes should be assigned to which field.
 
-It is not within the scope of this paper to resolve these theoretical concerns.
-Instead, we suggest that the data structure for a universal corpus should be even more lightweight than Abney and Bird suggest.
-We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
-To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource reference language as a convenient surrogate of meaning.
-By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules,\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} and as practised by documentary linguists.
+One possible resolution is to move all elements of the gloss written in capital letters to the \textsc{morph} field (as functional elements are usually annotated in this way), and all remaining elements to the \textsc{gloss} field.
+However, this loses information, since we no longer no which morpheme has which meaning.
+It would be ridiculous to throw away information in the IGTs, but to keep all information, we need to modify the \newcite{abney2011data}'s proposal.
 
-Then, only difference between analysed and aligned texts is that the text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
-The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies are difficult to raise.
-We admit that highly detailed annotations, such as trees and dependency graphs, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
-To make it indicate what reference language is being used, we propose prefixing \texttt{gloss-} to the language code of the reference language - for example \texttt{gloss-eng} and \texttt{gloss-eng} if a text has been glossed into English or Spanish, respectively.
-Using a prefix rather than a suffix makes it clear that this is not a subvariety of the given language.
+The simplest solution we can see is to allow morphemes to be a level of structure in the universal corpus, just as documents, sentences, and words already are.
+The overall architecture remains unchanged.
+We must then decide how to represent the glosses.
 
-Finally, it is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
-\newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining "seven dimensions" along which this can vary.
-[more detail?] Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
-We have written an API to allow access to this data according to the guidelines of Abney and Bird, who remain agnostic as to the specific form of data storage.
-If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.
+Even though glosses in ODIN are based on English, having been extracted from English-language documents, this is not true of IGTs in general.
+In particular, it is common for documentary linguists working on indigenous languages of the Americas to provide glosses and translations based on Spanish.
+For this reason, we believe it would be wise to specific the language used to produce the gloss.
+Since it is not quite the language itself, but a metalanguage, one solution would be to use new language codes that make it clear both that a metalanguage is being used, and also what natural language it is based on.
+The five-letter code \texttt{gloss} cannot be confused with any code in any version of ISO 639 (which includes codes of length two to four).
+Following the convention that subvarieties of a language are indicated with suffixes, we can append the code of the natural language being used.
+For example, glosses into English and Spanish-based metalanguages would be given the codes \texttt{gloss-eng} and \texttt{gloss-spa}, respectively.
+
+One benefit of this approach is that glossed texts are treated in exactly the same way as parallel texts.
+There is a unique identifier for each morpheme, and glosses are stored under this identifier and the corresponding gloss code.
+Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as a convenient surrogates of meaning.
+By the same reasoning, we can use glosses to provide a more detailed surrogate of meaning, only written in a metalanguage instead of a natural one.

### 2014-03-20 docs/computEL/data.tex
new heatmap file, added to data.tex

@@
 Across all our data sources, we manage to cover 1347 languages in 106 families, which represents 19.0\% of the world's languages.
-To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) in figure \textcolor{blue}{[insert reference here when it's ready]}.
+To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) in figure \ref{fig:heatmap}.
 The values in each cell have been colored according to proportion of languages represented, with green indicated good coverage and red poor.
@@
 \begin{table}[h!]
+\small
 \centering
@@
 
-\textcolor{blue}{[insert heatmap here]}
+\begin{figure*}[t]
+\begin{centering}
+\includegraphics[scale=0.6]{heatmap-numbers.png}
+
+\caption{Heatmap of languages in HLPSC according to endangerment status}\label{fig:heatmap}
+\end{centering}
+\end{figure*}
 

### 2014-03-20 docs/computEL/clean.tex
Update clean.tex

@@
 
-In the following subsections, we describe each source, and the processing required to convert the data into a standardised form.
+In the following subsections, we describe the processing required to convert the data into a standardised form.  We then discuss standardisation of language codes and file formats.
 
@@
 
-The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages.
-Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
+Although there is a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution is laden with encoding problems during their conversion from pdf to plaintext format.
 Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d}, which are free of encoding issues.
@@
 
-At the time of writing, Wikipedia contains 30.8 million articles in 287 languages, which provides a sizeable amount of clean monolingual text in a fairly wide range of languages.
-Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.
 One major issue with using the Wikipedia dump extracting text from a glut of Wikipedia markup.
@@
 
-Wikipedia uses its own set of language codes, most of which are in ISO~639-1 or ISO~639-2/3.
-We converted all of them into ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{http://meta.wikimedia.org/wiki/Special\_language\_codes}
-
 \subsection{Omniglot}
 
-The Omniglot website\footnote{htp://www.omniglot.com} is an online encyclopedia of writing systems and languages.
-We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
-The urls for these resources are:
-
-\begin{itemize}[noitemsep]
-\item \url{www.omniglot.com/language/phrases/*} 
-\item \url{www.omniglot.com/babel/*}
-\end{itemize}
-
-\noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
-To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
+The main issue with extracting this data is that the pages are designed to be human-readable, not machine-readable.
+Cleaning this data required parsing the html source, and extracting the relevant content, which required different code for the two types of page we considered (useful phrases, and the Tower of Babel story).
+Even after automatic extraction, some noise in the data remained, such as explanatory notes given in parentheses, which are written in English and not the target language.
+Even though the total amount of data here is small compared to our ther sources, the amount of effort required to process it was not, because of these idiosyncracies.
+We expect that researchers seeking to convert data from human-readable to machine-readable formats will encounter similar problems, but unfortunately there is unlikely to be a one-size-fits-all solution to this problem.
 
@@
 
-ODIN (The Online Database of Interlinear Text) is a repository of IGTs extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
-Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation; however, the data requires further processing to bring it in line with the proposed format for a universal corpus.
-
-The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
-Each XML file is saved under a filename that matches the respective language code.
-While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.
-[Huh?
-Should this really happen?
-- the codes are given on their website and look valid to me...]
+The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the IGTs (described in section \ref{sec:odin}) are encoded in tags of the form \texttt{<igt><example>...</example></igt>}.
+For example, the IGT in Figure 1 is represented by the XML snippet in Figure 2.
 
@@
 % \end{minipage}
-
-%%
-% Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
-% Also, some of the lines go outside of the column!!
-%%
 
@@
 % \end{minipage}
-
-\noindent The IGTs from an academic publication usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
-From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2.
-Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
 
@@
 
-\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
+The primary problem in extracting the data is a lack of consistency in the IGTs.
 In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
-In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.
+In addition, the source line in Figure 3 includes two types of metadata: the language name, and a citation, both of which introduce noise.
+Finally, extraneous punctuation such as the quotation marks in the translation line need to be removed.
 
-Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness':
-
-[This hasn't been implemented, but I think we could try this: Split source and gloss lines into tokens according to whitespace.
-Check the source has more tokens.
-Count the number of hyphens and equal signs in each token (since they denote morphemes).
-Match the gloss line with a substring of the source line, according to numbers of hyphens and equal signs.
-If there's exactly one match, keep it, and discard tokens at the ends of the line.
-If there's more than one match, look for other punctuation in the initial and final tokens.]
+To clean the source lines, we used the following regular expressions:
 
@@
 \end{itemize}
-\caption{Do we want this still as a figure?}
+\caption{I don't think we want this as a figure.}
 \end{figure}
@@
 
-\noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
-The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
+The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
+\textcolor{blue}{Maybe we should just remove this next sentence:} The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
+
+In future work, we intend to explore more sophisticated methods of cleaning.
+For example, we could leverage the fact that the number of hyphens and equal signs in each word should match between the source and gloss lines.
+However, it must be stressed that no method is likely to be foolproof.
+As people start to incorporate large sets of IGTs from documentary linguistic projects into a universal corpus, we can expect problems with inconsistencies to grow, and more time will need to be devoted to resolve this issue.
+
+\textcolor{blue}{[Feel free to modify the above paragraph]}
+
+\subsection{Language Codes}
+
+We chose to use ISO~639-3 as our standard set of language codes, since it aims for universal coverage, and has widespread acceptance in the community.
+As a result, we needed to make sure that all our data was labelled with these codes.
+Data from ODIN and the UDHR already used this standard, and hence did not pose problems.
+
+Wikipedia uses its own set of language codes, most of which are in ISO~639-1 or ISO~639-3.
+The older ISO~639-1 codes are easy to recognise, being two letters long instead of three, and can be straightforwardly converted.
+However, a small number of Wikipedia codes are not ISO codes at all - we converted these to ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{\url{http://meta.wikimedia.org/wiki/Special\_language\_codes}}
+
+Omniglot does not give codes at all, but only the language name.
+To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{\url{http://www-01.sil.org/iso639-3/iso-639-3.tab}}
+
+\subsection{File Formats}
+
+It is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
+\newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining "seven dimensions" along which this can vary.
+Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
+Metadata is stored separately.
+We have written an API to allow access to the data according to the guidelines of \newcite{abney2010universal}, who remain agnostic as to the specific form of data storage.
+If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.

### 2014-03-20 docs/computEL/intro.tex
new heatmap file, added to data.tex
various small mods

@@
 
-At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century.
-However, only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
+At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{krauss1992crisis} estimated 50\% of the extant languages at the time were not being learned by new generations of speakers, risking extinction by the end of the century.
+Only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
 
-Several years ago, Abney and Bird \shortcite{abney2010universal,abney2011data} posed the challenge of building a Universal Corpus (UC), calling it the Human Language Project.
+Several years ago, Abney and Bird \shortcite{abney2010universal,abney2011data} posed the challenge of building a Universal Corpus, calling it the Human Language Project.
 Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
-The challenge was issued to the computational linguistics (CL) community, from the perspective that the data management, language processing, and machine learning tools\footnote{\textcolor{blue}{am I missing anything?}} well-known in CL must be brought to bear on the problems of documentary linguistics (DL), if we are to make any serious progress toward building such a resource.
-The UC as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages (LRLs), which in turn drives developments in automated analysis, supporting the work of DL and increasing the efficiency of the language documentation process.
+The challenge was issued to the computational linguistics (CompLing) community, from the perspective that the language processing, machine learning, and data management tools\footnote{\textcolor{blue}{am I missing anything?}} well-known in CompLing must be brought to bear on the problems of documentary linguistics (DocLing), if we are to make any serious progress toward building such a resource.
+The Universal Corpus as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages, which in turn drives developments in automated analysis, supporting the work of DocLing and increasing the efficiency of the language documentation process.
 \textcolor{blue}{[cite our AL work?
@@
 \textcolor{blue}{[kind of clunky]} Of course there are many relevant efforts toward producing cross-linguistic resources, which we survey in section \ref{sec:related}.
-To the best of our knowledge, though, no existing effort meets these two desiderata to the extent of our corpus, which we will refer to throughout as the Human Language Project Seed Corpus (HLPSC).
-\textcolor{blue}{[other naming ideas?]}
+To the best of our knowledge, though, no existing effort meets these two desiderata to the extent of our corpus, which we name SeedLing: a seed corpus for the Human Language Project.
 
-To produce the HLPSC, we have drawn on four web sources, described in section \ref{sec:sources}.
+To produce SeedLing, we have drawn on four web sources, described in section \ref{sec:sources}.
 To bring the four resources into a single common format and data structure (section \ref{sec:structure}), each required different degrees and types of cleaning and standardisation.
@@
 By doing so we hope to make future resource conversion efforts more efficient, as they can be guided by the lessons we learned in assembling our seed corpus.
-To that end, many of the resources described in \ref{sec:related} are perhaps candidates for inclusion in the next stage of building a UC.
+To that end, many of the resources described in \ref{sec:related} are perhaps candidates for inclusion in the next stage of building a Universal Corpus.
 
-We believe the resulting corpus, which at present covers 1374 languages from 106 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing.
-To test this belief, we use the HLPSC in a test application (\ref{sec:cluster}): the task of language clustering.
-We use surface-level features (frequencies of character n-grams and words) extracted from our data to estimate the similarity of two languages.
+We believe the resulting corpus, which at present covers 1347 languages from 106 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing.
+Following this claim, we use SeedLing in a test application (\ref{sec:cluster}): the task of language clustering.
+With no additional pre-processing required, we extract surface-level features (frequencies of character n-grams and words) for estimating the similarity of two languages.
+Unlike most previous approaches to the task, we make no use of resources curated for aims of linguistic typology (e.g.
+values of typological features as in WALS \cite{wals}, Swadesh word lists).
 \textcolor{blue}{[START HERE WITH CLEARER DESCRIPTION OF CLUSTERING RESULTS]}

### 2014-03-20 docs/computEL/cluster.tex
added bib citations to cluster section

@@
 When working on low-resource or endangered languages, computational and documentary linguists face the issue of lack of resources and knowledge about the language.
-Often, having knowledge about related or similar language(s) provides useful lexical, syntactic or morphological minimal pairs across languages and also helps in bootstrapping language models in NLP (Yarowsky
-and Ngai, 2001; Xia and Lewis, 2007).
+Often, having knowledge about related or similar language(s) provides useful lexical, syntactic or morphological minimal pairs across languages and also helps in bootstrapping language models in NLP \cite{yarowsky:ngai:2001,xia2007multilingual}.
 
 The task of identifying related/similar language is also known as ``language classification" where (i) languages are mapped onto language families trees depending on their historical ancestry often measured by comparing string distance from a fix lexical list across languages, aka.
-\emph{genetic classification} or \emph{lexicostatistics} (e.g.
-Swadesh, 1952; Starostin, 2010) or (ii) languages are grouped by their degrees of similarity with regards to their typological features, aka.
-\emph{typological classification} (e.g.
-Daume, 2009; Georgi et al.
-2010).
+\emph{genetic classification} or \emph{lexicostatistics} \cite{swadesh1952,starostin2010} or (ii) languages are grouped by their degrees of similarity with regards to their typological features, aka.
+\emph{typological classification} \cite{georgi2010wals,daume2009}.
 

### 2014-03-20 docs/computEL/data.tex
various small mods
added bib citations to cluster section

@@
 \begin{centering}
-\includegraphics[scale=0.6]{heatmap-numbers.png}
-
-\caption{Heatmap of languages in HLPSC according to endangerment status}\label{fig:heatmap}
+\includegraphics[scale=0.63]{heatmap-numbers.png}
+%\includegraphics{heatmap-numbers.png}
+\caption{Heatmap of languages in SeedLing according to endangerment status}\label{fig:heatmap}
 \end{centering}
@@
 Although data size matters in general NLP, \emph{universality} is the utmost priority for a universal corpus.
-We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured):
+We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): The Online Database of Interlinear Text (ODIN), The Omniglot website, The Universal Declaration of Human Rights (UHDR), and Wikipedia.
 
-\begin{enumerate}
-\item The Online Database of Interlinear Text (ODIN)
-\item The Omniglot website
-\item The Universal Declaration of Human Rights (UDHR)
-\item Wikipedia
-\end{enumerate}
+% \begin{enumerate}
+% \item The Online Database of Interlinear Text (ODIN)
+% \item The Omniglot website
+% \item The Universal Declaration of Human Rights (UDHR)
+% \item Wikipedia
+% \end{enumerate}
 

### 2014-03-20 docs/computEL/paper.tex
Major edits to intro
new heatmap file, added to data.tex
various small mods
added bib citations to cluster section

@@
 \usepackage[utf8]{inputenc}
+\usepackage{graphicx}
 
@@
 
-\title{Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages}
+\title{SeedLing: Building and using a seed corpus \\ for the Human Language Project}
+%Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages}
 
@@
 \begin{abstract}
-\textcolor{blue}{[revisit title, revisit abstract]}
+\textcolor{blue}{[revisit abstract]}
 The study of endangered languages is constrained by a lack of data.
@@
 \input{cluster}
-\input{corpus}
+%\input{corpus}
 \input{concl}

### 2014-03-20 docs/computEL/intro.tex
modified discussion of clustering in intro

@@
 values of typological features as in WALS \cite{wals}, Swadesh word lists).
-\textcolor{blue}{[START HERE WITH CLEARER DESCRIPTION OF CLUSTERING RESULTS]}
-Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
+Despite our approach being highly dependent on orthography, we achieve a reasonable clustering performance over the languages in our dataset, demonstrating SeedLing's utility in cross-linguistic research.
+
+% \textcolor{blue}{[START HERE WITH CLEARER DESCRIPTION OF CLUSTERING RESULTS]}
+% Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
 Finally, we discuss future work in section \ref{sec:future}, and conclude in section \ref{sec:conclusion}.

### 2014-03-20 docs/computEL/paper.tex
modified discussion of clustering in intro

@@
 \textcolor{blue}{[revisit abstract]}
-The study of endangered languages is constrained by a lack of data.
-Existing corpora collections are limited in the range of languages covered, in standardisation, or in machine-readability.
+Computational study of endangered languages is constrained by the lack of digitally-available data.
+Existing corpora are limited in the range of languages covered, in standardisation, or in machine-readability.
 This makes the situation even worse for the computational linguist, especially one who would like to take a cross-linguistic or typological approach.

### 2014-03-20 docs/computEL/cluster.tex
Update cluster.tex
Discussion of clustering results

@@
 When working on low-resource or endangered languages, computational and documentary linguists face the issue of lack of resources and knowledge about the language.
-Often, having knowledge about related or similar language(s) provides useful lexical, syntactic or morphological minimal pairs across languages and also helps in bootstrapping language models in NLP \cite{yarowsky:ngai:2001,xia2007multilingual}.
+Often, having knowledge about related or similar languages provides useful lexical, syntactic or morphological minimal pairs across languages and also helps in bootstrapping language models in NLP \cite{yarowsky:ngai:2001,xia2007multilingual}.
 
-The task of identifying related/similar language is also known as ``language classification" where (i) languages are mapped onto language families trees depending on their historical ancestry often measured by comparing string distance from a fix lexical list across languages, aka.
-\emph{genetic classification} or \emph{lexicostatistics} \cite{swadesh1952,starostin2010} or (ii) languages are grouped by their degrees of similarity with regards to their typological features, aka.
-\emph{typological classification} \cite{georgi2010wals,daume2009}.
+Language classification can be carried out either: geneologically, mapping languages onto families trees depending on their historical ancenstry \cite{swadesh1952,starostin2010}; or typologically, grouping languages according to typological features \cite{georgi2010wals,daume2009}.
 
-To exemplify the use of the universal corpus for research on low-resource languages, we experimented with automatic detection of similar languages using K-means clustering with character ngrams and word unigrams features.
+To exemplify the use of the universal corpus for research on low-resource languages, we experimented with automatic detection of similar languages using hierarchical clustering with character ngrams and word unigrams features.
 Each language is represented by a vector of character bigrams and trigrams and word unigram features from the universal corpus.
@@
 \end{centering}
+\label{tab:cluster}
 \end{table*}
+
+\textcolor{blue}{[Use the above figures as a random baseline, and insert the new figures]}
 
@@
 see full table in http://goo.gl/bs2vPo
+
+There are many possible metrics to evaluate the quality of a clustering compared to a gold standar.
+\newcite{amigo2009metrics} propose a set of criteria which a clustering evaluation metric should satisfy, and demonstrate that most popular metrics fail to satisfy at least one of these criteria.
+However, they prove that they are satisfied by the BCubed metric, which we adopt for this reason.
+To calculate this, we find the induced cluster and gold standard class for each language, and calculate the F-score of the cluster compared to the class.
+These F-scores are then averaged across all languages.
+
+As well as performing clustering, we can view this as an information retrieval task: given a language, what are similar languages?
+To do this, we found the languages with the closest vectors of n-grams and words.
+Since BCubed is calculated averaging across languages and not clusters, we can use exactly the same calculation, using the set of nearby languages in place of a cluster.
+
+In table \ref{tab:cluster}, we give results of our clustering experiments.
+The f-scores are much higher than the random baseline, and comparable to the values reported by \newcite{georgi2010wals}, even though we have only used surface features, while they used typological features taken from WALS.
+
+It is also worth noting that precision is higher than recall.
+This is perhaps to be expected, given that related languages using wildly differing orthographies will appear to be very different.
+Nonetheless, our system is reasonably capable of identifying those languages which are both related and also written similarly.

### 2014-03-20 docs/computEL/cluster.tex
Update cluster.tex
Typo
Update cluster.tex
Shallowness

@@
 
-There are many possible metrics to evaluate the quality of a clustering compared to a gold standar.
+There are many possible metrics to evaluate the quality of a clustering compared to a gold standard.
 \newcite{amigo2009metrics} propose a set of criteria which a clustering evaluation metric should satisfy, and demonstrate that most popular metrics fail to satisfy at least one of these criteria.
@@
 In table \ref{tab:cluster}, we give results of our clustering experiments.
-The f-scores are much higher than the random baseline, and comparable to the values reported by \newcite{georgi2010wals}, even though we have only used surface features, while they used typological features taken from WALS.
+The F-scores are much higher than the random baseline, and comparable to the values reported by \newcite{georgi2010wals}, even though we have only used surface features, while they used typological features taken from WALS.
+This demonstrates that it possible for cross-linguistic research to be conducted even based on extremely shallow features.
 

### 2014-03-20 universalcorpus/simlang.py
added the corrected version of the clustering code and the simlang.py (finding similar languages by k-nearest neighbor).

@@ -1,0 +1,45 @@
+# -*- coding: utf-8 -*-
+
+from collections import defaultdict
+
+from crawlandclean import ethnologue
+from miniethnologue import ISO2LANG, MACRO2LANG, RETIRED2ISO
+
+
+dead = {"osp":"Old Spanish", "odt":"Old Dutch", "goh": "Old High German",
+        "got":"Gothic","wlm":"Middle Welsh","oge":"Old Georgian",
+        "tpn":"Tupinamba","ojp":"Old Japanese","sga":"Old Irish",
+        "hit":"Hittite","tkm":"Takelma","dum":"Middle Dutch",
+        "fro":"Old French","nci":"Classical Nahuatl","gmh":"Middle High German",
+        "mxi":"Mozarabic"}
+macro_split = {"nob":"Norwegian, Bokmaal","nno":"Norwegian Nynorsk"} # nor.
+constructed = {"ido":"Ido","tlh":"Klingon","tzl":"Talossan","jbo":"Lojban",
+               "ina":"Interlingua"}
+
+living_languages = set(ISO2LANG.keys()) - set(RETIRED2ISO.keys()) \
+- set(dead.keys()) - set(constructed.keys()) \
+- set(MACRO2LANG.keys()) - set(macro_split.keys())
+
+
+x = defaultdict(list)
+for line in open('distance.out'):
+  l1,l2, dist = line.strip().split('\t')
+  x[l1].append((dist,l2))
+  
+precisions, recalls, fscores = [], [], []
+for lang in x:
+  gold_class = ethnologue.FAMILIES2ISO[ethnologue.ISO2FAMILY[i]]
+  gold_class = [g for g in gold_class if g in living_languages]
+  induced_cluster = set([i for _,i in sorted(x['eng'][:21])])
+  
+  overlap = len(induced_cluster.intersection(gold_class))
+  rec = overlap /float(len(gold_class))
+  prec = overlap /float(len(induced_cluster))
+  
+  precisions.append(prec)
+  recalls.append(rec)
+  fscores.append((2*prec*rec)/float(prec+rec))
+
+print "{0:.5f}".format(avg(precisions)), "{0:.5f}".format(avg(recalls)), "{0:.5f}".format(avg(fscores))
+  
+  

### 2014-03-20 universalcorpus/langcluster.py
added simlang.py and cluster.eval.outputs

@@ -14,11 +14,12 @@ import numpy as np
 import pandas as pd
 ##from sklearn import cluster; import milk
 import scipy
+import scipy.spatial
+import scipy.cluster
 import matplotlib.pylab as plt
 
 from crawlandclean import odin, omniglot, udhr, ethnologue
 from miniethnologue import ISO2LANG, MACRO2LANG, RETIRED2ISO
-
 
 random.seed(0)
 
@@ -45,6 +46,7 @@ def normalize(featurevector, length=1):
     featurevector[feat] *= norm
   return featurevector
 
+
 dead = {"osp":"Old Spanish", "odt":"Old Dutch", "goh": "Old High German",
         "got":"Gothic","wlm":"Middle Welsh","oge":"Old Georgian",
         "tpn":"Tupinamba","ojp":"Old Japanese","sga":"Old Irish",
@@ -55,15 +57,17 @@ macro_split = {"nob":"Norwegian, Bokmaal","nno":"Norwegian Nynorsk"} # nor.
 constructed = {"ido":"Ido","tlh":"Klingon","tzl":"Talossan","jbo":"Lojban",
                "ina":"Interlingua"}
 
-living_languages = set(ISO2LANG.keys()) - set(RETIRED2ISO.keys()) - \
-set(dead.keys()) - set(constructed.keys()) - set(MACRO2LANG.keys()) - \
-set(macro_split.keys())
+living_languages = set(ISO2LANG.keys()) - set(RETIRED2ISO.keys()) \
+- set(dead.keys()) - set(constructed.keys()) \
+- set(MACRO2LANG.keys()) - set(macro_split.keys())
 
 def generate_ngrams(data_source):
   twograms = defaultdict(Counter) 
   threegrams = defaultdict(Counter)
   unigrams = defaultdict(Counter)
   for lang, sent in globals()[data_source].source_sents():
+    if lang not in living_languages:
+      continue
     if lang not in living_languages: continue;
     twograms[lang].update(Counter(sent2ngrams(sent,2))) 
     threegrams[lang].update(Counter(sent2ngrams(sent,3)))
@@ -107,14 +111,15 @@ def load_ngram_array():
     unigrams[lang] = normalize(unigrams[lang], 1/sqrt(3))
     
   allgrams = defaultdict(Counter)
-  for k,v in chain(twograms.iteritems(), threegrams.iteritems(), 
-                   unigrams.iteritems()):
-    allgrams[k].update(v)  
[…]

### 2014-03-20 universalcorpus/langcluster.py
added killswitch if evaluation outputs are already pulled in local repo.

@@ -225,6 +225,9 @@ condensed = scipy.spatial.distance.squareform(matrix)
 methods = 'single complete average weighted'.split()
 criterion = 'inconsistent distance maxclust'.split()
 
+if os.path.exist('cluster.eval.outputs'):
+  exit()
+  
 fouteval = open('cluster.eval.outputs','w')
 numclust = [10,20,30,40,50,60,70,80,90,100,110,120,130,147,148]
 for me, cr, nc  in product(methods,criterion, numclust):

### 2014-03-21 docs/computEL/liling.tex
Mostly reorganizational edits
added liling.tex with mini-section

--- 2014-03-19 docs/computEL/liling.tex
@@
 
-\section{Corpus Coverage and Universality, something, something numbers...}
-
+\section{Corpus Coverage and Universality, something, something numbers, heatmap, somehow, somewhere endangered...}
+% I think guy already wrote a small paragraph to describe the numbers in the table and the heatmap.
 \begin{table}[h!]
@@
 
-According to ethnologue
-Total Living languages: 7105
-Total Living language families: 147
+\noindent
+According to ethnologue \\
+Total Living languages: 7105 \\
+Total Living language families: 147 \\
 
-6,749
-33,117
+\newpage
+\section{Clustering}
+When working on low-resource or endangered languages, computational and documentary linguists face the issue of lack of resources and knowledge about the language.
+Often, having knowledge about related or similar languages provides useful lexical, syntactic or morphological minimal pairs across languages and also helps in bootstrapping language models in NLP \cite{yarowsky:ngai:2001,xia2007multilingual}.
+
+Language classification can be carried out either: geneologically, mapping languages onto families trees depending on their historical ancenstry \cite{swadesh1952,starostin2010}; or typologically, grouping languages according to typological features \cite{georgi2010wals,daume2009}.
+
+To exemplify the use of the universal corpus for research on low-resource languages, we experimented with automatic detection of similar languages using hierarchical clustering with character ngrams and word unigrams features.
+Each language is represented by a vector of frequencies of  character bigrams, character trigrams and word unigram features.
+Each of the three components in the vector is normalized by unit length.
+
+\subsection{Experimental Setup}
+We implemented hierarchical/agglomerative clustering using a variety of linkage methods: (i) \texttt{single}, (ii) \texttt{complete} and (iii) \texttt{weighted} (WPGMA, weighted pair group method with averaging).
+The \texttt{single} method calculates the distance between the newly formed clusters by assigning the minimal distance between the clusters and the \texttt{complete} method assigns the maximal distance between the newly formed clusters.
+The \texttt{weighted} method assigns the averaged distance between the newly formed cluster and its intermediate clusters.
+
+The \texttt{maxclust} criterion for flattening clusters from the hierarchy is achieved by finding the minimum threshold, \emph{r}, such that the cophenetic distance between any two original observations in the same flat cluster is no more than \emph{r} and not more than the number of clusters.
+
+\newpage
+\subsection{Results}
+\begin{table*}[h!]
+\begin{centering}
+
+    \begin{tabular}{l|ccc}
+    ~        & Precision & Recall       & F-score    \\ \hline
+    \texttt{single} & 0.1958	& 0.6755	 & 0.1240  \\
+	\texttt{complete} & \textbf{0.3153}	& 0.1699	 & \textbf{0.1420} \\
+	\texttt{weighted} & 0.0614	& \textbf{0.8565}	 & 0.1099 \\ \hline
+	\emph{random} & 0.1925 &	0.0927 & 	0.0692 \\
+    \end{tabular}
+\caption{Comparing of clustering algorithms where number of clusters is set to genetic grouping on Ethnologue.}
+\end{centering}
+\label{tab:cluster}
+\end{table*}
+
+\newpage
+\section{Conclusion}
+\subsection{Deliverables/API/something something to say that we have script to access the data easily and CompLing and DocLing people can use them with NLTK}
+Other than the clean corpus data, SeedLing comes with a python API to access the data programmatically and it is easily compatible with the analysis tools available from the Natural Language ToolKit (NLTK).
+In addition, the clustering script used in section X will also be made free available with an open source license.
+
+The API also includes a feature to access information on a language from Ethnologue and/or WALS.
+This feature queries the Ethnologue and/or WALS site both in real-time when internet connection is available and from a local copy of the information offline.
 

### 2014-03-21 docs/computEL/data.tex

@@
 Across all our data sources, we manage to cover 1347 languages in 106 families, which represents 19.0\% of the world's languages.
-To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) in figure \ref{fig:heatmap}.
+To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) \cite{lewis2010assessing} in figure \ref{fig:heatmap}.
 The values in each cell have been colored according to proportion of languages represented, with green indicated good coverage and red poor.

### 2014-03-22 docs/computEL/cluster.tex
finally working on final version

@@
 \noindent
-see full table in http://goo.gl/bs2vPo
 
@@
 Nonetheless, our system is reasonably capable of identifying those languages which are both related and also written similarly.
+
+%%%%%%%%%%%
+
+\section{Clustering}
+When working on low-resource or endangered languages, computational and documentary linguists face the issue of lack of resources and knowledge about the language.
+Often, having knowledge about related or similar languages provides useful lexical, syntactic or morphological minimal pairs across languages and also helps in bootstrapping language models in NLP \cite{yarowsky:ngai:2001,xia2007multilingual}.
+
+Language classification can be carried out either: geneologically, mapping languages onto families trees depending on their historical ancenstry \cite{swadesh1952,starostin2010}; or typologically, grouping languages according to typological features \cite{georgi2010wals,daume2009}.
+
+To exemplify the use of the universal corpus for research on low-resource languages, we experimented with automatic detection of similar languages using hierarchical clustering with character ngrams and word unigrams features.
+Each language is represented by a vector of frequencies of  character bigrams, character trigrams and word unigram features.
+Each of the three components in the vector is normalized by unit length.
+
+\subsection{Experimental Setup}
+We implemented hierarchical/agglomerative clustering using a variety of linkage methods: (i) \texttt{single}, (ii) \texttt{complete} and (iii) \texttt{weighted} (WPGMA, weighted pair group method with averaging).
+The \texttt{single} method calculates the distance between the newly formed clusters by assigning the minimal distance between the clusters and the \texttt{complete} method assigns the maximal distance between the newly formed clusters.
+The \texttt{weighted} method assigns the averaged distance between the newly formed cluster and its intermediate clusters.
+
+The \texttt{maxclust} criterion for flattening clusters from the hierarchy is achieved by finding the minimum threshold, \emph{r}, such that the cophenetic distance between any two original observations in the same flat cluster is no more than \emph{r} and not more than the number of clusters.
+
+\newpage
+\subsection{Results}
+\begin{table*}[h!]
+\begin{centering}
+
+    \begin{tabular}{l|ccc}
+    ~        & Precision & Recall       & F-score    \\ \hline
+    \texttt{single} & 0.1958	& 0.6755	 & 0.1240  \\
+	\texttt{complete} & \textbf{0.3153}	& 0.1699	 & \textbf{0.1420} \\
+	\texttt{weighted} & 0.0614	& \textbf{0.8565}	 & 0.1099 \\ \hline
+	\emph{random} & 0.1925 &	0.0927 & 	0.0692 \\
+    \end{tabular}
+\caption{Comparing of clustering algorithms where number of clusters is set to genetic grouping on Ethnologue.}
+\end{centering}
+\label{tab:cluster}
+\end{table*}

### 2014-03-22 docs/computEL/concl.tex
Mostly reorganizational edits
Major edits to intro
finally working on final version

@@
 We believe that this is the first corpus of its kind, which we hope will act as a seed corpus for the Human Language Project.
+
+\subsection{General recommendations for future sources?}
+
+\begin{itemize}
+\item next steps for DL?
+\item next steps for CL?
+\item uses of clustering?
+\end{itemize}
+
+\subsection{Deliverables/API/something something to say that we have script to access the data easily and CompLing and DocLing people can use them with NLTK}
+Other than the clean corpus data, SeedLing comes with a python API to access the data programmatically and it is easily compatible with the analysis tools available from the Natural Language ToolKit (NLTK).
+In addition, the clustering script used in section X will also be made free available with an open source license.
+
+The API also includes a feature to access information on a language from Ethnologue and/or WALS.
+This feature queries the Ethnologue and/or WALS site both in real-time when internet connection is available and from a local copy of the information offline.

### 2014-03-22 docs/computEL/paper.tex
updates to abstract

@@
 
+\newcommand{\bluenote}[1]{\textcolor{blue}{[#1]}}
+
 \setlength\titlebox{5cm}
@@
 \begin{abstract}
-\textcolor{blue}{[revisit abstract]}
-Computational study of endangered languages is constrained by the lack of digitally-available data.
-Existing corpora are limited in the range of languages covered, in standardisation, or in machine-readability.
-This makes the situation even worse for the computational linguist, especially one who would like to take a cross-linguistic or typological approach.
-We first survey existing efforts to compile cross-linguistic resources, then describe our own approach and give an example application - language clustering.
-To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contain data from a large number languages, and converted them into a standardised form consistent with the guidelines set out by \newcite{abney2010universal}.
-The resulting corpus is more easily-accessible and machine-readable than any of the underlying data sources, and represents a significant base corpus for researchers to draw on and add to in the future.
+  A bread-coverage corpus such as the Human Language Project
+  envisioned by \newcite{abney2010universal} would be a powerful
+  resource for the study of endangered languages.
+%, \bluenote{keep? from
+ %   the perspective of both the documentary and the computational
+  %  linguist}. 
+Existing corpora are limited in the range of languages
+  covered, in standardisation, or in machine-readability. In this
+  paper we present SeedLing, a seed corpus for the Human Language
+  Project. We first survey existing efforts to compile
+  cross-linguistic resources, then describe our own approach. To build
+  the foundation text for a universal corpus, we crawl and clean
+  texts from several web sources that contain data from a large number
+  of languages, and convert them into a standardised form consistent
+  with the guidelines set out by \shortcite{abney2010universal}. The
+  resulting corpus is more easily-accessible and machine-readable than
+  any of the underlying data sources, and represents a significant
+  base corpus for researchers to draw on and add to in the future.
+To show the utility of SeedLing for cross-lingual computational research, we use our data in the test application of clustering similar languages.
 \end{abstract}

### 2014-03-22 docs/computEL/data.tex
edits to intro, related work, and data

--- 2014-03-20 docs/computEL/data.tex
@@
 \section{The Data}\label{sec:data}
+
+\subsection{Universal Corpus and Data Structure} \label{sec:structure}
+
+Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}.
+Aligned texts consist of multiple parallel documents, aligned at the document, sentence, or word level.
+The collection of parallel documents as a whole is assigned a unique identifier, with each individual document labelled by this identifier and the corresponding language code.
+A monolingual document is simply one with only a single associated language code.
+Sentences and words within documents are also assigned identifiers (relative to the document), if alignment has been performed at that level.
+
+Analysed texts, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.
+This is stored as a table, where each row represents a word, and each column a type of information.
+They propose the following columns: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
+
+Out of our data sources, three can be straightforwardly represented in their aligned text structure.
+However, ODIN contains richer annotations, which are in fact difficult to fit into their proposal, and which we discuss in section \ref{sec:odin} below.
+
+\subsection{Data Sources} \label{sec:sources}
+
+Although data size matters in general NLP, \emph{universality} is the top priority for a universal corpus.
+We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): Online Database of Interlinear Text (ODIN), Omniglot website, Universal Declaration of Human Rights (UHDR), and Wikipedia.
+
+% \begin{enumerate}
+% \item The Online Database of Interlinear Text (ODIN)
+% \item The Omniglot website
+% \item The Universal Declaration of Human Rights (UDHR)
+% \item Wikipedia
+% \end{enumerate}
+
+Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN).
+Table~\ref{table:corpus} gives some coverage statistics, and we describe each source in the following subsections.
+
+\begin{table}[t]
+\small
+\centering
+    \begin{tabular}{l|rr|rr}
+    ~         				& \#Languages & \#Families 	&\#tokens		& Size	\\ \hline
+    ODIN      				& 1,217      & 101       		& 58,556		& 39 MB		\\
+    Omniglot  				& 134        & 21        		&	6,749			& 677 KB	\\
+    UDHR      				& 355        & 47        		&	33,117		& 5.2 MB	\\
+    Wikipedia 				& 209        & 22       		&						& 37 GB		\\ \hline
+    \textbf{Combined}	& 1,347			 & 106 
+    \end{tabular}
+\caption{Corpus Coverage}
+\label{table:corpus}
+\end{table}
+
+\paragraph{Universal Declaration of Human Rights.}
+
+The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations in 1948, and represents the first global expression of human rights.
+It consists of 30 articles, amounting to about four pages of text.
+This is a useful document for CL, since it has been translated into a wide variety of languages, providing a highly parallel text.
+
+\paragraph{Wikipedia.}
+
+Wikipedia\footnote{url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability.
+At the time of writing, it contained 30.8 million articles in 287 languages, which provides a sizeable amount of monolingual text in a fairly wide range of languages.
+Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.
+
+\paragraph{Omniglot.}
+
+The Omniglot website\footnote{url{http://www.omniglot.com}} is an online encyclopedia of writing systems and languages.
+We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us parallel data in a reasonably large number of languages.
+% The urls for these resources are:
+
+% \begin{itemize}[noitemsep]
+% \item \url{www.omniglot.com/language/phrases/*} 
+% \item \url{www.omniglot.com/babel/*}
+% \end{itemize}
+
+\paragraph{ODIN.} \label{sec:odin}
+
+ODIN (The Online Database of Interlinear Text) is a repository of
+interlinear glossed texts (IGTs) extracted from scholarly documents
+\cite{lewis2006odin,lewis2010odin}.  Compared to other resources, it
+is notable for the breadth of languages included and the level of
+linguistic annotation.  An IGT canonically consists of three lines:
+(i) the source, a sentence in a target language, (ii) the gloss, an
+analysis of each source element, and (iii) the translation,
+done at the sentence level. The gloss line can additionally include a
+number of linguistic terms, which means that the gloss 
+%can more
+%properly described as being written in a metalanguage, rather than a
+is written in metalanguage rather than natural language.  In ODIN, translations are into English, and glosses
+are written in an English-based metalanguage.  An accepted set of
+guidelines are given by the Leipzig Glossing
+Rules,\footnote{\url{http://www.eva.mpg.de/lingua/resources/glossing-rules.php}}
+where morphemes within words are separated by hyphens (or equal signs,
+for clitics), and the same number of hyphens should appear in each
+word of the source and gloss.
+
+The data from ODIN poses the first obstacle to straightforwardly adopting Abney and Bird's data structure. 
+The proposed data structure for the universal corpus is aligned at the word level, and includes a specific list of relevant features which should be used to annotate words.
+When we try to adapt IGTs into this format, we run into certain problems.
+Firstly, there is the problem that the most fundamental unit of analysis according to the Leipzig Glossing Rules is the morpheme, not the word.
+Ideally, we should encode this information explicitly in a universal corpus, assigning a unique identifier to each morpheme (instead of, or in addition to each word).
+Indeed, \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of \textit{word}, which undermines the central position of words in the proposed data structure.
+
+Secondly, it is unclear how to represent the gloss.
+Since the gloss line is not written in a natural language, we cannot treat it as a simple translation.
+However, it is not straightforward to incorporate it into the proposed structure for analysed texts, either.
+% We could simply always map the gloss to the \textsc{gloss} field, but this would run against the suggestion that this field should "contain symbols from controlled vocabularies".
+Indeed, this information might be more suited to the \textsc{morph, pos,} or \textsc{rel} fields, but it would be difficult to automatically process ODIN to determine which morphemes should be assigned to which field.
+One possible resolution is to move all elements of the gloss written in capital letters to the \textsc{morph} field (as functional elements are usually annotated in this way), and all remaining elements to the \textsc{gloss} field.
+However, this loses information, since we no longer know which morpheme has which meaning.
+To keep all information encoded in the IGT, we need to modify the \newcite{abney2011data}'s proposal.
+
+The simplest solution we can see is to allow morphemes to be a level of structure in the universal corpus, just as documents, sentences, and words already are.
+The overall architecture remains unchanged.
+We must then decide how to represent the glosses.
+
+Even though glosses in ODIN are based on English, having been
+extracted from English-language documents, this is not true of IGTs in
+general.  In particular, it is common for documentary linguists
+working on indigenous languages of the Americas to provide glosses and
+translations based on Spanish.  For this reason, we believe it would
+be wise to specify the language used to produce the gloss.  Since it
+is not quite the language itself, but a metalanguage, one solution
+would be to use new language codes that make it clear both that a
+metalanguage is being used, and also what natural language it is based
+on.  The five-letter code \texttt{gloss} cannot be confused with any
+code in any version of ISO 639 (with codes of length two to
+four).  Following the convention that subvarieties of a language are
+indicated with suffixes, we can append the code of the natural
+language.  For example, glosses into English and
+Spanish-based metalanguages would be given the codes
+\texttt{gloss-eng} and \texttt{gloss-spa}, respectively.
+
+One benefit of this approach is that glossed texts are treated in exactly the same way as parallel texts.
+There is a unique identifier for each morpheme, and glosses are stored under this identifier and the corresponding gloss code.
+Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as a convenient surrogates of meaning.
+By the same reasoning, we can use glosses to provide a more detailed surrogate of meaning, only written in a metalanguage instead of a natural one.
 
@@
 
-According to Ethnologue, there are 7105 living languages, and 147 living language families.
-Across all our data sources, we manage to cover 1347 languages in 106 families, which represents 19.0\% of the world's languages.
-To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) in figure \ref{fig:heatmap}.
-The values in each cell have been colored according to proportion of languages represented, with green indicated good coverage and red poor.
-It's interesting to note that vigorous languages (6a) are poorly represented across all data sources, and worse than more endangered categories.
-In terms of language documentation, vigorous languages are less urgent goals than those in categories 6b and up, but this highlights an unexpected gap in linguistic resources.
+According to Ethnologue, there are 7105 living languages, and 147
+living language families. Across all our data sources, we manage to
+cover 1347 languages in 106 families, which represents 19.0\% of the
+world's languages. To get a better idea of the kinds of languages
+represented, we give a breakdown according to their EGIDS scores
+(Expanded Graded Intergenerational Disruption Scale)
+\cite{lewis2010assessing} in Figure \ref{fig:heatmap}. The values in
+each cell have been colored according to proportion of languages
+represented, with green indicating good coverage and red poor. It's
+interesting to note that vigorous languages (6a) are poorly
+represented across all data sources, and worse than more endangered
+categories. In terms of language documentation, vigorous languages are
+less urgent goals than those in categories 6b and up, but this
+highlights an unexpected gap in linguistic resources.
 
@@
 
-\begin{table}[h!]
-\small
-\centering
-    \begin{tabular}{l|rr|rr}
-    ~         				& \#Languages & \#Families 	&\#tokens		& Size	\\ \hline
-    ODIN      				& 1,217      & 101       		& 58,556		& 39 MB		\\
-    Omniglot  				& 134        & 21        		&	6,749			& 677 KB	\\
-    UDHR      				& 355        & 47        		&	33,117		& 5.2 MB	\\
-    Wikipedia 				& 209        & 22       		&						& 37 GB		\\ \hline
-    \textbf{Combined}	& 1,347			 & 106 
-    \end{tabular}
-\caption{Corpus Coverage}
-\label{table:corpus}
-\end{table}
-
 \begin{figure*}[t]
@@
 \end{figure*}
-
-\subsection{Universal Corpus and Data Structure} \label{sec:structure}
-
-Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}.
-Aligned texts consist of multiple parallel documents, aligned at the document, sentence, or word level.
-The collection of parallel documents as a whole is assigned a unique identifier, with each individual document labelled by this identifier and the corresponding language code.
-A monolingual document is simply one for which the document ID only has a single language code associated with it.
-Sentences and words within documents are also assigned identifiers (relative to the document), if alignment has been performed at that level.
-
-Analyses texts, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.
-This is stored as a table, where each row represents a word, and each column a type of information.
-They propose the following columns: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
-
-Out of our data sources, three can be straightforwardly represented in their aligned text structure.
-However, ODIN contains richer annotations, which are in fact difficult to fit into their proposal, which we discuss in section \ref{sec:odin} below.
-
-\subsection{Data Sources} \label{sec:sources}
-
-Although data size matters in general NLP, \emph{universality} is the utmost priority for a universal corpus.
-We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): The Online Database of Interlinear Text (ODIN), The Omniglot website, The Universal Declaration of Human Rights (UHDR), and Wikipedia.
-
-% \begin{enumerate}
-% \item The Online Database of Interlinear Text (ODIN)
-% \item The Omniglot website
-% \item The Universal Declaration of Human Rights (UDHR)
-% \item Wikipedia
-% \end{enumerate}
-
-Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN).
-We describe each source in the following subsections.
-
-\subsubsection{Universal Declaration of Human Rights}
-
-The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations in 1948, and represents the first global expression of human rights.
-It consists of 30 articles, amounting to about four pages of text.
-This is a useful document for CL, since it has been translated into a wide variety of languages, providing a highly parallel text.
-
-\subsubsection{Wikipedia}
-
-Wikipedia\footnote{url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability.
-At the time of writing, it contained 30.8 million articles in 287 languages, which provides a sizeable amount of clean monolingual text in a fairly wide range of languages.
-Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.
-
-\subsubsection{Omniglot}
-
-The Omniglot website\footnote{url{http://www.omniglot.com}} is an online encyclopedia of writing systems and languages.
-We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
-The urls for these resources are:
-
-\begin{itemize}[noitemsep]
-\item \url{www.omniglot.com/language/phrases/*} 
-\item \url{www.omniglot.com/babel/*}
-\end{itemize}
-
-\subsubsection{ODIN} \label{sec:odin}
-
-ODIN (The Online Database of Interlinear Text) is a repository of interlinear glossed texts (IGTs) extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
-Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation.
-An IGT consists of three lines: (i) the source, a sentence in a target language, (ii) the gloss, an analysis of each element in the source, and (iii) the translation, done at the sentence level.
-The gloss line can additionally include a number of linguistic terms, which means that the gloss can more properly described as being written in a metalanguage, rather than a natural language.
-In ODIN, translations are into English, and glosses are written in an English-based metalanguage.
-An accepted set of guidelines are given by the Leipzig Glossing Rules,\footnote{\url{http://www.eva.mpg.de/lingua/resources/glossing-rules.php}}, where morphemes within words are separated by hyphens (or equal signs, for clitics), and the same number of hyphens should appear in each word of the source and gloss.
-
-In contrast, the proposed data structure for the universal corpus is aligned at the word level, and includes a specific list of relevant features which should be used to annotate words.
-When we try to adapt IGTs into this format, we run into certain problems.
-Firstly, there is the problem that the most fundamental unit of analysis according to the Leipzig Glossing Rules is the morpheme, not the word.
-Ideally, we should encode this information explicitly in a universal corpus, assigning a unique identifier to each morpheme (instead of, or in addition to each word).
-Indeed, \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which undermines the central position of words in the proposed data structure.
-
-Secondly, it is unclear how to represent the gloss.
-Since the gloss line is written in a natural language, we cannot treat it as a simple translation.
-However, it is not straightforward to incorporate it into the proposed structure for analysed texts, either.
-We could simply always map the gloss to the \textsc{gloss} field, but this would run against the suggestion that this field should "contain symbols from controlled vocabularies".
-Indeed, this information might be more suited to the \textsc{morph, pos,} or \textsc{rel} fields, but it would be difficult to automatically process ODIN to determine which morphemes should be assigned to which field.
-
-One possible resolution is to move all elements of the gloss written in capital letters to the \textsc{morph} field (as functional elements are usually annotated in this way), and all remaining elements to the \textsc{gloss} field.
-However, this loses information, since we no longer no which morpheme has which meaning.
-It would be ridiculous to throw away information in the IGTs, but to keep all information, we need to modify the \newcite{abney2011data}'s proposal.
-
-The simplest solution we can see is to allow morphemes to be a level of structure in the universal corpus, just as documents, sentences, and words already are.
-The overall architecture remains unchanged.
-We must then decide how to represent the glosses.
-
-Even though glosses in ODIN are based on English, having been extracted from English-language documents, this is not true of IGTs in general.
-In particular, it is common for documentary linguists working on indigenous languages of the Americas to provide glosses and translations based on Spanish.
-For this reason, we believe it would be wise to specific the language used to produce the gloss.
-Since it is not quite the language itself, but a metalanguage, one solution would be to use new language codes that make it clear both that a metalanguage is being used, and also what natural language it is based on.
-The five-letter code \texttt{gloss} cannot be confused with any code in any version of ISO 639 (which includes codes of length two to four).
-Following the convention that subvarieties of a language are indicated with suffixes, we can append the code of the natural language being used.
-For example, glosses into English and Spanish-based metalanguages would be given the codes \texttt{gloss-eng} and \texttt{gloss-spa}, respectively.
-
-One benefit of this approach is that glossed texts are treated in exactly the same way as parallel texts.
-There is a unique identifier for each morpheme, and glosses are stored under this identifier and the corresponding gloss code.
-Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as a convenient surrogates of meaning.
-By the same reasoning, we can use glosses to provide a more detailed surrogate of meaning, only written in a metalanguage instead of a natural one.

### 2014-03-22 docs/computEL/intro.tex
edits to intro, related work, and data

@@
 
-At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{krauss1992crisis} estimated 50\% of the extant languages at the time were not being learned by new generations of speakers, risking extinction by the end of the century.
+At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{simons2011update} calculate that 37\% of the extant languages at the time were at various stages of losing transmisson to new generations of speakers.
 Only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
@@
 Several years ago, Abney and Bird \shortcite{abney2010universal,abney2011data} posed the challenge of building a Universal Corpus, calling it the Human Language Project.
-Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
-The challenge was issued to the computational linguistics (CompLing) community, from the perspective that the language processing, machine learning, and data management tools\footnote{\textcolor{blue}{am I missing anything?}} well-known in CompLing must be brought to bear on the problems of documentary linguistics (DocLing), if we are to make any serious progress toward building such a resource.
-The Universal Corpus as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages, which in turn drives developments in automated analysis, supporting the work of DocLing and increasing the efficiency of the language documentation process.
-\textcolor{blue}{[cite our AL work?
-is this too strong?
-also, run-on sentence.
-also, perhaps mention this workshop, which has precisely these goals]}
+Such a corpus would include data from all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
+The challenge was issued to the computational linguistics community, from the perspective that the language processing, machine learning, and data manipulations and management tools well-known in computational linguistics must be brought to bear on the problems of documentary linguistics, if we are to make any serious progress toward building such a resource.
+The Universal Corpus as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages, which in turn drives developments in automated analysis, supporting the work and increasing the efficiency of the language documentation process.
+%\textcolor{blue}{[cite our AL work? is this too strong? also, run-on sentence. also, perhaps mention this workshop, which has precisely these goals]}
 
@@
 We aim for a collection of resources that includes data: (a) from as many languages as possible, and (b) in a format which is both in accordance with best practice archiving recommendations and readily accessible as training data for machine learning and for other computational methods.
-\textcolor{blue}{[kind of clunky]} Of course there are many relevant efforts toward producing cross-linguistic resources, which we survey in section \ref{sec:related}.
+Of course there are many relevant efforts toward producing cross-linguistic resources, which we survey in section \ref{sec:related}.
 To the best of our knowledge, though, no existing effort meets these two desiderata to the extent of our corpus, which we name SeedLing: a seed corpus for the Human Language Project.
@@
 To bring the four resources into a single common format and data structure (section \ref{sec:structure}), each required different degrees and types of cleaning and standardisation.
-We describe the steps required in detail in section \ref{sec:case_studies}, presenting each resource as a separate case study.
+We describe the steps required in section \ref{sec:case_studies}, presenting each resource as a separate case study.
 By doing so we hope to make future resource conversion efforts more efficient, as they can be guided by the lessons we learned in assembling our seed corpus.
-To that end, many of the resources described in \ref{sec:related} are perhaps candidates for inclusion in the next stage of building a Universal Corpus.
+To that end, many of the resources described in section~\ref{sec:related} are candidates for inclusion in the next stage of building a Universal Corpus.
 
 We believe the resulting corpus, which at present covers 1347 languages from 106 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing.
-Following this claim, we use SeedLing in a test application (\ref{sec:cluster}): the task of language clustering.
+To test this claim, we use SeedLing in a sample application (section \ref{sec:cluster}): the task of language clustering.
 With no additional pre-processing required, we extract surface-level features (frequencies of character n-grams and words) for estimating the similarity of two languages.

### 2014-03-22 docs/computEL/paper.tex
edits to intro, related work, and data

@@
   of languages, and convert them into a standardised form consistent
-  with the guidelines set out by \shortcite{abney2010universal}. The
+  with the guidelines of \newcite{abney2011data}. The
   resulting corpus is more easily-accessible and machine-readable than
@@
   base corpus for researchers to draw on and add to in the future.
-To show the utility of SeedLing for cross-lingual computational research, we use our data in the test application of clustering similar languages.
+To show the utility of SeedLing for cross-lingual computational research, we use our data in the test application of clustering languages.
 \end{abstract}

### 2014-03-22 docs/computEL/relwork.tex
edits to intro, related work, and data

@@
 
-\subsection{Traditional Archives}
-
+\paragraph{Traditional archives.}
 Many archives exist to store the wealth of traditional resources produced by the documentary linguistics community.
@@
 
-Some archives aim for a universal scope, including languages from all parts of the world.  Of particular note are:
+Some archives aim for a universal scope, including languages from all parts of the world.
+Of particular note are: The Language Archive, maintained by the Max Planck Institute of Psycholinguistics, which includes DoBeS (Dokumentation Bedrohter Sprachen) and many other projects;
+Collection Pangloss, maintained by LACITO (Langues et Civilisations à Tradition Orale);
+The Endangered Languages Archive (ELAR), maintained by the School of Oriental and African Studies, which forms one part of the Hans Rausing Endangered Languages Project (HRELP).
 
-The Language Archive, maintained by the Max Planck Institute of Psycholinguistics, which includes DoBeS (Dokumentation Bedrohter Sprachen), among many other projects.
-Collection Pangloss, maintained by LACITO (Langues et Civilisations à Tradition Orale).
-The Endangered Languages Archive (ELAR), maintained by the School of Oriental and African Studies, London, which forms one part of the Hans Rausing Endangered Languages Project (HRELP).
-
-Regional archives include:
-
-AILLA (Archive of Indigenous Languages of Latin America), at the University of Texas at Austin. (Not in OLAC)
-AIATSIS collections (Australian Institute of Aboriginal and Torres Strait Islander Studies) (Not in OLAC)
-California Language Archive, managed by the Survey of California and Other Indian Languages
-ANLA (Alaska Native Language Archive), at the University of Alaska Fairbanks.
-PARADISEC (Pacific and Regional Archive for Digital Sources in Endangered Cultures)
+Regional archives include: AILLA (Archive of Indigenous Languages of Latin America), University of Texas at Austin;
+AIATSIS collections (Australian Institute of Aboriginal and Torres Strait Islander Studies);
+California Language Archive, managed by the Survey of California and Other Indian Languages;
+ANLA (Alaska Native Language Archive), at the University of Alaska Fairbanks;
+PARADISEC (Pacific and Regional Archive for Digital Sources in Endangered Cultures).
 
@@
 Firstly, the data is not always machine readable.
-For instance, in their own words, ANLA consists of "mostly paper records".
-Even where the data is available digitally, these often take the form of scanned images, or audio files.
-While both can provide invaluable information, they are extremely difficult to process with a computer, requiring an impractical level of pre-processing of the image or audio, before linguistic analysis can begin.
-Even textual data, which avoids these issues, may not be available in a machine-readable form, being stored as pdfs or
-
+For instance, in their own words, ANLA consists of ``mostly paper records''.
+Even where the data is available digitally, these often take the form of scanned images or audio files.
+While both can provide invaluable information, they are extremely difficult to process with a computer, requiring an impractical level of image or video pre-processing before linguistic analysis can begin.
+Even textual data, which avoids these issues, may not be available in a machine-readable form, being stored as pdfs or other opaque formats.
 Secondly, when data is machine readable, the format can vary wildly.
@@
 It gives best practice recommendations, illustrated with eleven case studies, and provides input tools which link to the GOLD ontology language, and the OLAC metadata set.
-Further recommendations are given by \newcite{bird2003port}, who describes seven dimensions along which the portability of linguistic data can vary.
+Further recommendations are given by \newcite{bird2003port}, who describe seven dimensions along which the portability of linguistic data can vary.
 Various tools are available from The Language Archive at the Max Planck Institute for Psycholinguistics.
@@
 However, centralising access to disparate resources, while of course extremely helpful, does not solve the problem of inconsistent standards.
-Indeed, it can become difficult even to answer simple questions like "how many languages are represented?"
+Indeed, it can be difficult even to answer simple questions like ``how many languages are represented?''
 
-We do of course acknowledge that such resources have their purpose - for instance, there can be plenty of material to work with for a researcher who is looking to find data on a specific language, who is willing to put in a little time browsing through different catalogues, and who only needs human-readable data.
-However, for large-scale machine processing, they leave much to be desired.
+Such resources are invaluable for many purposes
+for instance, there can be plenty of material to work with for a
+researcher who is looking to find data on a specific language, who is
+willing to put in a little time browsing through different catalogues,
+and who only needs human-readable data.  However, for large-scale
+machine processing, they leave much to be desired.
 
-\subsection{Generic corpora collections}
-
-There are some corpora collections which do not take endangered languages to be a main aim, but which nonetheless cover an increasing number of languages.
+\paragraph{Generic corpus collections.}
+Some corpus collections exist which do not focus on endangered languages, but which nonetheless cover an increasing number of languages.
 
@@
 While language codes are used, they have not been standardised, so that multiple codes are used for the same language.
-
 Linguistic Data Consortium (LDC) and the European Language Resources Association (ELRA) both offer data in multiple languages.
@@
 
-It seems likely that resources which do not specifically aim to cover endangered languages will continue to be skewed heavily towards a handful of high-resource languages.
+% It seems likely that resources which do not specifically aim to cover endangered languages will continue to be skewed heavily towards a handful of high-resource languages.
 
-\subsection{Parallel Corpora}
+\paragraph{Parallel corpora.}
 
@@
 The OPUS corpus \cite{tiedemann2012opus} subsumes a number of other well-known parallel corpora, such as Europarl, and covers documents from 350 languages, with various language pairs.
-However, not all of the data has been manually checked, which introduces noise.
 
-\subsection{Web Corpora}
+%However, not all of the data has been manually checked, which introduces noise.
+
+\paragraph{Web corpora.}
 
@@
 HC Corpora currently provides download of corpora in 68 different language varieties, which vary in size from 2M to 150M words.
-The corpora are thus of a respectable size, but only 1% of the world's languages are represented.
+The corpora are thus of a respectable size, but only 1\% of the world's languages are represented.
 A further difficulty is that languages are named, without the corresponding ISO codes.
@@
 The Leipzig Corpora Collection (LCC)\footnote{http://corpora.uni-leipzig.de} \cite{biemann2007leipzig} provides download of corpora in 117 languages, and dictionaries in a number of others, bringing the total number of represented languages up to 230.
-The corpora are large, readily available, in plain text, and labelled with ISO codes.
+The corpora are large, readily available, in plain text, and labelled with ISO language codes.
 
@@
 
-While the above efforts are commendable, and allow cross-linguistic analysis to a certain extent, there is still a long way to go to producing a universal corpus.
+% While the above efforts are commendable, and allow cross-linguistic analysis to a certain extent, there is still a long way to go to producing a universal corpus.
 
-\subsection{Cross-linguistic projects}
+\paragraph{Cross-linguistic projects.}
 
@@
 Of particular note are the Endangered Languages Project, and the Rosetta Project.
+These projects are to be praised for their commitment to universality, but in their current forms it is difficult to use their data to perform large-scale NLP.
 
-However, as much as these projects are to be praised for their commitment to universality, it is difficult to use their data to perform large-scale NLP.
-A researcher looking for a data on a specific language can easily navigate to the language in question, and download the available resources - but trying to access all the data can be awkward.
+% A researcher looking for a data on a specific language can easily navigate to the language in question, and download the available resources - but trying to access all the data can be awkward.
 While the sleek online interface used by the Endangered Languages Project may make it more accessible to those unfamiliar with language processing, it is not ideal for a computational linguistics researcher.

### 2014-03-22 docs/computEL/clean.tex
edits to clean

@@
 \section{Data source case studies} \label{sec:case_studies}
-
-In this section we describe in detail the work required to merge each of our four data sources into a single common source. \textcolor{blue}{[Universal Corpus???]}
-
-In the following subsections, we describe the processing required to convert the data into a standardised form.  We then discuss standardisation of language codes and file formats.
+Consistency in data structures and formatting is essential to facilitate use of data in computational linguistics research \cite{palmer2010lilt}.
+In the following subsections, we describe the processing required to convert the data into a standardised form.
+We then discuss standardisation of language codes and file formats.
 
@@
 
-Although there is a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution is laden with encoding problems during their conversion from pdf to plaintext format.
-Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d}, which are free of encoding issues.
-The first four lines of each file records metadata, and the rest is the translation of the UDHR.
-This dataset was extremely clean, and simply required segmentation into sentences.
+Although there is a pre-compiled version of the UDHR data from the
+Natural Language ToolKit (NLTK) corpora
+distribution,\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}
+the distribution is laden with encoding problems 
+%during their
+%conversion from pdf to plaintext format. 
+Instead, we use the
+plaintext files available from the Unicode
+website,\footnote{http://unicode.org/udhr/d} which are free of
+encoding issues. The first four lines of each file record metadata,
+and the rest is the translation of the UDHR. This dataset is
+extremely clean, and simply required segmentation into sentences.
 
@@
 
-One major issue with using the Wikipedia dump extracting text from a glut of Wikipedia markup.
-To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
-After conversion into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
+One major issue with using the Wikipedia dump is the problem of separating text from abundant source-specific markup. To convert compressed
+Wikipedia dumps to textfiles, we used the WikiExtractor
+\footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor}
+tool. After conversion into textfiles, we used several regular expressions to
+delete residual Wikipedia markup and so-called ``magic
+words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
 
@@
 
-The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the IGTs (described in section \ref{sec:odin}) are encoded in tags of the form \texttt{<igt><example>...</example></igt>}.
-For example, the IGT in Figure 1 is represented by the XML snippet in Figure 2.
-
-\begin{figure}
-\quad 21 a.\quad o lesu mai \\
-\indent \qquad\qquad 2sg return here \\
-\indent \qquad\qquad `\emph{You return here.}' \\
-\caption{Fijian IGT from ODIN.} \label{fig:odin_fijian}
-\end{figure}
+The ODIN data is easily accessible in XML format from the online database,\footnote{http://odin.linguistlist.org/download} where data for each language is saved in a separate XML file and the IGTs are encoded in tags of the form \texttt{<igt><example>...</example></igt>}.
+For example, the IGT in figure~\ref{fig:odin_fijian} is represented by the XML snippet in figure~\ref{fig:odin_fijian_xml}.
 
@@
 
-\begin{figure}
+The primary problem in extracting the data is a lack of consistency in
+the IGTs. In the above examples, the sentence is introduced by a
+letter or number, which needs to be removed; however, the form of such
+indexing elements varies. In addition, the source line in figure~\ref{fig:odin_yimas_xml}
+includes two types of metadata: the language name, and a citation,
+both of which introduce noise.  Finally, extraneous punctuation such
+as the quotation marks in the translation line need to be removed. We used regular expressions for cleaning lines within the IGTs.
+
+\begin{figure}[t]
+\quad 21 a.\quad o lesu mai \\
+\indent \qquad\qquad 2sg return here \\
+\indent \qquad\qquad `\emph{You return here.}' \\
+\caption{Fijian IGT from ODIN.} \label{fig:odin_fijian}
+\end{figure}
+
+\begin{figure}[t]
+\small
 \begin{lstlisting}[language=XML]
@@
 
-\begin{figure}
+\begin{figure}[t]
+\small
 \begin{lstlisting}[language=XML]
@@
 
-The primary problem in extracting the data is a lack of consistency in the IGTs.
-In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
-In addition, the source line in Figure 3 includes two types of metadata: the language name, and a citation, both of which introduce noise.
-Finally, extraneous punctuation such as the quotation marks in the translation line need to be removed.
+% To clean the source lines, we used the following regular expressions:
 
-To clean the source lines, we used the following regular expressions:
-
-\begin{figure}
-\begin{itemize}
-\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
-\begin{itemize}
-\item[(i)]
-\begin{Verbatim}
-^(?\s?\w{1,5}\s*[):.]\s*
-\end{Verbatim}
-\item[(ii)] 
-\begin{Verbatim}
-[\[\(].{1,}[\]\)]
-\end{Verbatim}
-\end{itemize}
-\item \emph{Cleanest}: Only source lines without punctuation.
-\end{itemize}
-\caption{I don't think we want this as a figure.}
-\end{figure}
+% \begin{figure}
+% \begin{itemize}
+% \item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
+% \begin{itemize}
+% \item[(i)]
+% \begin{Verbatim}
+% ^(?\s?\w{1,5}\s*[):.]\s*
+% \end{Verbatim}
+% \item[(ii)] 
+% \begin{Verbatim}
+% [\[\(].{1,}[\]\)]
+% \end{Verbatim}
+% \end{itemize}
+% \item \emph{Cleanest}: Only source lines without punctuation.
+% \end{itemize}
+% \caption{I don't think we want this as a figure.}
+% \end{figure}
 
@@
 
-The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
+% The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
 \textcolor{blue}{Maybe we should just remove this next sentence:} The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
@@
 For example, we could leverage the fact that the number of hyphens and equal signs in each word should match between the source and gloss lines.
-However, it must be stressed that no method is likely to be foolproof.
-As people start to incorporate large sets of IGTs from documentary linguistic projects into a universal corpus, we can expect problems with inconsistencies to grow, and more time will need to be devoted to resolve this issue.
-
-\textcolor{blue}{[Feel free to modify the above paragraph]}
+However, no method is likely to be foolproof.
+As people start to incorporate large sets of IGTs from documentary linguistic projects into a universal corpus, we can expect problems with inconsistencies to grow.
 
@@
 It is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
-\newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining "seven dimensions" along which this can vary.
+\newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining seven dimensions along which this can vary.
 Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
@@
 We have written an API to allow access to the data according to the guidelines of \newcite{abney2010universal}, who remain agnostic as to the specific form of data storage.
-If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.
+If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according to these guidelines.

### 2014-03-22 docs/computEL/relwork.tex
submitted version

@@
 
-The Leipzig Corpora Collection (LCC)\footnote{http://corpora.uni-leipzig.de} \cite{biemann2007leipzig} provides download of corpora in 117 languages, and dictionaries in a number of others, bringing the total number of represented languages up to 230.
+The Leipzig Corpora Collection (LCC)\footnote{\url{http://corpora.uni-leipzig.de}} \cite{biemann2007leipzig} provides download of corpora in 117 languages, and dictionaries in a number of others, bringing the total number of represented languages up to 230.
 The corpora are large, readily available, in plain text, and labelled with ISO language codes.

### 2014-04-17 docs/computEL/data.tex

@@
 
-Wikipedia\footnote{url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability.
+Wikipedia\footnote{\url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability.
 At the time of writing, it contained 30.8 million articles in 287 languages, which provides a sizeable amount of monolingual text in a fairly wide range of languages.
@@
 
-The Omniglot website\footnote{url{http://www.omniglot.com}} is an online encyclopedia of writing systems and languages.
+The Omniglot website\footnote{\url{http://www.omniglot.com}} is an online encyclopedia of writing systems and languages.
 We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us parallel data in a reasonably large number of languages.

### 2014-04-17 docs/computEL/intro.tex

@@
 Such a corpus would include data from all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
-The challenge was issued to the computational linguistics community, from the perspective that the language processing, machine learning, and data manipulations and management tools well-known in computational linguistics must be brought to bear on the problems of documentary linguistics, if we are to make any serious progress toward building such a resource.
+The challenge was issued to the computational linguistics community, from the perspective that the language processing, machine learning, and data manipulation and management tools well-known in computational linguistics must be brought to bear on the problems of documentary linguistics, if we are to make any serious progress toward building such a resource.
 The Universal Corpus as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages, which in turn drives developments in automated analysis, supporting the work and increasing the efficiency of the language documentation process.
@@
 To bring the four resources into a single common format and data structure (section \ref{sec:structure}), each required different degrees and types of cleaning and standardisation.
-We describe the steps required in section \ref{sec:case_studies}, presenting each resource as a separate case study.
-By doing so we hope to make future resource conversion efforts more efficient, as they can be guided by the lessons we learned in assembling our seed corpus.
-To that end, many of the resources described in section~\ref{sec:related} are candidates for inclusion in the next stage of building a Universal Corpus.
+We describe the steps required in section \ref{sec:case_studies}, presenting each resource as a separate mini-case study.
+We hope to make future resource conversion efforts more efficient, as they can be guided by the lessons we learned in assembling our seed corpus.
+To that end, many of the resources described in section~\ref{sec:related} are candidates for inclusion in the next stage of building a universal corpus.
 

### 2014-04-17 docs/computEL/paper.tex

@@
 \begin{abstract}
-  A bread-coverage corpus such as the Human Language Project
+  A broad-coverage corpus such as the Human Language Project
   envisioned by \newcite{abney2010universal} would be a powerful
@@
   %  linguist}. 
-Existing corpora are limited in the range of languages
-  covered, in standardisation, or in machine-readability. In this
-  paper we present SeedLing, a seed corpus for the Human Language
-  Project. We first survey existing efforts to compile
-  cross-linguistic resources, then describe our own approach. To build
-  the foundation text for a universal corpus, we crawl and clean
-  texts from several web sources that contain data from a large number
-  of languages, and convert them into a standardised form consistent
-  with the guidelines of \newcite{abney2011data}. The
-  resulting corpus is more easily-accessible and machine-readable than
-  any of the underlying data sources, and represents a significant
-  base corpus for researchers to draw on and add to in the future.
-To show the utility of SeedLing for cross-lingual computational research, we use our data in the test application of clustering languages.
+  Existing corpora are limited in the range of languages covered, in
+  standardisation, or in machine-readability. In this paper we present
+  SeedLing, a seed corpus for the Human Language Project. We first
+  survey existing efforts to compile cross-linguistic resources, then
+  describe our own approach. To build the foundation text for a
+  universal corpus, we crawl and clean texts from several web sources
+  that contain data from a large number of languages, and convert them
+  into a standardised form consistent with the guidelines of
+  \newcite{abney2011data}. The resulting corpus is more
+  easily-accessible and machine-readable than any of the underlying
+  data sources, and, with data from 1347 languages covering 106
+  language families, represents a significant base corpus for
+  researchers to draw on and add to in the future. To show the utility
+  of SeedLing for cross-lingual computational research, we use our
+  data in the test application of detecting similar languages.
 \end{abstract}

### 2014-04-17 docs/computEL-final/clean.tex
copied computel to computel-final

--- 2014-03-22 docs/computEL/clean.tex
@@
-\section{Data source case studies} \label{sec:case_studies}
+\section{Data Clean-Up, Consistency, and Standardisation} \label{sec:case_studies}
 Consistency in data structures and formatting is essential to facilitate use of data in computational linguistics research \cite{palmer2010lilt}.
@@
 
-\subsection{UDHR}
+\subsection{Case Studies}
 
-Although there is a pre-compiled version of the UDHR data from the
+\paragraph{UDHR.} Although there is a pre-compiled version of the UDHR data from the
 Natural Language ToolKit (NLTK) corpora
-distribution,\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}
+distribution,\footnote{\url{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}}
 the distribution is laden with encoding problems 
@@
 plaintext files available from the Unicode
-website,\footnote{http://unicode.org/udhr/d} which are free of
+website,\footnote{\url{http://unicode.org/udhr/d}} which are free of
 encoding issues. The first four lines of each file record metadata,
@@
 
-\subsection{Wikipedia}
-
+\paragraph{Wikipedia.}
 One major issue with using the Wikipedia dump is the problem of separating text from abundant source-specific markup. To convert compressed
 Wikipedia dumps to textfiles, we used the WikiExtractor
-\footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor}
+\footnote{\url{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor}}
 tool. After conversion into textfiles, we used several regular expressions to
 delete residual Wikipedia markup and so-called ``magic
-words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
+words"\footnote{\url{http://en.wikipedia.org/wiki/Help:Magic\_words}}.
 
-\subsection{Omniglot}
-
+\paragraph{Omniglot.}
 The main issue with extracting this data is that the pages are designed to be human-readable, not machine-readable.
@@
 
-\subsection{ODIN}
-
-The ODIN data is easily accessible in XML format from the online database,\footnote{http://odin.linguistlist.org/download} where data for each language is saved in a separate XML file and the IGTs are encoded in tags of the form \texttt{<igt><example>...</example></igt>}.
+\paragraph{ODIN.}
+The ODIN data is easily accessible in XML format from the online database,\footnote{\url{http://odin.linguistlist.org/download}} where data for each language is saved in a separate XML file and the IGTs are encoded in tags of the form \texttt{<igt><example>...</example></igt>}.
 For example, the IGT in figure~\ref{fig:odin_fijian} is represented by the XML snippet in figure~\ref{fig:odin_fijian_xml}.
@@
 
-In future work, we intend to explore more sophisticated methods of cleaning.
+% In future work, we intend to explore more sophisticated methods of cleaning.
 For example, we could leverage the fact that the number of hyphens and equal signs in each word should match between the source and gloss lines.
-However, no method is likely to be foolproof.
-As people start to incorporate large sets of IGTs from documentary linguistic projects into a universal corpus, we can expect problems with inconsistencies to grow.
+%However, no method is likely to be foolproof.  
+%As people incorporate large sets of IGTs from documentary linguistic projects into a universal corpus, we can expect problems with inconsistencies to grow.
 
 \subsection{Language Codes}
-
-We chose to use ISO~639-3 as our standard set of language codes, since it aims for universal coverage, and has widespread acceptance in the community.
-As a result, we needed to make sure that all our data was labelled with these codes.
+We use ISO~639-3 as our standard set of language codes, since it aims for universal coverage, and has widespread acceptance in the community.
+% As a result, we needed to make sure that all our data was labelled with these codes.
 Data from ODIN and the UDHR already used this standard, and hence did not pose problems.

### 2014-04-17 docs/computEL-final/cluster.tex
copied computel to computel-final

% \section{Detecting Similar Languages} \label{sec:cluster}

% When working on low-resource or endangered languages, computational and documentary linguists face the issue of lack of resources and knowledge about the language. Often, having knowledge about related or similar languages provides useful lexical, syntactic or morphological minimal pairs across languages and also helps in bootstrapping language models in NLP \cite{yarowsky:ngai:2001,xia2007multilingual}.

% Language classification can be carried out either: geneologically, mapping languages onto families trees depending on their historical ancenstry \cite{swadesh1952,starostin2010}; or typologically, grouping languages according to typological features \cite{georgi2010wals,daume2009}.

% To exemplify the use of the universal corpus for research on low-resource languages, we experimented with automatic detection of similar languages using hierarchical clustering with character ngrams and word unigrams features. Each language is represented by a vector of character bigrams and trigrams and word unigram features from the universal corpus. 
% \newline \newline
% \noindent \textbf{SHOW FANCY FIGURE OF CLUSTER}
% \newline \newline
% \textbf{Evaluate the cluster (quantitatively/qualitatively)}

%%%%%%%%%%%

\section{Detecting Similar Languages}

To exemplify the use of SeedLing for computational research that is relevant for low-resource languages, we experiment  with automatic detection of similar languages on the basis of our data. 
\label{sec:cluster} When working on low-resource or endangered 
languages, documentary and computational linguists alike face the
issue of lack of resources and knowledge about the language. Often,
having knowledge about related or similar languages provides useful
lexical, syntactic or morphological minimal pairs across
languages. What's more, information about relatedness between
languages can be useful for identifying data sources for high-resource
languages to be used in bootstrapping approaches such as those
described in \newcite{yarowsky:ngai:2001} or
\newcite{xia2007multilingual}.

Language classification can be carried out in a number of ways. Two common approaches are genealogical classification, mapping languages onto family trees according to their historical relatedness \cite{swadesh1952,starostin2010}; and typological classification, grouping languages according to certain linguistic features \cite{georgi2010wals,daume2009}. These approaches require linguistic analysis. By contrast, we use very simple features (character ngrams and word unigrams) extracted from SeedLing and an off-the-shelf hierarchical clustering algorithm.\footnote{\url{http://glaros.dtc.umn.edu/gkhome/views/cluto}} Specifically, each language is represented by a vector of frequencies of character bigrams, character trigrams, and word unigrams. Each of the three types of vector components is normalized by unit length.

% To exemplify the use of the universal corpus for research on low-resource languages, we experimented with automatic detection of similar languages using hierarchical clustering with character ngrams and word unigrams features. Each language is represented by a vector of frequencies of  character bigrams, character trigrams and word unigram features. Each of the three components in the vector is normalized by unit length.

% \begin{table*}[t]
% \begin{centering}

%     \begin{tabular}{l|ccc|ccc}
%     ~        & complete & ~       & ~       & ward    & ~       & ~       \\ \cline{2-7}
%     ~        & precision & recall       & f-score       & precision    & recall       & f-score      \\ \hline
%     distance & 0.0614   & 0.8565 & \textbf{0.1099} & 0.06140  & 0.8565 & \textbf{0.1099} \\
%     maxclust & 0.1925  & \textbf{0.0927} & 0.0692 & \textbf{0.1963} & 0.0905 & 0.0686  \\
%     \end{tabular}
% \caption{Comparing of clustering algorithms where number of clusters is set to genetic grouping on Ethnologue.}
% \end{centering}
% \label{tab:cluster}
% \end{table*}

\paragraph{Experimental Setup.}
We perform hierarchical/agglomerative clustering using a variety of linkage methods: (i) \texttt{single}, (ii) \texttt{complete} and (iii) \texttt{weighted}.
%(WPGMA, weighted pair group method with averaging). 
The \texttt{single} method calculates the distance between the newly formed clusters by assigning the minimal distance between the clusters and the \texttt{complete} method assigns the maximal distance between the newly formed clusters. The \texttt{weighted} method assigns the averaged distance between the newly formed cluster and its intermediate clusters. We set the number of clusters to 147, the number of top-level genetic groupings in Ethnologue (presumably the maximum number of language family clusters we would seek to induce).

% The \texttt{maxclust} criterion for flattening clusters from the hierarchy is achieved by finding the minimum threshold, \emph{r}, such that the cophenetic distance between any two original observations in the same flat cluster is no more than \emph{r} and not more than the number of clusters.

\begin{table}[t]
\begin{centering}

    \begin{tabular}{l|ccc}
    ~        & Precision & Recall       & F-score    \\ \hline
    \texttt{single} & 0.1958	& 0.6755	 & 0.1240  \\
	\texttt{complete} & \textbf{0.3153}	& 0.1699	 & \textbf{0.1420} \\
	\texttt{weighted} & 0.0614	& \textbf{0.8565}	 & 0.1099 \\ \hline
	\emph{random} & 0.1925 &	0.0927 & 	0.0692 \\
    \end{tabular}
\caption{Comparison of clustering algorithms} %where number of clusters is set to genetic grouping on Ethnologue.} 
\end{centering}
 \label{table:cluster}
\end{table}

\paragraph{Evaluation.}
There are many possible metrics to evaluate the quality of a clustering compared to a gold standard. 
\newcite{amigo2009metrics} propose a set of criteria which a clustering evaluation metric should satisfy, and demonstrate that most popular metrics fail to satisfy at least one of these criteria.  However, they prove that they are satisfied by the BCubed metric, which we adopt for this reason.  To calculate this, we find the induced cluster and gold standard class for each language, and calculate the F-score of the cluster compared to the class.  These F-scores are then averaged across all languages.

In table 2, we give results of our clustering experiments, comparing three clustering approaches to a random baseline.  The F-scores are comparable to those reported by \newcite{georgi2010wals}, even though we have only used surface features, while they used typological features taken from WALS.  This demonstrates that it possible for cross-linguistic research to be conducted even based on extremely shallow features.

% As well as performing clustering, we can view this as an information retrieval task: given a language, what are similar languages?  To do this, we found the languages with the closest vectors of n-grams and words.  Since BCubed is calculated averaging across languages and not clusters, we can use exactly the same calculation, using the set of nearby languages in place of a cluster.

% In table \ref{table:cluster}, we give results of our clustering experiments.  The F-scores are much higher than the random baseline, and comparable to the values reported by \newcite{georgi2010wals}, even though we have only used surface features, while they used typological features taken from WALS.  This demonstrates that it possible for cross-linguistic research to be conducted even based on extremely shallow features.

It is also worth noting that precision is higher than recall.  This is perhaps to be expected, given that related languages using wildly differing orthographies will appear to be very different.  Nonetheless, our system is reasonably capable of identifying those languages which are both related and also written similarly.

### 2014-04-17 docs/computEL-final/concl.tex
copied computel to computel-final

\section{Conclusion and Outlook} \label{sec:conclusion}

In this paper, we have described the creation of SeedLing, a foundation text for a universal corpus, following the guidelines of Abney and Bird (2010; 2011). To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia. The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}. At present, SeedLing has data from 19\% of the world's languages, covering 72\% of language families. We believe that a corpus with such high degrees of language diversity, uniformity and cleanliness of data format, and ease of access provides an excellent seed for the universal corpus. It is our hope that taking steps toward creating this resource will spur both further data contributions and interesting computational research with cross-linguistic or typological perspectives; we have here demonstrated SeedLing's utility for such research by using the data to perform language clustering, with promising results.

SeedLing will be made available initially through the websites of the authors, as we have yet to properly address the question of long-term access; we welcome ideas or collaborations along these lines. In addition, we are releasing a Python API which allows programmatic access to the data and is easily compatible with the analysis tools available from the Natural Language ToolKit \cite{BirdKleinLoper09}. The API also facilitates access to information from both Ethnologue and WALS: a user can extract information about languages, language classification, endangerment status, typological features, and so on. Finally, we will make our clustering scripts freely available.

% To demonstrate the utility of this resource, we used the data to perform language clustering, which gave promising results.

% The API also includes a feature to access information on a language from Ethnologue and/or WALS. This feature queries the Ethnologue and/or WALS site both in real-time when internet connection is available and from a local copy of the information offline.

### 2014-04-17 docs/computEL-final/corpus.tex
copied computel to computel-final

--- 2014-03-04 docs/acl-2014/paper.tex
@@
-\documentclass[11pt]{article}
-\usepackage{acl2014}
-\usepackage{times}
-\usepackage{url}
-\usepackage{latexsym}
-
-\bibliographystyle{acl}
-
-\usepackage[utf8]{inputenc}
-
-\usepackage{enumitem}
-\usepackage{listings}
-\usepackage{fancyvrb}
-\usepackage{color}
-\definecolor{gray}{rgb}{0.4,0.4,0.4}
-\definecolor{darkblue}{rgb}{0.0,0.0,0.6}
-\definecolor{cyan}{rgb}{0.0,0.6,0.6}
-\definecolor{orange}{rgb}{1,0.5,0}
-
-\lstset{
-  basicstyle=\ttfamily,
-  columns=fullflexible,
-  showstringspaces=false,
-  commentstyle=\color{gray}\upshape
-}
-
-\lstdefinelanguage{XML}
-{
-  morestring=[b]",
-  morestring=[s]{>}{<},
-  morecomment=[s]{<?}{?>},
-  stringstyle=\color{black},
-  identifierstyle=\color{darkblue},
-  keywordstyle=\color{cyan},
-  morekeywords={xmlns,version,type}% list your attributes here
-}
-
-\setlength\titlebox{5cm}
-
-\title{Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages}
-
-\author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
-  Universität des Saarlandes \\
-  Campus, 66123 Saarbrücken, Germany \\
-  {\tt liling.tan@uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-  {\tt susfert@coli.uni-saarland.de}}
-
-\date{}
-
-\begin{document}
-\maketitle
-\begin{abstract}
-The study of endangered languages is constrained by a lack of data.
-Existing corpora collections are limited in the range of languages covered, in standardisation, or in machine-readability.
-This makes the situation even worse for the computational linguist, especially one who would like to take a cross-linguistic or typological approach.
-We first survey existing efforts to compile cross-linguistic resources, then describe our own approach and give an example application - language clustering.
-To build the foundation text for a Universal Corpus, we crawled and cleaned texts from several web sources that contain data from a large number languages, and converted them into a standardised form consistent with the guidelines set out by \newcite{abney2010universal}.
-The resulting corpus is more easily-accessible and machine-readable than any of the underlying data sources, and represents a significant base corpus for researchers to draw on and add to in the future.
-\end{abstract}
-
-\section{Introduction}
-
-At the time of writing, 7105 living languages are documented in Ethnologue\footnote{http://www.ethnologue.com}, but \newcite{krauss1992crisis} estimated 50\% of these are not being learnt by new generations of speakers, and risk extinction by the end of the century.
-However, only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
-
-Although there have been some previous attempts to produce cross-linguistic resources, there are few which both cover a wide range of languages, and are also machine-readable.
-We survey existing efforts in section \ref{sec:related}, and discuss their limitations in more detail.
-
-Abney and Bird \shortcite{abney2010universal,abney2011data} posed the grand challenge of building a Universal Corpus, calling it the Human Language Project.
-Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
-They propose a specific data structure, which we describe and discuss in section \ref{sec:structure}.
-We have accepted their challenge, and have begun converting existing resources into a format consistent with their specifications for a universal corpus.
-We have drawn on four [or five?] web sources, cleaning and standardising them as described in section \ref{sec:sources}, to produce a seed corpus for the Human Language Project.
-In sections \ref{sec:stats} and \ref{sec:copyright}, we respectively give a summary of the data contained, and discuss copyright and distribution.
-
-[Note: remove this section if we don't do the work!] We believe the resulting corpus is the first of its kind: large enough and consistent enough to allow language processing on a grand scale.
-In section \ref{sec:cluster}, we give an example application of this corpus: language clustering.
-We use the frequencies of character n-grams and words to estimate the similarity of two languages.
-Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
-Finally, we discuss future work in section \ref{sec:future}, and conclude in section \ref{sec:conclusion}.
-
-\section{Related Work} \label{sec:related}
-
-[There is a lot more stuff to be added here...]
-Currently, multilingual corpora efforts have limited coverage in the number of languages and the number of language families represented.
-For instance, the OPUS corpus \cite{tiedemann2012opus} covers over 90 languages, only 1.27\% of the world's languages; The Leipzig Corpora Collection\footnote{corpora.uni-leipzig.de} \cite{biemann2007leipzig} contains corpora and dictionaries in 230 languages, 3.24\% of the world's languages.
-Even corpora that boast of linguistic diversity include only a small number of language families, e.g.
-the linguistically diverse NTU-Multilingual Corpus \cite{tan2011ntu} covers only 7 out of 136 language families.
-Producing a universal corpus requires the linguistic community to merge existing corpora and provide an ubiquitous access interface.
-To compile the foundation text for the Universal Corpus, we crawled and cleaned web data that contains multilingual texts and merged them with existing corpora collections to form the foundation text for the Universal Corpus.
+\section{Topics related to data release, etc.}
 
 \subsection{Web as Corpus}
+
+\textcolor{blue}{[This might want to go into related work...]}
 
@@
 
-\section{Data Structure} \label{sec:structure}
-
-\newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}: the former consists of multiple parallel texts, aligned at the document, sentence, or word level [add more detail]; the latter contains more detailed annotations including parts of speech, morphological information, and syntactic relations.
-In our work, we have strived to maintain a data structure consistent with their recommendations for aligned texts.
-
-However, we disagree that analysed texts should be accommodated in this way.
-If such a universal corpus is to succeed at all, it must enjoy support from a substantial part, if not all, of the linguistics community, and doing this requires theory-neutrality.
-Although their data structure is fairly lightwight, it is not theory neutral, explicitly encoding support for dependency grammars but not constituency grammars, and positing a specific list of relevant features which should be used to annotate words.
-Other linguists might wish to use a different set of properties, which could threaten to fragment a universal corpus before it truly gets off the ground.
-More fundamentally, they assume that the corpus should be segmented into words, but \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of "word", which would render such an endeavour impossible from the start.
-
-It is not within the scope of this paper to resolve these theoretical concerns.
-Instead, we suggest that the data structure for a universal corpus should be even more lightweight than Abney and Bird suggest.
-We agree with their characterisation of aligned texts, but propose an alternative for analysed texts.
-To motivate the role of parallel texts in a universal corpus, they propose using translations into a high-resource reference language as a convenient surrogate of meaning.
-By the same reasoning, we can use the glosses in an IGT to provide a more detailed surrogate of meaning.
-We propose that, just as we can align texts at the document, sentence, and word levels, we can also align texts at the morpheme level, as recommended by the Leipzig Glossing Rules,\footnote{http://www.eva.mpg.de/lingua/resources/glossing-rules.php} and as practised by documentary linguists.
-
-Then, only difference between analysed and aligned texts is that the text is aligned with descriptions in a metalanguage,\footnote{While we would urge researchers to keep such annotations simple, as usually done in IGTs, there is nothing in principle to stop someone from using a complex metalanguage capable of encoding the full set of properties proposed by Abney and Bird, or even more complicated data structures.} rather than a natural language.
-The benefits, however, are that we can use the same data structure to represent both aligned and analysed texts, and the representation is simple enough that linguistic controversies are difficult to raise.
-We admit that highly detailed annotations, such as trees and dependency graphs, are more awkward to encode, but such resources are unlikely to be available in more than a small handful of languages, and will remain outside the scope of a universal corpus, at least for the foreseeable future.
-To make it indicate what reference language is being used, we propose prefixing \texttt{gloss-} to the language code of the reference language - for example \texttt{gloss-eng} and \texttt{gloss-eng} if a text has been glossed into English or Spanish, respectively.
-Using a prefix rather than a suffix makes it clear that this is not a subvariety of the given language.
-
-Finally, it is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
-\newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining "seven dimensions" along which this can vary.
-[more detail?] Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
-We have written an API to allow access to this data according to the guidelines of Abney and Bird, who remain agnostic as to the specific form of data storage.
-If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according these guidelines.
-
-\section{Data Sources} \label{sec:sources}
-
-Although data size matters in general NLP \cite{banko2001scaling} [is this the best reference?], \emph{universality} is the utmost priority for a universal corpus.
-We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured):
-
-\begin{enumerate}
-\item The Online Database of Interlinear Text (ODIN)
-\item The Omniglot website
-\item The Universal Declaration of Human Rights (UDHR)
-\item Wikipedia
-\item The Leipzig Corpora Collection (LCC) [do we include this?]
-\end{enumerate}
-
-In the following subsections, we describe each source, and the processing required to convert the data into a standardised form.
-Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia and LCC) to parallel text (UDHR and Omniglot) to IGTs (ODIN).
-
-\subsection{ODIN}
-
-ODIN (The Online Database of Interlinear Text) is a repository of IGTs extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
-Compared to other resources, it is notable for the breadth of languages included, and the level of linguistic annotation; however, the data requires further processing to bring it in line with the proposed format for a universal corpus.
-
-The ODIN data is easily accessible in XML format from the online database\footnote{http://odin.linguistlist.org/download}, where data for each language is saved in a separate XML file and the Interlinear Glossed Texts (IGTs) are encoded in the \texttt{<igt><example>...</example></igt>} tags.
-Each XML file is saved under a filename that matches the respective language code.
-While cleaning the data, filenames that does not adhere to ISO 639-3\footnote{http://www-01.sil.org/iso639-3} were excluded in the uniWaC compilation.
-[Huh?
-Should this really happen?
-- the codes are given on their website and look valid to me...]
-\\ \\
-\begin{minipage}{\columnwidth}
-
-\quad 21 a.\quad o lesu mai \\
-\indent \qquad\qquad 2sg return here \\
-\indent \qquad\qquad `\emph{You return here.}' \\
-\centerline{Figure 1: Fijian IGT from ODIN.}
-\end{minipage}
-
-%%
-% Is there a reason we need to use minipages rather than figures? This doesn't seem to allow automatic numbering.
-% Also, some of the lines go outside of the column!!
-%%
-
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>21 a. o lesu mai</line>
-    <line>2sg return here</line>
-    <line>`You return here.'</line>
-  </example>
-</igt>
-\end{lstlisting} 
-\centerline{Figure 2: Fijian IGT in ODIN's XML format.}
-\end{minipage}
-\\ \\
-\noindent The IGTs from an academic publication usually follow the Leipzig Glossing Rules, where the first line is the source language text, the second line contains the word/morphemic equivalence and the third line is an English gloss.
-From the ODIN XML format, an IGT as of Figure 1 will be represented as in an XML snippet as in Figure 2.
-Eventually, we need to clean and extract (i) the source text `\emph{o lesu mai}' without the preceding index `\emph{a.}' and (ii) the target language gloss without the quotation marks `\emph{You return here}'.
-\begin{minipage}{\columnwidth}
-\begin{lstlisting}[language=XML]
-<igt>
-  <example>
-    <line>(69) na-Na-tmi-kwalca-t 
-    Yimas (Foley 1991)</line>
-    <line>3sgA-1sgO-say-rise-PERF
-    </line>
-    <line>`She woke me up' 
-    (by verbal action)</line>
-  </example>
-</igit>
-\end{lstlisting} 
-\smallskip
-\centerline{Figure 3: Yimas IGT in ODIN's XML format.}
-\end{minipage}
-\\ \\
-\noindent The primary problem in extracting the source text is a lack of consistency in the IGTs.
-In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
-In addition, the source line in Figure 3 includes the language name, and a citation for the example, which introduces noise into the corpus.
-
-Lewis and Xia (2010) noted that it is not uncommon for IGTs found in documents to deviate from the standard convention; when compiling the ODIN data, they attempted a regex-based approached and reported 59\% F-score.
-Similar to their regex approaches for extracting IGT, we cleaned the source line with regexes and kept the ODIN data with two levels of `cleanliness':
-
-[This hasn't been implemented, but I think we could try this: Split source and gloss lines into tokens according to whitespace.
-Check the source has more tokens.
-Count the number of hyphens and equal signs in each token (since they denote morphemes).
-Match the gloss line with a substring of the source line, according to numbers of hyphens and equal signs.
-If there's exactly one match, keep it, and discard tokens at the ends of the line.
-If there's more than one match, look for other punctuation in the initial and final tokens.]
-\\ \\
-\begin{minipage}{\columnwidth}
-\begin{itemize}
-\item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
-\begin{itemize}
-\item[(i)]
-\begin{Verbatim}
-^(?\s?\w{1,5}\s*[):.]\s*
-\end{Verbatim}
-\item[(ii)] 
-\begin{Verbatim}
-[\[\(].{1,}[\]\)]
-\end{Verbatim}
-\end{itemize}
-\item \emph{Cleanest}: Only source lines without punctuation.
-\end{itemize}
-\end{minipage}
-\\ \\
-
-\noindent The original version of the ODIN data contains XML files for 1275 languages, while the cleaner version of ODIN contains IGTs for 1042 languages and the cleanest version contains IGTs for 402 languages.
-The drop from 1275 to 1042 languages was largely because {\color{red} XXXX} XML files from the original ODIN data had used language codes that were not in ISO 639-3 and for {\color{red} XXXX} other files, the \texttt{<igt>...</igt>} tags were missing.
-
-\subsection{Omniglot}
-
-The Omniglot website\footnote{htp://www.omniglot.com} is an online encyclopedia of writing systems and languages.
-We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us a parallel data in a reasonably large number of languages.
-The urls for these resources are:
-
-\begin{itemize}[noitemsep]
-\item \url{www.omniglot.com/language/phrases/*} 
-\item \url{www.omniglot.com/babel/*}
-\end{itemize}
-
-\noindent One problem with this data is that only the language name is given, not the ISO 639-3 code.
-To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{http://www-01.sil.org/iso639-3/iso-639-3.tab}
-
-\subsection{Universal Declaration of Human Rights}
-
-The Universal Declaration of Human Rights (UDHR) is a document released by the United Nations, which has been translated into a wide variety of languages.
-Although there was a pre-compiled version of the UDHR data from the Natural Language ToolKit (NLTK) corpora distribution\footnote{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}, the distribution was laden with encoding problems during their conversion from pdf to plaintext format.
-Instead, we used the plaintext files available from the Unicode website\footnote{http://unicode.org/udhr/d}, which are free of encoding issues.
-The first four lines of each file records metadata, and the rest is the translation of the UDHR.
-This dataset was extremely clean, and simply required segmentation into sentences.
-
-\subsection{Wikipedia}
-
-At the time of writing, Wikipedia contains 30.8 million articles in 287 languages, which provides a sizeable amount of clean monolingual text in a fairly wide range of languages.
-Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.
-One major issue with using the Wikipedia dump extracting text from a glut of Wikipedia markup.
-To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor \footnote{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor} tool.
-After conversion into textfiles, we used the several regexes to delete residual Wikipedia markup and so-called ``magic words"\footnote{http://en.wikipedia.org/wiki/Help:Magic\_words}.
-
-Wikipedia uses its own set of language codes, most of which are in ISO~639-1 or ISO~639-2/3.
-We converted all of them into ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{http://meta.wikimedia.org/wiki/Special\_language\_codes}
-
-\section{Copyright Issues} \label{sec:copyright}
+\subsection{Copyright Issues} \label{sec:copyright}
 
@@
 %%
-
-\section{Representation and Universality} \label{sec:stats}
-
-Number of languages, size of corpus, etc.
-
-\section{Language Clustering} \label{sec:cluster}
-
-This would be an example application, if we have time...
-
-\section{Future Work} \label{sec:future}
-
-Specific suggestions on how to work on resources mentioned in section \ref{sec:related}...
-
-\section{Conclusion} \label{sec:conclusion}
-
-In this paper, we have described the creation of a foundation text for a universal corpus, following the guidelines of Abney and Bird (2010; 2011).
-To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia, and the LCC.
-The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}.
-[Give a summary statistic of size.] To demonstrate the utility of this resource, we used the data to perform language clustering, which gave promising results.
-We believe that this is the first corpus of its kind, which we hope will act as a seed corpus for the Human Language Project.
-
-\bibliography{references}
-
-\end{document}

### 2014-04-21 docs/computEL-final/cluster.tex
Rewritten description of clustering experiments

@@
-% \section{Detecting Similar Languages} \label{sec:cluster}
+\section{Detecting Similar Languages} \label{sec:cluster}
 
-% When working on low-resource or endangered languages, computational and documentary linguists face the issue of lack of resources and knowledge about the language.
-Often, having knowledge about related or similar languages provides useful lexical, syntactic or morphological minimal pairs across languages and also helps in bootstrapping language models in NLP \cite{yarowsky:ngai:2001,xia2007multilingual}.
+To exemplify the use of SeedLing for computational research on low-resource languages, we experiment with automatic detection of similar languages.
+When working on endangered languages, documentary and computational linguists alike face the issue of lack of resources.
+It is often helpful to exploit lexical, syntactic or morphological knowledge about related or similar languages.
+For example, information about relatedness can be useful for identifying high-resource languages to be used in bootstrapping approaches such as those described in \newcite{yarowsky:ngai:2001} or \newcite{xia2007multilingual}.
 
-% Language classification can be carried out either: geneologically, mapping languages onto families trees depending on their historical ancenstry \cite{swadesh1952,starostin2010}; or typologically, grouping languages according to typological features \cite{georgi2010wals,daume2009}.
-
-% To exemplify the use of the universal corpus for research on low-resource languages, we experimented with automatic detection of similar languages using hierarchical clustering with character ngrams and word unigrams features.
-Each language is represented by a vector of character bigrams and trigrams and word unigram features from the universal corpus.
-% \newline \newline
-% \noindent \textbf{SHOW FANCY FIGURE OF CLUSTER}
-% \newline \newline
-% \textbf{Evaluate the cluster (quantitatively/qualitatively)}
-
-%%%%%%%%%%%
-
-\section{Detecting Similar Languages}
-
-To exemplify the use of SeedLing for computational research that is relevant for low-resource languages, we experiment  with automatic detection of similar languages on the basis of our data. 
-\label{sec:cluster} When working on low-resource or endangered 
-languages, documentary and computational linguists alike face the
-issue of lack of resources and knowledge about the language. Often,
-having knowledge about related or similar languages provides useful
-lexical, syntactic or morphological minimal pairs across
-languages. What's more, information about relatedness between
-languages can be useful for identifying data sources for high-resource
-languages to be used in bootstrapping approaches such as those
-described in \newcite{yarowsky:ngai:2001} or
-\newcite{xia2007multilingual}.
-
-Language classification can be carried out in a number of ways.
-Two common approaches are genealogical classification, mapping languages onto family trees according to their historical relatedness \cite{swadesh1952,starostin2010}; and typological classification, grouping languages according to certain linguistic features \cite{georgi2010wals,daume2009}.
-These approaches require linguistic analysis.
-By contrast, we use very simple features (character ngrams and word unigrams) extracted from SeedLing and an off-the-shelf hierarchical clustering algorithm.\footnote{\url{http://glaros.dtc.umn.edu/gkhome/views/cluto}} Specifically, each language is represented by a vector of frequencies of character bigrams, character trigrams, and word unigrams.
-Each of the three types of vector components is normalized by unit length.
-
-% To exemplify the use of the universal corpus for research on low-resource languages, we experimented with automatic detection of similar languages using hierarchical clustering with character ngrams and word unigrams features.
-Each language is represented by a vector of frequencies of  character bigrams, character trigrams and word unigram features.
-Each of the three components in the vector is normalized by unit length.
-
-% \begin{table*}[t]
-% \begin{centering}
-
-%     \begin{tabular}{l|ccc|ccc}
-%     ~        & complete & ~       & ~       & ward    & ~       & ~       \\ \cline{2-7}
-%     ~        & precision & recall       & f-score       & precision    & recall       & f-score      \\ \hline
-%     distance & 0.0614   & 0.8565 & \textbf{0.1099} & 0.06140  & 0.8565 & \textbf{0.1099} \\
-%     maxclust & 0.1925  & \textbf{0.0927} & 0.0692 & \textbf{0.1963} & 0.0905 & 0.0686  \\
-%     \end{tabular}
-% \caption{Comparing of clustering algorithms where number of clusters is set to genetic grouping on Ethnologue.}
-% \end{centering}
-% \label{tab:cluster}
-% \end{table*}
+Language classification can be carried out in various ways.
+Two common approaches are genealogical classification, mapping languages onto family trees according to their historical relatedness \cite{swadesh1952,starostin2010}; and typological classification, grouping languages according to linguistic features \cite{georgi2010wals,daume2009}.
+Both of these approaches require linguistic analysis.
+By contrast, we use surface features (character n-gram and word unigram frequencies) extracted from SeedLing, and apply an off-the-shelf hierarchical clustering algorithm.\footnote{\url{http://www.scipy.org/}} Specifically, each language is represented as a vector of frequencies of character bigrams, character trigrams, and word unigrams.
+Each of these three components is normalized to unit length.
 
 \paragraph{Experimental Setup.}
-We perform hierarchical/agglomerative clustering using a variety of linkage methods: (i) \texttt{single}, (ii) \texttt{complete} and (iii) \texttt{weighted}.
-%(WPGMA, weighted pair group method with averaging). 
-The \texttt{single} method calculates the distance between the newly formed clusters by assigning the minimal distance between the clusters and the \texttt{complete} method assigns the maximal distance between the newly formed clusters.
-The \texttt{weighted} method assigns the averaged distance between the newly formed cluster and its intermediate clusters.
-We set the number of clusters to 147, the number of top-level genetic groupings in Ethnologue (presumably the maximum number of language family clusters we would seek to induce).
+We performed hierarchical clustering, which produces a tree structure: each leaf represents a language, and each node a cluster.
+We used linkage methods, which recursively build the tree starting from the leaves.
+Initially, each language is in a separate cluster, then we iteratively find the closest two clusters and merge them.
+Each time we do this, we take the two corresponding subtrees, and introduce a new node to join them.
 
-% The \texttt{maxclust} criterion for flattening clusters from the hierarchy is achieved by finding the minimum threshold, \emph{r}, such that the cophenetic distance between any two original observations in the same flat cluster is no more than \emph{r} and not more than the number of clusters.
+We tested three methods, which differ in how they define the distance between two clusters.
+The \texttt{single} method considers all pairs of languages with one from each cluster, and finds the smallest distance.
+The \texttt{complete} method instead finds the largest distance.
+The \texttt{weighted} method calculates an average across all the pairwise differences, weighted according to the languages' positions in the tree.
+
+While a tree structure might be linguistically useful, evaluating the structure is more awkward.
+However, the above methods can be easily used to produce partitional clusters, by stopping when we reach a certain number of clusters.
+We set the number of clusters to 147, the number of top-level genetic groupings in Ethnologue.
 
@@
     \end{tabular}
-\caption{Comparison of clustering algorithms} %where number of clusters is set to genetic grouping on Ethnologue.} 
+\caption{Comparison of clustering algorithms}
+\label{table:cluster}
 \end{centering}
- \label{table:cluster}
 \end{table}
@@
 However, they prove that they are satisfied by the BCubed metric, which we adopt for this reason.
-To calculate this, we find the induced cluster and gold standard class for each language, and calculate the F-score of the cluster compared to the class.
+To calculate this, we take the induced cluster and gold standard class for each language, and calculate the F-score of the cluster compared to the class.
 These F-scores are then averaged across all languages.
 
-In table 2, we give results of our clustering experiments, comparing three clustering approaches to a random baseline.
+In table \ref{table:cluster}, we give results of our clustering experiments, comparing the three clustering methods to a random baseline (clustering based on random distances, averaged over several runs).
 The F-scores are comparable to those reported by \newcite{georgi2010wals}, even though we have only used surface features, while they used typological features taken from WALS.
@@
 
-% As well as performing clustering, we can view this as an information retrieval task: given a language, what are similar languages?
+%%% [ADD THIS!] As well as performing clustering, we can view this as an information retrieval task: given a language, what are similar languages?
 To do this, we found the languages with the closest vectors of n-grams and words.
@@
 
-% In table \ref{table:cluster}, we give results of our clustering experiments.
-The F-scores are much higher than the random baseline, and comparable to the values reported by \newcite{georgi2010wals}, even though we have only used surface features, while they used typological features taken from WALS.
-This demonstrates that it possible for cross-linguistic research to be conducted even based on extremely shallow features.
+It is worth noting that precision is higher than recall.
+This is perhaps to be expected, given that related languages using wildly differing orthographies will appear different.
+Nonetheless, our system is reasonably capable of identifying languages which are both related and written similarly.
 
-It is also worth noting that precision is higher than recall.
-This is perhaps to be expected, given that related languages using wildly differing orthographies will appear to be very different.
-Nonetheless, our system is reasonably capable of identifying those languages which are both related and also written similarly.
+\paragraph{Qualitative Inspection.}

### 2014-04-22 universalcorpus/langcluster.py
Recoded some of langcluster in a more modular form.
Fixed a couple of typos.

@@ -61,6 +61,8 @@ living_languages = set(ISO2LANG.keys()) - set(RETIRED2ISO.keys()) \
 - set(dead.keys()) - set(constructed.keys()) \
 - set(MACRO2LANG.keys()) - set(macro_split.keys())
 
+### Note: We should not just ignore Norwegian!
+
 def generate_ngrams(data_source):
   twograms = defaultdict(Counter) 
   threegrams = defaultdict(Counter)
@@ -201,36 +203,53 @@ for l1, l2, dist in dmatrix:
 matrix = [v.values() for k,v in matrix.iteritems()]
 matrix = pd.lib.to_object_array(matrix).astype(float) 
 '''
+
+condensed_file = 'distance_condensed.pk'
+if not os.path.exists(condensed_file):
+  labels = []
+  index = {}
+  for l1, l2, dist in dmatrix:
+    if l1 not in labels:
+      index[l1] = len(labels)
+      labels.append(l1)
+      
   
-labels = []
-index = {}
-for l1, l2, dist in dmatrix:
-  if l1 not in labels:
-    index[l1] = len(labels)
-    labels.append(l1)
-    
-
-matrix = np.zeros((len(labels),len(labels)))
-
-for l1, l2, dist in dmatrix:
-  index1 = index[l1]
-  index2 = index[l2]
-  if l1 == l2:
-    continue
-  matrix[index1,index2] = dist
-  matrix[index1,index2] = dist
-
-condensed = scipy.spatial.distance.squareform(matrix)
+  matrix = np.zeros((len(labels),len(labels)))
+  
+  for l1, l2, dist in dmatrix:
+    index1 = index[l1]
+    index2 = index[l2]
+    if l1 == l2:
+      continue
+    matrix[index1,index2] = dist
+    matrix[index1,index2] = dist
+  
+  condensed = scipy.spatial.distance.squareform(matrix)
+  with open(condensed_file,'wb') as f:
+    pickle.dump((condensed, labels), f)
+  
+else:
+  with open(condensed_file,'rb') as f:
+    condensed, labels = pickle.load(f)
 
 methods = 'single complete average weighted'.split()
 criterion = 'inconsistent distance maxclust'.split()
 
-if os.path.exist('cluster.eval.outputs'):
+output_file = 'cluster2.eval.outputs'
+if os.path.exists(output_file):
+  print('Output file already exists! Exiting.')
   exit()
-  
[…]

### 2014-04-22 universalcorpus/langcluster_modular.py

@@ -1,0 +1,154 @@
+from __future__ import print_function, division
+from scipy.cluster.hierarchy import linkage, fcluster
+import cPickle as pickle
+from collections import defaultdict
+from miniethnologue import ISO2LANG, LANG2ISO
+from crawlandclean.ethnologue import ISO2FAMILY, FAMILIES2ISO
+
+def bcubed(induced, gold, verbose=False):
+	"""
+	Calculates the B-Cubed score for an induced clustering compared to a gold standard
+	Both inputs should dictionaries, mapping from items to sets of items
+	"""
+	
+	if set(induced.keys()) ^ set(gold.keys()):
+		print(induced.keys())
+		print(gold.keys())
+		raise KeyError("The induced clusters and gold classes must be defined over the same set")
+	
+	N_items = len(induced.keys())
+	sum_fscore = 0
+	sum_precision = 0
+	sum_recall = 0
+	
+	for key in induced.keys():
+		cluster = induced[key]
+		klass = gold[key]
+		intersect = cluster & klass
+		
+		precision = len(intersect) / len(cluster)
+		recall = len(intersect) / len(klass)
+		fscore = 2 * len(intersect) / (len(cluster) + len(klass))
+		
+		sum_precision += precision
+		sum_recall += recall
+		sum_fscore += fscore
+	
+	av_precision = sum_precision/N_items
+	av_recall = sum_recall/N_items
+	av_fscore = sum_fscore/N_items
+	
+	if verbose:
+		print('Precision: {}'.format(av_precision))
+		print('Recall:    {}'.format(av_recall))
+		print('F-score:   {}'.format(av_fscore))
+	
+	return (av_precision, av_recall, av_fscore)
+
+
+def flatten(hierarchy, labels, N_clust=147):
+	"""
+	Takes a hierarchical clustering and flattens it.
+	Returns a dictionary from items to the set of elements in the cluster
+	Note that the scipy function fcluster does not let you fix the number of clusters
+	"""
+	N_items = len(labels)
+	members = dict() # Map from the index for a cluster to its members
+	for i in range(N_items):
+		members[i] = frozenset({labels[i]})
+	for n in range(N_items - N_clust):
+		j = hierarchy[n,0]
+		k = hierarchy[n,1]
+		members[N_items + n] = members.pop(j) | members.pop(k)
[…]

### 2014-04-22 universalcorpus/simlang.py

@@ -11,15 +11,16 @@ dead = {"osp":"Old Spanish", "odt":"Old Dutch", "goh": "Old High German",
         "tpn":"Tupinamba","ojp":"Old Japanese","sga":"Old Irish",
         "hit":"Hittite","tkm":"Takelma","dum":"Middle Dutch",
         "fro":"Old French","nci":"Classical Nahuatl","gmh":"Middle High German",
-        "mxi":"Mozarabic"}
+        "mxi":"Mozarabic","grc":"Ancient Greek"}
 macro_split = {"nob":"Norwegian, Bokmaal","nno":"Norwegian Nynorsk"} # nor.
 constructed = {"ido":"Ido","tlh":"Klingon","tzl":"Talossan","jbo":"Lojban",
-               "ina":"Interlingua"}
+               "ina":"Interlingua","epo":"Esperanto"}
 
 living_languages = set(ISO2LANG.keys()) - set(RETIRED2ISO.keys()) \
 - set(dead.keys()) - set(constructed.keys()) \
 - set(MACRO2LANG.keys()) - set(macro_split.keys())
 
+### Note: We should not just ignore Norwegian!
 
 x = defaultdict(list)
 for line in open('distance.out'):
@@ -28,7 +29,7 @@ for line in open('distance.out'):
   
 precisions, recalls, fscores = [], [], []
 for lang in x:
-  gold_class = ethnologue.FAMILIES2ISO[ethnologue.ISO2FAMILY[i]]
+  gold_class = ethnologue.FAMILIES2ISO[ethnologue.ISO2FAMILY[lang]]
   gold_class = [g for g in gold_class if g in living_languages]
   induced_cluster = set([i for _,i in sorted(x['eng'][:21])])
   

### 2014-04-23 universalcorpus/crawlandclean/ethnologue.py
Added pairwise clustering evaluation. Started tidying langcluster.py

@@ -66,7 +66,8 @@ def read_language(filehandle):
   """
   Extracts languages names, the language family, and dialect information 
   from a page on Ethnologue.
-  """ 
+  """
+  from bs4 import BeautifulSoup as bs
   soup = bs(filehandle.read())
   primary_name = soup.find("meta", property="og:title")["content"]
   alternate_names = soup.find("div", class_="field-name-field-alternate-names"       ).find("div", class_=["field-item", "even"]).string.split(", ")
@@ -174,6 +175,7 @@ def lang_families():
 ##lang_families()
 
 def download_lang_families():
+  import time
   _lf = {k:[j[0] for j in v] for k,v in load_language_families().items()}
   for i in _lf:
     for j in _lf[i]:

### 2014-04-23 universalcorpus/langcluster.py

@@ -1,7 +1,7 @@
 # -*- coding: utf-8 -*-
 
 import cPickle as pickle
-import time, os, ast, random,codecs
+import time, os, ast, random, codecs
 
 from math import sqrt, acos
 from collections import Counter, defaultdict
@@ -20,6 +20,7 @@ import matplotlib.pylab as plt
 
 from crawlandclean import odin, omniglot, udhr, ethnologue
 from miniethnologue import ISO2LANG, MACRO2LANG, RETIRED2ISO
+from evaluate import normalise
 
 random.seed(0)
 
@@ -30,21 +31,6 @@ def word2ngrams(text, n=3):
 def sent2ngrams(text, n=3):
   """ Convert sentence into character ngrams. """
   return list(chain(*[word2ngrams(i,n) for i in text.lower().split()]))
-    
-def normalize(featurevector, length=1):
-  """
-  Normalises a feature vector to a specific length.
-  Zero vectors are left zero.
-  NOTE: featurevector will be updated with normalized count.
-  """
-  from math import sqrt
-  try:
-    norm = length / sqrt(sum([x*x for x in featurevector.values()]))
-  except ZeroDivisionError:
-    return featurevector
-  for feat in featurevector:
-    featurevector[feat] *= norm
-  return featurevector
 
 
 dead = {"osp":"Old Spanish", "odt":"Old Dutch", "goh": "Old High German",
@@ -108,9 +94,9 @@ def load_ngram_array():
     threegrams[k].update(v)
   
   for lang in twograms:
-    twograms[lang] = normalize(twograms[lang], 1/sqrt(3))
-    threegrams[lang] = normalize(threegrams[lang], 1/sqrt(3))
-    unigrams[lang] = normalize(unigrams[lang], 1/sqrt(3))
+    normalise(twograms[lang], 1/sqrt(3))
+    normalise(threegrams[lang], 1/sqrt(3))
+    normalise(unigrams[lang], 1/sqrt(3))
     
   allgrams = defaultdict(Counter)
   for k,v in chain(twograms.iteritems(), threegrams.iteritems(), unigrams.iteritems()):
@@ -125,29 +111,11 @@ def distance(c1, c2):
     return 0
   return acos(dimsum)
 
-def plot_tree(P, pos=None):
-  import matplotlib.pylab as plt
[…]

### 2014-04-23 universalcorpus/langcluster_modular.py

@@ -1,19 +1,20 @@
 from __future__ import print_function, division
-from scipy.cluster.hierarchy import linkage, fcluster
+from scipy.cluster.hierarchy import linkage
 import cPickle as pickle
 from collections import defaultdict
-from miniethnologue import ISO2LANG, LANG2ISO
+from miniethnologue import ISO2LANG
 from crawlandclean.ethnologue import ISO2FAMILY, FAMILIES2ISO
 
 def bcubed(induced, gold, verbose=False):
 	"""
 	Calculates the B-Cubed score for an induced clustering compared to a gold standard
-	Both inputs should dictionaries, mapping from items to sets of items
+	Both inputs should be dictionaries, mapping from items to sets of items
 	"""
 	
 	if set(induced.keys()) ^ set(gold.keys()):
-		print(induced.keys())
-		print(gold.keys())
+		if verbose:
+			print(induced.keys())
+			print(gold.keys())
 		raise KeyError("The induced clusters and gold classes must be defined over the same set")
 	
 	N_items = len(induced.keys())
@@ -45,6 +46,36 @@ def bcubed(induced, gold, verbose=False):
 	
 	return (av_precision, av_recall, av_fscore)
 
+def pairwise(induced, gold, verbose=False):
+	"""
+	Calculates pairwise precision, recall, and f-scores for an induced clustering compared to a gold standard
+	Both inputs should be dictionaries, mapping from items to sets of items
+	"""
+	items = set(induced.keys())
+	if items ^ set(gold.keys()):
+		if verbose:
+			print(induced.keys())
+			print(gold.keys())
+		raise KeyError("The induced clusters and gold classes must be defined over the same set")
+	
+	induced_pairs = {(x,y) for cluster in set(induced.values()) for x in cluster for y in cluster if x<y}
+	gold_pairs    = {(x,y) for klass   in set(gold.values())    for x in klass   for y in klass if x<y}
+	
+	both = len(induced_pairs & gold_pairs)
+	false_pos = len(induced_pairs) - both
+	false_neg = len(gold_pairs)    - both
+	
+	precision = both / (both + false_pos)
+	recall = both / (both + false_neg)
+	fscore = 2 * both / (2 * both + false_pos + false_neg)
+	
+	if verbose:
+		print('Precision: {}'.format(precision))
+		print('Recall:    {}'.format(recall))
[…]

### 2014-04-24 docs/computEL-final/clean.tex
Minor changes after readthrough (typos, etc.)

@@
 Natural Language ToolKit (NLTK) corpora
-distribution,\footnote{\url{http://nltk.googlecode.com/svn/trunk/nltk\_data/index.xml}}
-the distribution is laden with encoding problems 
+distribution,\footnote{\url{http://nltk.googlecode.com/svn/trunk/nltk_data/index.xml}}
+the distribution is laden with encoding problems. 
 %during their
@@
 One major issue with using the Wikipedia dump is the problem of separating text from abundant source-specific markup. To convert compressed
-Wikipedia dumps to textfiles, we used the WikiExtractor
-\footnote{\url{http://medialab.di.unipi.it/wiki/Wikipedia\_Extractor}}
+Wikipedia dumps to textfiles, we used the WikiExtractor\footnote{\url{http://medialab.di.unipi.it/wiki/Wikipedia_Extractor}}
 tool. After conversion into textfiles, we used several regular expressions to
 delete residual Wikipedia markup and so-called ``magic
-words"\footnote{\url{http://en.wikipedia.org/wiki/Help:Magic\_words}}.
+words"\footnote{\url{http://en.wikipedia.org/wiki/Help:Magic_words}}.
 
 \paragraph{Omniglot.}
-The main issue with extracting this data is that the pages are designed to be human-readable, not machine-readable.
-Cleaning this data required parsing the html source, and extracting the relevant content, which required different code for the two types of page we considered (useful phrases, and the Tower of Babel story).
+The main issue with extracting the Omniglot data is that the pages are designed to be human-readable, not machine-readable.
+Cleaning this data required parsing the HTML source, and extracting the relevant content, which required different code for the two types of page we considered ('\emph{Useful foreign phrases}' and '\emph{Tower of Babel}').
 Even after automatic extraction, some noise in the data remained, such as explanatory notes given in parentheses, which are written in English and not the target language.
-Even though the total amount of data here is small compared to our ther sources, the amount of effort required to process it was not, because of these idiosyncracies.
+Even though the total amount of data here is small compared to our other sources, the amount of effort required to process it was not, because of these idiosyncracies.
 We expect that researchers seeking to convert data from human-readable to machine-readable formats will encounter similar problems, but unfortunately there is unlikely to be a one-size-fits-all solution to this problem.
@@
 \paragraph{ODIN.}
-The ODIN data is easily accessible in XML format from the online database,\footnote{\url{http://odin.linguistlist.org/download}} where data for each language is saved in a separate XML file and the IGTs are encoded in tags of the form \texttt{<igt><example>...</example></igt>}.
+The ODIN data is easily accessible in XML format from the online database\footnote{\url{http://odin.linguistlist.org/download}}.
+Data for each language is saved in a separate XML file and the IGTs are encoded in tags of the form \texttt{<igt><example>...</example></igt>}.
 For example, the IGT in figure~\ref{fig:odin_fijian} is represented by the XML snippet in figure~\ref{fig:odin_fijian_xml}.
@@
 \indent \qquad\qquad `\emph{You return here.}' \\
-\caption{Fijian IGT from ODIN.} \label{fig:odin_fijian}
+\caption{Fijian IGT from ODIN} \label{fig:odin_fijian}
 \end{figure}
@@
 \end{lstlisting} 
-\caption{Fijian IGT in ODIN's XML format.} \label{fig:odin_fijian_xml}
+\caption{Fijian IGT in ODIN's XML format} \label{fig:odin_fijian_xml}
 \end{figure}
@@
 \smallskip
-\caption{Yimas IGT in ODIN's XML format.}\label{fig:odin_yimas_xml}
+\caption{Yimas IGT in ODIN's XML format}\label{fig:odin_yimas_xml}
 \end{figure}
@@
 The older ISO~639-1 codes are easy to recognise, being two letters long instead of three, and can be straightforwardly converted.
-However, a small number of Wikipedia codes are not ISO codes at all - we converted these to ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{\url{http://meta.wikimedia.org/wiki/Special\_language\_codes}}
+However, a small number of Wikipedia codes are not ISO codes at all - we converted these to ISO~639-3, following documentation from the Wikimedia Foundation.\footnote{\url{http://meta.wikimedia.org/wiki/Special_language_codes}}
 
@@
 It is important to make sure that the data we have compiled will be available to future researchers, regardless of how the surrounding infrastructure changes.
-\newcite{bird2003port} describes a set of best practices for maintaining portability of digital information, outlining seven dimensions along which this can vary.
+\newcite{bird2003port} describe a set of best practices for maintaining portability of digital information, outlining seven dimensions along which this can vary.
 Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.

### 2014-04-24 docs/computEL-final/data.tex
Minor changes after readthrough (typos, etc.)

--- 2014-04-17 docs/computEL/data.tex
@@
 It consists of 30 articles, amounting to about four pages of text.
-This is a useful document for CL, since it has been translated into a wide variety of languages, providing a highly parallel text.
+This is a useful document for NLP, since it has been translated into a wide variety of languages, providing a highly parallel text.
 
@@
 Wikipedia\footnote{\url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability.
-At the time of writing, it contained 30.8 million articles in 287 languages, which provides a sizeable amount of monolingual text in a fairly wide range of languages.
-Text dumps are made regularly, and can be downloaded from \url{www.dumps.wikimedia.org}.
+At the time of writing, it contained 30.8 million articles in 286 languages, which provides a sizeable amount of monolingual text in a fairly wide range of languages.
+Text dumps are made regularly available, and can be downloaded from \url{www.dumps.wikimedia.org}.
 
@@
 However, this loses information, since we no longer know which morpheme has which meaning.
-To keep all information encoded in the IGT, we need to modify the \newcite{abney2011data}'s proposal.
+To keep all information encoded in the IGT, we need to modify \newcite{abney2011data}'s proposal.
 
@@
 There is a unique identifier for each morpheme, and glosses are stored under this identifier and the corresponding gloss code.
-Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as a convenient surrogates of meaning.
+Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as convenient surrogates of meaning.
 By the same reasoning, we can use glosses to provide a more detailed surrogate of meaning, only written in a metalanguage instead of a natural one.

### 2014-04-24 docs/computEL-final/relwork.tex
Minor changes after readthrough (typos, etc.)

--- 2014-03-22 docs/computEL/relwork.tex
@@
 
-Such resources are invaluable for many purposes
-for instance, there can be plenty of material to work with for a
+Such resources are invaluable for many purposes.
+For instance, there can be plenty of material to work with for a
 researcher who is looking to find data on a specific language, who is
@@
 The corpora are thus of a respectable size, but only 1\% of the world's languages are represented.
-A further difficulty is that languages are named, without the corresponding ISO codes.
+A further difficulty is that languages are named, without the corresponding ISO language codes.
 This problem may be small when all 68 varieties are well-known, but will become a serious concern if this corpus collection is to grow and include smaller languages.

### 2014-04-25 docs/computEL-final/clean.tex
Tidied tex files
(I have removed some commented sections that are unlikely to be included
again.  They are of course still available in the old commits.)

@@
 
-\paragraph{UDHR.} Although there is a pre-compiled version of the UDHR data from the
-Natural Language ToolKit (NLTK) corpora
-distribution,\footnote{\url{http://nltk.googlecode.com/svn/trunk/nltk_data/index.xml}}
+\paragraph{UDHR.} Although there is a pre-compiled version of the UDHR data in the Natural Language ToolKit (NLTK) corpora distribution,\footnote{\url{http://nltk.googlecode.com/svn/trunk/nltk_data/index.xml}}
 the distribution is laden with encoding problems.
-%during their
-%conversion from pdf to plaintext format. 
-Instead, we use the
-plaintext files available from the Unicode
-website,\footnote{\url{http://unicode.org/udhr/d}} which are free of
-encoding issues. The first four lines of each file record metadata,
-and the rest is the translation of the UDHR. This dataset is
-extremely clean, and simply required segmentation into sentences.
+Instead, we used the plaintext files available from the Unicode website,\footnote{\url{http://unicode.org/udhr/d}} which are free from encoding issues.
+The first four lines of each file record metadata, and the rest is the translation of the UDHR.
+This dataset is extremely clean, and simply required segmentation into sentences.
 
 \paragraph{Wikipedia.}
-One major issue with using the Wikipedia dump is the problem of separating text from abundant source-specific markup. To convert compressed
-Wikipedia dumps to textfiles, we used the WikiExtractor\footnote{\url{http://medialab.di.unipi.it/wiki/Wikipedia_Extractor}}
-tool. After conversion into textfiles, we used several regular expressions to
-delete residual Wikipedia markup and so-called ``magic
-words"\footnote{\url{http://en.wikipedia.org/wiki/Help:Magic_words}}.
+One major issue with using the Wikipedia dump is the problem of separating text from abundant source-specific markup.
+To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor\footnote{\url{http://medialab.di.unipi.it/wiki/Wikipedia_Extractor}} tool.
+After conversion into textfiles, we used several regular expressions to delete residual Wikipedia markup and so-called ``magic words"\footnote{\url{http://en.wikipedia.org/wiki/Help:Magic_words}}.
 
@@
 
-% \begin{minipage}{\columnwidth}
-% \quad 21 a.\quad o lesu mai \\
-% \indent \qquad\qquad 2sg return here \\
-% \indent \qquad\qquad `\emph{You return here.}' \\
-% %\caption{Fijian IGT from ODIN.} 
-% \end{minipage}
-
-The primary problem in extracting the data is a lack of consistency in
-the IGTs. In the above examples, the sentence is introduced by a
-letter or number, which needs to be removed; however, the form of such
-indexing elements varies. In addition, the source line in figure~\ref{fig:odin_yimas_xml}
-includes two types of metadata: the language name, and a citation,
-both of which introduce noise.  Finally, extraneous punctuation such
-as the quotation marks in the translation line need to be removed. We used regular expressions for cleaning lines within the IGTs.
+The primary problem in extracting the data is a lack of consistency in the IGTs.
+In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
+In addition, the source line in figure~\ref{fig:odin_yimas_xml} includes two types of metadata: the language name, and a citation, both of which introduce noise.
+Finally, extraneous punctuation such as the quotation marks in the translation line need to be removed.
+We used regular expressions for cleaning lines within the IGTs.
 
@@
 
-% \begin{minipage}{\columnwidth}
-% \begin{lstlisting}[language=XML]
-% <igt>
-%   <example>
-%     <line>21 a. o lesu mai</line>
-%     <line>2sg return here</line>
-%     <line>`You return here.'</line>
-%   </example>
-% </igt>
-% \end{lstlisting} 
-% \centerline{Figure 2: Fijian IGT in ODIN's XML format.}
-% \end{minipage}
-
 \begin{figure}[t]
@@
 \end{figure}
-
-% \begin{minipage}{\columnwidth}
-% \begin{lstlisting}[language=XML]
-% <igt>
-%   <example>
-%     <line>(69) na-Na-tmi-kwalca-t 
-%     Yimas (Foley 1991)</line>
-%     <line>3sgA-1sgO-say-rise-PERF
-%     </line>
-%     <line>`She woke me up' 
-%     (by verbal action)</line>
-%   </example>
-% </igit>
-% \end{lstlisting} 
-% \smallskip
-% \centerline{Figure 3: Yimas IGT in ODIN's XML format.}
-% \end{minipage}
 
@@
 % \end{figure}
-
-% \begin{minipage}{\columnwidth}
-% \begin{itemize}
-% \item \emph{Cleaner}: Removed (i) all heading and trailing text embedded in square or rounded brackets and (ii) heading double character token ending with bracket or fullstop.
-% \begin{itemize}
-% \item[(i)]
-% \begin{Verbatim}
-% ^(?\s?\w{1,5}\s*[):.]\s*
-% \end{Verbatim}
-% \item[(ii)] 
-% \begin{Verbatim}
-% [\[\(].{1,}[\]\)]
-% \end{Verbatim}
-% \end{itemize}
-% \item \emph{Cleanest}: Only source lines without punctuation.
-% \end{itemize}
-% \end{minipage}
 

### 2014-04-25 docs/computEL-final/data.tex
Tidied tex files
(I have removed some commented sections that are unlikely to be included
again.  They are of course still available in the old commits.)

@@
 We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): Online Database of Interlinear Text (ODIN), Omniglot website, Universal Declaration of Human Rights (UHDR), and Wikipedia.
-
-% \begin{enumerate}
-% \item The Online Database of Interlinear Text (ODIN)
-% \item The Omniglot website
-% \item The Universal Declaration of Human Rights (UDHR)
-% \item Wikipedia
-% \end{enumerate}
 
@@
 % The urls for these resources are:
-
-% \begin{itemize}[noitemsep]
-% \item \url{www.omniglot.com/language/phrases/*} 
-% \item \url{www.omniglot.com/babel/*}
-% \end{itemize}
 
@@
 
-Even though glosses in ODIN are based on English, having been
-extracted from English-language documents, this is not true of IGTs in
-general.  In particular, it is common for documentary linguists
-working on indigenous languages of the Americas to provide glosses and
-translations based on Spanish.  For this reason, we believe it would
-be wise to specify the language used to produce the gloss.  Since it
-is not quite the language itself, but a metalanguage, one solution
-would be to use new language codes that make it clear both that a
-metalanguage is being used, and also what natural language it is based
-on.  The five-letter code \texttt{gloss} cannot be confused with any
-code in any version of ISO 639 (with codes of length two to
-four).  Following the convention that subvarieties of a language are
-indicated with suffixes, we can append the code of the natural
-language.  For example, glosses into English and
-Spanish-based metalanguages would be given the codes
-\texttt{gloss-eng} and \texttt{gloss-spa}, respectively.
+Even though glosses in ODIN are based on English, having been extracted from English-language documents, this is not true of IGTs in general.
+In particular, it is common for documentary linguists working on indigenous languages of the Americas to provide glosses and translations based on Spanish.
+For this reason, we believe it would be wise to specify the language used to produce the gloss.
+Since it is not quite the language itself, but a metalanguage, one solution
+would be to use new language codes that make it clear both that a metalanguage is being used, and also what natural language it is based on.
+The five-letter code \texttt{gloss} cannot be confused with any code in any version of ISO 639 (with codes of length two to four).
+Following the convention that subvarieties of a language are indicated with suffixes, we can append the code of the natural language.
+For example, glosses into English and Spanish-based metalanguages would be given the codes \texttt{gloss-eng} and \texttt{gloss-spa}, respectively.
 
@@
 There is a unique identifier for each morpheme, and glosses are stored under this identifier and the corresponding gloss code.
-Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as convenient surrogates of meaning.
+Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as a convenient surrogate of meaning.
 By the same reasoning, we can use glosses to provide a more detailed surrogate of meaning, only written in a metalanguage instead of a natural one.
@@
 
-According to Ethnologue, there are 7105 living languages, and 147
-living language families. Across all our data sources, we manage to
-cover 1347 languages in 106 families, which represents 19.0\% of the
-world's languages. To get a better idea of the kinds of languages
-represented, we give a breakdown according to their EGIDS scores
-(Expanded Graded Intergenerational Disruption Scale)
-\cite{lewis2010assessing} in Figure \ref{fig:heatmap}. The values in
-each cell have been colored according to proportion of languages
-represented, with green indicating good coverage and red poor. It's
-interesting to note that vigorous languages (6a) are poorly
-represented across all data sources, and worse than more endangered
-categories. In terms of language documentation, vigorous languages are
-less urgent goals than those in categories 6b and up, but this
-highlights an unexpected gap in linguistic resources.
-
-%\begin{table}
-%\centering
-%    \begin{tabular}{l|cc}
-%    ~         				& \#Languages & \#Families \\ \hline
-%    ODIN      				& 1218       & 101       \\
-%    Omniglot  				& 134        & 21        \\
-%    UDHR      				& 355        & 47        \\
-%    Wikipedia 				& 209        & 22       \\ \hline
-%    \textbf{Combined}	& 1347			 & 106 
-%    \end{tabular}
-%\caption{Corpus Coverage}
-%\end{table}
+According to Ethnologue, there are 7105 living languages, and 147 living language families.
+Across all our data sources, we manage to cover 1347 languages in 106 families, which represents 19.0\% of the world's languages.
+To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) \cite{lewis2010assessing} in Figure \ref{fig:heatmap}.
+The values in each cell have been colored according to proportion of languages represented, with green indicating good coverage and red poor.
+It's interesting to note that vigorous languages (6a) are poorly represented across all data sources, and worse than more endangered categories.
+In terms of language documentation, vigorous languages are less urgent goals than those in categories 6b and up, but this highlights an unexpected gap in linguistic resources.
 

### 2014-04-25 docs/computEL-final/cluster.tex
Updated results, added discussion

@@
 
-We tested three methods, which differ in how they define the distance between two clusters.
-The \texttt{single} method considers all pairs of languages with one from each cluster, and finds the smallest distance.
-The \texttt{complete} method instead finds the largest distance.
-The \texttt{weighted} method calculates an average across all the pairwise differences, weighted according to the languages' positions in the tree.
+We tested four methods, which differ in how they define the distance between two clusters.
+The \texttt{complete} method considers all pairs of languages with one from each cluster, and finds the largest distance.
+The \texttt{single} method instead finds the smallest distance.
+The \texttt{average} and \texttt{weighted} methods calculate averages across all the pairwise differences, either unweighted, or weighted according to the languages' positions in the tree.
 
@@
     ~        & Precision & Recall       & F-score    \\ \hline
-    \texttt{single} & 0.1958	& 0.6755	 & 0.1240  \\
-	\texttt{complete} & \textbf{0.3153}	& 0.1699	 & \textbf{0.1420} \\
-	\texttt{weighted} & 0.0614	& \textbf{0.8565}	 & 0.1099 \\ \hline
-	\emph{random} & 0.1925 &	0.0927 & 	0.0692 \\
+	\texttt{complete} & \textbf{0.3142}	& 0.1938	 & \textbf{0.1556} \\
+	\texttt{single} & 0.1949	& \textbf{0.7850}	 & 0.1282 \\
+	\texttt{average} & 0.2216	& 0.5163	& 0.1348 \\
+	\texttt{weighted} & 0.2326	& 0.3343	& 0.1271 \\ \hline
+	\emph{random} & 0.1836	& 0.0917	& 0.0675 \\
     \end{tabular}
@@
 
-In table \ref{table:cluster}, we give results of our clustering experiments, comparing the three clustering methods to a random baseline (clustering based on random distances, averaged over several runs).
-The F-scores are comparable to those reported by \newcite{georgi2010wals}, even though we have only used surface features, while they used typological features taken from WALS.
+In table \ref{table:cluster}, we give results of our clustering experiments, comparing the three clustering methods to a random baseline (clustering based on random distances, averaged over 20 runs).
+It is worth noting that precision is higher than recall.
+This is perhaps to be expected, given that related languages using wildly differing orthographies will appear different.
+Nonetheless, our system is reasonably capable of identifying languages which are both related and written similarly.
+
+To allow a closer comparison with \newcite{georgi2010wals}, we can also calculate pairwise scores - i.e.
+considering when pairs of languages are in the same cluster or the same class.
+For 147 clusters, the \texttt{complete} method gives a pairwise f-score of 0.147, while they report 0.140.
+The figures are not quite comparable since we are evaluating over a different set of languages; nonetheless, this result is suggestive, given that we have only used surface features, while they used typological features taken from WALS.
 This demonstrates that it possible for cross-linguistic research to be conducted even based on extremely shallow features.
@@
 
-It is worth noting that precision is higher than recall.
-This is perhaps to be expected, given that related languages using wildly differing orthographies will appear different.
-Nonetheless, our system is reasonably capable of identifying languages which are both related and written similarly.
+\paragraph{Qualitative Inspection.}
+Ga: It's actually pretty difficult to interpret the clusters directly because there's so much noise. I will try to write something better, but here are a few thoughts I've had:
 
-\paragraph{Qualitative Inspection.}
+Using the \texttt{complete} method with 147 clusters, we obtain a mix of large and small clusters.
+There are 27 singleton clusters, while the largest cluster contains 123 languages.
+Although these singleton clusters belong to classes of various size, it is interesting that the distribution of sizes is similar to the distribution in the gold standard data, where we have 27 single classes, and the largest class contains 150 languages.
+This isn't true for the other clustering methods - the \texttt{single} method produces 138 singleton clusters, with almost all languages in a single cluster of size 968.

### 2014-04-25 docs/computEL-final/intro.tex
Updated results, added discussion

--- 2014-04-17 docs/computEL/intro.tex
@@
 values of typological features as in WALS \cite{wals}, Swadesh word lists).
-Despite our approach being highly dependent on orthography, we achieve a reasonable clustering performance over the languages in our dataset, demonstrating SeedLing's utility in cross-linguistic research.
+Despite our approach being highly dependent on orthography, our clustering performance matches the results obtained by \newcite{georgi2010wals} using typolological features, which demonstrates SeedLing's utility in cross-linguistic research.
 

### 2014-04-25 universalcorpus/langcluster_modular.py

@@ -51,8 +51,8 @@ def pairwise(induced, gold, verbose=False):
 	Calculates pairwise precision, recall, and f-scores for an induced clustering compared to a gold standard
 	Both inputs should be dictionaries, mapping from items to sets of items
 	"""
-	items = set(induced.keys())
-	if items ^ set(gold.keys()):
+	
+	if set(induced.keys()) ^ set(gold.keys()):
 		if verbose:
 			print(induced.keys())
 			print(gold.keys())
@@ -93,8 +93,21 @@ def flatten(hierarchy, labels, N_clust=147):
 		members[N_items + n] = members.pop(j) | members.pop(k)
 	return {item:cluster for cluster in members.values() for item in cluster}
 
+def flatten_scipy(hierarchy, labels, N_clust=147):
+	"""
+	Takes a hierarchical clustering and flattens it, using scipy.
+	Returns a dictionary from items to the set of elements in the cluster
+	Note that this does not let you precisely fix the number of clusters
+	"""
+	from scipy.cluster.hierarchy import fcluster
+	members = defaultdict(set) # Map from the index for a cluster to its members
+	flat = fcluster(hierarchy, t=N_clust, criterion='maxclust')
+	for i in range(len(flat)):
+		members[flat[i]].add(labels[i])
+	frozen_members = {key:frozenset(value) for key, value in members.items()}
+	return {item:cluster for cluster in frozen_members.values() for item in cluster}
 
-def evaluate(filename='distance_condensed.pk', option='once', method='complete', N_clust=147, metric=bcubed, verbose=True):
+def evaluate(filename='distance_condensed.pk', option='once', method='complete', N_clust=147, metric=bcubed, flat_func=flatten, verbose=True):
 	if verbose: print("Loading data...")
 	with open(filename,'rb') as f:
 		condensed, labels = pickle.load(f)
@@ -130,7 +143,7 @@ def evaluate(filename='distance_condensed.pk', option='once', method='complete',
 			for _ in range(length):
 				condensed.append(random())
 			hierarchy = linkage(condensed, method=method)
-			flat = flatten(hierarchy, labels, N_clust)
+			flat = flat_func(hierarchy, labels, N_clust)
 			p, r, f = metric(flat, gold)
 			if verbose: print(p, r, f)
 			sum_precision += p
[…]

### 2014-04-25 docs/computEL-final/clean.tex
added API information

@@
 Omniglot does not give codes at all, but only the language name.
-To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{\url{http://www-01.sil.org/iso639-3/iso-639-3.tab}}
+To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{\url{http://www-01.sil.org/iso639-3/iso-639-3.tab}}.
+
+To facilitate the standardization of language code, we have written a small python API that can be used in query information about a language and its code(s).
+Also, a function to convert language names to their appropriate ISO 639-3 codes.
+If internet connection is available, the API automatically updates the ISO 639-3 file stored locally on the users harddrive before responding to the user's query.
+An example usage of the API is as such:
+
+\small\begin{verbatim}
+>>> from seedling import miniethnologue
+>>> sil = miniethnologue.miniSIL()
+>>> sil.ISO6393['eng']
+{'iso6391': 'en', 
+'iso6392t': 'eng',
+'iso6392b': 'eng',
+'name': 'English', 
+'ismacro': False, 
+'scope': 'Indvidual', 
+'type': 'Living'}
+>>> sil.name2code('German')
+deu
+\end{verbatim}
+\normalsize
+
+Similarly, we provide an API to access language information from the World Atlas of Language Structures (WALS).
+The language clustering experiment in section 5 uses this API to access the language family information.
+
+\small\begin{verbatim}
+>>> from seedling import miniwals
+>>> wals = miniwals.MiniWALS()
+>>> wals['eng']
+{'glottocode': 'stan1293', 
+'name': 'English', 
+'family': 'Indo-European', 
+'longitude': '0.0', 
+'latitude': '52.0', 
+'genus': 'Germanic', 
+'macroarea': u'Eurasia'}
+\end{verbatim}
+\normalsize
 
@@
 If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according to these guidelines.
+SeedLing's python API can be invoked as such:
+
+\small\begin{verbatim}
+>>> from seedling import udhr
+>>> for lang, sent in udhr.sents():
+>>>   print lang, sent
+...
+eng Universal Declaration of 
+Human Rights
+eng Preamble
+eng Whereas recognition of the 
+inherent dignity and of the equal 
+...
+\end{verbatim}
+\normalsize
+
+Without programmatically accessing the corpus using the API, users are able to download a copy tab-delimited plaintext files and process it easily in the programming language or software of their choice.

### 2014-04-25 docs/computEL-final/concl.tex
Tidied tex files
(I have removed some commented sections that are unlikely to be included
again.  They are of course still available in the old commits.)
added link to new repo

@@
 SeedLing will be made available initially through the websites of the authors, as we have yet to properly address the question of long-term access; we welcome ideas or collaborations along these lines.
-In addition, we are releasing a Python API which allows programmatic access to the data and is easily compatible with the analysis tools available from the Natural Language ToolKit \cite{BirdKleinLoper09}.
+In addition, we are releasing a Python API which allows programmatic access to the data and is easily compatible with the analysis tools available from the Natural Language ToolKit \cite{BirdKleinLoper09} \footnote{\url{https://github.com/alvations/SeedLing}}.
 The API also facilitates access to information from both Ethnologue and WALS: a user can extract information about languages, language classification, endangerment status, typological features, and so on.
@@
 
-% To demonstrate the utility of this resource, we used the data to perform language clustering, which gave promising results.
-
 % The API also includes a feature to access information on a language from Ethnologue and/or WALS.

### 2014-04-27 universalcorpus/convert.py
Added new lists of dead, constructed, and macro languages

@@ -1,0 +1,686 @@
+# -*- coding: utf-8 -*-
+
+constructed = {"ido":"Ido",
+               "tlh":"Klingon",
+               "tzl":"Talossan",
+               "jbo":"Lojban",
+               "ina":"Interlingua",
+               "epo":"Esperanto",
+               # Additional constructed languages from http://linguistlist.org/forms/langs/GetListOfConstructedLgs.cfm
+               "bzt": "Brithenig",
+               "dws": "Dutton",
+               "igs": "Interglossa",
+               "ldn": "Láadan",
+               "nov": "Novial",
+               "occ": "Occidental",
+               "qya": "Quenya",
+               "rmv": "Romanova",
+               "sjn": "Sindarin",
+               "vol": "Volapük"}
+
+dead = {"aaq": "Abnaki, Eastern",
+        "abj": "Aka-Bea",
+        "aci": "Aka-Cari",
+        "ack": "Aka-Kora",
+        "acl": "Akar-Bale",
+        "acs": "Acroá",
+        "aes": "Alsea",
+        "aga": "Aguano",
+        "aho": "Ahom",
+        "ait": "Arikem",
+        "ajw": "Ajawa",
+        "akj": "Aka-Jeru",
+        "akk": "Akkadian",
+        "akm": "Aka-Bo",
+        "akx": "Aka-Kede",
+        "aky": "Aka-Kol",
+        "ana": "Andaqui",
+        "anb": "Andoa",
+        "ang": "English, Old",
+        "ans": "Anserma",
+        "aoh": "Arma",
+        "aqp": "Atakapa",
+        "arc": "Aramaic, Official",
+        "aru": "Aruá",
+        "ash": "Abishira",
+        "atc": "Atsahuaca",
+        "auo": "Auyokawa",
+        "ave": "Avestan",
+        "avs": "Aushiri",
+        "awk": "Awabakal",
+        "axb": "Abipon",
+        "axg": "Arára, Mato Grosso",
+        "axm": "Armenian, Middle",
+        "ayy": "Ayta, Tayabas",
+        "bae": "Baré",
+        "bjb": "Banggarla",
+        "bjf": "Neo-Aramaic, Barzani Jewish",
+        "bjy": "Bayali",
+        "bll": "Biloxi",
+        "bmn": "Bina",
+        "boi": "Barbareño",
[…]

### 2014-04-27 universalcorpus/convert.py
Added dictionary to convert retired ISO 639-3 codes

@@ -1,5 +1,7 @@
 # -*- coding: utf-8 -*-
 
+# Reference: http://www-01.sil.org/iso639-3/download.asp
+
 constructed = {"ido":"Ido",
                "tlh":"Klingon",
                "tzl":"Talossan",
@@ -17,6 +19,254 @@ constructed = {"ido":"Ido",
                "rmv": "Romanova",
                "sjn": "Sindarin",
                "vol": "Volapük"}
+
+retired_headings = ('Id', 'Ref_Name', 'Ret_Reason', 'Change_To', 'Ret_Remedy', 'Effective')
+
+retired = [("fri", "Western Frisian", "C", "fry", "", "2007-02-01"),
+					 ("auv", "Auvergnat", "M", "oci", "", "2007-03-14"),
+					 ("gsc", "Gascon", "M", "oci", "", "2007-03-14"),
+					 ("lms", "Limousin", "M", "oci", "", "2007-03-14"),
+					 ("lnc", "Languedocien", "M", "oci", "", "2007-03-14"),
+					 ("prv", "Provençal", "M", "oci", "", "2007-03-14"),
+					 ("amd", "Amapá Creole", "N", "", "", "2007-07-18"),
+					 ("bgh", "Bogan", "D", "bbh", "", "2007-07-18"),
+					 ("bnh", "Banawá", "M", "jaa", "", "2007-07-18"),
+					 ("bvs", "Belgian Sign Language", "S", "", "Split into Langue des signes de Belgique Francophone [sfb], and Vlaamse Gebarentaal [vgt]", "2007-07-18"),
+					 ("ccy", "Southern Zhuang", "S", "", "Split into five languages: Nong Zhuang [zhn];  Yang Zhuang [zyg]; Yongnan Zhuang [zyn]; Zuojiang Zhuang [zzj]; Dai Zhuang [zhd].", "2007-07-18"),
+					 ("cit", "Chittagonian", "S", "", "Split into Rohingya [rhg], and Chittagonian (new identifier [ctg])", "2007-07-18"),
+					 ("flm", "Falam Chin", "S", "", "Split into Ranglong [rnl], and Falam Chin (new identifier [cfm]).", "2007-07-18"),
+					 ("jap", "Jaruára", "M", "jaa", "", "2007-07-18"),
[…]

### 2014-04-27 docs/computEL-final/data.tex
removed strikethrough words as suggested in alexis' comments

@@
 Although data size matters in general NLP, \emph{universality} is the top priority for a universal corpus.
-We chose to focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): Online Database of Interlinear Text (ODIN), Omniglot website, Universal Declaration of Human Rights (UHDR), and Wikipedia.
+We focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): Online Database of Interlinear Text (ODIN), Omniglot website, Universal Declaration of Human Rights (UHDR), and Wikipedia.
 
@@
 The Omniglot website\footnote{\url{http://www.omniglot.com}} is an online encyclopedia of writing systems and languages.
-We chose to extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us parallel data in a reasonably large number of languages.
+We extract information from pages on \emph{`Useful foreign phrases'} and the \emph{`Tower of Babel'} story, both of which give us parallel data in a reasonably large number of languages.
 % The urls for these resources are:
@@
 Even though glosses in ODIN are based on English, having been extracted from English-language documents, this is not true of IGTs in general.
-In particular, it is common for documentary linguists working on indigenous languages of the Americas to provide glosses and translations based on Spanish.
+For example, it is common for documentary linguists working on indigenous languages of the Americas to provide glosses and translations based on Spanish.
 For this reason, we believe it would be wise to specify the language used to produce the gloss.

### 2014-04-27 docs/computEL-final/relwork.tex
Tidied tex files
(I have removed some commented sections that are unlikely to be included
again.  They are of course still available in the old commits.)
removed strikethrough words as suggested in alexis' comments

@@
 In this section, we review existing efforts to compile multilingual machine-readable resources.
-Although some commercial resources are available, we restrict attention to freely accessible data, as this will be most relevant to documentary linguistics and computational linguists.
-All figures given below were correct at the time of writing, but it must be borne in mind that most resources discussed below are constantly growing.
+Although some commercial resources are available, we restrict attention to freely accessible data\footnote{All figures given below were correct at the time of writing, but it must be borne in mind that most resources discussed below are constantly growing.}.
 
@@
 Firstly, the data is not always machine readable.
-For instance, in their own words, ANLA consists of ``mostly paper records''.
 Even where the data is available digitally, these often take the form of scanned images or audio files.
@@
 
-Such resources are invaluable for many purposes.
-For instance, there can be plenty of material to work with for a
-researcher who is looking to find data on a specific language, who is
-willing to put in a little time browsing through different catalogues,
-and who only needs human-readable data.  However, for large-scale
-machine processing, they leave much to be desired.
+Such resources are invaluable for many purposes. For instance, there can be plenty of material to work with for a
+researcher who is looking to find data on a specific language, who is willing to put in a little time browsing through different catalogues, and who only needs human-readable data.
+However, for large-scale machine processing, they leave much to be desired.
 
@@
 A further difficulty is that languages are named, without the corresponding ISO language codes.
-This problem may be small when all 68 varieties are well-known, but will become a serious concern if this corpus collection is to grow and include smaller languages.
 

### 2014-04-27 docs/computEL-final/clean.tex
fixed superscript after punctuation.

--- 2014-04-25 docs/computEL-final/clean.tex
@@
 
-\paragraph{UDHR.} Although there is a pre-compiled version of the UDHR data in the Natural Language ToolKit (NLTK) corpora distribution,\footnote{\url{http://nltk.googlecode.com/svn/trunk/nltk_data/index.xml}}
-the distribution is laden with encoding problems.
-Instead, we used the plaintext files available from the Unicode website,\footnote{\url{http://unicode.org/udhr/d}} which are free from encoding issues.
+\paragraph{UDHR.} We used the plaintext UDHR files available from the Unicode website\footnote{\url{http://unicode.org/udhr/d}} which uses the standard UTF8 encoding for all languages.
 The first four lines of each file record metadata, and the rest is the translation of the UDHR.
@@
 To convert compressed Wikipedia dumps to textfiles, we used the WikiExtractor\footnote{\url{http://medialab.di.unipi.it/wiki/Wikipedia_Extractor}} tool.
-After conversion into textfiles, we used several regular expressions to delete residual Wikipedia markup and so-called ``magic words"\footnote{\url{http://en.wikipedia.org/wiki/Help:Magic_words}}.
+After conversion into textfiles, we used several regular expressions to delete residual Wikipedia markup and so-called ``\emph{magic words}".\footnote{\url{http://en.wikipedia.org/wiki/Help:Magic_words}}
 
@@
 Omniglot does not give codes at all, but only the language name.
-To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{\url{http://www-01.sil.org/iso639-3/iso-639-3.tab}}
+To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{\url{http://www-01.sil.org/iso639-3/iso-639-3.tab}}.
+
+To facilitate the standardization of language code, we have written a small python API that can be used to query information about a language and its code(s), as well as language family information.
 

### 2014-04-27 docs/computEL-final/concl.tex
edited final paragraph in conclusion as commented.

@@
 
-SeedLing will be made available initially through the websites of the authors, as we have yet to properly address the question of long-term access; we welcome ideas or collaborations along these lines.
-In addition, we are releasing a Python API which allows programmatic access to the data and is easily compatible with the analysis tools available from the Natural Language ToolKit \cite{BirdKleinLoper09} \footnote{\url{https://github.com/alvations/SeedLing}}.
-The API also facilitates access to information from both Ethnologue and WALS: a user can extract information about languages, language classification, endangerment status, typological features, and so on.
-Finally, we will make our clustering scripts freely available.
-
-% The API also includes a feature to access information on a language from Ethnologue and/or WALS.
-This feature queries the Ethnologue and/or WALS site both in real-time when internet connection is available and from a local copy of the information offline.
+SeedLing (and the API) is currently available via a GitHub repository.\footnote{https://github.com/alvations/SeedLing} We have yet to fully address questions of long-term access, and we welcome ideas or collaborations along these lines.

### 2014-04-27 docs/computEL-final/paper.tex
added author information

--- 2014-04-17 docs/computEL/paper.tex
@@
 \title{SeedLing: Building and using a seed corpus \\ for the Human Language Project}
-%Getting the Ball Rolling: Producing a Foundation Text for a Universal Corpus of the World's Languages}
 
-% \author{Liling Tan, Guy Emerson, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
-%   Universität des Saarlandes \\
-%   Campus, 66123 Saarbrücken, Germany \\
-%   {\tt liling.tan@uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-%   {\tt susfert@coli.uni-saarland.de}}
+\author{Guy Emerson, Liling Tan, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
+Universität des Saarlandes \\
+Campus, 66123 Saarbrücken, Germany \\
+{\tt liling.tan@uni-saarland.de, emerson@coli.uni-saarland.de,} \\
+{\tt susfert@coli.uni-saarland.de, apalmer@coli.uni-sb.de,}\\
+{\tt regneri@coli.uni-saarland.de}}
 

### 2014-04-28 docs/computEL-final/paper.tex
Edit email addresses.

@@
 Universität des Saarlandes \\
-Campus, 66123 Saarbrücken, Germany \\
-{\tt liling.tan@uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-{\tt susfert@coli.uni-saarland.de, apalmer@coli.uni-sb.de,}\\
-{\tt regneri@coli.uni-saarland.de}}
-
+66123 Saarbrücken, Germany \\
+{\tt \{emerson, liling, susfert, apalmer, regneri\}@coli.uni-saarland.de} \\
+%{\tt susfert@coli.uni-saarland.de, apalmer@coli.uni-sb.de,}\\
+%{\tt regneri@coli.uni-saarland.de}}
+}
 \date{}

### 2014-04-28 docs/computEL-final/data.tex
Wikipedia links.

@@
 
-Wikipedia\footnote{\url{http://www.wikipedia.org}} is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability.
+Wikipedia is a collaboratively-edited encyclopedia, appealing to use for NLP because of its large size and easy availability.
 At the time of writing, it contained 30.8 million articles in 286 languages, which provides a sizeable amount of monolingual text in a fairly wide range of languages.
-Text dumps are made regularly available, and can be downloaded from \url{www.dumps.wikimedia.org}.
+Text dumps are made regularly available, and can be downloaded from \url{http://dumps.wikimedia.org}.
 

### 2014-04-28 docs/computEL-final/data.tex
Edit data section.

@@
 Analysed texts, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.
-This is stored as a table, where each row represents a word, and each column a type of information.
-They propose the following columns: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
+This is stored as a table, where each row represents a word, and columns can represent the following types of information: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
 

### 2014-04-28 docs/computEL-final/cluster.tex
Formatting of links.

@@
 Both of these approaches require linguistic analysis.
-By contrast, we use surface features (character n-gram and word unigram frequencies) extracted from SeedLing, and apply an off-the-shelf hierarchical clustering algorithm.\footnote{\url{http://www.scipy.org/}} Specifically, each language is represented as a vector of frequencies of character bigrams, character trigrams, and word unigrams.
+By contrast, we use surface features (character n-gram and word unigram frequencies) extracted from SeedLing, and apply an off-the-shelf hierarchical clustering algorithm.\footnote{\url{http://www.scipy.org}} Specifically, each language is represented as a vector of frequencies of character bigrams, character trigrams, and word unigrams.
 Each of these three components is normalized to unit length.

### 2014-04-28 docs/computEL-final/concl.tex
Formatting of links.

@@
 
-SeedLing (and the API) is currently available via a GitHub repository.\footnote{https://github.com/alvations/SeedLing} We have yet to fully address questions of long-term access, and we welcome ideas or collaborations along these lines.
+SeedLing (and the API) is currently available via a GitHub repository.\footnote{\url{https://github.com/alvations/SeedLing}} We have yet to fully address questions of long-term access, and we welcome ideas or collaborations along these lines.

### 2014-04-28 docs/computEL-final/clean.tex
Added reference to Xia Lewis Lewis paper, tidied some discussion

--- 2014-04-25 docs/computEL-final/clean.tex
@@
 
-\paragraph{UDHR.} Although there is a pre-compiled version of the UDHR data in the Natural Language ToolKit (NLTK) corpora distribution,\footnote{\url{http://nltk.googlecode.com/svn/trunk/nltk_data/index.xml}}
-the distribution is laden with encoding problems.
-Instead, we used the plaintext files available from the Unicode website,\footnote{\url{http://unicode.org/udhr/d}} which are free from encoding issues.
+\paragraph{UDHR.} We used the plaintext UDHR files available from the Unicode website\footnote{\url{http://unicode.org/udhr/d}} which uses UTF-8 encoding for all languages.
 The first four lines of each file record metadata, and the rest is the translation of the UDHR.
@@
 \subsection{Language Codes}
-We use ISO~639-3 as our standard set of language codes, since it aims for universal coverage, and has widespread acceptance in the community.
-% As a result, we needed to make sure that all our data was labelled with these codes.
-Data from ODIN and the UDHR already used this standard, and hence did not pose problems.
+As \newcite{xia2010multilingual} explain, language names do not always suffice to identify languages, since many names are ambiguous.
+For this reason, sets of language codes exist to more accurately identify languages.
+We use ISO~639-3\footnote{url{http://www-01.sil.org/iso639-3/default.asp}} as our standard set of codes, since it aims for universal coverage, and has widespread acceptance in the community.
+The data from ODIN and the UDHR already used this standard.
+
+To facilitate the standardization of language codes, we have written a python API that can be used to query information about a language or a code, fetching up-to-date information from SIL International (which maintains the ISO~639-3 code set), as well as from Ethnologue.
 
@@
 
-Omniglot does not give codes at all, but only the language name.
-To resolve this issue, we automatically converted language names to codes using information from the SIL website.\footnote{\url{http://www-01.sil.org/iso639-3/iso-639-3.tab}}
+Omniglot does not give codes at all, but only the language name. To resolve this issue, we automatically converted language names to codes using information from the SIL website.
 
@@
 \newcite{bird2003port} describe a set of best practices for maintaining portability of digital information, outlining seven dimensions along which this can vary.
-Following this advice, we have ensured that all our data is available as plain text files, with utf-8 encoding, labelled with the relevant ISO 639-3 code.
+Following this advice, we have ensured that all our data is available as plain-text files, with UTF-8 encoding, labelled with the relevant ISO 639-3 code.
 Metadata is stored separately.
-We have written an API to allow access to the data according to the guidelines of \newcite{abney2010universal}, who remain agnostic as to the specific form of data storage.
-If, for reasons of space or speed, an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according to these guidelines.
+This allows users to easily process the data using the programming language or software of their choice.
+
+To allow access to the data following Abney and Bird's guidelines, as discussed in section \ref{sec:data}, we have written an API, which we distribute along with the data.
+Abney and Bird remain agnostic to the specific file format used, but if an alternative format would be preferred, the data would be straightfoward to convert since it can be accessed according to these guidelines.
+As examples of functionality, our API allows a user to fetch all sentences in a given language, all sentences from a given source.

### 2014-04-28 docs/computEL-final/clean.tex
Further edits

@@
 
-\paragraph{UDHR.} We used the plaintext UDHR files available from the Unicode website\footnote{\url{http://unicode.org/udhr/d}} which uses UTF-8 encoding for all languages.
+\paragraph{UDHR.} We used the plain-text UDHR files available from the Unicode website\footnote{\url{http://unicode.org/udhr/d}} which uses UTF-8 encoding for all languages.
 The first four lines of each file record metadata, and the rest is the translation of the UDHR.
@@
 For this reason, sets of language codes exist to more accurately identify languages.
-We use ISO~639-3\footnote{url{http://www-01.sil.org/iso639-3/default.asp}} as our standard set of codes, since it aims for universal coverage, and has widespread acceptance in the community.
+We use ISO~639-3\footnote{\url{http://www-01.sil.org/iso639-3/default.asp}} as our standard set of codes, since it aims for universal coverage, and has widespread acceptance in the community.
 The data from ODIN and the UDHR already used this standard.

### 2014-04-28 docs/computEL-final/cluster.tex
Further edits

@@
 \paragraph{Experimental Setup.}
-We performed hierarchical clustering, which produces a tree structure: each leaf represents a language, and each node a cluster.
-We used linkage methods, which recursively build the tree starting from the leaves.
+We first perform hierarchical clustering, which produces a tree structure: each leaf represents a language, and each node a cluster.
+We use linkage methods, which recursively build the tree starting from the leaves.
 Initially, each language is in a separate cluster, then we iteratively find the closest two clusters and merge them.
@@
 
-We tested four methods, which differ in how they define the distance between two clusters.
-The \texttt{complete} method considers all pairs of languages with one from each cluster, and finds the largest distance.
+We test four methods, which differ in how they define the distance between two clusters.
+The \texttt{complete} method considers all possible pairs with one language from each cluster, and finds the largest distance.
 The \texttt{single} method instead finds the smallest distance.
@@
 
-While a tree structure might be linguistically useful, evaluating the structure is more awkward.
-However, the above methods can be easily used to produce partitional clusters, by stopping when we reach a certain number of clusters.
+To ease evaluation, we produce a partitional clustering, by stopping when we reach a certain number of clusters.
 We set the number of clusters to 147, the number of top-level genetic groupings in Ethnologue.
@@
     ~        & Precision & Recall       & F-score    \\ \hline
+	\emph{random} & 0.1836	& 0.0917	& 0.0675 \\ \hline
 	\texttt{complete} & \textbf{0.3142}	& 0.1938	 & \textbf{0.1556} \\
@@
 	\texttt{average} & 0.2216	& 0.5163	& 0.1348 \\
-	\texttt{weighted} & 0.2326	& 0.3343	& 0.1271 \\ \hline
-	\emph{random} & 0.1836	& 0.0917	& 0.0675 \\
+	\texttt{weighted} & 0.2326	& 0.3343	& 0.1271 \\
     \end{tabular}
@@
 \paragraph{Evaluation.}
-There are many possible metrics to evaluate the quality of a clustering compared to a gold standard. 
+We compare our induced clusters to the language families given in Ethnologue.
+However, there are many possible metrics to evaluate the quality of a clustering.
 \newcite{amigo2009metrics} propose a set of criteria which a clustering evaluation metric should satisfy, and demonstrate that most popular metrics fail to satisfy at least one of these criteria.
-However, they prove that they are satisfied by the BCubed metric, which we adopt for this reason.
-To calculate this, we take the induced cluster and gold standard class for each language, and calculate the F-score of the cluster compared to the class.
+However, they prove that all criteria are satisfied by the BCubed metric, which we adopt for this reason.
+To calculate the BCubed score, we take the induced cluster and gold standard class for each language, and calculate the F-score of the cluster compared to the class.
 These F-scores are then averaged across all languages.
@@
 considering when pairs of languages are in the same cluster or the same class.
-For 147 clusters, the \texttt{complete} method gives a pairwise f-score of 0.147, while they report 0.140.
-The figures are not quite comparable since we are evaluating over a different set of languages; nonetheless, this result is suggestive, given that we have only used surface features, while they used typological features taken from WALS.
+For 147 clusters, the \texttt{complete} method gives a pairwise f-score of 0.147, while Georgi et al.
+report 0.140.
+The figures are not quite comparable since we are evaluating over a different set of languages; nonetheless, this result is suggestive, given that we only use surface features, while Georgi et al.
+use typological features taken from WALS.
 This demonstrates that it possible for cross-linguistic research to be conducted even based on extremely shallow features.
@@
 \paragraph{Qualitative Inspection.}
-Ga: It's actually pretty difficult to interpret the clusters directly because there's so much noise. I will try to write something better, but here are a few thoughts I've had:
-
+Interpreting the clusters directly is difficult, because of a large amount of noise in the data.
 Using the \texttt{complete} method with 147 clusters, we obtain a mix of large and small clusters.

### 2014-04-28 docs/computEL-final/data.tex
Further edits

@@
 
-Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, distinguishing between \textbf{aligned texts} and \textbf{analysed texts}.
-Aligned texts consist of multiple parallel documents, aligned at the document, sentence, or word level.
+Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, and we aim to adopt this structure where possible.
+Two types of text are distinguished: \textbf{aligned texts} consist of multiple parallel documents, aligned at the document, sentence, or word level.
 The collection of parallel documents as a whole is assigned a unique identifier, with each individual document labelled by this identifier and the corresponding language code.
@@
 Analysed texts, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.
-This is stored as a table, where each row represents a word, and columns can represent the following types of information: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
+This is stored as a table, where rows represent words, and columns represent the following types of information: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
 
@@
     \begin{tabular}{l|rr|rr}
-    ~         				& \#Languages & \#Families 	&\#tokens		& Size	\\ \hline
+    ~         				& Langs. 	& Families 	& Tokens		& Size	\\ \hline
     ODIN      				& 1,217      & 101       		& 58,556		& 39 MB		\\

### 2014-04-28 docs/computEL-final/paper.tex
Further edits

--- 2014-04-27 docs/computEL-final/paper.tex
@@
 Universität des Saarlandes \\
-Campus, 66123 Saarbrücken, Germany \\
-{\tt liling.tan@uni-saarland.de, emerson@coli.uni-saarland.de,} \\
-{\tt susfert@coli.uni-saarland.de, apalmer@coli.uni-sb.de,}\\
-{\tt regneri@coli.uni-saarland.de}}
-
+66123 Saarbrücken, Germany \\
+{\tt \{emerson, liling, susfert, apalmer, regneri\}} \\
+{\tt @coli.uni-saarland.de} \\
+}
 \date{}

### 2014-04-28 docs/computEL-final/relwork.tex
Further edits

@@
 In this section, we review existing efforts to compile multilingual machine-readable resources.
-Although some commercial resources are available, we restrict attention to freely accessible data\footnote{All figures given below were correct at the time of writing, but it must be borne in mind that most resources discussed below are constantly growing.}.
+Although some commercial resources are available, we restrict attention to freely accessible data.\footnote{All figures given below were correct at the time of writing, but it must be borne in mind that most resources discussed below are constantly growing.}
 
@@
 
-Such resources are invaluable for many purposes. For instance, there can be plenty of material to work with for a
-researcher who is looking to find data on a specific language, who is willing to put in a little time browsing through different catalogues, and who only needs human-readable data.
-However, for large-scale machine processing, they leave much to be desired.
+In short, while traditional archives are invaluable for many purposes, for large-scale machine processing, they leave much to be desired.
 
@@
 The Leipzig Corpora Collection (LCC)\footnote{\url{http://corpora.uni-leipzig.de}} \cite{biemann2007leipzig} provides download of corpora in 117 languages, and dictionaries in a number of others, bringing the total number of represented languages up to 230.
-The corpora are large, readily available, in plain text, and labelled with ISO language codes.
+The corpora are large, readily available, in plain-text, and labelled with ISO language codes.
 

### 2014-04-28 docs/computEL-final/concl.tex
Added heatmap spreadsheet. Also small changes.

--- 2014-04-27 docs/computEL-final/concl.tex
@@
 At present, SeedLing has data from 19\% of the world's languages, covering 72\% of language families.
-We believe that a corpus with such high degrees of language diversity, uniformity and cleanliness of data format, and ease of access provides an excellent seed for the universal corpus.
+We believe that a corpus with such diversity of languages, uniformity of format, cleanliness of data, and ease of access provides an excellent seed for a universal corpus.
 It is our hope that taking steps toward creating this resource will spur both further data contributions and interesting computational research with cross-linguistic or typological perspectives; we have here demonstrated SeedLing's utility for such research by using the data to perform language clustering, with promising results.
 
-SeedLing (and the API) is currently available via a GitHub repository.\footnote{https://github.com/alvations/SeedLing} We have yet to fully address questions of long-term access, and we welcome ideas or collaborations along these lines.
+SeedLing (both the data and the API) is currently available via a GitHub repository.\footnote{\url{https://github.com/alvations/SeedLing}} We have yet to fully address questions of long-term access, and we welcome ideas or collaborations along these lines.

### 2014-04-28 docs/computEL-final/data.tex
Edit data section.

@@
 
-Analysed texts, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.
+\textbf{Analysed} texts, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.
 This is stored as a table, where rows represent words, and columns represent the following types of information: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
@@
 Our resulting corpus runs the full gamut of text types outlined by Abney and Bird, ranging from single-language text (Wikipedia) to parallel text (UDHR and Omniglot) to IGTs (ODIN).
+For 332 languages the corpus contains data from more than one source.
 Table~\ref{table:corpus} gives some coverage statistics, and we describe each source in the following subsections.
@@
     ~         				& Langs. 	& Families 	& Tokens		& Size	\\ \hline
-    ODIN      				& 1,217      & 101       		& 58,556		& 39 MB		\\
-    Omniglot  				& 134        & 21        		&	6,749			& 677 KB	\\
-    UDHR      				& 355        & 47        		&	33,117		& 5.2 MB	\\
-    Wikipedia 				& 209        & 22       		&						& 37 GB		\\ \hline
-    \textbf{Combined}	& 1,347			 & 106 
+    ODIN      				& 1,270      & 100       		& 58,556		& 39 MB		\\
+    Omniglot  				& 129        & 20        		& 6,749		& 677 KB	\\
+    UDHR      				& 352        & 46        		& 33,117		& 5.2 MB	\\
+    Wikipedia 				& 271        & 21       		&				& 37 GB\\ \hline
+    \textbf{Combined}	    & 1,451	    & 105 
     \end{tabular}

### 2014-04-28 docs/computEL-final/data.tex
Updated with Liling's new token figures...

--- 2014-04-28 docs/computEL-final/data.tex
@@
 Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, and we aim to adopt this structure where possible.
-Two types of text are distinguished: \textbf{aligned texts} consist of multiple parallel documents, aligned at the document, sentence, or word level.
+Two types of text are distinguished:
+
+\textbf{Aligned texts} consist of multiple parallel documents, aligned at the document, sentence, or word level.
 The collection of parallel documents as a whole is assigned a unique identifier, with each individual document labelled by this identifier and the corresponding language code.
@@
 
-Analysed texts, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.
+\textbf{Analysed texts}, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.
 This is stored as a table, where rows represent words, and columns represent the following types of information: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
@@
 Table~\ref{table:corpus} gives some coverage statistics, and we describe each source in the following subsections.
+For 332 languages, the corpus contains data from more than one source.
 
@@
     ~         				& Langs. 	& Families 	& Tokens		& Size	\\ \hline
-    ODIN      				& 1,217      & 101       		& 58,556		& 39 MB		\\
-    Omniglot  				& 134        & 21        		&	6,749			& 677 KB	\\
-    UDHR      				& 355        & 47        		&	33,117		& 5.2 MB	\\
-    Wikipedia 				& 209        & 22       		&						& 37 GB		\\ \hline
-    \textbf{Combined}	& 1,347			 & 106 
+    ODIN      				& 1,270      & 100       		& 351,161		& 39 MB		\\
+    Omniglot  				& 129        & 20        		& 31,318		& 677 KB	\\
+    UDHR      				& 352        & 46        		& 640,588		& 5.2 MB	\\
+    Wikipedia 				& 271        & 21       		&				& 37 GB\\ \hline
+    \textbf{Combined}	    & 1,451	    & 105 
     \end{tabular}

### 2014-04-28 docs/computEL-final/cluster.tex
Updated with Liling's new token figures...
changed table -> Table when referencing

@@
 By contrast, we use surface features (character n-gram and word unigram frequencies) extracted from SeedLing, and apply an off-the-shelf hierarchical clustering algorithm.\footnote{\url{http://www.scipy.org}} Specifically, each language is represented as a vector of frequencies of character bigrams, character trigrams, and word unigrams.
-Each of these three components is normalized to unit length.
+Each of these three components is normalised to unit length.
+Data was taken from ODIN, Omniglot, and the UDHR.
 
@@
 
-In table \ref{table:cluster}, we give results of our clustering experiments, comparing the three clustering methods to a random baseline (clustering based on random distances, averaged over 20 runs).
+In Table \ref{table:cluster}, we give results of our clustering experiments, comparing the three clustering methods to a random baseline (clustering based on random distances, averaged over 20 runs).
 It is worth noting that precision is higher than recall.

### 2014-04-28 docs/computEL-final/clean.tex
New new heatmap

@@
 
+For some languages, there is more than one standard orthography.
+For example, Mandarin Chinese is written with either traditional and simplified characters; Serbian is written with either the Cyrillic or the Roman alphabet.
+For cross-linguistic NLP, it would be helpful to have codes for orthographies, but unfortunately no standards exist for this purpose.
+
 \subsection{File Formats}

### 2014-04-28 docs/computEL-final/clean.tex
figure\ref to Figure\ref

--- 2014-04-28 docs/computEL-final/clean.tex
@@
 Data for each language is saved in a separate XML file and the IGTs are encoded in tags of the form \texttt{<igt><example>...</example></igt>}.
-For example, the IGT in figure~\ref{fig:odin_fijian} is represented by the XML snippet in figure~\ref{fig:odin_fijian_xml}.
+For example, the IGT in Figure~\ref{fig:odin_fijian} is represented by the XML snippet in Figure~\ref{fig:odin_fijian_xml}.
 
@@
 In the above examples, the sentence is introduced by a letter or number, which needs to be removed; however, the form of such indexing elements varies.
-In addition, the source line in figure~\ref{fig:odin_yimas_xml} includes two types of metadata: the language name, and a citation, both of which introduce noise.
+In addition, the source line in Figure~\ref{fig:odin_yimas_xml} includes two types of metadata: the language name, and a citation, both of which introduce noise.
 Finally, extraneous punctuation such as the quotation marks in the translation line need to be removed.

### 2014-04-28 docs/computEL-final/concl.tex
Update number of languages and families.

@@
 The resulting corpus is more easily machine-readable than any of the underlying data sources, and has been stored according to the best practices suggested by \newcite{bird2003port}.
-At present, SeedLing has data from 19\% of the world's languages, covering 72\% of language families.
+At present, SeedLing has data from 19\% of the world's living languages, covering 72\% of language families.
 We believe that a corpus with such diversity of languages, uniformity of format, cleanliness of data, and ease of access provides an excellent seed for a universal corpus.
@@
 
-SeedLing (both the data and the API) is currently available via a GitHub repository.\footnote{\url{https://github.com/alvations/SeedLing}} We have yet to fully address questions of long-term access, and we welcome ideas or collaborations along these lines.
+SeedLing (data, API and documentation) is currently available via a GitHub repository.\footnote{\url{https://github.com/alvations/SeedLing}} We have yet to fully address questions of long-term access, and we welcome ideas or collaborations along these lines.

### 2014-04-28 docs/computEL-final/intro.tex
Update number of languages and families.

@@
 
-We believe the resulting corpus, which at present covers 1347 languages from 106 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing.
+We believe the resulting corpus, which at present covers 1451 languages from 105 language families, is the first of its kind: large enough and consistent enough to allow broadly multilingual language processing.
 To test this claim, we use SeedLing in a sample application (section \ref{sec:cluster}): the task of language clustering.

### 2014-04-28 docs/computEL-final/paper.tex
Update number of languages and families.

@@
   easily-accessible and machine-readable than any of the underlying
-  data sources, and, with data from 1347 languages covering 106
+  data sources, and, with data from 1451 languages covering 105
   language families, represents a significant base corpus for

### 2014-04-28 docs/computEL-final/clean.tex
Reworked discussion of experiments, trimmed down to 8 pages.

--- 2014-04-28 docs/computEL-final/clean.tex
@@
 
+Some languages have more than one orthography.
+For example, Mandarin Chinese is written with either traditional or simplified characters; Serbian is written with either the Cyrillic or the Roman alphabet.
+For cross-linguistic NLP, it could be helpful to have standard codes to identify orthographies, but at present none exist.
+
 \subsection{File Formats}

### 2014-04-28 docs/computEL-final/cluster.tex
Reworked discussion of experiments, trimmed down to 8 pages.

@@
 To exemplify the use of SeedLing for computational research on low-resource languages, we experiment with automatic detection of similar languages.
-When working on endangered languages, documentary and computational linguists alike face the issue of lack of resources.
-It is often helpful to exploit lexical, syntactic or morphological knowledge about related or similar languages.
-For example, information about relatedness can be useful for identifying high-resource languages to be used in bootstrapping approaches such as those described in \newcite{yarowsky:ngai:2001} or \newcite{xia2007multilingual}.
+When working on endangered languages, documentary and computational linguists alike face a lack of resources.
+It is often helpful to exploit lexical, syntactic or morphological knowledge of related languages.
+For example, similar high-resource languages can be used in bootstrapping approaches, such as described by \newcite{yarowsky:ngai:2001} or \newcite{xia2007multilingual}.
 
@@
 
-We test four methods, which differ in how they define the distance between two clusters.
-The \texttt{complete} method considers all possible pairs with one language from each cluster, and finds the largest distance.
-The \texttt{single} method instead finds the smallest distance.
-The \texttt{average} and \texttt{weighted} methods calculate averages across all the pairwise differences, either unweighted, or weighted according to the languages' positions in the tree.
+We define the distance between two clusters by considering all possible pairs of languages, with one from each cluster, and taking the largest distance.
+We experimented with other ways to define the distance between clusters, but results were poor and we omit results for brevity.
 
-To ease evaluation, we produce a partitional clustering, by stopping when we reach a certain number of clusters.
-We set the number of clusters to 147, the number of top-level genetic groupings in Ethnologue.
+To ease evaluation, we produce a partitional clustering, by stopping when we reach a certain number of clusters, set in advance.
 
 \begin{table}[t]
-\begin{centering}
-
+\centering
     \begin{tabular}{l|ccc}
     ~        & Precision & Recall       & F-score    \\ \hline
-	\emph{random} & 0.1836	& 0.0917	& 0.0675 \\ \hline
-	\texttt{complete} & \textbf{0.3142}	& 0.1938	 & \textbf{0.1556} \\
-	\texttt{single} & 0.1949	& \textbf{0.7850}	 & 0.1282 \\
-	\texttt{average} & 0.2216	& 0.5163	& 0.1348 \\
-	\texttt{weighted} & 0.2326	& 0.3343	& 0.1271 \\
+	Base. 1: random	& 0.184	& 0.092	& 0.068 \\
+	Base. 2: together	& 0.061	& 1.000	& 0.112 \\
+	Base. 3: separate	& 1.000	& 0.086	& 0.122 \\ \hline
+	SeedLing	& 0.255	& 0.205	& 0.150 
     \end{tabular}
-\caption{Comparison of clustering algorithms}
+\caption{Clustering compared with baselines}
 \label{table:cluster}
-\end{centering}
 \end{table}
 
+\begin{figure}[t]
+\begin{centering}
+\includegraphics[scale=0.75]{performance.png}
+\caption{Performance against number of clusters}
+\label{fig:performance}
+\end{centering}
+\end{figure}
+
 \paragraph{Evaluation.}
-We compare our induced clusters to the language families given in Ethnologue.
-However, there are many possible metrics to evaluate the quality of a clustering.
+We compare our clustering to the language families in Ethnologue.
+However, there are many ways to evaluate clustering quality.
 \newcite{amigo2009metrics} propose a set of criteria which a clustering evaluation metric should satisfy, and demonstrate that most popular metrics fail to satisfy at least one of these criteria.
-However, they prove that all criteria are satisfied by the BCubed metric, which we adopt for this reason.
+However, they prove that all criteria are satisfied by the BCubed metric, which we therefore adopt.
 To calculate the BCubed score, we take the induced cluster and gold standard class for each language, and calculate the F-score of the cluster compared to the class.
@@
 
-In Table \ref{table:cluster}, we give results of our clustering experiments, comparing the three clustering methods to a random baseline (clustering based on random distances, averaged over 20 runs).
-It is worth noting that precision is higher than recall.
-This is perhaps to be expected, given that related languages using wildly differing orthographies will appear different.
-Nonetheless, our system is reasonably capable of identifying languages which are both related and written similarly.
+In Table \ref{table:cluster}, we set the number of clusters to be 105, the number of language families in our data, and compare this with three baselines: a random baseline (averaged over 20 runs); putting all languages in a single cluster; and putting all languages in separate clusters.
+Our clustering outperforms all baselines.
+It is worth noting that precision is higher than recall, which is perhaps expected, given that related languages using wildly differing orthographies will appear distinct.
 
-To allow a closer comparison with \newcite{georgi2010wals}, we can also calculate pairwise scores - i.e.
-considering when pairs of languages are in the same cluster or the same class.
-For 147 clusters, the \texttt{complete} method gives a pairwise f-score of 0.147, while Georgi et al.
+To allow a closer comparison with \newcite{georgi2010wals}, we calculate pairwise scores - i.e.
+considering if pairs of languages are in the same cluster or the same class.
+For 105 clusters, we achieve a pairwise f-score of 0.147, while Georgi et al.
 report 0.140.
-The figures are not quite comparable since we are evaluating over a different set of languages; nonetheless, this result is suggestive, given that we only use surface features, while Georgi et al.
-use typological features taken from WALS.
-This demonstrates that it possible for cross-linguistic research to be conducted even based on extremely shallow features.
+The figures are not quite comparable since we are evaluating over a different set of languages; nonetheless, we only use surface features, while Georgi et al.
+use typological features from WALS.
+This suggests the possibility for cross-linguistic research to be conducted based on shallow features.
 
-%%% [ADD THIS!] As well as performing clustering, we can view this as an information retrieval task: given a language, what are similar languages?
-To do this, we found the languages with the closest vectors of n-grams and words.
-Since BCubed is calculated averaging across languages and not clusters, we can use exactly the same calculation, using the set of nearby languages in place of a cluster.
+In Figure \ref{fig:performance}, we vary the number of clusters.
+The highest f-score is obtained for 199 clusters.
+There is a notable jump in performance between 98 and 99, just before the true number of families, 105.
 
-\paragraph{Qualitative Inspection.}
-Interpreting the clusters directly is difficult, because of a large amount of noise in the data.
-Using the \texttt{complete} method with 147 clusters, we obtain a mix of large and small clusters.
-There are 27 singleton clusters, while the largest cluster contains 123 languages.
-Although these singleton clusters belong to classes of various size, it is interesting that the distribution of sizes is similar to the distribution in the gold standard data, where we have 27 single classes, and the largest class contains 150 languages.
-This isn't true for the other clustering methods - the \texttt{single} method produces 138 singleton clusters, with almost all languages in a single cluster of size 968.
+Interpreting the clusters directly is difficult, because they are noisy.
+However, the distribution of cluster sizes mirrors the true distribution - for 105 clusters, we have 48 clusters of size 1 or 2, with the largest cluster of size 130; while in fact there are 51 families with only 1 or 2 languages in the data, with the largest of size 150.

### 2014-04-28 docs/computEL-final/concl.tex
Reworked discussion of experiments, trimmed down to 8 pages.

@@
 
-In this paper, we have described the creation of SeedLing, a foundation text for a universal corpus, following the guidelines of Abney and Bird (2010; 2011).
+In this paper, we have described the creation of SeedLing, a foundation text for a Universal Corpus, following the guidelines of Abney and Bird (2010; 2011).
 To do this, we cleaned and standardised data from several multilingual data sources: ODIN, Omniglot, the UDHR, Wikipedia.

### 2014-04-28 docs/computEL-final/data.tex
New new heatmap
Update number of languages and families.
Reworked discussion of experiments, trimmed down to 8 pages.

@@
 
-\textbf{Aligned texts} consist of multiple parallel documents, aligned at the document, sentence, or word level.
-The collection of parallel documents as a whole is assigned a unique identifier, with each individual document labelled by this identifier and the corresponding language code.
-A monolingual document is simply one with only a single associated language code.
-Sentences and words within documents are also assigned identifiers (relative to the document), if alignment has been performed at that level.
+\textbf{Aligned texts} consist of parallel documents, aligned at the document, sentence, or word level. Note that monolingual documents are viewed as aligned texts only tied to a single language.
 
 \textbf{Analysed texts}, in addition to the raw text, contain more detailed annotations including parts of speech, morphological information, and syntactic relations.
-This is stored as a table, where rows represent words, and columns represent the following types of information: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
+This is stored as a table, where rows represent words, and columns represent: document ID, language code, sentence ID, word ID, wordform, lemma, morphological information, part of speech, gloss, head/governor, and relation/role.
 
@@
 Although data size matters in general NLP, \emph{universality} is the top priority for a universal corpus.
-We focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): Online Database of Interlinear Text (ODIN), Omniglot website, Universal Declaration of Human Rights (UHDR), and Wikipedia.
+We focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): the Online Database of Interlinear Text (ODIN), the Omniglot website, the Universal Declaration of Human Rights (UHDR), and Wikipedia.
 
@@
     Wikipedia 				& 271        & 21       		&				& 37 GB\\ \hline
-    \textbf{Combined}	    & 1,451	    & 105 
+    Combined	    & 1,451	    & 105 
     \end{tabular}
@@
 \end{table}
+
+\begin{figure*}[t]
+\begin{centering}
+\includegraphics[scale=0.5]{heatmap2.png}
+\caption{Heatmap of languages in SeedLing according to endangerment status}\label{fig:heatmap}
+\end{centering}
+\end{figure*}
 
@@
 According to Ethnologue, there are 7105 living languages, and 147 living language families.
-Across all our data sources, we manage to cover 1347 languages in 106 families, which represents 19.0\% of the world's languages.
+Across all our data sources, we manage to cover 1451 languages in 105 families, which represents 19.0\% of the world's languages.
 To get a better idea of the kinds of languages represented, we give a breakdown according to their EGIDS scores (Expanded Graded Intergenerational Disruption Scale) \cite{lewis2010assessing} in Figure \ref{fig:heatmap}.
@@
 In terms of language documentation, vigorous languages are less urgent goals than those in categories 6b and up, but this highlights an unexpected gap in linguistic resources.
-
-\begin{figure*}[t]
-\begin{centering}
-\includegraphics[scale=0.63]{heatmap-numbers.png}
-%\includegraphics{heatmap-numbers.png}
-\caption{Heatmap of languages in SeedLing according to endangerment status}\label{fig:heatmap}
-\end{centering}
-\end{figure*}

### 2014-04-28 docs/computEL-final/intro.tex
Reworked discussion of experiments, trimmed down to 8 pages.

@@
 
-At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{simons2011update} calculate that 37\% of the extant languages at the time were at various stages of losing transmisson to new generations of speakers.
+At the time of writing, 7105 living languages are documented in Ethnologue,\footnote{\url{http://www.ethnologue.com}} but \newcite{simons2011update} calculated that 37\% of extant languages were at various stages of losing transmisson to new generations.
 Only a fraction of the world's languages are well documented, fewer have machine-readable resources, and fewer again have resources with linguistic annotations \cite{maxwell2006annotation} - so the time to work on compiling these resources is now.
 
-Several years ago, Abney and Bird \shortcite{abney2010universal,abney2011data} posed the challenge of building a Universal Corpus, calling it the Human Language Project.
-Such a corpus would include data from all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
+Several years ago, Abney and Bird \shortcite{abney2010universal,abney2011data} posed the challenge of building a Universal Corpus, naming it the Human Language Project.
+Such a corpus would include data from all the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
 The challenge was issued to the computational linguistics community, from the perspective that the language processing, machine learning, and data manipulation and management tools well-known in computational linguistics must be brought to bear on the problems of documentary linguistics, if we are to make any serious progress toward building such a resource.
-The Universal Corpus as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages, which in turn drives developments in automated analysis, supporting the work and increasing the efficiency of the language documentation process.
+The Universal Corpus as envisioned would facilitate broadly cross-lingual natural language processing (NLP), in particular driving innovation in research addressing NLP for low-resource languages, which in turn supports the language documentation process.
 %\textcolor{blue}{[cite our AL work? is this too strong? also, run-on sentence. also, perhaps mention this workshop, which has precisely these goals]}
 
-We have accepted this challenge and have begun converting existing resources into a format consistent with Abney and Bird's specifications for a universal corpus.
-We aim for a collection of resources that includes data: (a) from as many languages as possible, and (b) in a format which is both in accordance with best practice archiving recommendations and readily accessible as training data for machine learning and for other computational methods.
-Of course there are many relevant efforts toward producing cross-linguistic resources, which we survey in section \ref{sec:related}.
+We have accepted this challenge and have begun converting existing resources into a format consistent with Abney and Bird's specifications.
+We aim for a collection of resources that includes data: (a) from as many languages as possible, and (b) in a format both in accordance with best practice archiving recommendations and also readily accessible for computational methods.
+Of course, there are many relevant efforts toward producing cross-linguistic resources, which we survey in section \ref{sec:related}.
 To the best of our knowledge, though, no existing effort meets these two desiderata to the extent of our corpus, which we name SeedLing: a seed corpus for the Human Language Project.
@@
 We describe the steps required in section \ref{sec:case_studies}, presenting each resource as a separate mini-case study.
-We hope to make future resource conversion efforts more efficient, as they can be guided by the lessons we learned in assembling our seed corpus.
-To that end, many of the resources described in section~\ref{sec:related} are candidates for inclusion in the next stage of building a universal corpus.
+We hope that the lessons we learned in assembling our seed corpus can guide future resource conversion efforts.
+To that end, many of the resources described in section~\ref{sec:related} are candidates for inclusion in the next stage of building a Universal Corpus.
 
@@
 To test this claim, we use SeedLing in a sample application (section \ref{sec:cluster}): the task of language clustering.
-With no additional pre-processing required, we extract surface-level features (frequencies of character n-grams and words) for estimating the similarity of two languages.
-Unlike most previous approaches to the task, we make no use of resources curated for aims of linguistic typology (e.g.
+With no additional pre-processing, we extract surface-level features (frequencies of character n-grams and words) to estimate the similarity of two languages.
+Unlike most previous approaches to the task, we make no use of resources curated for linguistic typology (e.g.
 values of typological features as in WALS \cite{wals}, Swadesh word lists).
 Despite our approach being highly dependent on orthography, our clustering performance matches the results obtained by \newcite{georgi2010wals} using typolological features, which demonstrates SeedLing's utility in cross-linguistic research.
-
-% \textcolor{blue}{[START HERE WITH CLEARER DESCRIPTION OF CLUSTERING RESULTS]}
-% Despite this approach being highly dependent on orthography, we are able to reconstruct substantial parts of several language families, demonstrating the utility of this resource in cross-linguistic research.
-Finally, we discuss future work in section \ref{sec:future}, and conclude in section \ref{sec:conclusion}.
-
-% Abney and Bird \shortcite{abney2010universal,abney2011data} posed the grand challenge of building a Universal Corpus, calling it the Human Language Project.
-Such a corpus would include all of the world's languages, in a consistent structure, facilitating large-scale cross-linguistic processing.
-They propose a specific data structure, which we describe and discuss in section \ref{sec:structure}.
-We have accepted their challenge, and have begun converting existing resources into a format consistent with their specifications for a universal corpus.
-We have drawn on four [or five?] web sources, cleaning and standardising them as described in section \ref{sec:sources}, to produce a seed corpus for the Human Language Project.
-In sections \ref{sec:stats} and \ref{sec:copyright}, we respectively give a summary of the data contained, and discuss copyright and distribution.
-\textcolor{blue}{[needs rewriting following reorganization.]}
-
-% [Note: remove this section if we don't do the work!] We believe the resulting corpus is the first of its kind: large enough and consistent enough to allow language processing on a grand scale.
-In section \ref{sec:cluster}, we give an example application of this corpus: language clustering.
-We use the frequencies of character n-grams and words to estimate the similarity of two languages.

### 2014-04-28 docs/computEL-final/relwork.tex
Reworked discussion of experiments, trimmed down to 8 pages.

@@
 In this section, we review existing efforts to compile multilingual machine-readable resources.
-Although some commercial resources are available, we restrict attention to freely accessible data.\footnote{All figures given below were correct at the time of writing, but it must be borne in mind that most resources discussed below are constantly growing.}
+Although some commercial resources are available, we restrict attention to freely accessible data.\footnote{All figures given below were correct at the time of writing, but it must be borne in mind that most of these resources are constantly growing.}
 
@@
 Many archives exist to store the wealth of traditional resources produced by the documentary linguistics community.
-Such documents are increasingly being digitised, or produced in a digital form, and there are a number of archives which now offer free online access to their data.
+Such documents are increasingly being digitised, or produced in a digital form, and there are a number of archives which now offer free online access.
 
-Some archives aim for a universal scope, including languages from all parts of the world.
-Of particular note are: The Language Archive, maintained by the Max Planck Institute of Psycholinguistics, which includes DoBeS (Dokumentation Bedrohter Sprachen) and many other projects;
-Collection Pangloss, maintained by LACITO (Langues et Civilisations à Tradition Orale);
-The Endangered Languages Archive (ELAR), maintained by the School of Oriental and African Studies, which forms one part of the Hans Rausing Endangered Languages Project (HRELP).
-
-Regional archives include: AILLA (Archive of Indigenous Languages of Latin America), University of Texas at Austin;
-AIATSIS collections (Australian Institute of Aboriginal and Torres Strait Islander Studies);
-California Language Archive, managed by the Survey of California and Other Indian Languages;
-ANLA (Alaska Native Language Archive), at the University of Alaska Fairbanks;
-PARADISEC (Pacific and Regional Archive for Digital Sources in Endangered Cultures).
+Some archives aim for a universal scope, such as The Language Archive (maintained by the Max Planck Institute of Psycholinguistics), Collection Pangloss (maintained by LACITO), and The Endangered Languages Archive (maintained by SOAS).
+Most archives are regional, including AILLA, ANLA, PARADISEC, and many others.
 

### 2014-04-28 docs/computEL-final/concl.tex
"Universal Corpus" capitalisation to max Abney and Bird

@@
 At present, SeedLing has data from 19\% of the world's living languages, covering 72\% of language families.
-We believe that a corpus with such diversity of languages, uniformity of format, cleanliness of data, and ease of access provides an excellent seed for a universal corpus.
+We believe that a corpus with such diversity of languages, uniformity of format, cleanliness of data, and ease of access provides an excellent seed for a Universal Corpus.
 It is our hope that taking steps toward creating this resource will spur both further data contributions and interesting computational research with cross-linguistic or typological perspectives; we have here demonstrated SeedLing's utility for such research by using the data to perform language clustering, with promising results.

### 2014-04-28 docs/computEL-final/data.tex
"Universal Corpus" capitalisation to max Abney and Bird

@@
 
-Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for the universal corpus in more detail, and we aim to adopt this structure where possible.
+Building on their previous paper, \newcite{abney2011data} describe the data structure they envisage for a Universal Corpus in more detail, and we aim to adopt this structure where possible.
 Two types of text are distinguished:
@@
 
-Although data size matters in general NLP, \emph{universality} is the top priority for a universal corpus.
+Although data size matters in general NLP, \emph{universality} is the top priority for a Universal Corpus.
 We focus on the following data sources, because they include a large number of languages, include several parallel texts, and demonstrate a variety of data types which a linguist might encounter (structured, semi-structured, unstructured): the Online Database of Interlinear Text (ODIN), the Omniglot website, the Universal Declaration of Human Rights (UHDR), and Wikipedia.
@@
 
-ODIN (The Online Database of Interlinear Text) is a repository of
-interlinear glossed texts (IGTs) extracted from scholarly documents
-\cite{lewis2006odin,lewis2010odin}.  Compared to other resources, it
-is notable for the breadth of languages included and the level of
-linguistic annotation.  An IGT canonically consists of three lines:
-(i) the source, a sentence in a target language, (ii) the gloss, an
-analysis of each source element, and (iii) the translation,
-done at the sentence level. The gloss line can additionally include a
-number of linguistic terms, which means that the gloss 
-%can more
-%properly described as being written in a metalanguage, rather than a
-is written in metalanguage rather than natural language.  In ODIN, translations are into English, and glosses
-are written in an English-based metalanguage.  An accepted set of
-guidelines are given by the Leipzig Glossing
-Rules,\footnote{\url{http://www.eva.mpg.de/lingua/resources/glossing-rules.php}}
-where morphemes within words are separated by hyphens (or equal signs,
-for clitics), and the same number of hyphens should appear in each
-word of the source and gloss.
+ODIN (The Online Database of Interlinear Text) is a repository of interlinear glossed texts (IGTs) extracted from scholarly documents \cite{lewis2006odin,lewis2010odin}.
+Compared to other resources, it is notable for the breadth of languages included and the level of linguistic annotation.
+An IGT canonically consists of three lines: (i) the source, a sentence in a target language, (ii) the gloss, an analysis of each source element, and (iii) the translation, done at the sentence level.
+The gloss line can additionally include a number of linguistic terms, which means that the gloss is written in metalanguage rather than natural language.
+In ODIN, translations are into English, and glosses are written in an English-based metalanguage.
+An accepted set of guidelines are given by the Leipzig Glossing Rules,\footnote{\url{http://www.eva.mpg.de/lingua/resources/glossing-rules.php}} where morphemes within words are separated by hyphens (or equal signs, for clitics), and the same number of hyphens should appear in each word of the source and gloss.
 
-The data from ODIN poses the first obstacle to straightforwardly adopting Abney and Bird's data structure. 
-The proposed data structure for the universal corpus is aligned at the word level, and includes a specific list of relevant features which should be used to annotate words.
+The data from ODIN poses the first obstacle to straightforwardly adopting Abney and Bird's proposal.
+The suggested data structure is aligned at the word level, and includes a specific list of relevant features which should be used to annotate words.
 When we try to adapt IGTs into this format, we run into certain problems.
 Firstly, there is the problem that the most fundamental unit of analysis according to the Leipzig Glossing Rules is the morpheme, not the word.
-Ideally, we should encode this information explicitly in a universal corpus, assigning a unique identifier to each morpheme (instead of, or in addition to each word).
+Ideally, we should encode this information explicitly in a Universal Corpus, assigning a unique identifier to each morpheme (instead of, or in addition to each word).
 Indeed, \newcite{haspelmath2011segment} argues that there is no cross-linguistically valid definition of \textit{word}, which undermines the central position of words in the proposed data structure.
@@
 However, it is not straightforward to incorporate it into the proposed structure for analysed texts, either.
-% We could simply always map the gloss to the \textsc{gloss} field, but this would run against the suggestion that this field should "contain symbols from controlled vocabularies".
-Indeed, this information might be more suited to the \textsc{morph, pos,} or \textsc{rel} fields, but it would be difficult to automatically process ODIN to determine which morphemes should be assigned to which field.
 One possible resolution is to move all elements of the gloss written in capital letters to the \textsc{morph} field (as functional elements are usually annotated in this way), and all remaining elements to the \textsc{gloss} field.
@@
 
-The simplest solution we can see is to allow morphemes to be a level of structure in the universal corpus, just as documents, sentences, and words already are.
+The simplest solution we can see is to allow morphemes to be a level of structure in the Universal Corpus, just as documents, sentences, and words already are.
 The overall architecture remains unchanged.
@@
 There is a unique identifier for each morpheme, and glosses are stored under this identifier and the corresponding gloss code.
-Furthermore, to motivate the important place of parallel texts in a universal corpus, Abney and Bird view translations into a high-resource reference language as a convenient surrogate of meaning.
+Furthermore, to motivate the important place of parallel texts in a Universal Corpus, Abney and Bird view translations into a high-resource reference language as a convenient surrogate of meaning.
 By the same reasoning, we can use glosses to provide a more detailed surrogate of meaning, only written in a metalanguage instead of a natural one.

### 2014-04-28 docs/computEL-final/paper.tex
Reworked discussion of experiments, trimmed down to 8 pages.
"Universal Corpus" capitalisation to max Abney and Bird

@@
   describe our own approach. To build the foundation text for a
-  universal corpus, we crawl and clean texts from several web sources
+  Universal Corpus, we crawl and clean texts from several web sources
   that contain data from a large number of languages, and convert them
@@
   language families, represents a significant base corpus for
-  researchers to draw on and add to in the future. To show the utility
-  of SeedLing for cross-lingual computational research, we use our
+  researchers to draw on and add to in the future. To demonstrate the
+  utility of SeedLing for cross-lingual computational research, we use our
   data in the test application of detecting similar languages.
@@
 \input{relwork}
-%\input{liling}
 \input{data}
@@
 \input{cluster}
-%\input{corpus}
 \input{concl}

### 2014-04-28 docs/computEL-final/data.tex
Michaela's affiliation, and small edits

@@
 
-Out of our data sources, three can be straightforwardly represented in their aligned text structure.
-However, ODIN contains richer annotations, which are in fact difficult to fit into their proposal, and which we discuss in section \ref{sec:odin} below.
+Out of our data sources, three can be straightforwardly represented in the aligned text structure.
+However, ODIN contains richer annotations, which are in fact difficult to fit into Abney and Bird's proposal, and which we discuss in section \ref{sec:odin} below.
 

### 2014-04-28 docs/computEL-final/paper.tex
Michaela's affiliation, and small edits

@@
 
-\author{Guy Emerson, Liling Tan, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
+\author{Guy Emerson, Liling Tan, \\ \textbf{Susanne Fertmann, Alexis Palmer} \\
 Universität des Saarlandes \\
 66123 Saarbrücken, Germany \\
-{\tt \{emerson, liling, susfert, apalmer, regneri\}} \\
-{\tt @coli.uni-saarland.de} \\
-}
+{\tt \{emerson, liling, susfert,} \\ {\tt apalmer\} @coli.uni-saarland.de} \\\And
+Michaela Regneri \\
+IT-Department \\
+SPIEGEL-Verlag \\
+Hamburg, Germany \\
+{\tt michaela\_regneri@spiegel.de}}
+
 \date{}

### 2014-04-28 docs/computEL-final/paper.tex
Corrections, mainly in evaluation.

--- 2014-04-28 docs/computEL-final/paper.tex
@@
 
-\author{Guy Emerson, Liling Tan, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
+\author{Guy Emerson, Liling Tan, \\ \textbf{Susanne Fertmann, Alexis Palmer} \\
 Universität des Saarlandes \\
 66123 Saarbrücken, Germany \\
-{\tt \{emerson, liling, susfert, apalmer, regneri\}} \\
-{\tt @coli.uni-saarland.de} \\
-}
+{\tt \{emerson, liling, susfert,} \\ {\tt apalmer\}@coli.uni-saarland.de} \\\And
+Michaela Regneri \\
+IT-Department \\
+SPIEGEL-Verlag \\
+Hamburg, Germany \\
+{\tt michaela\_regneri@spiegel.de}}
+
 \date{}

### 2014-04-28 docs/computEL-final/paper.tex
Reset Michaela's affiliation, added acknowledgements.

--- 2014-04-28 docs/computEL-final/paper.tex
@@
 
-\author{Guy Emerson, Liling Tan, Susanne Fertmann, Alexis Palmer and Michaela Regneri \\
+\author{Guy Emerson, Liling Tan, Susanne Fertmann, Alexis Palmer, and Michaela Regneri \\
 Universität des Saarlandes \\
@@
 {\tt \{emerson, liling, susfert, apalmer, regneri\}} \\
-{\tt @coli.uni-saarland.de} \\
-}
+{\tt @coli.uni-saarland.de}}
+
 \date{}
@@
 
+\section*{Acknowledgements}
+We thank the three anonymous reviewers for their helpful comments.
+This research was supported in part by the Cluster of Excellence "Multi-modal Computing and Interaction" in the German Excellence Initiative.
+
 \bibliography{references}

### 2014-05-19 docs/computEL-final/cluster.tex

@@
     ~        & Precision & Recall       & F-score \\ \hline
+	SeedLing	& 0.255	& 0.205	& 0.150 \\ \hline
 	Base. 1: random	& 0.184	& 0.092	& 0.068 \\
 	Base. 2: together	& 0.061	& 1.000	& 0.112 \\
-	Base. 3: separate	& 1.000	& 0.086	& 0.122 \\ \hline
-	SeedLing	& 0.255	& 0.205	& 0.150 
+	Base. 3: separate	& 1.000	& 0.086	& 0.122
     \end{tabular}
@@
 
-In Table \ref{table:cluster}, we set the number of clusters to be 105, the number of language families in our data, and compare this with three baselines: a random baseline (averaged over 20 runs); putting all languages in a single cluster; and putting all languages in separate clusters.
+In Table \ref{table:cluster}, we set the number of clusters to be 105, the number of language families in our data, and compare this with three baselines: a random baseline (averaged over 20 runs); putting all languages in a single cluster; and putting each language in a separate cluster.
 Our clustering outperforms all baselines.
@@
 Interpreting the clusters directly is difficult, because they are noisy.
-However, the distribution of cluster sizes mirrors the true distribution - for 105 clusters, we have 48 clusters of size 1 or 2, with the largest cluster of size 130; while in fact there are 51 families with only 1 or 2 languages in the data, with the largest of size 150.
+However, the distribution of cluster sizes mirrors the true distribution - for 105 clusters, we have 48 clusters of size 1 or 2, with the largest cluster of size 130; while in our gold standard, there are 51 families with only 1 or 2 languages in the data, with the largest of size 150.

### 2014-05-19 docs/computEL-final/relwork.tex

@@
 
-Many of these archives are part of the Open Language Archive Community (OLAC), a subcommunity of the Open Archives Initiative.
+Many archives are part of the Open Language Archive Community (OLAC), a subcommunity of the Open Archives Initiative.
 OLAC maintains a metadata standard, based on the 15-element Dublin Core, which allows a user to search through all participating archives in a unified fashion.
 However, centralising access to disparate resources, while of course extremely helpful, does not solve the problem of inconsistent standards.
-Indeed, it can be difficult even to answer simple questions like ``how many languages are represented?''
+Indeed, it can even be hard to answer simple questions like ``how many languages are represented?''
 

### 2014-05-20 docs/computEL-final/intro.tex
Small typo

@@
 To produce SeedLing, we have drawn on four web sources, described in section \ref{sec:sources}.
-To bring the four resources into a single common format and data structure (section \ref{sec:structure}), each required different degrees and types of cleaning and standardisation.
+To bring the four resources into a common format and data structure (section \ref{sec:structure}), each required different degrees and types of cleaning and standardisation.
 We describe the steps required in section \ref{sec:case_studies}, presenting each resource as a separate mini-case study.
